---
title: "6A. Random number generators"
author: "Mingyang Lu"
old_id: "06A"
---

Stochastic models are driven by streams of random numbers, so Part 6 begins with
how those numbers are made. A random number generator (RNG) produces a sequence
that only *looks* random: it is fully deterministic (hence *pseudo*-random) yet
passes statistical tests for randomness. Good generators are fast, needing only a
handful of arithmetic and logical operations per value, and the samplers built on
them let us draw from any distribution we need [@devroye1986nonuniform]. This
chapter tests a generator visually and then builds samplers for the uniform,
exponential, and Gaussian distributions.

## 6A.1 A visual test for randomness

A quick visual check, the Pickover test [@pickover1995random], plots consecutive
pairs $(R_i, R_{i+k})$ for lags $k = 1, 2, 3$. A good uniform
generator scatters these points evenly over the unit square with no visible
structure; stripes, bands, or clusters would betray correlations between nearby
values.

### R implementation {.impl}

```{r}
#| fig-width: 6
#| fig-height: 5
#| out-width: "62%"
set.seed(10)
n <- 3000
r <- runif(n)                 # n pseudo-random numbers from R's default RNG

rng_test <- function(r) {
  n <- length(r)
  par(mar = c(5, 4, 1, 7), xpd = TRUE)   # room for the legend outside, on the right
  plot(r[1:(n-1)], r[2:n], pch = 16, cex = 0.4, asp = 1,
       col = adjustcolor(2, 0.5), xlab = bquote(R[i]), ylab = bquote(R[i+k]))
  points(r[1:(n-2)], r[3:n], pch = 16, cex = 0.4, col = adjustcolor(3, 0.5))
  points(r[1:(n-3)], r[4:n], pch = 16, cex = 0.4, col = adjustcolor(4, 0.5))
  legend("topright", inset = c(-0.28, 0), title = "k",
         legend = c("1", "2", "3"), col = 2:4, pch = 16)
}
rng_test(r)
```

### Python implementation {.impl}

```{python}
#| fig-width: 6
#| fig-height: 5
#| out-width: "62%"
import numpy as np
import matplotlib.pyplot as plt

np.random.seed(10)
n = 3000
r = np.random.rand(n)         # n pseudo-random numbers from NumPy's default RNG

def rng_test(r):
    n = len(r)
    plt.scatter(r[:n-1], r[1:], s=1, alpha=0.5, color="red",   label="1")
    plt.scatter(r[:n-2], r[2:], s=1, alpha=0.5, color="green", label="2")
    plt.scatter(r[:n-3], r[3:], s=1, alpha=0.5, color="blue",  label="3")
    plt.xlabel(r"$R_i$"); plt.ylabel(r"$R_{i+k}$")
    plt.gca().set_aspect("equal")
    plt.legend(title="k", loc="upper left", bbox_to_anchor=(1.02, 1))
    plt.show()

rng_test(r)
```

The points fill the unit square evenly, with no visible pattern, so the default
generator passes this test. The picture would look quite different for a
non-uniform distribution, or for a poor generator with short-range correlations.

## 6A.2 Uniform distribution of discrete numbers

Discrete uniform draws come from sampling **with replacement**; sampling without
replacement would exhaust the pool and correlate successive values. As an example
we draw $10^4$ values from $\{1, 2\}$.

### R implementation {.impl}

```{r}
n <- 10000
r_binary <- sample(x = 2, size = n, replace = TRUE)
hist(r_binary, freq = FALSE, breaks = 2, xlab = "r", main = "")
```

### Python implementation {.impl}

```{python}
n = 10000
r_binary = np.random.choice([1, 2], size=n, replace=True)
plt.hist(r_binary, bins=2, density=True, edgecolor="black", alpha=0.7)
plt.xlabel("r"); plt.ylabel("Density"); plt.show()
```

The two outcomes occur with roughly equal probability, as the histogram shows.

## 6A.3 Exponential distribution

Given uniform draws $x$ on $(0, 1)$, the *inverse-transform* method turns them into
draws $y$ from an exponential distribution by inverting the exponential
cumulative distribution [@devroye1986nonuniform],

$$ y = -\ln x . \tag{1} $$

::: {.callout-note title="Recipe 6A.3.1"}
**Objective.** Generate exponentially distributed random numbers from uniform ones.

**Model.** The target is the exponential distribution on `y >= 0`.

**Method.** Inverse-transform sampling of the exponential distribution: for a uniform `x` on (0, 1), take `y = -ln(x)`.

**Test.** `n = 10000` uniform draws on (0, 1), transformed and shown as a density histogram.

**Show.** A density histogram of the samples against the exponential density.

**Verification.** The flat uniform input becomes the characteristic decaying exponential density after the transform.
:::

### R implementation {.impl}

```{r}
n <- 10000
r <- runif(n)
y <- -log(r)                  # exponential draws by the inverse transform
hist(r, freq = FALSE, xlab = "r", main = "")
hist(y, freq = FALSE, xlab = "y", main = "")
```

### Python implementation {.impl}

```{python}
n = 10000
r = np.random.rand(n)
y = -np.log(r)                # exponential draws by the inverse transform
plt.hist(r, bins=30, density=True, color="blue", edgecolor="black", alpha=0.7)
plt.xlabel("r"); plt.ylabel("Density"); plt.show()
plt.hist(y, bins=30, density=True, color="green", edgecolor="black", alpha=0.7)
plt.xlabel("y"); plt.ylabel("Density"); plt.show()
```

The flat uniform input (first histogram) becomes the characteristic decaying
exponential (second) after the transform.

## 6A.4 Gaussian distribution

For Gaussian draws we use the polar form of the Box-Muller method
[@box1958note]. Sample a point $(x, y)$ uniformly in the square $[-1, 1]^2$, keep
it only if it falls inside the unit disk ($0 < R^2 < 1$ with $R^2 = x^2 + y^2$),
and otherwise resample. Each accepted point yields two independent
standard-normal values,

$$
\begin{cases}
r_1 = x \sqrt{-2 \ln(R^2) / R^2} \\
r_2 = y \sqrt{-2 \ln(R^2) / R^2}
\end{cases} \tag{2}
$$

In practice one simply calls `rnorm` or `np.random.randn`, but it is instructive
to see how such a sampler works.

::: {.callout-note title="Recipe 6A.4.1"}
**Objective.** Generate standard-normal random numbers from uniform ones.

**Model.** The target is the standard normal distribution `N(0, 1)`.

**Method.** The polar (Marsaglia) Box-Muller method: draw a point uniform in the square [-1, 1]^2, keep it if `R2 = x^2 + y^2` lies in (0, 1), and return `x*sqrt(-2*ln(R2)/R2)` and `y*sqrt(-2*ln(R2)/R2)` as two independent normals.

**Test.** `n = 10000` accepted draws, shown as a density histogram.

**Show.** A density histogram of the samples against the standard normal density.

**Verification.** Both output streams follow the bell-shaped normal density.
:::

### R implementation {.impl}

```{r}
ran_gaussian <- function() {
  repeat {
    x <- 2 * runif(2) - 1                  # two uniforms on [-1, 1]
    R2 <- sum(x * x)
    if (R2 < 1 && abs(R2) > 1e-6) break    # accept only inside the unit disk
  }
  return(x * sqrt(-2 * log(R2) / R2))
}

n <- 10000
y <- replicate(n, ran_gaussian())          # 2 x n: two normal streams
hist(y[1,], freq = FALSE, xlab = expression(r[1]), main = "")
hist(y[2,], freq = FALSE, xlab = expression(r[2]), main = "")
```

### Python implementation {.impl}

```{python}
def ran_gaussian():
    while True:
        x = 2 * np.random.rand(2) - 1            # two uniforms on [-1, 1]
        R2 = np.sum(x * x)
        if 0 < R2 < 1 and abs(R2) > 1e-6:        # accept only inside the unit disk
            break
    return x * np.sqrt(-2 * np.log(R2) / R2)

n = 10000
y = np.array([ran_gaussian() for _ in range(n)]).T   # 2 x n: two normal streams
plt.hist(y[0], bins=30, density=True, color="blue", edgecolor="black", alpha=0.7)
plt.xlabel("r_1"); plt.ylabel("Density"); plt.show()
plt.hist(y[1], bins=30, density=True, color="green", edgecolor="black", alpha=0.7)
plt.xlabel("r_2"); plt.ylabel("Density"); plt.show()
```

Both output streams follow the bell-shaped normal distribution. Because R and
Python draw from different generators, the individual values differ between the
two, but the distributions they sample are the same, which is all that matters for
a stochastic simulation.
