---
title: "6B. Brownian motion"
author: "Mingyang Lu"
old_id: "06B"
---

Brownian motion, the erratic wandering of a particle buffeted by random kicks, is
the prototypical stochastic process and the foundation for the stochastic
differential equations of the following chapters. We build it up from a simple
one-dimensional random walk, study the statistics of an ensemble of walks, and
arrive at the Wiener process.

## 6B.1 A one-dimensional random walk

A particle starts at the origin, $x(0) = 0$. At each time step it hops left or
right by a fixed distance $\Delta x$, with equal probability. We write two
implementations: one with an explicit loop (like an ODE integrator) and one using
only vector operations. The vectorized version is concise but less flexible for
more complex dynamics; both give statistically identical walks.

### R implementation {.impl}

```{r}
# 1D random walk, explicit loop
brownian <- function(t.total, dt, dx) {
  t_all <- seq(0, t.total, by = dt)
  n_all <- length(t_all)
  x_all <- numeric(n_all)          # starts at 0
  for (i in 1:(n_all - 1)) {
    x_all[i+1] <- x_all[i] + dx * sample(c(-1, 1), size = 1)
  }
  return(cbind(t_all, x_all))
}

# 1D random walk, vectorized (cumulative sum of random +/- steps)
brownian_concise <- function(t.total, dt, dx) {
  t_all <- seq(0, t.total, by = dt)
  x_all <- cumsum(sample(c(-1, 1), size = length(t_all), replace = TRUE)) * dx
  return(cbind(t_all, x_all))
}

set.seed(12)
t.total <- 1000; dt <- 1; dx <- 1; nrep <- 1000   # nrep independent walks
results_bm_1d <- replicate(nrep, brownian(t.total, dt, dx))   # array [time, 2, rep]

plot(NULL, xlab = "t", ylab = "x", xlim = c(0, 1000), ylim = c(-70, 70))
for (i in 1:10) lines(results_bm_1d[,1,i], results_bm_1d[,2,i], col = i)   # first 10 walks
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

# 1D random walk, explicit loop
def brownian(t_total, dt, dx):
    t_all = np.arange(0, t_total + dt, dt)
    n_all = len(t_all)
    x_all = np.zeros(n_all)          # starts at 0
    for i in range(1, n_all):
        x_all[i] = x_all[i-1] + dx * np.random.choice([-1, 1])
    return np.column_stack((t_all, x_all))

# 1D random walk, vectorized (cumulative sum of random +/- steps)
def brownian_concise(t_total, dt, dx):
    t_all = np.arange(0, t_total + dt, dt)
    x_all = np.cumsum(np.random.choice([-1, 1], size=len(t_all))) * dx
    return np.column_stack((t_all, x_all))

np.random.seed(12)
t_total, dt, dx, nrep = 1000, 1, 1, 1000          # nrep independent walks
results_bm_1d = np.array([brownian(t_total, dt, dx) for _ in range(nrep)])  # [rep, time, 2]

plt.xlabel("t"); plt.ylabel("x"); plt.xlim(0, 1000); plt.ylim(-70, 70)
for i in range(10):
    plt.plot(results_bm_1d[i, :, 0], results_bm_1d[i, :, 1])   # first 10 walks
plt.show()
```

Each walk wanders differently because it draws its own random steps, but as a
group they spread symmetrically about the origin, some ending far from it and
others near.

## 6B.2 Statistics of the ensemble

Across many walks the mean position stays at zero while the spread grows with
time. At any fixed $t$ the distribution of $x$ is (exactly) binomial and, after
many steps, approaches a Gaussian by the central limit theorem, with a width that
widens as $t$ increases. Theory gives the variance $\langle x^2 \rangle = 2 D t$,
with the one-dimensional diffusion constant [@einstein1905brownian]

$$ D = \lim_{\Delta t \rightarrow 0} \frac{\Delta x^2}{2 \Delta t} . \tag{3} $$

We view the spread with a box plot across sampled times, confirm the Gaussian
shape with histograms at a few times, and check the variance against the theory.
For our steps ($\Delta x = \Delta t = 1$), $D = 0.5$, so $\langle x^2 \rangle = t$.

### R implementation {.impl}

```{r}
t_ind <- seq(1, 1000, by = 100)                 # sampled time indices
t_sample <- results_bm_1d[t_ind, 1, 1]          # the corresponding times (0, 100, ...)

# box plot of x at each sampled time
bm_long <- data.frame(t = c(results_bm_1d[t_ind, 1, ]), x = c(results_bm_1d[t_ind, 2, ]))
boxplot(x ~ t, data = bm_long, xlab = "t", ylab = "x", ylim = c(-100, 100))
```

```{r}
#| fig-width: 9
#| fig-height: 3.2
#| out-width: "92%"
#| rmar: false
# histograms of x at three representative times (Gaussian, widening with t)
par(mfrow = c(1, 3), mar = c(4, 4, 2.5, 1))
for (i in c(1, 5, 10)) {
  x <- results_bm_1d[t_ind[i], 2, ]
  hist(x, freq = FALSE, xlab = "x", xlim = c(-100, 100),
       main = paste0("t = ", t_sample[i]))
}
```

```{r}
# mean(x) and mean(x^2) vs time, against theory
mean_x  <- sapply(t_ind, function(i) mean(results_bm_1d[i, 2, ]))
mean_x2 <- sapply(t_ind, function(i) mean(results_bm_1d[i, 2, ]^2))
plot(t_sample, mean_x, col = 1, xlab = "t", ylab = "Values", xlim = c(0, 910), ylim = c(-20, 1000))
points(t_sample, mean_x2, col = 2)
abline(a = 0, b = 0, col = 3); abline(a = 0, b = 1, col = 4)
legend("topleft", inset = 0.02, col = 1:4, pch = c(1, 1, NA, NA), lty = c(NA, NA, 1, 1),
       legend = c("Mean(x)", "Mean(x^2)", "y = 0 (theory)", "y = t (theory)"))
```

### Python implementation {.impl}

```{python}
t_ind = np.arange(0, 1000, 100)                 # sampled time indices
t_sample = results_bm_1d[0, t_ind, 0]           # the corresponding times (0, 100, ...)
data_bm_1d = np.array([results_bm_1d[:, k, 1] for k in t_ind]).T   # [rep, sampled-time]

plt.boxplot(data_bm_1d); plt.xticks(range(1, len(t_ind)+1), t_ind)
plt.xlabel("t"); plt.ylabel("x"); plt.ylim(-100, 100); plt.show()
```

```{python}
#| fig-width: 9
#| fig-height: 3.2
#| out-width: "92%"
# histograms of x at three representative times (Gaussian, widening with t)
fig, axs = plt.subplots(1, 3, figsize=(9, 3.2))
for ax, i in zip(axs, [0, 4, 9]):
    ax.hist(data_bm_1d[:, i], bins=15, density=True, color="w", edgecolor="b")
    ax.set_xlabel("x"); ax.set_xlim(-100, 100); ax.set_title(f"t = {t_sample[i]:.0f}")
plt.tight_layout(); plt.show()
```

```{python}
# mean(x) and mean(x^2) vs time, against theory
mean_x  = data_bm_1d.mean(axis=0)
mean_x2 = (data_bm_1d**2).mean(axis=0)
plt.plot(t_sample, mean_x,  "o", color="black", label="Mean(x)")
plt.plot(t_sample, mean_x2, "o", color="red",   label="Mean(x^2)")
plt.plot(t_sample, t_sample*0, color="green", label="y = 0 (theory)")
plt.plot(t_sample, t_sample,   color="blue",  label="y = t (theory)")
plt.xlabel("t"); plt.ylabel("Values"); plt.xlim(0, 910); plt.ylim(-20, 1000)
plt.legend(loc="upper left"); plt.show()
```

The box plot shows a mean pinned at zero with a spread that grows with time, and
the histograms confirm an approximately Gaussian shape that widens as $t$
increases. The measured $\langle x^2 \rangle$ tracks the line $y = t$, so
$D \approx 0.5$ as expected, with residual scatter that would shrink with more
walks.

## 6B.3 Gaussian move steps

Nothing forces the steps to be discrete $\pm \Delta x$. Drawing each step from a
Gaussian of mean $0$ and standard deviation $\Delta x$ gives the same large-scale
statistics.

### R implementation {.impl}

```{r}
brownian_gaussian_steps <- function(t.total, dt, dx) {
  t_all <- seq(0, t.total, by = dt)
  n_all <- length(t_all)
  x_all <- numeric(n_all)
  for (i in 1:(n_all - 1)) {
    x_all[i+1] <- x_all[i] + rnorm(1, mean = 0, sd = dx)   # Gaussian step
  }
  return(cbind(t_all, x_all))
}

set.seed(12)
results_bm_gauss <- replicate(nrep, brownian_gaussian_steps(t.total, dt, dx))
bm_gauss_long <- data.frame(t = c(results_bm_gauss[t_ind, 1, ]), x = c(results_bm_gauss[t_ind, 2, ]))
boxplot(x ~ t, data = bm_gauss_long, xlab = "t", ylab = "x", ylim = c(-100, 100))
```

### Python implementation {.impl}

```{python}
def brownian_gaussian_steps(t_total, dt, dx):
    t_all = np.arange(0, t_total + dt, dt)
    n_all = len(t_all)
    x_all = np.zeros(n_all)
    for i in range(1, n_all):
        x_all[i] = x_all[i-1] + np.random.normal(0, dx)      # Gaussian step
    return np.column_stack((t_all, x_all))

np.random.seed(12)
results_bm_gauss = np.array([brownian_gaussian_steps(t_total, dt, dx) for _ in range(nrep)])
data_gauss = np.array([results_bm_gauss[:, k, 1] for k in t_ind]).T
plt.boxplot(data_gauss); plt.xticks(range(1, len(t_ind)+1), t_ind)
plt.xlabel("t"); plt.ylabel("x"); plt.ylim(-100, 100); plt.show()
```

The box plot is statistically the same as before: the step distribution does not
matter for the long-time behavior, only its mean and variance.

## 6B.4 Brownian motion in 2D

The walk generalizes to a plane by drawing an independent Gaussian step for each
of $x$ and $y$ at every time.

### R implementation {.impl}

```{r}
#| fig-width: 5
#| fig-height: 5
#| out-width: "58%"
brownian_2d <- function(t.total, dt, dx) {
  # dx: length-2 vector of step standard deviations in x and y
  t_all <- seq(0, t.total, by = dt)
  n_all <- length(t_all)
  x_all <- matrix(0, nrow = n_all, ncol = 2)   # starts at (0, 0)
  for (i in 1:(n_all - 1)) {
    x_all[i+1,] <- x_all[i,] + rnorm(2, mean = 0, sd = 1) * dx
  }
  return(cbind(t_all, x_all))
}

set.seed(12)
results_bm_2d <- brownian_2d(t.total = 10000, dt = 1, dx = c(1, 1))
par(pty = "s")
plot(results_bm_2d[,2], results_bm_2d[,3], type = "l", col = 2, asp = 1,
     xlab = "x", ylab = "y", xlim = c(-100, 100), ylim = c(-100, 100))
```

### Python implementation {.impl}

```{python}
#| fig-width: 5
#| fig-height: 5
#| out-width: "58%"
def brownian_2d(t_total, dt, dx):
    # dx: length-2 array of step standard deviations in x and y
    t_all = np.arange(0, t_total + dt, dt)
    n_all = len(t_all)
    x_all = np.zeros((n_all, 2))                 # starts at (0, 0)
    for i in range(1, n_all):
        x_all[i] = x_all[i-1] + np.random.normal(0, 1, size=2) * dx
    return np.column_stack((t_all, x_all))

np.random.seed(12)
results_bm_2d = brownian_2d(10000, 1, np.array([1, 1]))
plt.plot(results_bm_2d[:, 1], results_bm_2d[:, 2], color="red", linewidth=0.4)
plt.gca().set_aspect("equal")
plt.xlabel("x"); plt.ylabel("y"); plt.xlim(-100, 100); plt.ylim(-100, 100); plt.show()
```

The path is a tangled two-dimensional walk that drifts away from the origin over
time, its typical distance again growing like $\sqrt{t}$.

## 6B.5 The Wiener process

A precise mathematical description of Brownian motion is the *Wiener process*
$W_t$, a continuous-time process with four defining properties
[@gardiner2009stochastic]:

**(1)** $W_0 = 0$.

**(2)** Independent increments: for $t > 0$ and $h \geq 0$, the increment
$W_{t+h} - W_t$ is independent of the past $\{W_s : s < t\}$.

**(3)** Gaussian increments: $W_{t+h} - W_t \sim \mathcal{N}(0, h)$.

**(4)** Continuous paths: $W_t$ is continuous in $t$.

Our Gaussian-step walk is essentially a Wiener process. To make the diffusion
constant explicit, we scale the step standard deviation as $\sqrt{2 D \Delta t}$,
so that the variance grows as $2 D t$ regardless of the step size. This scaling is
essential: a common mistake is to set the standard deviation to $\Delta t$.

### R implementation {.impl}

```{r}
brownian_wiener <- function(t.total, dt, D) {
  t_all <- seq(0, t.total, by = dt)
  n_all <- length(t_all)
  x_all <- numeric(n_all)
  step_sd <- sqrt(2 * D * dt)                    # critical scaling
  for (i in 1:(n_all - 1)) {
    x_all[i+1] <- x_all[i] + rnorm(1, mean = 0, sd = step_sd)
  }
  return(cbind(t_all, x_all))
}

set.seed(12)
results_wiener <- replicate(nrep, brownian_wiener(t.total = 1000, dt = 10, D = 0.5))
tw_ind <- seq(1, 100, by = 10)
tw_sample <- results_wiener[tw_ind, 1, 1]
wiener_long <- data.frame(t = c(results_wiener[tw_ind, 1, ]), x = c(results_wiener[tw_ind, 2, ]))
boxplot(x ~ t, data = wiener_long, xlab = "t", ylab = "x", ylim = c(-100, 100))
```

### Python implementation {.impl}

```{python}
def brownian_wiener(t_total, dt, D):
    t_all = np.arange(0, t_total + dt, dt)
    n_all = len(t_all)
    x_all = np.zeros(n_all)
    step_sd = np.sqrt(2 * D * dt)                # critical scaling
    for i in range(1, n_all):
        x_all[i] = x_all[i-1] + np.random.normal(0, step_sd)
    return np.column_stack((t_all, x_all))

np.random.seed(12)
results_wiener = np.array([brownian_wiener(1000, 10, 0.5) for _ in range(nrep)])
tw_ind = np.arange(0, 100, 10)
tw_sample = results_wiener[0, tw_ind, 0]
data_wiener = np.array([results_wiener[:, k, 1] for k in tw_ind]).T
plt.boxplot(data_wiener); plt.xticks(range(1, len(tw_ind)+1), tw_sample.astype(int))
plt.xlabel("t"); plt.ylabel("x"); plt.ylim(-100, 100); plt.show()
```

Even with the much larger step $\Delta t = 10$, the statistics come out right,
because the $\sqrt{2 D \Delta t}$ scaling keeps the variance on track. A large step
is fine for pure diffusion, but once a deterministic force is added the step must
again be small, just as for an ODE. That combination of deterministic drift and
random diffusion is the stochastic differential equation we take up in
[Part 6C](06c-sde-integrators.qmd).
