---
title: "6C. SDE integrators"
author: "Mingyang Lu"
old_id: "06C"
---

The random walk of [Part 6B](06b-brownian-motion.qmd) had $\langle x \rangle = 0$
and $\langle x^2 \rangle \sim t$, statistics captured by the Wiener process. That
framework generalizes to many other stochastic processes through *stochastic
differential equations* (SDEs) [@gardiner2009stochastic],

$$ dX = f(X)\,dt + \sqrt{2 D(X)}\,dW , \tag{1} $$

where $W$ is a Wiener process, $f(X)$ is the drift (the same derivative function as
in an ODE), and $D(X)$ is a state-dependent diffusion coefficient that sets the
noise strength. Physics often writes the same process as a *Langevin equation*,

$$ \frac{dX}{dt} = f(X) + \sqrt{2 D(X)}\,\eta(t) , \tag{2} $$

with a noise term $\eta(t)$. The usual choice is Gaussian white noise, which is
uncorrelated in time and has equal intensity at all frequencies,

$$
\begin{cases}
\langle \eta(t) \rangle = 0 \\
\langle \eta(t) \eta(t') \rangle = \delta(t - t')
\end{cases} \tag{3}
$$

where $\langle \cdot \rangle$ denotes an average and $\delta$ is the Dirac delta
function. This white noise is the continuous-time counterpart of the Wiener
increments.

## 6C.1 The Euler-Maruyama method

The Euler-Maruyama method extends the Euler method for ODEs to SDEs
[@kloeden1992numerical]. Solving Equation (1) step by step,

$$ X_{n+1} = X_n + f(X_n)\,\Delta t + \sqrt{2 D(X_n)}\,\Delta W_n , \tag{4} $$

where each $\Delta W_n$ is a normal random variable of mean zero and variance
$\Delta t$ (so $\langle \Delta W_n^2 \rangle = \Delta t$). A common beginner's
mistake is to fold $\sqrt{2 D}$ into the drift and use the plain ODE Euler step:
that is wrong, because it gives the noise a variance of $\Delta t^2$ rather than
$\Delta t$. We test the method first on pure Brownian motion, $f(X) = 0$.

### R implementation {.impl}

```{r}
# Euler-Maruyama method for a one-variable SDE
SDE_Euler <- function(derivs, noise, t0, X0, t.total, dt, ...){
  # derivs: drift f(t, X, ...); noise: s(t, X, ...) = sqrt(2 D)
  t_all <- seq(t0, t.total, by = dt)
  n_all <- length(t_all)
  X_all <- numeric(n_all); X_all[1] <- X0
  dW_all <- rnorm(n_all, mean = 0, sd = sqrt(dt))   # all Wiener increments at once
  for (i in 1:(n_all - 1)) {
    X_all[i+1] <- X_all[i] + dt * derivs(t_all[i], X_all[i], ...) +
                  dW_all[i] * noise(t_all[i], X_all[i], ...)
  }
  return(cbind(t_all, X_all))
}

# pure Brownian motion: zero drift, constant noise
f_bm <- function(t, X, D) 0
noise_bm <- function(t, X, D) sqrt(2 * D)

set.seed(1)
plot(NULL, xlab = "t", ylab = "X", xlim = c(0, 200), ylim = c(-150, 150))
for (i in 1:10) {
  res <- SDE_Euler(f_bm, noise_bm, t0 = 0, X0 = 0, t.total = 200, dt = 0.01, D = 10)
  lines(res[,1], res[,2], col = i)
}
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

# Euler-Maruyama method for a one-variable SDE
def SDE_Euler(derivs, noise, t0, X0, t_total, dt, **kwargs):
    # derivs: drift f(t, X, ...); noise: s(t, X, ...) = sqrt(2 D)
    t_all = np.arange(t0, t_total + dt, dt)
    n_all = len(t_all)
    X_all = np.zeros(n_all); X_all[0] = X0
    dW_all = np.random.normal(0, np.sqrt(dt), size=n_all - 1)   # all Wiener increments at once
    for i in range(n_all - 1):
        X_all[i+1] = (X_all[i] + dt * derivs(t_all[i], X_all[i], **kwargs)
                      + dW_all[i] * noise(t_all[i], X_all[i], **kwargs))
    return np.column_stack((t_all, X_all))

# pure Brownian motion: zero drift, constant noise
def f_bm(t, X, D): return 0
def noise_bm(t, X, D): return np.sqrt(2 * D)

np.random.seed(1)
plt.xlabel("t"); plt.ylabel("X"); plt.xlim(0, 200); plt.ylim(-150, 150)
for i in range(10):
    res = SDE_Euler(f_bm, noise_bm, t0=0, X0=0, t_total=200, dt=0.01, D=10)
    plt.plot(res[:, 0], res[:, 1], linewidth=1)
plt.show()
```

The trajectories reproduce the free Brownian motion of the previous chapter. We
generate all Wiener increments in one call, which is fast but memory-hungry for
long runs; a compiled Fortran or C routine (as in
[Part 6D](06d-stochastic-transitions.qmd)) is the better choice there.

## 6C.2 The Ornstein-Uhlenbeck process

Adding a restoring drift $f(X) = -kX$ (a particle tethered to a spring, by Hooke's
law) with constant noise gives the Ornstein-Uhlenbeck process
[@uhlenbeck1930theory],

$$ dX = -kX\,dt + \sqrt{2 D}\,dW . \tag{5} $$

Unlike free Brownian motion, $X$ now fluctuates around zero with a variance set by
the balance of drift and noise, $\langle x^2 \rangle = D/k$. We run three noise
levels and check the variance against this prediction.

### R implementation {.impl}

```{r}
f_ou <- function(t, X, k, D) -k * X
noise_ou <- function(t, X, k, D) sqrt(2 * D)

set.seed(1)
D_all <- c(100, 25, 1)
res_ou <- lapply(D_all, function(D)
  SDE_Euler(f_ou, noise_ou, t0 = 0, X0 = 0, t.total = 1000, dt = 0.01, k = 1, D = D))

plot(NULL, xlab = "t", ylab = "X", xlim = c(0, 100), ylim = c(-30, 30))
for (j in 1:3) lines(res_ou[[j]][,1], res_ou[[j]][,2], col = j + 1)
legend("bottomright", inset = 0.02, title = "D", legend = D_all, col = 2:4, lty = 1, cex = 0.8)
```

```{r}
# variance of X vs noise level D (theory: var = D / k, here k = 1)
var_ou <- sapply(res_ou, function(r) var(r[,2]))
plot(D_all, var_ou, xlab = "D", ylab = "Variance")
abline(a = 0, b = 1)
```

### Python implementation {.impl}

```{python}
def f_ou(t, X, k, D): return -k * X
def noise_ou(t, X, k, D): return np.sqrt(2 * D)

np.random.seed(1)
D_all = [100, 25, 1]
res_ou = [SDE_Euler(f_ou, noise_ou, t0=0, X0=0, t_total=1000, dt=0.01, k=1, D=D) for D in D_all]

plt.xlabel("t"); plt.ylabel("X"); plt.xlim(0, 100); plt.ylim(-30, 30)
for r, D, c in zip(res_ou, D_all, ["red", "green", "blue"]):
    plt.plot(r[:, 0], r[:, 1], color=c, label=f"{D}")
plt.legend(title="D", loc="lower right", fontsize="small"); plt.show()
```

```{python}
# variance of X vs noise level D (theory: var = D / k, here k = 1)
var_ou = [np.var(r[:, 1]) for r in res_ou]
plt.scatter(D_all, var_ou, color="black")
plt.plot(D_all, D_all, color="black")
plt.xlabel("D"); plt.ylabel("Variance"); plt.show()
```

The larger the diffusion $D$, the noisier the trajectory. The measured variances
fall close to the line $\langle x^2 \rangle = D$ (since $k = 1$), with small
deviations that would shrink for longer runs.

## 6C.3 The Milstein method

The Milstein method is a more accurate SDE integrator [@kloeden1992numerical]. For
an SDE written as

$$ dX = f(X)\,dt + s(X)\,dW , \tag{6} $$

it adds a correction term derived from a stochastic Taylor expansion,

$$ X_{n+1} = X_n + f(X_n)\,\Delta t + s(X_n)\,\Delta W_n + \tfrac{1}{2} s(X_n) s'(X_n) \left( \Delta W_n^2 - \Delta t \right) , \tag{7} $$

where $s'$ is the derivative of $s(X)$ with respect to $X$ and $s(X) = \sqrt{2 D(X)}$.
When $D$ is constant, $s' = 0$ and the method reduces to Euler-Maruyama; the extra
term only matters, and improves convergence, when the noise depends on the state.

### R implementation {.impl}

```{r}
# Milstein method for a one-variable SDE
SDE_Milstein <- function(derivs, noise, dnoise, t0, X0, t.total, dt, ...){
  # dnoise: derivative of the noise term s(X) with respect to X
  t_all <- seq(t0, t.total, by = dt)
  n_all <- length(t_all)
  X_all <- numeric(n_all); X_all[1] <- X0
  dW_all <- rnorm(n_all, mean = 0, sd = sqrt(dt))
  for (i in 1:(n_all - 1)) {
    X <- X_all[i]; s <- noise(t_all[i], X, ...); dW <- dW_all[i]
    X_all[i+1] <- X + dt * derivs(t_all[i], X, ...) + dW * s +
                  0.5 * s * dnoise(t_all[i], X, ...) * (dW^2 - dt)
  }
  return(cbind(t_all, X_all))
}
```

### Python implementation {.impl}

```{python}
# Milstein method for a one-variable SDE
def SDE_Milstein(derivs, noise, dnoise, t0, X0, t_total, dt, **kwargs):
    # dnoise: derivative of the noise term s(X) with respect to X
    t_all = np.arange(t0, t_total + dt, dt)
    n_all = len(t_all)
    X_all = np.zeros(n_all); X_all[0] = X0
    dW_all = np.random.normal(0, np.sqrt(dt), size=n_all - 1)
    for i in range(n_all - 1):
        X = X_all[i]; s = noise(t_all[i], X, **kwargs); dW = dW_all[i]
        X_all[i+1] = (X + dt * derivs(t_all[i], X, **kwargs) + dW * s
                      + 0.5 * s * dnoise(t_all[i], X, **kwargs) * (dW**2 - dt))
    return np.column_stack((t_all, X_all))
```

## 6C.4 Gene expression noise in a gene circuit

As a biological application, we add noise to the bistable
[self-activating gene](../02-odes/02a-modeling-gene-circuits.qmd) of Part 2, with a
state-dependent noise term $s(X) = b\sqrt{X}$,

$$ dX = \left[ g_0 + g_1 \frac{(X/X_0)^{n_X}}{1 + (X/X_0)^{n_X}} - kX \right] dt + b\sqrt{X}\,dW . \tag{8} $$

The noise standard deviation grows with $\sqrt{X}$, so it is large at high
expression and vanishes as $X \to 0$. A similar square-root noise underlies the
Cox-Ingersoll-Ross model of interest rates in finance [@cox1985theory]. Because the
noise depends on $X$, we integrate with the Milstein method.

### R implementation {.impl}

```{r}
hill_ex <- function(X, X0, n) { a <- (X/X0)^n; a / (1 + a) }
f_1g   <- function(t, X, b) 10 + 45 * hill_ex(X, 200, 4) - 0.15 * X   # self-activating gene
noise_1g  <- function(t, X, b) b * sqrt(X)
dnoise_1g <- function(t, X, b) b / sqrt(X) / 2

set.seed(1)
b_all <- c(5, 2, 0.5)
res_1g <- lapply(b_all, function(b)
  SDE_Milstein(f_1g, noise_1g, dnoise_1g, t0 = 0, X0 = 300, t.total = 1000, dt = 0.01, b = b))

plot(NULL, xlab = "t", ylab = "X", xlim = c(0, 1000), ylim = c(0, 700))
for (j in 1:3) lines(res_1g[[j]][,1], res_1g[[j]][,2], col = j + 1)
legend("topright", inset = 0.02, title = "b", legend = b_all, col = 2:4, lty = 1, cex = 0.8)
```

### Python implementation {.impl}

```{python}
def hill_ex(X, X0, n):
    a = (X / X0)**n
    return a / (1 + a)
def f_1g(t, X, b): return 10 + 45 * hill_ex(X, 200, 4) - 0.15 * X   # self-activating gene
def noise_1g(t, X, b): return b * np.sqrt(X)
def dnoise_1g(t, X, b): return b / np.sqrt(X) / 2

np.random.seed(1)
b_all = [5, 2, 0.5]
res_1g = [SDE_Milstein(f_1g, noise_1g, dnoise_1g, t0=0, X0=300, t_total=1000, dt=0.01, b=b) for b in b_all]

plt.xlabel("t"); plt.ylabel("X"); plt.xlim(0, 1000); plt.ylim(0, 700)
for r, b, c in zip(res_1g, b_all, ["red", "green", "blue"]):
    plt.plot(r[:, 0], r[:, 1], color=c, label=f"{b}")
plt.legend(title="b", loc="upper right", fontsize="small"); plt.show()
```

The circuit is bistable, with stable states near $X = 100$ and $X = 300$ (its
[effective potential](../02-odes/02d-effective-potential.qmd) has two wells). The
states persist under noise, and the striking new feature is *noise-induced
transitions* between them: at low noise ($b = 0.5$) $X$ just jitters around one
state; at $b = 2$ occasional jumps appear; and at $b = 5$ the circuit hops between
states frequently.

For comparison, the same circuit with a **constant** noise $s(X) = b$ (so
$s' = 0$, and Milstein equals Euler-Maruyama) behaves quite differently.

### R implementation {.impl}

```{r}
noise_1g_c  <- function(t, X, b) b
dnoise_1g_c <- function(t, X, b) 0

set.seed(1)
b_all <- c(30, 20, 10)
res_1gc <- lapply(b_all, function(b)
  SDE_Milstein(f_1g, noise_1g_c, dnoise_1g_c, t0 = 0, X0 = 300, t.total = 1000, dt = 0.1, b = b))

plot(NULL, xlab = "t", ylab = "X", xlim = c(0, 1000), ylim = c(0, 700))
for (j in 1:3) lines(res_1gc[[j]][,1], res_1gc[[j]][,2], col = j + 1)
legend("topright", inset = 0.02, title = "b", legend = b_all, col = 2:4, lty = 1, cex = 0.8)
```

### Python implementation {.impl}

```{python}
def noise_1g_c(t, X, b): return b
def dnoise_1g_c(t, X, b): return 0

np.random.seed(1)
b_all = [30, 20, 10]
res_1gc = [SDE_Milstein(f_1g, noise_1g_c, dnoise_1g_c, t0=0, X0=300, t_total=1000, dt=0.1, b=b) for b in b_all]

plt.xlabel("t"); plt.ylabel("X"); plt.xlim(0, 1000); plt.ylim(0, 700)
for r, b, c in zip(res_1gc, b_all, ["red", "green", "blue"]):
    plt.plot(r[:, 0], r[:, 1], color=c, label=f"{b}")
plt.legend(title="b", loc="upper right", fontsize="small"); plt.show()
```

The constant-noise dynamics look different in two ways. With state-independent
noise the low state now fluctuates as much as (indeed more than) the high state,
whereas the $\sqrt{X}$ noise damped fluctuations at low $X$. And at high noise the
low state can be driven below $X = 0$, an artifact of the finite time step that a
smaller $\Delta t$ would reduce.
