---
title: "3C. Modeling a chemostat"
author: "Mingyang Lu"
old_id: "03C"
---

This chapter applies the phase-plane toolkit of [Part 3A](03a-nullclines.qmd) and
[Part 3B](03b-steady-states-2d.qmd) to a different two-variable system, the
classic **chemostat** model of bacterial growth in continuous culture
[@sontag_systemsbio]. It introduces no new methods; the numerical steps are the
same as before (RK4 integration, vector fields, nullclines, steady-state
stability), so the focus is on setting up a new model and reusing that machinery
in both R and Python. Practice questions are embedded throughout.

## 3C.1 Chemostat model

A chemostat is a continuous culture device: fresh medium is pumped in at a
constant rate and an equal volume of culture (cells plus medium) is drawn off, so
the working volume stays fixed. Writing $N$ for the (scaled) bacterial population
and $C$ for the (scaled) limiting-nutrient concentration, the dimensionless model
is [@sontag_systemsbio]

\begin{equation}
\begin{cases}
\dfrac{dN}{dt} = f(N, C) = \alpha_1 \left(\dfrac{C}{C + 1}\right) N - N \\[2mm]
\dfrac{dC}{dt} = g(N, C) = -\left(\dfrac{C}{C + 1}\right) N - C + \alpha_2
\end{cases} \tag{1}
\end{equation}

The $\frac{C}{C+1}$ term is Michaelis-Menten nutrient uptake, $\alpha_1$ sets the
maximum growth rate relative to the washout rate, and $\alpha_2$ is the scaled
inflow nutrient concentration. Setting both derivatives to zero gives two steady
states,

$$ \bar{X}_1 = (0, \alpha_2) \qquad \text{and} \qquad
   \bar{X}_2 = \left(\alpha_1\left(\alpha_2 - \frac{1}{\alpha_1 - 1}\right),\ \frac{1}{\alpha_1 - 1}\right) . \tag{2} $$

$\bar{X}_1$ is the *washout* state (no bacteria); $\bar{X}_2$ is the *coexistence*
state. For both to be physical (non-negative coordinates) we need $\alpha_1 > 1$
and $\alpha_2 > \frac{1}{\alpha_1 - 1}$. Throughout we take $\alpha_1 = 2$ and
$\alpha_2 = 5$, for which $\bar{X}_1 = (0, 5)$ and $\bar{X}_2 = (8, 1)$.

## 3C.2 Simulation and the phase plane

We integrate with the same generic RK4 routine as
[Part 3A](03a-nullclines.qmd), now carrying $\alpha_1$ and $\alpha_2$ as extra
parameters. Starting exactly at $N = 0$ the population cannot grow, so the system
relaxes to the washout state $\bar{X}_1 = (0, 5)$. Seeding an arbitrarily small
population instead ($N(0) = 0.01$) sends the very same dynamics to the
coexistence state $\bar{X}_2 = (8, 1)$: the trajectory first drifts toward
$\bar{X}_1$, then peels away and ends at $\bar{X}_2$. The phase plane below shows
both trajectories.

::: {.callout-note title="Recipe 3C.2.1"}
**Objective.** Simulate a chemostat (population N and substrate) and see it reach either washout or coexistence.

**Method.** The generic multi-variable RK4 from Part 3A, integrating the chemostat ODEs with the feed and dilution parameters.

**Test.** The chemostat model from N(0) = 0 and from a tiny seed N(0) = 0.01.

**Verification.** Starting at N = 0 relaxes to the washout state (0, 5); a tiny seed instead ends at the coexistence state (8, 1), first drifting toward washout then peeling away.
:::

### R implementation {.impl}

```{r}
derivs_chemostat <- function(t, Xs, alpha1, alpha2) {   # dimensionless chemostat
  N <- Xs[1]; C <- Xs[2]
  mm <- N * C / (1 + C)                                 # Michaelis-Menten uptake
  c(alpha1 * mm - N, -mm - C + alpha2)
}

# generic multi-variable RK4 (same as Part 3A)
RK4_generic <- function(derivs, X0, t.total, dt, ...) {
  t_all <- seq(0, t.total, by = dt); n_all <- length(t_all)
  X_all <- matrix(0, n_all, length(X0)); X_all[1, ] <- X0
  for (i in 1:(n_all - 1)) {
    t0 <- t_all[i]
    k1 <- dt * derivs(t0,          X_all[i, ],          ...)
    k2 <- dt * derivs(t0 + dt / 2, X_all[i, ] + k1 / 2, ...)
    k3 <- dt * derivs(t0 + dt / 2, X_all[i, ] + k2 / 2, ...)
    k4 <- dt * derivs(t0 + dt,     X_all[i, ] + k3,     ...)
    X_all[i + 1, ] <- X_all[i, ] + (k1 + 2 * k2 + 2 * k3 + k4) / 6
  }
  cbind(t_all, X_all)
}
```

```{r}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
alpha1 <- 2; alpha2 <- 5
res1 <- RK4_generic(derivs_chemostat, c(0, 0),    20, 0.01, alpha1, alpha2)   # washout
res2 <- RK4_generic(derivs_chemostat, c(0.01, 0), 20, 0.01, alpha1, alpha2)   # tiny seed

plot(NULL, xlab = "N", ylab = "C", xlim = c(0, 10), ylim = c(0, 10), asp = 1)
lines(res1[, 2], res1[, 3], col = 2)   # -> washout state (0, 5)
lines(res2[, 2], res2[, 3], col = 3)   # -> coexistence state (8, 1)
legend("topright", inset = 0.02, legend = c("from (0, 0)", "from (0.01, 0)"),
       col = 2:3, lty = 1, cex = 0.8)
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

def derivs_chemostat(t, Xs, alpha1, alpha2):   # dimensionless chemostat
    N, C = Xs
    mm = N * C / (1 + C)                        # Michaelis-Menten uptake
    return [alpha1 * mm - N, -mm - C + alpha2]

# generic multi-variable RK4 (same as Part 3A)
def RK4_generic(derivs, X0, t_total, dt, **kw):
    t_all = np.arange(0, t_total + dt, dt); n = len(t_all)
    X_all = np.zeros((n, len(X0))); X_all[0] = X0
    for i in range(n - 1):
        t0 = t_all[i]
        k1 = dt * np.array(derivs(t0,          X_all[i],          **kw))
        k2 = dt * np.array(derivs(t0 + dt / 2, X_all[i] + k1 / 2, **kw))
        k3 = dt * np.array(derivs(t0 + dt / 2, X_all[i] + k2 / 2, **kw))
        k4 = dt * np.array(derivs(t0 + dt,     X_all[i] + k3,     **kw))
        X_all[i + 1] = X_all[i] + (k1 + 2 * k2 + 2 * k3 + k4) / 6
    return np.column_stack((t_all, X_all))
```

```{python}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
alpha1, alpha2 = 2.0, 5.0
res1 = RK4_generic(derivs_chemostat, [0, 0],    20, 0.01, alpha1=alpha1, alpha2=alpha2)
res2 = RK4_generic(derivs_chemostat, [0.01, 0], 20, 0.01, alpha1=alpha1, alpha2=alpha2)

plt.plot(res1[:, 1], res1[:, 2], color="red",   label="from (0, 0)")       # -> (0, 5)
plt.plot(res2[:, 1], res2[:, 2], color="green", label="from (0.01, 0)")    # -> (8, 1)
plt.gca().set_aspect("equal"); plt.xlim(0, 10); plt.ylim(0, 10)
plt.xlabel("N"); plt.ylabel("C"); plt.legend(loc="upper right"); plt.show()
```

::: {.callout-note title="Practice Q1"}
Repeat the simulation for the initial conditions (a) $X(0) = \bar{X}_1 = (0, 5)$;
(b) $(0.01, 5)$, slightly off $\bar{X}_1$ in $N$; and (c) $(0, 5.01)$, slightly
off in $C$. Which perturbations decay back to $\bar{X}_1$ and which grow away
toward $\bar{X}_2$? What does that already tell you about the stability of
$\bar{X}_1$?
:::

::: {.callout-note title="Practice Q2"}
Vary $\alpha_1$ and $\alpha_2$ and confirm that the coexistence state
$\bar{X}_2$ follows Equation (2). For which parameter values do both steady
states stay physical (non-negative)?
:::

::: {.callout-note title="Practice Q3"}
Let $Z(t) = N(t) + \alpha_1 C(t) - \alpha_1 \alpha_2$. Show analytically that
$Z(t) = Z(0)\, e^{-t}$, so $Z \to 0$ from any start. Verify it in a simulation,
and show that the line $Z(N, C) = 0$ passes through both $\bar{X}_1$ and
$\bar{X}_2$. This is one way to prove that $\bar{X}_2$ is the global attractor of
the chemostat.
:::

## 3C.3 Vector field and nullclines

Drawing the unit vector field over the phase plane shows the flow at a glance,
exactly as in [Part 3A](03a-nullclines.qmd). Overlaying the two nullclines
locates the steady states at their crossings. The chemostat's *N-nullcline*
($dN/dt = 0$) factors into two straight pieces, $N = 0$ and
$C = \frac{1}{\alpha_1 - 1}$, while the *C-nullcline* ($dC/dt = 0$) solves in
closed form for $N$ as a function of $C$,

$$ N(C) = \frac{\alpha_2}{C} + \alpha_2 - 1 - C . \tag{3} $$

Sweeping $C$ traces it directly. When separation of variables is not available,
the same nullcline can be found by root finding: for each fixed $N$, solve
$g(N, C) = 0$ for $C$. Both are shown below; the contour and continuation
methods of Part 3A apply here too and are left as practice.

### R implementation {.impl}

```{r}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
#| message: false
#| warning: false
library(ggplot2)

# unit vector field on a grid (arrows normalized to a fixed length)
grid <- expand.grid(N = seq(0, 10, 0.5), C = seq(0, 10, 0.5))
vfu <- t(apply(grid, 1, function(Xs) {
  v <- derivs_chemostat(0, Xs, alpha1, alpha2)
  c(Xs, v / sqrt(sum(v^2)))
}))
vfu <- as.data.frame(vfu); colnames(vfu) <- c("N", "C", "dN", "dC")
p_field <- ggplot(vfu, aes(N, C)) +
  geom_segment(aes(xend = N + 0.3 * dN, yend = C + 0.3 * dC),
               arrow = arrow(length = unit(0.05, "in"))) +
  coord_fixed()   # equal N/C scale: a square phase plane

# nullclines by separation of variables
C_grid <- seq(0.05, 10, 0.05); N_grid <- seq(0, 10, 0.05)
nullN_1 <- data.frame(N = 0, C = C_grid)                     # dN/dt=0: N = 0
nullN_2 <- data.frame(N = N_grid, C = 1 / (alpha1 - 1))      # dN/dt=0: C = 1/(alpha1-1)
nullC   <- data.frame(N = alpha2 / C_grid + alpha2 - 1 - C_grid, C = C_grid)   # dC/dt=0
nullC   <- nullC[nullC$N >= 0 & nullC$N <= 10, ]             # keep within the plotted range

p_field +
  geom_path(data = nullN_1, aes(N, C, colour = "dN/dt=0"), linewidth = 1) +
  geom_path(data = nullN_2, aes(N, C, colour = "dN/dt=0"), linewidth = 1) +
  geom_path(data = nullC,   aes(N, C, colour = "dC/dt=0"), linewidth = 1) +
  labs(colour = NULL) +
  theme(legend.position = c(0.99, 0.99), legend.justification = c(1, 1),
        legend.background = element_rect(fill = "white", colour = "grey80"))
```

```{r}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
#| message: false
# C-nullcline by root finding: solve g(N, C) = 0 for C at each N (rootSolve)
library(rootSolve)
g_C <- function(C, N, alpha2) -N * C / (1 + C) - C + alpha2

nullC_root <- do.call(rbind, lapply(seq(0, 10, 0.1), function(N) {
  cbind(N = N, C = uniroot.all(g_C, interval = c(0, 10), N = N, alpha2 = alpha2))
}))

p_field +
  geom_path(data = as.data.frame(nullC_root), aes(N, C, colour = "dC/dt=0"), linewidth = 1) +
  labs(colour = NULL) +
  theme(legend.position = c(0.99, 0.99), legend.justification = c(1, 1),
        legend.background = element_rect(fill = "white", colour = "grey80"))
```

### Python implementation {.impl}

```{python}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
# unit vector field on a grid (arrows normalized to a fixed length)
N_ax = np.arange(0, 10.01, 0.5); C_ax = np.arange(0, 10.01, 0.5)
grid = np.array(np.meshgrid(N_ax, C_ax)).T.reshape(-1, 2)

def unit_arrow(N, C, scale=0.3):
    v = np.array(derivs_chemostat(0, (N, C), alpha1, alpha2))
    n = np.sqrt(v @ v)
    return scale * v / n if n > 0 else np.zeros(2)

for N, C in grid:
    dN, dC = unit_arrow(N, C)
    plt.arrow(N, C, dN, dC, head_width=0.12, head_length=0.12, ec="black")

# nullclines by separation of variables
C_grid = np.arange(0.05, 10.01, 0.05); N_grid = np.arange(0, 10.01, 0.05)
plt.plot(np.zeros_like(C_grid), C_grid, color="red", lw=2, label="dN/dt=0")   # N = 0
plt.plot(N_grid, np.full_like(N_grid, 1 / (alpha1 - 1)), color="red", lw=2)   # C = 1/(alpha1-1)
nullC_N = alpha2 / C_grid + alpha2 - 1 - C_grid                               # dC/dt=0, Eq (3)
keep = (nullC_N >= 0) & (nullC_N <= 10)
plt.plot(nullC_N[keep], C_grid[keep], color="blue", lw=2, label="dC/dt=0")

plt.gca().set_aspect("equal"); plt.xlim(0, 10); plt.ylim(0, 10)
plt.xlabel("N"); plt.ylabel("C"); plt.legend(loc="upper right"); plt.show()
```

```{python}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
# C-nullcline by root finding: solve g(N, C) = 0 for C at each N.
# g(N, C) = 0 has a single positive root per N, so a bracketing solver (brentq)
# suffices; R's rootSolve::uniroot.all would instead return every root in [0, 10].
from scipy.optimize import brentq
def g_C(C, N, alpha2):
    return -N * C / (1 + C) - C + alpha2

N_grid = np.arange(0, 10.01, 0.1)
nullC_root = np.array([[N, brentq(g_C, 0.01, 10, args=(N, alpha2))] for N in N_grid])

plt.plot(nullC_root[:, 0], nullC_root[:, 1], color="blue", lw=2, label="dC/dt=0")
plt.gca().set_aspect("equal"); plt.xlim(0, 10); plt.ylim(0, 10)
plt.xlabel("N"); plt.ylabel("C"); plt.legend(loc="upper right"); plt.show()
```

::: {.callout-note title="Practice Q4"}
Compute the C-nullcline by ODE relaxation instead of root finding: for each fixed
$N$, treat $\frac{dC}{dt} = g(N, C)$ as a one-variable ODE, start from $(N, 0)$,
and integrate to steady state to get $C$. Sweep $N$ to trace the whole nullcline.
How does this compare to root finding in robustness and generality?
:::

::: {.callout-note title="Practice Q5"}
A related, warm-started sweep: begin at $(0, 0)$, relax $C$ to steady state at
fixed $N$, record $(N, C)$, then step $N \to N + \Delta N$ using the previous $C$
as the new starting point, and repeat up to $N = 15$. Implement it and compare
with Q4.
:::

::: {.callout-note title="Practice Q6"}
Implement arc-length numerical continuation for the C-nullcline, as in
[Part 3A](03a-nullclines.qmd) and [Part 2F](../02-odes/02f-bifurcation-curves.qmd).
Parameterize the curve by arc length $s$ so the tracing survives points where
$\frac{dC}{dN} \to \infty$, with a Newton correction to stay on $g(N, C) = 0$.
:::

## 3C.4 Steady states and stability

The steady states are the intersections of the two nullclines, and their
stability follows from the eigenvalues of the Jacobian. Both the line-intersection
routines (`find_intersection`, `find_intersection_all_fast`) and the numerical
Jacobian classifier (`stability_2D`) are exactly the ones built in
[Part 3B](03b-steady-states-2d.qmd), so rather than repeat them we apply them
directly to the chemostat's nullclines.

Doing so returns the two steady states of Equation (2). The Jacobian eigenvalues
classify $\bar{X}_2 = (8, 1)$ as a **stable node** and $\bar{X}_1 = (0, 5)$ as a
**saddle**. This is consistent with the simulations of 3C.2, where a trajectory
started at the washout state $\bar{X}_1$ was driven away by the smallest seed
population and settled on the coexistence state $\bar{X}_2$.

::: {.callout-note title="Practice Q7"}
The N-nullcline is stored as two separate line pieces ($N = 0$ and
$C = \frac{1}{\alpha_1 - 1}$), so running the all-pairs intersection search of
[Part 3B](03b-steady-states-2d.qmd) can report the same steady state more than
once. Why does that happen, and how would you deduplicate the results? Then apply
`find_intersection_all_fast` and `stability_2D` to reproduce the classification
above.
:::
