---
title: "3E. Bifurcation for two-variable systems"
author: "Mingyang Lu"
old_id: "03E"
---

[Part 3A](03a-nullclines.qmd) and [Part 3B](03b-steady-states-2d.qmd) built the
tools to locate a two-variable system's steady states and classify their stability
at a fixed set of parameters. This chapter puts them to work to trace a
**bifurcation diagram**: how the steady states, and their stability, change as a
control parameter varies. As in [Part 2E](../02-odes/02e-bifurcation.qmd) for
one-variable systems, a qualitative change in the number or stability of steady
states is a bifurcation [@strogatz2015nonlinear]. Here two classic types appear in
the same two-gene toggle switch [@gardner2000toggle], depending on which parameter
we sweep.

We use a slightly modified toggle switch with two variables $X$ and $Y$ and two
tunable parameters, the maximum production rate $g_X$ and the common degradation
rate $k$,

\begin{equation}
\begin{cases}
\dfrac{dX}{dt} = f_X(X, Y, g_X, k) = 5 + g_X\dfrac{1}{1 + (Y/100)^4} - kX \\[2mm]
\dfrac{dY}{dt} = f_Y(X, Y, g_X, k) = 5 + 50\dfrac{1}{1 + (X/100)^4} - kY
\end{cases} \tag{1}
\end{equation}

At $g_X = 50$ the two genes have identical parameters, so the circuit is symmetric
under swapping $X \leftrightarrow Y$; moving $g_X$ away from $50$ breaks that
symmetry. We build a bifurcation diagram by varying either $g_X$ or $k$.

## 3E.1 The bifurcation scan

The recipe reuses the whole Part 3A/3B pipeline at each parameter value: (1) build
the two nullclines by separation of variables (Part 3A), (2) find their
intersections, the steady states (Part 3B), and (3) classify each by the
eigenvalues of its Jacobian (Part 3B). Sweeping the control parameter and plotting
the steady-state $X$ against it, colored by stability, traces the diagram. The
function `find_steady_states` packages the three steps; `find_intersection`,
`find_intersection_all_fast`, and `stability_2D` are exactly the routines from
Part 3B.

::: {.callout-note title="Recipe 3E.1.1"}
**Objective.** Trace a bifurcation diagram: the steady states of the toggle switch versus a control parameter, colored by stability.

**Method.** At each parameter value, build the nullclines (separation of variables), find their intersections (steady states), and classify each by its Jacobian eigenvalues; sweep the parameter and plot steady-state X against it.

**Test.** The toggle switch, sweeping a control parameter across a range.

**Verification.** Stable and unstable branches appear and merge at bifurcation points, so the number and stability of steady states change with the parameter.
:::

### R implementation {.impl}

```{r}
# modified toggle switch, with gX and k as tunable parameters
fx <- function(X, Y, gX, k) 5 + gX / (1 + (Y / 100)^4) - k * X   # dX/dt
fy <- function(X, Y, gX, k) 5 + 50 / (1 + (X / 100)^4) - k * Y   # dY/dt
derivs_ts <- function(t, Xs, gX, k) c(fx(Xs[1], Xs[2], gX, k), fy(Xs[1], Xs[2], gX, k))
```

```{r}
# nullclines by separation of variables (as in Part 3A), as point series within range
null_fx <- function(gX, k, X_range, Y_range) {                    # dX/dt = 0
  Y_all <- seq(Y_range[1], Y_range[2], length.out = 1000)
  X_all <- (5 + gX / (1 + (Y_all / 100)^4)) / k
  keep <- (X_all - X_range[1]) * (X_all - X_range[2]) <= 0
  cbind(X = X_all[keep], Y = Y_all[keep])
}
null_fy <- function(gX, k, X_range, Y_range) {                    # dY/dt = 0
  X_all <- seq(X_range[1], X_range[2], length.out = 1000)
  Y_all <- (5 + 50 / (1 + (X_all / 100)^4)) / k
  keep <- (Y_all - Y_range[1]) * (Y_all - Y_range[2]) <= 0
  cbind(X = X_all[keep], Y = Y_all[keep])
}
```

```{r}
# line-segment intersection finders (same as Part 3B)
find_intersection <- function(lineA, lineB, nA, nB) {
  dAX <- lineA[nA + 1, 1] - lineA[nA, 1]; dAY <- lineA[nA + 1, 2] - lineA[nA, 2]
  dBX <- lineB[nB + 1, 1] - lineB[nB, 1]; dBY <- lineB[nB + 1, 2] - lineB[nB, 2]
  dABX <- lineB[nB, 1] - lineA[nA, 1];    dABY <- lineB[nB, 2] - lineA[nA, 2]
  d <- dAX * dBY - dAY * dBX
  alpha <- (dABX * dBY - dABY * dBX) / d; beta <- (dABX * dAY - dABY * dAX) / d
  if (alpha * (1 - alpha) >= 0 && beta * (1 - beta) >= 0)
    c((1 - alpha) * lineA[nA, 1] + alpha * lineA[nA + 1, 1],
      (1 - alpha) * lineA[nA, 2] + alpha * lineA[nA + 1, 2], 1)
  else c(alpha, beta, 0)
}

find_intersection_all_fast <- function(lineA, lineB, dxdt, dydt) {
  small <- 1e-3
  lineA_ind <- which(dydt * c(dydt[-1], NA) <= small)
  lineB_ind <- which(dxdt * c(dxdt[-1], NA) <= small)
  lines_all <- expand.grid(lineA_ind, lineB_ind)
  results <- apply(lines_all, 1, function(i) find_intersection(lineA, lineB, i[1], i[2]))
  t(results[1:2, results[3, ] == 1, drop = FALSE])
}
```

```{r}
# stability from the numerical Jacobian eigenvalues (same as Part 3B)
stability_2D <- function(derivs, ss, ...) {
  delta <- 0.001
  f0  <- derivs(0, ss, ...)
  fdx <- derivs(0, ss + c(delta, 0), ...)
  fdy <- derivs(0, ss + c(0, delta), ...)
  jacobian <- matrix(c((fdx[1] - f0[1]) / delta, (fdx[2] - f0[2]) / delta,
                       (fdy[1] - f0[1]) / delta, (fdy[2] - f0[2]) / delta), 2)
  lambda <- eigen(jacobian)$values
  if (is.complex(lambda)) { if (Re(lambda[1]) < 0) 4 else 5 }
  else if (lambda[1] < 0 && lambda[2] < 0) 1
  else if (lambda[1] >= 0 && lambda[2] >= 0) 2 else 3
}
```

```{r}
# steady states + stability at one parameter setting
find_steady_states <- function(gX, k, X_range, Y_range) {
  line_1 <- null_fx(gX, k, X_range, Y_range)     # dX/dt = 0 nullcline
  line_2 <- null_fy(gX, k, X_range, Y_range)     # dY/dt = 0 nullcline
  dxdt <- fx(line_2[, 1], line_2[, 2], gX, k)    # dX/dt along the dY/dt = 0 nullcline
  dydt <- fy(line_1[, 1], line_1[, 2], gX, k)    # dY/dt along the dX/dt = 0 nullcline
  ss <- find_intersection_all_fast(line_1, line_2, dxdt, dydt)
  t(apply(ss, 1, function(s) c(s, stability_2D(derivs_ts, s, gX, k))))
}
```

### Python implementation {.impl}

```{python}
import numpy as np

# modified toggle switch, with gX and k as tunable parameters
def fx(X, Y, gX, k): return 5 + gX / (1 + (Y / 100)**4) - k * X   # dX/dt
def fy(X, Y, gX, k): return 5 + 50 / (1 + (X / 100)**4) - k * Y   # dY/dt
def derivs_ts(t, Xs, gX, k): return [fx(Xs[0], Xs[1], gX, k), fy(Xs[0], Xs[1], gX, k)]
```

```{python}
# nullclines by separation of variables (as in Part 3A), as point series within range
def null_fx(gX, k, X_range, Y_range):                            # dX/dt = 0
    Y_all = np.linspace(Y_range[0], Y_range[1], 1000)
    X_all = (5 + gX / (1 + (Y_all / 100)**4)) / k
    keep = (X_all - X_range[0]) * (X_all - X_range[1]) <= 0
    return np.column_stack((X_all[keep], Y_all[keep]))
def null_fy(gX, k, X_range, Y_range):                            # dY/dt = 0
    X_all = np.linspace(X_range[0], X_range[1], 1000)
    Y_all = (5 + 50 / (1 + (X_all / 100)**4)) / k
    keep = (Y_all - Y_range[0]) * (Y_all - Y_range[1]) <= 0
    return np.column_stack((X_all[keep], Y_all[keep]))
```

```{python}
# line-segment intersection finders (same as Part 3B)
def find_intersection(lineA, lineB, nA, nB):
    dAX = lineA[nA+1,0]-lineA[nA,0]; dAY = lineA[nA+1,1]-lineA[nA,1]
    dBX = lineB[nB+1,0]-lineB[nB,0]; dBY = lineB[nB+1,1]-lineB[nB,1]
    dABX = lineB[nB,0]-lineA[nA,0];  dABY = lineB[nB,1]-lineA[nA,1]
    d = dAX*dBY - dAY*dBX
    alpha = (dABX*dBY - dABY*dBX)/d; beta = (dABX*dAY - dABY*dAX)/d
    if (0 <= alpha <= 1) and (0 <= beta <= 1):
        return np.array([(1-alpha)*lineA[nA,0]+alpha*lineA[nA+1,0],
                         (1-alpha)*lineA[nA,1]+alpha*lineA[nA+1,1], 1])
    return np.array([alpha, beta, 0])

def find_intersection_all_fast(lineA, lineB, dxdt, dydt):
    small = 1e-3
    lineA_ind = np.where(dydt * np.append(dydt[1:], np.nan) <= small)[0]
    lineB_ind = np.where(dxdt * np.append(dxdt[1:], np.nan) <= small)[0]
    lines_all = np.array(np.meshgrid(lineA_ind, lineB_ind)).T.reshape(-1, 2)
    if len(lines_all) == 0:
        return np.empty((0, 2))
    results = np.array([find_intersection(lineA, lineB, i, j) for i, j in lines_all])
    return results[results[:, 2] == 1, :2]
```

```{python}
# stability from the numerical Jacobian eigenvalues (same as Part 3B)
def stability_2D(derivs, ss, *args):
    delta = 0.001
    f0  = np.array(derivs(0, ss, *args))
    fdx = np.array(derivs(0, ss + np.array([delta, 0]), *args))
    fdy = np.array(derivs(0, ss + np.array([0, delta]), *args))
    J = np.array([[(fdx[0]-f0[0])/delta, (fdy[0]-f0[0])/delta],
                  [(fdx[1]-f0[1])/delta, (fdy[1]-f0[1])/delta]])
    lam = np.linalg.eigvals(J)
    if np.any(np.iscomplex(lam)):
        return 4 if np.real(lam[0]) < 0 else 5
    if lam[0] < 0 and lam[1] < 0:
        return 1
    return 2 if lam[0] >= 0 and lam[1] >= 0 else 3
```

```{python}
# steady states + stability at one parameter setting
def find_steady_states(gX, k, X_range, Y_range):
    line_1 = null_fx(gX, k, X_range, Y_range)      # dX/dt = 0 nullcline
    line_2 = null_fy(gX, k, X_range, Y_range)      # dY/dt = 0 nullcline
    dxdt = fx(line_2[:, 0], line_2[:, 1], gX, k)   # dX/dt along the dY/dt = 0 nullcline
    dydt = fy(line_1[:, 0], line_1[:, 1], gX, k)   # dY/dt along the dX/dt = 0 nullcline
    ss = find_intersection_all_fast(line_1, line_2, dxdt, dydt)
    return np.array([np.append(s, stability_2D(derivs_ts, s, gX, k)) for s in ss])
```

## 3E.2 Saddle-node bifurcation

Sweeping $g_X$ from $0$ to $100$ (at fixed $k = 0.1$) breaks the $X \leftrightarrow Y$
symmetry. The diagram colors each steady state by stability (stable node or
saddle). As $g_X$ increases past a critical value a stable node and a saddle
appear together, so the circuit turns from monostable to bistable. Two branches
being created (or destroyed) at a point in this way is a **saddle-node**
bifurcation.

### R implementation {.impl}

```{r}
X_range <- c(0, 1000); Y_range <- c(0, 1000); k <- 0.1
ss_all <- numeric()
for (gX in seq(0, 100, by = 1)) {
  ss_all <- rbind(ss_all, cbind(gX, find_steady_states(gX, k, X_range, Y_range)))
}
colnames(ss_all) <- c("gX", "X", "Y", "Stability")
plot(ss_all[, "gX"], ss_all[, "X"], col = ss_all[, "Stability"], pch = 16, cex = 0.4,
     xlab = "gX", ylab = "X", xlim = c(0, 100), ylim = c(0, 1000))
legend("topleft", inset = 0.02, legend = c("Stable", "Saddle"), col = c(1, 3), pch = 16, cex = 0.8)
```

### Python implementation {.impl}

```{python}
import matplotlib.pyplot as plt
X_range, Y_range, k = [0, 1000], [0, 1000], 0.1
ss_all = []
for gX in range(0, 101):
    for s in find_steady_states(gX, k, X_range, Y_range):
        ss_all.append([gX, *s])
ss_all = np.array(ss_all)

colors = {1: "black", 3: "green"}   # stability code -> color
for code, lab in [(1, "Stable"), (3, "Saddle")]:
    m = ss_all[:, 3] == code
    plt.scatter(ss_all[m, 0], ss_all[m, 1], color=colors[code], s=8, label=lab)
plt.xlabel("gX"); plt.ylabel("X"); plt.xlim(0, 100); plt.ylim(0, 1000)
plt.legend(loc="upper left"); plt.show()
```

## 3E.3 Pitchfork bifurcation

Now fix $g_X = 50$, where the circuit is symmetric under $X \leftrightarrow Y$,
and sweep the degradation rate $k$. Below a critical value near $k = 0.03$ a
single steady state is in view; above it the diagram splits into three, a central
saddle flanked by two stable branches. Because the two stable states are related
by the $X \leftrightarrow Y$ symmetry, they emerge together as a matched pair when
the symmetric state turns into the saddle. This splitting of one state into a
symmetric pair is a **pitchfork** bifurcation, and it is the symmetry (absent when
$g_X \neq 50$) that makes it a pitchfork rather than a saddle-node.

### R implementation {.impl}

```{r}
X_range <- c(0, 1000); Y_range <- c(0, 1000); gX <- 50
ss_all <- numeric()
for (k in seq(0.01, 0.15, by = 0.001)) {
  ss_all <- rbind(ss_all, cbind(k, find_steady_states(gX, k, X_range, Y_range)))
}
colnames(ss_all) <- c("k", "X", "Y", "Stability")
plot(ss_all[, "k"], ss_all[, "X"], col = ss_all[, "Stability"], pch = 16, cex = 0.4,
     xlab = "k", ylab = "X", xlim = c(0, 0.15), ylim = c(0, 800))
legend("topleft", inset = 0.02, legend = c("Stable", "Saddle"), col = c(1, 3), pch = 16, cex = 0.8)
```

### Python implementation {.impl}

```{python}
X_range, Y_range, gX = [0, 1000], [0, 1000], 50
ss_all = []
for k in np.arange(0.01, 0.151, 0.001):
    for s in find_steady_states(gX, k, X_range, Y_range):
        ss_all.append([k, *s])
ss_all = np.array(ss_all)

for code, lab in [(1, "Stable"), (3, "Saddle")]:
    m = ss_all[:, 3] == code
    plt.scatter(ss_all[m, 0], ss_all[m, 1], color=colors[code], s=8, label=lab)
plt.xlabel("k"); plt.ylabel("X"); plt.xlim(0, 0.15); plt.ylim(0, 800)
plt.legend(loc="upper left"); plt.show()
```

Both diagrams were cheap to compute because the nullclines followed from
separation of variables. When they cannot be written in closed form, the general
tool is numerical continuation, the arc-length predictor-corrector of
[Part 2F](../02-odes/02f-bifurcation-curves.qmd) generalized from one equation in
two variables to $n$ equations in $n + 1$ variables (here $n = 2$). Reproducing
these bifurcation curves by numerical continuation is left as an exercise in
[Part 3I](03i-exercises.qmd).

::: {.callout-note title="Practice Q1"}
Both diagrams come from the same circuit. Varying $k$ keeps the
$X \leftrightarrow Y$ symmetry of Equation (1) at $g_X = 50$ (both genes then have
identical parameters), while varying $g_X$ breaks it. Explain how that difference
produces a pitchfork in one case and a saddle-node in the other. (Hint: a
pitchfork needs a symmetry for the two outer branches to appear as a matched
pair.)
:::
