---
title: "3F. Separatrix"
author: "Mingyang Lu"
old_id: "03F"
---

[Part 3B](03b-steady-states-2d.qmd) classified the toggle switch's three steady
states: two stable nodes and, between them, a saddle. Each stable node has a
**basin of attraction**, the set of initial conditions that flow to it, and the
curve dividing the two basins is the **separatrix** [@strogatz2015nonlinear]. A
trajectory that starts on one side of the separatrix always ends at that side's
attractor. Computing a separatrix is hard in general, but for a saddle it has a
clean description: the separatrix is the saddle's **stable manifold**, the pair of
trajectories that flow *into* the saddle. We recover it with a simple reverse-time
trick.

Reversing time negates every derivative,

$$
\begin{cases}
\dfrac{dX}{dt} = -f_X(X, Y) \\[2mm]
\dfrac{dY}{dt} = -f_Y(X, Y)
\end{cases}
$$

which reverses the direction of every trajectory. Under this reversal the saddle
keeps being a saddle, but its stable manifold becomes an *unstable* manifold: the
trajectories that used to flow into the saddle now flow out of it, along the
separatrix. So we integrate the reversed system from points just beside the saddle
and let them run out along the separatrix.

## 3F.1 The reversed vector field

We first draw the unit vector field of the reversed toggle switch (the same
circuit as [Part 3A](03a-nullclines.qmd), every derivative negated). Near the
saddle the arrows now point *inward* along the line joining the three steady
states and *outward* along the separatrix, the reverse of the original flow.

### R implementation {.impl}

```{r}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
#| message: false
library(ggplot2)
hill_inh <- function(X, X_th, n) 1 / (1 + (X / X_th)^n)      # same as Part 3A

# the toggle switch with every derivative negated: the reverse-time system
derivs_ts_rev <- function(t, Xs) {
  X <- Xs[1]; Y <- Xs[2]
  c(-(5 + 50 * hill_inh(Y, 100, 4) - 0.1 * X),
    -(4 + 40 * hill_inh(X, 150, 4) - 0.12 * Y))
}

grid <- expand.grid(X = seq(0, 600, 20), Y = seq(0, 650, 20))
vfu <- t(apply(grid, 1, function(Xs) {
  v <- derivs_ts_rev(0, Xs); c(Xs, v / sqrt(sum(v^2)))
}))
vfu <- as.data.frame(vfu); colnames(vfu) <- c("X", "Y", "dX", "dY")

p1 <- ggplot(vfu, aes(X, Y)) +
  geom_segment(aes(xend = X + 15 * dX, yend = Y + 15 * dY),
               arrow = arrow(length = unit(0.05, "in"))) +
  coord_fixed(xlim = c(0, 600), ylim = c(0, 650))
p1
```

### Python implementation {.impl}

```{python}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
import numpy as np
import matplotlib.pyplot as plt

def hill_inh(X, X_th, n):                                    # same as Part 3A
    return 1 / (1 + (X / X_th)**n)

# the toggle switch with every derivative negated: the reverse-time system
def derivs_ts_rev(t, Xs):
    X, Y = Xs
    return [-(5 + 50 * hill_inh(Y, 100, 4) - 0.1 * X),
            -(4 + 40 * hill_inh(X, 150, 4) - 0.12 * Y)]

X_all = np.arange(0, 601, 20); Y_all = np.arange(0, 651, 20)
grid = np.array(np.meshgrid(X_all, Y_all)).T.reshape(-1, 2)

def unit_arrow(X, Y, scale=15):
    v = np.array(derivs_ts_rev(0, (X, Y)))
    return scale * v / np.sqrt(v @ v)

for X, Y in grid:
    dX, dY = unit_arrow(X, Y)
    plt.arrow(X, Y, dX, dY, head_width=8, head_length=8, ec="black")
plt.gca().set_aspect("equal")
plt.xlabel("X"); plt.ylabel("Y"); plt.xlim(0, 600); plt.ylim(0, 650); plt.show()
```

## 3F.2 Tracing the separatrix

Now we seed the reversed system with ten points just beside the saddle
(approximately $(190, 127)$, from [Part 3B](03b-steady-states-2d.qmd)) and
integrate: they flow outward along the separatrix, which we draw in red. Overlaid
in blue are trajectories of the *original* system from ten random starts. Each
blue trajectory settles onto one of the two attractors, and none of them crosses
the red separatrix, confirming that it is the boundary between the two basins.

::: {.callout-note title="Recipe 3F.2.1"}
**Objective.** Trace the separatrix dividing the toggle switch's two basins of attraction.

**Model.** Genes X and Y repress each other; each gene's transcription is a basal rate plus a repressive Hill function of the other, with linear degradation: `dX/dt = gX0 + gX1/(1 + (Y/Yth)^nY) - kX*X`, `dY/dt = gY0 + gY1/(1 + (X/Xth)^nX) - kY*Y`.

**Method.** Integrate the time-reversed system `dX/dt = -f(X)` from points just beside the saddle; the reversed flow carries them outward along the separatrix.

**Test.** The toggle switch (`gX0 = 5, gX1 = 50, Yth = 100, nY = 4, kX = 0.1`; `gY0 = 4, gY1 = 40, Xth = 150, nX = 4, kY = 0.12`), seeding ten points near the saddle (about (190, 127)).

**Verification.** The reversed trajectories trace a curve that no forward trajectory from random starts ever crosses, confirming it is the boundary between the two basins.
:::

### R implementation {.impl}

```{r}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
#| warning: false
# RK4 for a generic multi-variable system (same as Part 3A)
RK4_generic <- function(derivs, X0, t.total, dt, ...) {
  t_all <- seq(0, t.total, by = dt); n_all <- length(t_all)
  X_all <- matrix(0, n_all, length(X0)); X_all[1, ] <- X0
  for (i in 1:(n_all - 1)) {
    t0 <- t_all[i]
    k1 <- dt * derivs(t0,          X_all[i, ],          ...)
    k2 <- dt * derivs(t0 + dt / 2, X_all[i, ] + k1 / 2, ...)
    k3 <- dt * derivs(t0 + dt / 2, X_all[i, ] + k2 / 2, ...)
    k4 <- dt * derivs(t0 + dt,     X_all[i, ] + k3,     ...)
    X_all[i + 1, ] <- X_all[i, ] + (k1 + 2 * k2 + 2 * k3 + k4) / 6
  }
  cbind(t_all, X_all)
}

# the original (forward-time) toggle switch
derivs_ts <- function(t, Xs) {
  X <- Xs[1]; Y <- Xs[2]
  c(5 + 50 * hill_inh(Y, 100, 4) - 0.1 * X,
    4 + 40 * hill_inh(X, 150, 4) - 0.12 * Y)
}

set.seed(77)
t.total <- 100; dt <- 0.01
X_range <- c(0, 600); Y_range <- c(0, 650)
saddle <- c(189.96888, 126.64398)          # the saddle from Part 3B

# original system, 10 random starts -> the two attractors (blue)
X_init_1 <- matrix(runif(20, 0, 600), ncol = 2)
for (i in seq_len(nrow(X_init_1))) {
  r <- as.data.frame(RK4_generic(derivs_ts, X_init_1[i, ], t.total, dt))
  colnames(r) <- c("t", "X", "Y")
  p1 <- p1 + geom_path(data = r, aes(X, Y), color = "blue")
}

# reversed system, 10 points beside the saddle -> the separatrix (red)
X_init_2 <- matrix(runif(20, -1, 1), ncol = 2)
for (i in seq_len(nrow(X_init_2))) {
  r <- as.data.frame(RK4_generic(derivs_ts_rev, X_init_2[i, ] + saddle, t.total, dt))
  colnames(r) <- c("t", "X", "Y")
  r <- r[(r$X - X_range[1]) * (r$X - X_range[2]) < 0 &
         (r$Y - Y_range[1]) * (r$Y - Y_range[2]) < 0, ]   # keep points in view
  p1 <- p1 + geom_path(data = r, aes(X, Y), color = "red")
}
p1
```

### Python implementation {.impl}

```{python}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
#| warning: false
# RK4 for a generic multi-variable system (same as Part 3A)
def RK4_generic(derivs, X0, t_total, dt, **kw):
    t_all = np.arange(0, t_total + dt, dt); n = len(t_all)
    X_all = np.zeros((n, len(X0))); X_all[0] = X0
    for i in range(n - 1):
        t0 = t_all[i]
        k1 = dt * np.array(derivs(t0,          X_all[i],          **kw))
        k2 = dt * np.array(derivs(t0 + dt / 2, X_all[i] + k1 / 2, **kw))
        k3 = dt * np.array(derivs(t0 + dt / 2, X_all[i] + k2 / 2, **kw))
        k4 = dt * np.array(derivs(t0 + dt,     X_all[i] + k3,     **kw))
        X_all[i + 1] = X_all[i] + (k1 + 2 * k2 + 2 * k3 + k4) / 6
    return np.column_stack((t_all, X_all))

# the original (forward-time) toggle switch
def derivs_ts(t, Xs):
    X, Y = Xs
    return [5 + 50 * hill_inh(Y, 100, 4) - 0.1 * X,
            4 + 40 * hill_inh(X, 150, 4) - 0.12 * Y]

np.random.seed(75)
t_total, dt = 100, 0.01
X_range, Y_range = [0, 600], [0, 650]
saddle = np.array([189.96888, 126.64398])   # the saddle from Part 3B

for X, Y in grid:                            # redraw the reversed vector field
    dX, dY = unit_arrow(X, Y)
    plt.arrow(X, Y, dX, dY, head_width=8, head_length=8, ec="black")

# original system, 10 random starts -> the two attractors (blue)
for X0 in np.random.uniform(0, 600, (10, 2)):
    r = RK4_generic(derivs_ts, X0, t_total, dt)
    plt.plot(r[:, 1], r[:, 2], color="blue")

# reversed system, 10 points beside the saddle -> the separatrix (red)
for d in np.random.uniform(-2, 2, (10, 2)):
    r = RK4_generic(derivs_ts_rev, saddle + d, t_total, dt)
    inside = (((r[:, 1] - X_range[0]) * (r[:, 1] - X_range[1]) < 0) &
              ((r[:, 2] - Y_range[0]) * (r[:, 2] - Y_range[1]) < 0))
    plt.plot(r[inside, 1], r[inside, 2], color="red")

plt.gca().set_aspect("equal")
plt.xlabel("X"); plt.ylabel("Y"); plt.xlim(0, 600); plt.ylim(0, 650); plt.show()
```

::: {.callout-note title="Practice Q1"}
Take two initial conditions on opposite sides of the separatrix but very close to
it, integrate the *original* system, and confirm they end at different attractors.
How does the time a trajectory lingers near the saddle depend on how close its
start is to the separatrix?
:::
