---
title: "5A. Integrators for second-order ODEs"
author: "Mingyang Lu"
old_id: "05A"
---

Molecular dynamics (MD) simulates the motion of interacting particles by
integrating Newton's equation of motion forward in time [@frenkel2023understanding].
Before adding the complications of many particles and realistic forces, this
chapter develops the integrators themselves on the simplest mechanical system, a
one-dimensional harmonic oscillator.

Newton's second law relates a particle's acceleration to the force acting on it,

$$ \frac{d^2 x(t)}{dt^2} = \frac{F(x, t)}{m} = f(x, t) . \tag{1} $$

Here $x$ is the position, $t$ is time, $F$ is the force, and $m$ is the mass. We
write $f = F/m$ (the acceleration, or force per unit mass) to keep the notation
compact. A second-order ODE can always be recast as a coupled pair of first-order
ODEs by introducing the velocity $v = dx/dt$,

$$
\begin{cases}
\dfrac{dv}{dt} = f \\[2mm]
\dfrac{dx}{dt} = v
\end{cases} \tag{2}
$$

The integrators below advance this pair step by step. Throughout, $h$ is the time
step size, and a subscript $n$ denotes a quantity evaluated at time
$t_n = t_0 + n h$, so $f_n = f(x_n, t_n)$.

## Modeling a harmonic oscillator

We test each integrator on the harmonic oscillator, a mass on a spring of
stiffness $k$, whose force is linear in the displacement,

$$ f(x) = -k x . $$

Its exact motion is a pure sinusoid with a constant energy, so any growth in
amplitude or drift in energy that appears in a simulation is a numerical
artifact we can spot at a glance.

### R implementation {.impl}

```{r}
# force (per unit mass) of a harmonic oscillator
f_ho <- function(t, x, k){
  return(-k * x)
}
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

# force (per unit mass) of a harmonic oscillator
def f_ho(t, x, k):
    return -k * x
```

## Euler method

The Euler method carries over directly from the [first-order ODE version](../02-odes/02b-numerical-integration.qmd): advance both the
velocity and the position by one step using their current derivatives,

$$
\begin{aligned}
v_{n+1} &= v_n + h f_n \\
x_{n+1} &= x_n + h v_n .
\end{aligned}
$$

To judge a simulation we track the *massic energy* (energy per unit mass),
$e = \tfrac{1}{2} k x^2 + \tfrac{1}{2} v^2$, which is exactly conserved for the
true oscillator. We run each integrator with a small step ($h = 0.01$) and a
larger one ($h = 0.1$) from the same initial condition.

### R implementation {.impl}

```{r}
# Euler integrator for Newton's equation of motion (explicit v, then x)
euler_md <- function(f, t0, x0, v0, t.total, dt, ...){
  # f: acceleration function f(t, x, ...); x0, v0: initial position and velocity
  t_all <- seq(t0, t.total, by = dt)
  n_all <- length(t_all)
  x_all <- numeric(n_all); v_all <- numeric(n_all)
  x_all[1] <- x0; v_all[1] <- v0
  for (i in 1:(n_all - 1)) {
    v_all[i+1] <- v_all[i] + dt * f(t_all[i], x_all[i], ...)
    x_all[i+1] <- x_all[i] + dt * v_all[i]
  }
  return(cbind(t_all, x_all, v_all))   # columns: t, x, v
}

# small step: reasonable oscillation
dt1 <- 0.01; dt2 <- 0.1; k <- 0.1
result_ho_euler <- euler_md(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt1, k = k)
plot(result_ho_euler[,1], result_ho_euler[,2], type = "l", col = 2,
     xlab = "t", ylab = "values", xlim = c(0,100), ylim = c(-10,10))
lines(result_ho_euler[,1], result_ho_euler[,3], col = 4)
legend("topleft", inset = 0.02, legend = c("x","v"), col = c(2,4), lty = 1, cex = 0.8)
```

```{r}
# larger step: the trajectory drifts outward
result_ho_euler2 <- euler_md(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt2, k = k)
plot(result_ho_euler2[,1], result_ho_euler2[,2], type = "l", col = 2,
     xlab = "t", ylab = "values", xlim = c(0,100), ylim = c(-10,10))
lines(result_ho_euler2[,1], result_ho_euler2[,3], col = 4)
legend("topleft", inset = 0.02, legend = c("x","v"), col = c(2,4), lty = 1, cex = 0.8)
```

```{r}
# massic energy over time for both step sizes
energy_ho <- function(result, k) {
  # result: matrix with columns (t, x, v); k: spring constant
  potential_e <- 0.5 * k * result[,2]^2
  kinetic_e   <- 0.5 * result[,3]^2
  return(potential_e + kinetic_e)
}

plot(result_ho_euler[,1], energy_ho(result_ho_euler, k), type = "l", col = 2,
     xlab = "t", ylab = "E", xlim = c(0,100), ylim = c(0,6))
lines(result_ho_euler2[,1], energy_ho(result_ho_euler2, k), col = 4)
legend("topleft", inset = 0.02, legend = c("dt = 0.01","dt = 0.1"), col = c(2,4), lty = 1, cex = 0.8)
```

### Python implementation {.impl}

```{python}
# Euler integrator for Newton's equation of motion (explicit v, then x)
def euler_md(f, t0, x0, v0, t_total, dt, **kwargs):
    # f: acceleration function f(t, x, ...); x0, v0: initial position and velocity
    t_all = np.arange(t0, t_total + dt, dt)
    n_all = len(t_all)
    x_all = np.zeros(n_all); v_all = np.zeros(n_all)
    x_all[0] = x0; v_all[0] = v0
    for i in range(n_all - 1):
        v_all[i+1] = v_all[i] + dt * f(t_all[i], x_all[i], **kwargs)
        x_all[i+1] = x_all[i] + dt * v_all[i]
    return np.column_stack((t_all, x_all, v_all))   # columns: t, x, v

# small step: reasonable oscillation
dt1, dt2, k = 0.01, 0.1, 0.1
result_ho_euler = euler_md(f_ho, 0, 1, 2, 100, dt1, k=k)
plt.plot(result_ho_euler[:,0], result_ho_euler[:,1], color="red", label="x")
plt.plot(result_ho_euler[:,0], result_ho_euler[:,2], color="blue", label="v")
plt.xlabel("t"); plt.ylabel("values"); plt.xlim(0,100); plt.ylim(-10,10)
plt.legend(loc="upper left"); plt.show()
```

```{python}
# larger step: the trajectory drifts outward
result_ho_euler2 = euler_md(f_ho, 0, 1, 2, 100, dt2, k=k)
plt.plot(result_ho_euler2[:,0], result_ho_euler2[:,1], color="red", label="x")
plt.plot(result_ho_euler2[:,0], result_ho_euler2[:,2], color="blue", label="v")
plt.xlabel("t"); plt.ylabel("values"); plt.xlim(0,100); plt.ylim(-10,10)
plt.legend(loc="upper left"); plt.show()
```

```{python}
# massic energy over time for both step sizes
def energy_ho(result, k):
    # result: matrix with columns (t, x, v); k: spring constant
    potential_e = 0.5 * k * result[:,1]**2
    kinetic_e   = 0.5 * result[:,2]**2
    return potential_e + kinetic_e

plt.plot(result_ho_euler[:,0], energy_ho(result_ho_euler, k), color="red", label="dt = 0.01")
plt.plot(result_ho_euler2[:,0], energy_ho(result_ho_euler2, k), color="blue", label="dt = 0.1")
plt.xlabel("t"); plt.ylabel("E"); plt.xlim(0,100); plt.ylim(0,6)
plt.legend(loc="upper left"); plt.show()
```

With a small step the simulation reproduces a clean harmonic oscillation, but
at the larger step the trajectory swings ever wider. The energy makes the problem
precise: the Euler method does not conserve it. At $h = 0.1$ the energy grows
steadily, and even at $h = 0.01$ it drifts slowly upward rather than staying
constant. A good MD integrator should instead keep the energy bounded over long
runs.

### Symplectic Euler method

Two properties separate a good MD integrator from a poor one: it should preserve
the *symplectic form* (which keeps the energy bounded) and be *time-reversible*
(so the motion can be run backward). The plain Euler method has neither. A tiny
change, evaluating the position update with the *new* velocity instead of the old
one,

$$
\begin{aligned}
v_{n+1} &= v_n + h f_n \\
x_{n+1} &= x_n + h v_{n+1} ,
\end{aligned}
$$

makes the method symplectic (though still not time-reversible). Its
implementation and analysis are left as an exercise.

## Leapfrog method

The leapfrog method is a second-order integrator that stores the velocity at
half-integer time steps, so that position and velocity "leap" over each other,

$$
\begin{aligned}
x_{n+1} &= x_n + h v_{n+1/2} \\
v_{n+3/2} &= v_{n+1/2} + h f_{n+1} .
\end{aligned}
$$

It needs the position $x_0$ and the half-step velocity $v_{1/2}$ to start; the
latter is obtained from the initial velocity by a half Euler step,
$v_{1/2} = v_0 + \tfrac{1}{2} h f_0$.

### R implementation {.impl}

```{r}
# Leapfrog integrator: v is stored at half steps (n + 1/2)
leapfrog <- function(f, t0, x0, v0, t.total, dt, ...){
  t_all <- seq(t0, t.total, by = dt)
  n_all <- length(t_all)
  x_all <- numeric(n_all); v_all <- numeric(n_all)
  x_all[1] <- x0
  v_all[1] <- v0 + 0.5 * dt * f(t_all[1], x_all[1], ...)   # half step to v_{1/2}
  for (i in 1:(n_all - 1)) {
    x_all[i+1] <- x_all[i] + dt * v_all[i]
    v_all[i+1] <- v_all[i] + dt * f(t_all[i+1], x_all[i+1], ...)
  }
  return(cbind(t_all, x_all, v_all))
}

# trajectories at the two step sizes (v is shifted by half a step when plotted)
result_ho_leapfrog  <- leapfrog(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt1, k = k)
result_ho_leapfrog2 <- leapfrog(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt2, k = k)

plot(result_ho_leapfrog[,1], result_ho_leapfrog[,2], type = "l", col = 2,
     xlab = "t", ylab = "values", xlim = c(0,100), ylim = c(-10,10))
lines(result_ho_leapfrog[,1] + 0.5 * dt1, result_ho_leapfrog[,3], col = 4)
legend("topleft", inset = 0.02, legend = c("x","v"), col = c(2,4), lty = 1, cex = 0.8)
```

```{r}
# massic energy: bounded even at the large step size
plot(result_ho_leapfrog[,1], energy_ho(result_ho_leapfrog, k), type = "l", col = 2,
     xlab = "t", ylab = "E", xlim = c(0,100), ylim = c(0,6))
lines(result_ho_leapfrog2[,1], energy_ho(result_ho_leapfrog2, k), col = 4)
legend("topleft", inset = 0.02, legend = c("dt = 0.01","dt = 0.1"), col = c(2,4), lty = 1, cex = 0.8)
```

### Python implementation {.impl}

```{python}
# Leapfrog integrator: v is stored at half steps (n + 1/2)
def leapfrog(f, t0, x0, v0, t_total, dt, **kwargs):
    t_all = np.arange(t0, t_total + dt, dt)
    n_all = len(t_all)
    x_all = np.zeros(n_all); v_all = np.zeros(n_all)
    x_all[0] = x0
    v_all[0] = v0 + 0.5 * dt * f(t0, x0, **kwargs)   # half step to v_{1/2}
    for i in range(1, n_all):
        x_all[i] = x_all[i-1] + dt * v_all[i-1]
        v_all[i] = v_all[i-1] + dt * f(t_all[i], x_all[i], **kwargs)
    return np.column_stack((t_all, x_all, v_all))

# trajectories at the two step sizes (v is shifted by half a step when plotted)
result_ho_leapfrog  = leapfrog(f_ho, 0, 1, 2, 100, dt1, k=k)
result_ho_leapfrog2 = leapfrog(f_ho, 0, 1, 2, 100, dt2, k=k)

plt.plot(result_ho_leapfrog[:,0], result_ho_leapfrog[:,1], color="red", label="x")
plt.plot(result_ho_leapfrog[:,0] + 0.5 * dt1, result_ho_leapfrog[:,2], color="blue", label="v")
plt.xlabel("t"); plt.ylabel("values"); plt.xlim(0,100); plt.ylim(-10,10)
plt.legend(loc="upper left"); plt.show()
```

```{python}
# massic energy: bounded even at the large step size
plt.plot(result_ho_leapfrog[:,0], energy_ho(result_ho_leapfrog, k), color="red", label="dt = 0.01")
plt.plot(result_ho_leapfrog2[:,0], energy_ho(result_ho_leapfrog2, k), color="blue", label="dt = 0.1")
plt.xlabel("t"); plt.ylabel("E"); plt.xlim(0,100); plt.ylim(0,6)
plt.legend(loc="upper left"); plt.show()
```

Unlike the Euler method, the leapfrog energy does not drift, even at the large
step size. It is a symplectic, time-reversible method. One caveat affects the
energy plot: leapfrog stores $x_n$ and $v_{n+1/2}$ at staggered times rather than
at the same instant, so the massic energy computed from them shows a small
oscillatory ripple when the step size is large.

## Velocity Verlet method

The leapfrog method never has $x$ and $v$ available at the same time. The
velocity Verlet method fixes this by splitting the velocity update into two half
steps that bracket the position update [@swope1982computer],

$$
\begin{aligned}
v_{n+1/2} &= v_n + \tfrac{1}{2} h f_n \\
x_{n+1} &= x_n + h v_{n+1/2} \\
v_{n+1} &= v_{n+1/2} + \tfrac{1}{2} h f_{n+1} .
\end{aligned}
$$

### R implementation {.impl}

```{r}
# Velocity Verlet: half-step v, full-step x, half-step v
velocity_verlet <- function(f, t0, x0, v0, t.total, dt, ...){
  t_all <- seq(t0, t.total, by = dt)
  n_all <- length(t_all)
  x_all <- numeric(n_all); v_all <- numeric(n_all)
  x_all[1] <- x0; v_all[1] <- v0
  for (i in 1:(n_all - 1)) {
    v_half     <- v_all[i] + 0.5 * dt * f(t_all[i], x_all[i], ...)
    x_all[i+1] <- x_all[i] + dt * v_half
    v_all[i+1] <- v_half + 0.5 * dt * f(t_all[i+1], x_all[i+1], ...)
  }
  return(cbind(t_all, x_all, v_all))
}

# trajectory at the small step size (x and v now share time points)
result_ho_vv  <- velocity_verlet(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt1, k = k)
result_ho_vv2 <- velocity_verlet(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt2, k = k)

plot(result_ho_vv[,1], result_ho_vv[,2], type = "l", col = 2,
     xlab = "t", ylab = "values", xlim = c(0,100), ylim = c(-10,10))
lines(result_ho_vv[,1], result_ho_vv[,3], col = 4)
legend("topleft", inset = 0.02, legend = c("x","v"), col = c(2,4), lty = 1, cex = 0.8)
```

```{r}
# massic energy: conserved at both step sizes
plot(result_ho_vv[,1], energy_ho(result_ho_vv, k), type = "l", col = 2,
     xlab = "t", ylab = "E", xlim = c(0,100), ylim = c(0,6))
lines(result_ho_vv2[,1], energy_ho(result_ho_vv2, k), col = 4)
legend("topleft", inset = 0.02, legend = c("dt = 0.01","dt = 0.1"), col = c(2,4), lty = 1, cex = 0.8)
```

### Python implementation {.impl}

```{python}
# Velocity Verlet: half-step v, full-step x, half-step v
def velocity_verlet(f, t0, x0, v0, t_total, dt, **kwargs):
    t_all = np.arange(t0, t_total + dt, dt)
    n_all = len(t_all)
    x_all = np.zeros(n_all); v_all = np.zeros(n_all)
    x_all[0] = x0; v_all[0] = v0
    for i in range(1, n_all):
        v_half = v_all[i-1] + 0.5 * dt * f(t_all[i-1], x_all[i-1], **kwargs)
        x_all[i] = x_all[i-1] + dt * v_half
        v_all[i] = v_half + 0.5 * dt * f(t_all[i], x_all[i], **kwargs)
    return np.column_stack((t_all, x_all, v_all))

# trajectory at the small step size (x and v now share time points)
result_ho_vv  = velocity_verlet(f_ho, 0, 1, 2, 100, dt1, k=k)
result_ho_vv2 = velocity_verlet(f_ho, 0, 1, 2, 100, dt2, k=k)

plt.plot(result_ho_vv[:,0], result_ho_vv[:,1], color="red", label="x")
plt.plot(result_ho_vv[:,0], result_ho_vv[:,2], color="blue", label="v")
plt.xlabel("t"); plt.ylabel("values"); plt.xlim(0,100); plt.ylim(-10,10)
plt.legend(loc="upper left"); plt.show()
```

```{python}
# massic energy: conserved at both step sizes
plt.plot(result_ho_vv[:,0], energy_ho(result_ho_vv, k), color="red", label="dt = 0.01")
plt.plot(result_ho_vv2[:,0], energy_ho(result_ho_vv2, k), color="blue", label="dt = 0.1")
plt.xlabel("t"); plt.ylabel("E"); plt.xlim(0,100); plt.ylim(0,6)
plt.legend(loc="upper left"); plt.show()
```

Velocity Verlet conserves the massic energy at both step sizes and returns $x$
and $v$ at the same time points, which is why it is the standard workhorse
integrator in molecular dynamics.

### Position Verlet method

By symmetry, one can instead split the *position* update into two half steps
around a full velocity step, giving the position Verlet method. Working out its
update equations is left as an exercise.

## Verlet method

If the velocity is not needed as an output, the original Verlet method integrates
the position alone [@verlet1967computer]. Adding the Taylor expansions of
$x_{n+2}$ and $x_n$ about $x_{n+1}$ cancels the velocity and the odd-order terms,
leaving

$$ x_{n+2} = 2 x_{n+1} - x_n + h^2 f_{n+1} . $$

The scheme needs two starting positions; the second is obtained from the initial
position and velocity by

$$ x_1 = x_0 + h v_0 + \tfrac{1}{2} h^2 f_0 . $$

### R implementation {.impl}

```{r}
# Verlet integrator: position only, no explicit velocity
verlet <- function(f, t0, x0, v0, t.total, dt, ...){
  t_all <- seq(t0, t.total, by = dt)
  n_all <- length(t_all)
  x_all <- numeric(n_all)
  x_all[1] <- x0
  x_all[2] <- x0 + dt * v0 + 0.5 * dt * dt * f(t_all[1], x_all[1], ...)
  for (i in 2:(n_all - 1)) {
    x_all[i+1] <- 2 * x_all[i] - x_all[i-1] + dt * dt * f(t_all[i], x_all[i], ...)
  }
  return(cbind(t_all, x_all))   # columns: t, x
}

result_ho_v  <- verlet(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt1, k = k)
result_ho_v2 <- verlet(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt2, k = k)

plot(result_ho_v[,1], result_ho_v[,2], type = "l", col = 2,
     xlab = "t", ylab = "x", xlim = c(0,100), ylim = c(-10,10))
lines(result_ho_v2[,1], result_ho_v2[,2], col = 4)
legend("topleft", inset = 0.02, legend = c("dt = 0.01","dt = 0.1"), col = c(2,4), lty = 1, cex = 0.8)
```

### Python implementation {.impl}

```{python}
# Verlet integrator: position only, no explicit velocity
def verlet(f, t0, x0, v0, t_total, dt, **kwargs):
    t_all = np.arange(t0, t_total + dt, dt)
    n_all = len(t_all)
    x_all = np.zeros(n_all)
    x_all[0] = x0
    x_all[1] = x0 + dt * v0 + 0.5 * dt * dt * f(t_all[0], x0, **kwargs)
    for i in range(1, n_all - 1):
        x_all[i+1] = 2 * x_all[i] - x_all[i-1] + dt * dt * f(t_all[i], x_all[i], **kwargs)
    return np.column_stack((t_all, x_all))   # columns: t, x

result_ho_v  = verlet(f_ho, 0, 1, 2, 100, dt1, k=k)
result_ho_v2 = verlet(f_ho, 0, 1, 2, 100, dt2, k=k)

plt.plot(result_ho_v[:,0], result_ho_v[:,1], color="red", label="dt = 0.01")
plt.plot(result_ho_v2[:,0], result_ho_v2[:,1], color="blue", label="dt = 0.1")
plt.xlabel("t"); plt.ylabel("x"); plt.xlim(0,100); plt.ylim(-10,10)
plt.legend(loc="upper left"); plt.show()
```

The Verlet method reproduces the oscillation at both step sizes while storing
only positions, which is why it and its velocity Verlet variant underlie most
practical MD codes. We extend these integrators to systems of many interacting
particles in [Part 5B](05b-orbital-motion.qmd).
