---
title: "Introduction to R Programming"
author: "Mingyang Lu"
date: "10/6/2021"
output: 
  github_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Why Using R?
R is a programming language designed for statistical computing and graphics. Unlike a compiled language, such as C and Fortran, R is typically used as an interpreted language. There are several reasons why I choose R here.

* R is simple to use. If you have some experience with any other programming language, it should be pretty easy to learn R.

* R comes with an interactive environment for programming. Many users also use [RStudio](https://www.rstudio.com) for R. It provides an integrated environment for coding and scripting and a user-friendly Graphic User Interface (GUI), very similar to Matlab. [Here is an article comparing R and Matlab.](https://hackr.io/blog/r-vs-matlab)  

* R is cross-platform. Both R and RStudio are available in Windows, Mac, and Linux. 

* Many R packages have been developed and are freely accessible for statistical analyses and biomedical research. R packages can usually be found via (1) [Github](https://github.com); (2) The Comprehensive R Archive Network ([CRAN](https://cran.r-project.org)); (3) [BioConductor](https://www.bioconductor.org) for bioinformatics.

* It is not difficult to develop R packages. R also provides interface for other languages, such as C++ and Fortran, for computational efficiency. 

* R packages can also be easily converted into interactive web app using [Shiny](https://shiny.rstudio.com).

To get started, one needs to download and install R and RStudio from the following websites:  [CRAN](https://cran.r-project.org) and [RStudio](https://www.rstudio.com/). I recommend R version 3.6.3 for this tutorial. 

From the R Console, type the following command to get help.

```{r}
help(runif)     # show documentation for the topic -- runif in this example
```

## A crash course of R
I will not provide a detailed R tutorial, as they are already available. I recommend the following resources for learning R.

* [An Introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html) from [CRAN](https://cran.r-project.org)

* [R & Bioconductor Manual](http://manuals.bioinformatics.ucr.edu/home/R_BioCondManual) by Thomas Girke at UC Riverside

* [Efficient R programming](https://csgillespie.github.io/efficientR/) by Colin Gillespie & Robin Lovelace

Here, I will provide a brief overview of R syntax and highlight a few R's unique features. 

#### Typical data types & math operations
```{r}
a = 10 + 3.14    # integer, real number; assignment to a
a*a # a multiplies a
a**2 # square of a, which gives the same value as the previous one
b = 3.0 + 4.0i  # complex number; assignment to b
Mod(b) # The modulus of b
c = (3 + 4 == 7)   # logical; c is TRUE
!c   # not c, which is FALSE
d = "Monday"  # character
class(d)  # showing the data type of d
nchar(d)  # number of characters in a string
```
#### Vectors
```{r}
vec = c("Man","Woman","Woman","Man","Woman")   # a vector of characters
length(vec) #length of a vector
vec == "Woman" # logic values comparing each element to "Woman"
which(vec == "Woman") # identifying the indices of the elements being "Woman"
```
#### Factors
Factor type is an important data type in R. It is commonly used to represent categorical data. 
```{r}
vec_factor = as.factor(vec)   # converting the character vector to a factor vector
class(vec);class(vec_factor)  # checking the data types
levels(vec_factor); nlevels(vec_factor)  # levels and the number of levels
```
#### If statement
```{r}
a = 0
if(a == 1){   # conditions
  print("a equals to 1")
} else {
  print("a is not 1")
}
```
#### For loops
```{r}
for (i in 1:5) print(i)   # version 1
```

```{r}
for (i in seq_len(5)) print(i)   # version 2
```

```{r}
v = c(1,2,5,7)
for (i in v){
  print(i)   # version 3
}
```

#### While statement
```{r}
i = 0
while(i < 5){
  i = i + 2
  print(i)
}
```

#### Apply
Another unique feature in R. Apply performs a function iteratively through an array or matrix. It's usually more efficient than a for-loop. Syntax: apply(X, MARGIN, FUN). Here, X is an array or matrix. MARGIN=1 performs the specified function by rows, MARGIN=2 by columns. FUN is the function to apply. 
```{r}
mat = matrix(rnorm(16), nrow = 4)    # generating a random matrix of 4 x 4
mat
apply(mat, 1, sum)   # row sum
rowSums(mat)   # An alternative way to do row sum
apply(mat, 2, sum)   # column sum
colSums(mat)  # An alternative way to do column sum
```
There are other apply-like functions, such as lappy, sappy, tapply, *etc.* [Check here for more details.](https://www.datacamp.com/community/tutorials/r-tutorial-apply-family)

#### Functions
```{r}
myfunction <- function(x) {    # a simple function to perform x square
  x*x
}
b = myfunction(4)
print(paste0("b = ",b))
```

#### List
Lists can contain elements of different types. 
```{r}
my_list = list("a", c(1,2,3), FALSE, 3.14)  # Defining a new list
my_list[[1]]   # the first element
my_list[[2]]   # the second element
names(my_list) = c("letters", "array", "TF", "pi")  # naming list elements
my_list$TF   # retrieve elements by names
my_vector = unlist(my_list)   # convert the list to a vector
my_vector
```

#### Data frame
Data frame is widely used in R. It's similar to a matrix but allows mixed data types in it.
```{r}
my_data = data.frame(
          id = c(1:5),
          atom = c("N","CA","CB","C","O"),
          mass = c(14, 12, 12, 12, 16),
          size = c(1.2, 1.4, 1.4, 1.4, 1.1)
)
print(my_data)
colnames(my_data)
str(my_data)      # get the structure 
my_data$mass      # retrive a column
my_data$backbone = c(T, T, F, T, T)   # add a column
print(my_data)
my_data = rbind(my_data, c(6, "S", 32, 1.6, F))   # add a row
print(my_data)
```

#### Input/output
```{r}
save(my_data,file = "my_data.Rdata")    # save the data frame my_data to file
saveRDS(my_data, file = "my_data.RDS")  # save my_data to an RDS file
load("my_data.Rdata")   # read data from an Rdata file; my_data, if exist, would be overwritten
my_data2 = readRDS("my_data.RDS")   # read data from an RDS file, assign it to another variable
```

#### Package
The following code works to install/remove packages (not in the R markdown format to avoid adding/removing packages)
```
if(! "umap" %in% rownames(installed.packages())) {      # check if the package exists
  install.packages("umap")   # if not, install "umap"
}
library("umap")    # load the umap library
help(umap)
remove.packages("umap")    # remove it ...
```

#### Basic plotting
``` {r}
x = seq_len(5)
y = x**2
plot(x, y)       # plot points
curve(x**2)      # plot f(x) = x*x
y = data.frame(xvalues = rnorm(1000), yvalues = rnorm(1000))    # 1000 random points in 2D
plot(y$xvalues, y$yvalues)     # scatter plot
library(ggplot2)
ggplot(y, aes(x = xvalues, y = yvalues)) +    # density map
  stat_density2d(aes(fill = ..density..), geom = "raster", contour = FALSE)
``` 

## A few notes

#### Factor v.s. character

#### Matrix v.s. data frame

#### List v.s. vector

## Efficient R programming

#### Vectors

#### Vactors

#### C/Fortran codes

#### Compilation

#### Parallelization
