---
title: "Part 1: Introduction to R Programming"
subtitle: "A. Basics of R"
author: "Mingyang Lu"
date: "08/16/2023"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Why Using R?
R is a programming language designed for statistical computing and graphics. Unlike a compiled language, such as C and Fortran, R is typically used as an interpreted language. There are several reasons why I choose R here.

* R is simple to use. If you have some experience with any other programming language, it should be pretty easy to learn R.

* R comes with an interactive environment for programming. Many users also use [RStudio](https://www.rstudio.com) for R. It provides an integrated environment for coding and scripting and a user-friendly Graphic User Interface (GUI), very similar to Matlab. [Here is an article comparing R and Matlab.](https://hackr.io/blog/r-vs-matlab)  

* R is cross-platform. Both R and RStudio are available in Windows, Mac, and Linux. 

* Many R packages have been developed and are freely accessible for statistical analyses and biomedical research. R packages can usually be found via (1) [Github](https://github.com); (2) The Comprehensive R Archive Network ([CRAN](https://cran.r-project.org)); (3) [BioConductor](https://www.bioconductor.org) for bioinformatics.

* It is not difficult to develop R packages. R also provides interface for other languages, such as C++ and Fortran, for computational efficiency. 

* R packages can also be easily converted into interactive web app using [Shiny](https://shiny.rstudio.com).

To get started, one needs to download and install R and RStudio from the following websites:  [CRAN](https://cran.r-project.org) and [RStudio](https://www.rstudio.com/). I recommend R version 3.6.3 for this tutorial. 

From the R console, type the following command to get help.

```{r}
help(runif)     # show documentation for the topic -- runif in this example
```

# A crash course of R
I will not provide a detailed R tutorial, as they are already available. I recommend the following resources for learning R.

* [An Introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html) from [CRAN](https://cran.r-project.org)

* [R & Bioconductor Manual](http://manuals.bioinformatics.ucr.edu/home/R_BioCondManual) by Thomas Girke at UC Riverside

* [Efficient R programming](https://csgillespie.github.io/efficientR/) by Colin Gillespie & Robin Lovelace

Here, I will provide a brief overview of R syntax and highlight a few R's unique features. 

## Typical data types & math operations
```{r}
a = 10 + 3.14    # integer, real number; assignment to a
a*a # a multiplies a
a**2 # square of a, which gives the same value as the previous one
b = 3.0 + 4.0i  # complex number; assignment to b
Mod(b) # The modulus of b
c = (3 + 4 == 7)   # logical; c is TRUE
!c   # not c, which is FALSE
d = "Monday"  # character
class(d)  # show the data type of d
nchar(d)  # number of characters in a string
```

## Vectors
```{r}
vec = c("Man","Woman","Woman","Man","Woman")   # a vector of characters
length(vec) #length of a vector
vec == "Woman" # logic values, compare each element to "Woman"
which(vec == "Woman") # identify the indices of the elements being "Woman"
```

## Factors
Factor type is an important data type in R. It is commonly used to represent categorical data. 
```{r}
vec_factor = as.factor(vec)   # convert the character vector to a factor vector
class(vec);class(vec_factor)  # check the data types
levels(vec_factor); nlevels(vec_factor)  # levels and the number of levels
```

## If statement
```{r}
a = 0
if(a == 1){   # conditions
  print("a equals to 1")
} else {
  print("a is not 1")
}
```

## For loops
```{r}
for (i in 1:5) print(i)   # version 1
```

```{r}
for (i in seq_len(5)) print(i)   # version 2
```

```{r}
v = c(1,2,5,7)
for (i in v){
  print(i)   # version 3
}
```

## While statement
```{r}
i = 0
while(i < 5){
  i = i + 2
  print(i)
}
```

## Apply
Another unique feature in R. Apply performs a function iteratively through an array or matrix. It's usually more efficient than a for-loop. Syntax: apply(X, MARGIN, FUN). Here, X is an array or matrix. MARGIN=1 performs the specified function by rows, MARGIN=2 by columns. FUN is the function to apply. 
```{r}
mat = matrix(rnorm(16), nrow = 4)    # generate a random matrix of 4 x 4
mat
apply(mat, 1, sum)   # row sum
rowSums(mat)   # An alternative way to do row sum
apply(mat, 2, sum)   # column sum
colSums(mat)  # An alternative way to do column sum
```
There are other apply-like functions, such as lappy, sappy, tapply, *etc.* [Check here for more details.](https://www.datacamp.com/community/tutorials/r-tutorial-apply-family)

## Functions
```{r}
myfunction <- function(x) {    # a simple function to perform x square
  return(x*x)
}
b = myfunction(4)
print(paste0("b = ",b))
```

## List
Lists can contain elements of different types. 
```{r}
my_list = list("a", c(1,2,3), FALSE, 3.14)  # define a new list
my_list[[1]]   # the first element
my_list[[2]]   # the second element
names(my_list) = c("letters", "array", "TF", "pi")  # name list elements
my_list$TF   # retrieve elements by names
my_vector = unlist(my_list)   # convert the list to a vector
my_vector
```

## Data frame
Data frame is widely used in R. It's similar to a matrix but allows mixed data types in it.
```{r}
my_data = data.frame(
          id = c(1:5),
          atom = c("N","CA","CB","C","O"),
          mass = c(14, 12, 12, 12, 16),
          size = c(1.2, 1.4, 1.4, 1.4, 1.1),
          stringsAsFactors = F
)
print(my_data)
colnames(my_data)
str(my_data)      # get the structure 
my_data$mass      # retrieve a column
my_data$backbone = c(T, T, F, T, T)   # add a column
print(my_data)
my_data = rbind(my_data, c(6, "S", 32, 1.6, F))   # add a row
print(my_data)
```

## Input/output
```{r}
save(my_data,file = "extra/data/01A/my_data.Rdata")    # save the data frame my_data to file
saveRDS(my_data, file = "extra/data/01A/my_data.RDS")  # save my_data to an RDS file
load("extra/data/01A/my_data.Rdata")   # read data from an Rdata file; my_data, if exist, would be overwritten
my_data2 = readRDS("extra/data/01A/my_data.RDS")   # read data from an RDS file, assign it to another variable
```

## Package
The following code works to install/remove packages 
```
if(! "umap" %in% rownames(installed.packages())) {      # check if the package exists
  install.packages("umap")   # if not, install "umap"
}
library("umap")    # load the umap library
help(umap)
remove.packages("umap")    # remove it ...
```

## Basic plotting
``` {r}
x = seq_len(5)
y = x**2
plot(x, y)       # plot points
curve(x**2)      # plot f(x) = x*x
y = data.frame(xvalues = rnorm(1000), yvalues = rnorm(1000))    # 1000 random points in 2D
plot(y$xvalues, y$yvalues)     # scatter plot
library(ggplot2)
ggplot(y, aes(x = xvalues, y = yvalues)) +    # density map
  stat_density2d(aes(fill = ..density..), geom = "raster", contour = FALSE)
``` 

# A few notes
The previous section highlights the basic usage of R. Please refer to "[An Introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html)" and [R & Bioconductor Manual](http://manuals.bioinformatics.ucr.edu/home/R_BioCondManual) for more details. In the following, I will emphasize a few R features that new learners may overlook or be confused about.

## Factor vs. string
Factors may look similar to Strings, but they are very different in R scripting. What makes things worse is that, in R versions before 4.0, stringsAsFactors = TRUE by default. That means, for R functionality using the parameter stringsAsFactors, any string would be interpreted as a factor . However, in R versions 4.0 or later, the default stringsAsFactors = FALSE. If a package is developed for R before v4.0 and by assuming default treatment of strings (as factors), the same package may not work properly for R v.4.0 or later. A good practice for R programming is (1) Use R version 3.6.3 if many legacy R packages are required; (2) always specify how string would be treated. Here is an example to create a data frame containing characters/strings.

```{r}
my_data = data.frame(
          id = c(1:5),
          atom = c("N","CA","CB","C","O"),
          mass = c(14, 12, 12, 12, 16),
          size = c(1.2, 1.4, 1.4, 1.4, 1.1)
)
print(my_data)
class(my_data$atom)   # factor if R version < 4.0; character if R version >=4.0
my_data = data.frame(
          id = c(1:5),
          atom = c("N","CA","CB","C","O"),
          mass = c(14, 12, 12, 12, 16),
          size = c(1.2, 1.4, 1.4, 1.4, 1.1),
          stringsAsFactors = F
)

class(my_data$atom)  # always character type
```
Here is an example to read a table from a file.
```{r}
write.table(my_data, file = "extra/data/01A/my_data.txt", quote = F, sep="\t", row.names = F) #The table is saved to a file
data_read = read.table("extra/data/01A/my_data.txt", header = T, sep = "\t")   #Read the table from the saved file
class(data_read$atom)   # factor if R version < 4.0; character if R version >=4.0
data_read = read.table("extra/data/01A/my_data.txt", header = T, sep = "\t", as.is = T)   # as.is = T reads strings as characters
class(data_read$atom)
```

## Data frame vs. matrix
Matrix is more suitable for linear algebra analysis; while data frame is good to save heterogeneous data and ideal for statistics analysis, plotting, *etc.* Sometimes, data can be saved as either a data frame or a matrix. It is important to be aware of the data type, so that appropriate functions/operators are applied. Refer to [the following article](https://www.geeksforgeeks.org/matrix-vs-dataframe-in-r/) for the comparison of matrix and data frame.
[Basic plotting] shows an example of plotting with data frame. Here is an example of linear regression with data frame.
```{r}
plot(my_data$mass, my_data$size, xlab = "Size", ylab = "Mass")    # size vs. mass in my_data
model = lm(size~mass, data = my_data)   # linear regression
abline(model)   # add the fitted line
summary(model)   # summary of the linear model

```

## List vs. vector
Lists are recursive type of vectors -- lists can have values of different types. A data frame is a list where column elements have the same data type. In the above-mentioned example of linear regression, model is a list containing results of the linear model.
```{r}
str(model)   # show the structure of the list
```
Below, mean value is computed for each column of the matrix using lapply. 
```{r}
means = lapply(mat, mean)    # this doesn't work
length(means)
mat_list =  lapply(seq_len(ncol(mat)), function(i) mat[,i])  # convert to a list by column vectors
means = lapply(mat_list,mean)  # calculate means for a list
means
means2 = colMeans(mat)   # an alternative way to compute column means
means2
```

## Usage of ellipsis in R funcitons
When defining an R function, you may use ellipsis (*i.e.*, ...) to pass any argument to the function. This is especially useful when the function calls another function with varying number of arguments. For example, 

```{r}
## a function without using ellipsis
func_main <- function(a, b, c){
  return(a + b + c)
}

func_main(a = 1, b = 2.4, c = 0.6)
 
## a function where an argument of the function is the name of another function, func_2nd
## ellipsis ... is used to pass arguments for func_2nd
func_main2 <- function(a, func_2nd, ...){
  return(a + func_2nd(...))
}

## Usage case 1
func1 <- function(d, e){
  return(d + e)
}
func_main2(a = 1, func_2nd = func1, d = 2.4, e = 0.6)

## Usage case 2
func2 <- function(f, g, h, i){
  return(f+g+h*i)
}
func_main2(a = 1, func_2nd = func2, f = 10, g = 20, h = 1, i = -3)

## Wrong usages
#func_main2(a = 1, func_2nd = func2, f = 10, g = 20, h = 1) 
#func_main2(a = 1, func_2nd = func2, d = 2.4, e = 0.6) 
```

Make sure that all arguments passed through ellipsis are the same whenever ellipsis is used in body of the function. Thus, in the usage case 2 above, the code doesn't work if only f, g, h are passed to func2. It doesn't work either if, for example, d and e are passed to func2.
