---
title: "Part 1: Introduction to R Programming"
subtitle: "C. Exercises"
author: "Mingyang Lu"
date: "10/12/2021"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Radius of gyration
In this exercise, we will compute radius of gyration of a protein structure. First, install a package named bio3d.  
```{r}
if(! "bio3d" %in% rownames(installed.packages())) {      # check if the package exists
  install.packages("bio3d")   # if not, install "bio3d"
}
library(bio3d)
```
The package bio3d allows users to download and process protein structures from Protein Data Bank.
```{r}
mystructure <- read.pdb("7l3u")  # load online data
coord <- mystructure$atom
class(coord)
head(coord)
```
The above code reads the PDB file for the protein named 7L3U and accesses the atomic coordinate information. Start from the data frame coord, and compute radius of gyration of the structure.
Here, the radius of gyration $R_g$ is defined as:

$$R_g=\sqrt{\frac{\sum_{j=1}^{N}m_j(\mathbf{r_j}-\mathbf{\overline{r}})^2}{\sum_{j=1}^{N}m_j}}$$
where $\mathbf{\overline{r}}$ is defined as

$$\mathbf{\overline{r}}=\frac{\sum_{j=1}^{N}m_j\mathbf{r_j}}{\sum_{j=1}^{N}m_j}$$

Hint: try the following for the usage of atom2mass
```{r}
  ?atom2mass
```

# Stirling's approximation
Part B shows how to benchmark code performance with microbenchmark. Here, we will benchmark more using the same approach. We would like to compute the following quantity for large n:

$$\ln n!=\sum_{j=1}^{n}\ln j$$

Create R functions using the following methods: 

* (1) Iteration with a for loop in R; 
* (2) Use vector operations in R; 
* (3) Use apply-family functions (such as apply, lapply, *etc.*) in R; 
* (4) Use C or Fortran subroutines/functions. 

Benchmark these methods with microbenchmark.

In mathematics, there is famous formula called Stirling's approximation.

$$\ln n! \approx n\ln n - n$$
Plot the difference of the left hand side and the right hand side, *.i.e.*,
$$\ln n! - n\ln n + n$$
as the function $n$
