---
title: "Part 1: Introduction to R Programming"
subtitle: "C. Numerical methods"
author: "Mingyang Lu"
date: "08/16/2023"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A [Wikipedia](https://en.wikipedia.org/wiki/Numerical_method) definition of numerical method says: 
*In numerical analysis, a numerical method is a mathematical tool designed to solve numerical problems. The implementation of a numerical method with an appropriate convergence check in a programming language is called a numerical algorithm.*

Here are two examples of numerical methods.

# Babylonian Method for finding square roots

Starting from an initial guess $x_1 > 0$, we perform a series of improved guesses iteratively according to the following formula:

$$x_{n+1} = \frac{1}{2} (x_n + \frac{a}{x_{n}})$$
The stopping criteria is:

$$ |x_{n+1}^2 - a| < \epsilon$$
,where \espilon is a small positive number, representing the tolerance. 

Here is an implementation of the Babylonian method.

```{r}
# The Babylonian method for finding square root
# a: a positive real number; we will solve the square root of a.
# x1: the initial guess of the solution. x1 should be positive too.
# epsilon: a small positive number for the tolerance.

babylonian <- function(a, x1, epislon = 1e-6) {
  if(a <= 0 || x1 <= 0 ){
    print("a and x1 should be both positive!")
    return(0)
  }else{
    x = x1
    i = 0
    while(abs(x*x - a) > epislon){
      i = i + 1
      x = (x + a/x)/2
    }
    print(paste0("Number of iterations: ", i))
    return(x)
  }
}

babylonian(2,1)
babylonian(2,100)
babylonian(2,200)
```

# Cubic spline interpolation

Consider a series $n+1$ points $(x_0, y_0)$, $(x_1, y_1)$, ... , $(x_n, y_n)$ to interpolate. For two points $(x_i, y_i)$ and $(x_{i+1}, y_{i+1})$, we define a cubic polynomial

$$y_i(x) = a_i  x^3 + b_i x^2 + c_i x + d_i \tag {1}$$
To satisfy the boundary conditions, we have

$$ y_i(x_i) = y_{i+1}(x_i) \tag {2}$$
$$ \frac{dy_i}{dx}(x_i) = \frac{dy_{i+1}}{dx}(x_i) \tag {3}$$
$$ \frac{d^2y_i}{dx^2}(x_i) = \frac{d^2y_{i+1}}{dx^2}(x_i) \tag {4}$$
Usually we also need another two boundary conditions, *e.g.*, for zero second derivatives for the end points (aka natural boundary condition).

$$ \frac{d^2y_0}{dx^2}(x_0) = \frac{d^2y_{n-1}}{dx^2}(x_n) = 0 \tag {5}$$

Below is a small data set of three points for average temperatures at different latitudes. 

```{r}
x = c(5, 15, 25) 
y = c(-3.02,-3.07, -3.17)
plot(x,y, xlab = "Latitude", ylab = "Average temperature", type = "b", col = "red", lwd = 0.5)
```

In this simple case, $n = 3$. From Equations (1) and (2),

$$y_0 = a_0  x_0^3 + b_0 x_0^2 + c_0 x_0 + d_0 $$

$$y_1 = a_0  x_1^3 + b_0 x_1^2 + c_0 x_1 + d_0 $$

$$y_1 = a_1  x_1^3 + b_1 x_1^2 + c_1 x_1 + d_1 $$
$$y_2 = a_1  x_2^3 + b_1 x_2^2 + c_1 x_2 + d_1 $$
From Equations (1) and (3),

$$3 a_0  x_1^2 + 2 b_0 x_1 + c_0 = 3 a_1  x_1^2 + 2 b_1 x_1 + c_1 $$
From Equations (1) and (4),

$$6 a_0  x_1 + 2 b_0 = 6 a_1  x_1 + 2 b_1 $$
From Equations (1) and (5),

$$6 a_0  x_0 + 2 b_0 = 6 a_1  x_2 + 2 b_1 = 0 $$


These all lead to a system of linear equations

\def\A{
\begin{pmatrix}
      a_0 \\ b_0 \\ c_0 \\ d_0 \\ a_1 \\ b_1 \\ c_1 \\ d_1
\end {pmatrix}}

\def\C{
\begin{pmatrix}
       y_0 \\ y_1 \\ y_1 \\ y_2 \\ 0 \\ 0 \\ 0 \\ 0
\end {pmatrix}}

\def\B{
\begin{pmatrix}
       x_0^3 & x_0^2 & x_0 & 1 & 0 & 0 & 0 & 0 \\
       x_1^3 & x_1^2 & x_1 & 1 & 0 & 0 & 0 & 0 \\
       0 & 0 & 0 & 0 & x_1^3 & x_1^2 & x_1 & 1 \\
       0 & 0 & 0 & 0 & x_2^3 & x_2^2 & x_2 & 1 \\
       3 x_1^2 & 2 x_1 & 1 & 0 & -3 x_1^2 & -2 x_1 & -1 & 0 \\
       6 x_1 & 2 & 0 & 0 & -6 x_1 & -2 & 0 & 0 \\
       6 x_0 & 2 & 0 & 0 & 0 & 0 & 0 & 0 \\
       0 & 0 & 0 & 0 & 6 x_2 & 2 & 0 & 0 
\end {pmatrix}}

\begin{equation}
\B \A = \C  \tag{6}
\end{equation}

Here's an implementation of Equation (6) in R

```{r}
vec_c = c(y[1], y[2], y[2], y[3], 0, 0, 0, 0) # the vector from the right hand side of Equation (6)

a = matrix(c(x[1]**3, x[1]**2, x[1], 1, 0, 0, 0, 0,
             x[2]**3, x[2]**2, x[2], 1, 0, 0, 0, 0,
             0, 0, 0, 0, x[2]**3, x[2]**2, x[2], 1,
             0, 0, 0, 0, x[3]**3, x[3]**2, x[3], 1,
             3*x[2]**2, 2*x[2], 1, 0, -3*x[2]**2, -2*x[2],-1, 0,
             6*x[2], 2, 0, 0, -6*x[2], -2, 0, 0,
             6*x[1], 2, 0, 0, 0, 0, 0, 0,
             0, 0, 0, 0, 6*x[3], 2, 0, 0), nrow = 8, byrow = TRUE)

vec_p = solve(a, vec_c) #  the solution vector (parameters for cubic spline)

plot(x,y, xlab = "Latitude", ylab = "Average temperature", type = "b", col = "red", lwd = 0.5)
curve(vec_p[1]*x**3+ vec_p[2]*x**2 + vec_p[3]*x + vec_p[4], from = x[1], to = x[2], add = TRUE, lwd = 1.5, col = "black")
curve(vec_p[5]*x**3+ vec_p[6]*x**2 + vec_p[7]*x + vec_p[8], from = x[2], to = x[3], add = TRUE, lwd = 1.5, col = "blue")
legend("bottomleft", inset=.02, legend = c("Linear", "Cubic spline 1", "Cubic spline 2"), 
       col = c("red", "black", "blue"), lty = 1, cex = 0.8)
```
 