---
title: "Part 3: Phase plane"
subtitle: "A. Nullclines"
author: "Mingyang Lu"
date: "11/03/2021"
output: 
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In Part 3, we will generalize what we learn from Part 2 to ODEs with two variables and apply them to model a gene circuit with two nodes. We will focus on phase plane analysis for a visual characterization of nonlinear dynamical systems.

# A system with two variables
We first model the toggle switch circuit with two genes $X$ and $Y$, whose dynamics can be described by a set of two ODEs below (*i.e.*, equation (13) in Part 2A). 

\begin{equation}
\begin{cases} \frac{dX}{dt} = g_{X_0} + g_{X1}\frac{1}{1+(Y/Y_0)^{n_Y}} - k_XX \\
              \frac{dY}{dt} = g_{Y_0} + g_{Y1}\frac{1}{1+(X/X_0)^{n_X}} - k_YY \end{cases}
\end{equation}

### ODE simulations

Below shows functions to compute the derivatives of the toggle switch circuit. For illustration, the kinetic parameters of the circuit are provided in the script. 
``` {r}
hill_inh <- function(X,X0,n) {
  a = (X/X0)**n
  return(1/(1+a))
}

derivs_ts <- function(t, Xs) {
  X = Xs[1]
  Y = Xs[2]
  dxdt = 5 + 50 * hill_inh(Y, 100, 4) - 0.1 * X
  dydt = 4 + 40 * hill_inh(X, 150, 4) - 0.12 * Y
  return(c(dxdt, dydt))
}
```

We can simualte the dynamics of $X$ and $Y$ using 4th order Runge-Kutta (RK4). The function RK4_2D for a system with two variables is very similar to the function RK4 in Part 2B. Now, the time trajectory is stored in a matrix called X_all.

```{r}
# 4th order Runge-Kutta (RK4)
RK4_2D <- function(derivs, init, t.total, dt){
  # derivs: the function of the derivatives 
  # init: initial condition in c(X0, Y0)
  # t.total: total simulation time, assuming t starts from 0 at the beginning
  # dt: time step size 
  t_all = seq(0, t.total, by=dt)
  n_all = length(t_all)
  X_all = matrix(0, nrow = n_all, ncol = 2)
  X_all[1,] = init 
  for (i in 1:(n_all-1)) {
    t_0= t_all[i]
    t_0.5 = t_0 + 0.5*dt
    t_1 = t_0 + dt
    k1 = dt * derivs(t_0,X_all[i,])
    k2 = dt * derivs(t_0.5,X_all[i,] + k1/2)
    k3 = dt * derivs(t_0.5,X_all[i,] + k2/2)
    k4 = dt * derivs(t_1,X_all[i,] + k3)
    X_all[i+1,] = X_all[i,] + (k1+2*k2+2*k3+k4)/6
  }
  return(cbind(t_all, X_all))   # the output is a matrix of t & X(t) for all time steps
}

set.seed(77)   # set the seed for the random number generator, remove this line to generate more random initial conditions
X_init_all = array(runif(20, 0, 600), dim = c(10,2)) # generate 10 random initial conditions

t.total = 100
dt  = 0.01
plot(NULL, xlab="t (Minute)", ylab="Levels (nM)", xlim=c(0,80), ylim=c(0,650)) 
for(i in seq_len(nrow(X_init_all))){
  results = RK4_2D(derivs_ts, X_init_all[i,], t.total, dt)
  lines(results[,1], results[,2], col = 1)
  lines(results[,1], results[,3], col = 2)
}
legend("top", inset=0.02, legend = c("X", "Y"),
       col=1:2, lty=1, cex=0.8)
```

We randomly selected 10 initial conditions and simulate the ODEs starting from each. As shown in the above figure, the system can reach to two stable steady states, *i.e.*, two sets of $X$ and $Y$. 

### Phase plane

A powerful way to characterize the dynamic behavior of such a two-variable system is the phase plane analysis. A phase plane describes the states of a system by two axes representing, *e.g.* in this case, the levels of the two variables. Here, we plot the time trajectories from the previous simulations to the phase plane of $X$ and $Y$.

``` {r, fig.width = 5, fig.height = 5}
plot(NULL, xlab="X (nM)", ylab="Y (nM)", xlim=c(0,600), ylim=c(0,650)) 
for(i in seq_len(nrow(X_init_all))){
  results = RK4_2D(derivs_ts, X_init_all[i,], t.total, dt)
  lines(results[,2], results[,3], col = 1)
}
``` 
 
The trajectories shown in the phase plane illustrate again the existence of two stable steady states. 

### Vector field

We can also plot arrows for points in the phase plane to represent the direction of changes.

\begin{equation}
\begin{cases} V_X = \frac{dX}{dt}  \\
              V_Y = \frac{dY}{dt} \end{cases}
\end{equation}

``` {r,fig.width = 5, fig.height = 5}
library(ggplot2)

X_all = seq(0, 600, by=20)   # all X grids
Y_all = seq(0, 650, by=20)   # all Y grids
XY_all = expand.grid(X_all, Y_all)   # all combinations of X and Y

results = t(apply(XY_all, MARGIN=1, function(Xs) {return(c(Xs, derivs_ts(0, Xs)))})) # generate all vector field data
colnames(results) = c("X", "Y", "dX", "dY")

ggplot(data=as.data.frame(results), aes(x=X, y=Y)) + 
  geom_segment(aes(xend=X+dX, yend=Y+dY), arrow = arrow(length = unit(0.05,"in")))

```

In the vector filed, vectors are pointing towards to stable steady states. There is also another steady state in the middle of the plot, where the nearby vectors point toward it from one direction and point outwards from another direction. This steady state is a saddle point, a type of unstable steady statThere is also another steady state in the middle of the plot, where the nearby vectors point toward it from one direction and point outwards from another direction. This steady state is a saddle point, a type of unstable steady state. 

We can choose to plot the unit vectors of the vector field to highlight the directions. 

``` {r,fig.width = 5, fig.height = 5}

results_unit = t(apply(XY_all, MARGIN=1, function(Xs) {
  v = derivs_ts(0, Xs)
  v_norm = v / sqrt(v[1]**2 + v[2]**2)
  return(c(Xs, v_norm))})) # generate all vector field data
colnames(results_unit) = c("X", "Y", "dX", "dY")

p1 = ggplot(data=as.data.frame(results_unit), aes(x=X, y=Y)) + 
  geom_segment(aes(xend=X+20*dX, yend=Y+20*dY), arrow = arrow(length = unit(0.05,"in")))
p1
```

# Nullclines

Another convenient way to evaluate the behavior of a nonlinear dynamical system is the techniques of nullcline analysis. A nullcline corresponds to the steady state condition of one ODE. Thus, there are two nullclines for (1) $\frac{dX}{dt} = 0$ and (2) $\frac{dY}{dt} = 0$. In the current example, the first nullcline satisfies 

$$f_X(X, Y) = g_{X0} + g_{X1}\frac{1}{1+(Y/Y_0)^{n_Y}} - k_XX = 0  \tag{1} $$
The second one satisfies

$$ f_Y(X, Y) = g_{Y0} + g_{Y1}\frac{1}{1+(X/X_0)^{n_X}} - k_YY = 0 \tag{2} $$
The curve for an equation like $f_X(X, Y) = 0$ can be obtain numerically, as we will discussed very soon. But for this relatively simple example, we can get the curve right away by *separation of variables*.

From equation (1), we can move $k_XX$ to the right hand side, and we have

$$X(Y)  = \frac{g_{X0} + g_{X1}\frac{1}{1+(Y/Y_0)^{n_Y}}}{k_X} \tag{3}$$

Similarly, from equation (2), we can also get

$$ Y(X) = \frac{g_{Y0} + g_{Y1}\frac{1}{1+(X/X_0)^{n_X}}}{k_Y} \tag{4} $$

The first nullcline (equation (3)) can also be expression as $Y(X)$. Although we don't have to, as $X(Y)$ is sufficent to obtain the nullcline curve numerically. 

``` {r,fig.width = 6, fig.height = 5}
X_all = seq(0, 600, by=1)   # all X grids
Y_all = seq(0, 650, by=1)   # all Y grids

nullcline1 <- function(Y) {
  X= (5 + 50 * hill_inh(Y, 100, 4))/0.1 
  return(cbind(X, Y))
}
nullcline2 <- function(X) {
  Y = (4 + 40 * hill_inh(X, 150, 4))/0.12
  return(cbind(X, Y))
}

null1 = nullcline1(Y_all)
null2 = nullcline2(X_all)

p1 +
  geom_line(data = as.data.frame(null1), aes(x=X, y=Y, colour = "dX/dt=0"), size=1) + 
  geom_line(data = as.data.frame(null2), aes(x=X, y=Y, colour = "dY/dt=0"), size=1) 

``` 

The intersections of the two nullclines are the steady states of the system. By checking the number of intersections, we can identify the multi-stability property easily. The separation of variables should be applied whenever possible. However, it may not be feasible in many situations. We then have to reply on other approaches, such as contour method and numerical continuation.

### Contour method

One of such approaches relies on the contour analysis. We consider $Z = f_X(X,Y)$ and construct a landscape of a plane of $X$ and $Y$, with $Z$ as the height. The nullcline would be the contour line when $Z=0$. This can be easily done with the *contour* function in base R or other plotting packages, such as *geom_contour* from *ggplot2*. In this approach, $Z$ values for all 2D grid points are required to compute the contour. This can be computational intensive when high resolution grid points are needed. The contour method is usually robust, as long as the $Z$ values are smooth enough for interpolation.

``` {r,fig.width = 5, fig.height = 5}
X_all = seq(0, 600, by=10)   # all X grids
Y_all = seq(0, 650, by=10)   # all Y grids
nX = length(X_all)
nY = length(Y_all)
XY_all = expand.grid(X_all, Y_all)   # all combinations of X and Y

results = t(apply(XY_all, MARGIN=1, function(Xs) {return(c(Xs, derivs_ts(0, Xs)))})) # generate all vector field data
z_X = array(results[,3], dim = c(nX, nY))
z_Y = array(results[,4], dim = c(nX, nY))

contour(X_all, Y_all, z_X, levels = 0, col = 2, xlab = "X", ylab = "Y", drawlabels = F)
contour(X_all, Y_all, z_Y, levels = 0, col = 3, add = TRUE, drawlabels = F)
```

### Numerical continuation

Another approach is the numerical continuation, which we have already applied in the bifurcation analysis in Part 2D. In that previous analysis, we identify the bifurcation curve that satisfies $f(X, k) = 0$. Similarly, we will identify the nullcline curve that satisfies $f_X(X, Y) = 0$ (and the other one that obeys $f_Y(X, Y) = 0$). Here, we can choose $Y$ as the control parameter and find $X(Y)$ (that's what we will do below). Of course, it is also fine to choose $X$ as the control parameter and find $Y(X)$. 

``` {r,fig.width = 5, fig.height = 5}
f_null1 <- function(X, Y) {       # f_X(X, Y)
  return(5 + 50 * hill_inh(Y, 100, 4) - 0.1 * X)
}
dfdX_null1 <- function(X, Y) {   # Jacobian (1st order partial derivatives) for f_X(X, Y)
  return(-0.1)
}
dfdY_null1 <- function(X, Y) {
  y_frac = (Y/100)**4
  return(-200/Y*y_frac/(1+y_frac)**2)
}
f_null2 <- function(X, Y) {       # f_Y(X, Y)
  return(4 + 40 * hill_inh(X, 150, 4) - 0.12 * Y)
}
dfdX_null2 <- function(X, Y) {   # Jacobian for f_Y(X, Y)
  x_frac = (X/150)**4
  return(-160/X*x_frac/(1+x_frac)**2)
}
dfdY_null2 <- function(X, Y) {   
  return(-0.12)
}

# Newton's Method
find_root_Newton <- function(X, func, dfunc, parms, X_range, error = 10^-3) {
  while(abs(func(X, parms)) > error){
    f = func(X, parms)
    X = X - f/dfunc(X, parms)
    if((X-X_range[1])*(X-X_range[2]) > 0) break 
  }
  return(X)
}

# 4th order Runge-Kutta (RK4) for 1D
# simplified to (1) only output steady-state X or final X, 
# (2) derivative is not the function of t in this implementation, thus some t-dependent parts are simplified
RK4_1D_steady <- function(derivs, X0, params, t.total, dt){
  # derivs: the function of the derivatives 
  # X0: initial condition
  # t.total: total simulation time, assuming t starts from 0 at the beginning
  # dt: time step size 
  error = 10^-8
  t = 0
  X = X0
  while(t <= t.total){
    df = derivs(X, params)
    if(abs(df) < error) break
  
    k1 = dt * df
    k2 = dt * derivs(X + k1/2, params)
    k3 = dt * derivs(X + k2/2, params)
    k4 = dt * derivs(X + k3, params)
    
    X = X + (k1+2*k2+2*k3+k4)/6
    t = t + dt
  }
  return(X)
}

nullcline_numcont <- function(func, dfdX, dfdY, X_range, Y_range, X_init, Y_init, direct = 1, step_arc) {
  nmax_cycle = 10000
  
  results = matrix(NA, nrow = nmax_cycle, ncol = 2)
  cycle = 1
  step_Y_previous = direct 
  step_X_previous = 0
  
  X_new = X_init
  Y = Y_init 
  results[cycle,] = c(X_new, Y)
  while(((X_new - X_range[1])*(X_new - X_range[2]) <= 0) & # boundary check
        ((Y - Y_range[1]) * (Y - Y_range[2]) <= 0) & 
        (cycle < nmax_cycle)) {
    h = -dfdY(X_new, Y)/dfdX(X_new, Y) 
    
    step_Y = step_arc/sqrt(1+h**2)
    step_X = step_Y*h
    
    if((step_Y_previous * step_Y + step_X_previous * step_X) < 0){ 
      step_Y = - step_Y
      step_X = - step_X
    }
    step_Y_previous = step_Y
    step_X_previous = step_X
    
    Y = Y + step_Y
    X_init = X_new + step_X
    
    X_new = find_root_Newton(X=X_init, func=func, dfunc=dfdX, parms=Y, X_range = X_range) 
    
    cycle = cycle + 1
    results[cycle,] = c(X_new, Y)
  }
  results = na.omit(results)
  colnames(results) = c("X","Y")
  return(results)
}

X_range = c(0, 600)
Y_range = c(0, 650)

# the initial X,Y need to be provided, here we manually selected them with an initial simulation
X_init_guess_null1 = 500
Y_init_guess_null1 = 1
X_init_null1 = RK4_1D_steady(f_null1, X_init_guess_null1, Y_init_guess_null1, 1000, 0.01)  # start from ~ (550, 1)
null1_cont = nullcline_numcont(f_null1, dfdX_null1, dfdY_null1, X_range, Y_range, 
                               X_init = X_init_null1, Y_init = Y_init_guess_null1, direct = 1, step_arc = 3)

X_init_guess_null2 = 1
Y_init_guess_null2 =366.6 
X_init_null2 = RK4_1D_steady(f_null2, X_init_guess_null2, Y_init_guess_null2, 1000, 0.01)   # start from ~ (8,366.6)
null2_cont = nullcline_numcont(f_null2, dfdX_null2, dfdY_null2, X_range, Y_range, 
                               X_init = X_init_null2, Y_init = Y_init_guess_null2, direct = -1, step_arc = 3)
plot(NULL, xlab="X", ylab="Y", xlim=X_range, ylim=Y_range)
points(null1_cont[,1], null1_cont[,2], pch = 1, cex = 0.3, col = 2)
points(null2_cont[,1], null2_cont[,2], pch = 1, cex = 0.3, col = 3)
legend("topright", inset=0.02, legend = c("dX/dt=0", "dY/dt=0"), col=2:3, lty=1, cex=0.8)
```

The core code is equivalent to the code for 1-variable bifurcation analysis in Part 2D. It looks lengthy and somewhat cumbersome, as the boundary of two variables and the initial direction of nullclines need to be specified. Again, the readers can try the algorithm without the correction method by *find_root_Newton*. One issue about the above code is the lack of divergent check of the functions and Jacobians. 

``` {r}
save(null1_cont, null2_cont, file = "./extra/data/03A/null_03A.RData")    # we save the data for 03B

```
