---
title: "Part 3: Phase plane"
subtitle: "C. Bifurcation"
author: "Mingyang Lu"
date: "11/08/2021"
output: 
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this part, we will integrate what we have discussed in the previous two parts. The goal is to generate bifurcation diagrams to characterize the dynamical behavior of the toggle switch circuit. To generate a bifurcation diagram, we first need to specify a control parameter. We will plot bifurcation for the steady states of the circuit as the function of the control parameter. We will need to (1) choose the value of the control parameter; (2) identify nullclines; (3) find the steady states by the intersections of the two nullclines; (4) find the stability; (5) repeat 1 - 4 until we reconstruct the bifurcation diagram.

Here is script to achieve this.

```{r}


```

In this example, nullclines can be determined easily. Otherwise, we have to reply on other approaches, such as numerical continuation. We previously discussed the numerical continuation to identify the curve satisfying the condition of a equation with two variables. This can be generalized to the conditions of a set of $n$ equations with $n+1$ variables. In this case, $n=2$. We will discuss numerical continuation for toggle switch bifurcation in Part 3F.