---
title: "Part 3: Phase plane"
subtitle: "E. Bifurcation for two-variable systems"
author: "Mingyang Lu"
date: "11/08/2021"
output: 
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Generating bifurcation diagram

In this part, we will integrate what we have discussed in the Parts 3AB. The goal is to generate bifurcation diagrams to characterize the dynamical behavior of the toggle switch circuit. To generate a bifurcation diagram, we first need to specify a control parameter. We will plot bifurcation for the steady states of the circuit as the function of the control parameter. We will need to (1) choose the value of the control parameter; (2) identify nullclines; (3) find the steady states by the intersections of the two nullclines; (4) find the stability; (5) repeat 1 - 4 until we reconstruct the bifurcation diagram.

Here, we consider the same toggle switch circuit but with slightly modified parameters. The rate equations have two variables $X$ and $Y$ and two parameters $g_X$ and $k$.


\begin{equation}
\begin{cases} \frac{dX}{dt} = f_X(X, Y, g_X, k) = 5 + g_X\frac{1}{1+(Y/100)^4} - kX \\
              \frac{dY}{dt} = f_Y(X, Y, g_X, k) = 5 + 50\frac{1}{1+(X/100)^4} - kY \tag{1} \end{cases}
\end{equation}

We will identify bifurcation diagrams by varying either $g_X$ or $k$.

```{r}
# ODE terms
fx <- function(X, Y, gX, k) {       # dX/dt
  return(5 + gX/(1 + (Y/100)**4) - k * X)
}
fy <- function(X, Y, gX, k) {       # dY/dt
  return(5 + 50/(1 + (X/100)**4) - k * Y)
}

# nullcline can be determined by separation of variables. Here, we obtain points along the bifurcation curve. 
null_fx <- function(gX, k, X_range, Y_range) {
  Y_all = seq(Y_range[1], Y_range[2], length.out = 1000)
  X_all = (5 + gX/(1 + (Y_all/100)**4))/k
  ind = which((X_all - X_range[1]) *(X_all - X_range[2]) <= 0)
  results = cbind(X_all, Y_all)
  colnames(results) = c("X", "Y")
  return(results)
}

null_fy <- function(gX, k, X_range, Y_range) {
  X_all = seq(X_range[1], X_range[2], length.out = 1000)
  Y_all = (5 + 50/(1 + (X_all/100)**4)) / k 
  ind = which((Y_all - Y_range[1]) *(Y_all - Y_range[2]) <= 0)
  results = cbind(X_all, Y_all)
  colnames(results) = c("X", "Y")
  return(results)
}

# find all intersections
find_intersection_all_fast <- function(lineA, lineB, dxdt, dydt) {
  small = 10^-3   # use a small positive number instead of 0 to allow some numerical errors in the following step; 
  lineA_ind = which(dydt * c(dydt[-1],NA)<=small)  # find sign flipping between two consecutive points
  lineB_ind = which(dxdt * c(dxdt[-1],NA)<=small) 
  
  lines_all = expand.grid(lineA_ind, lineB_ind)   # all combinations of the above
  results = apply(lines_all, MARGIN = 1, function(inds) {
    return(find_intersection(lineA, lineB, inds[1], inds[2]))})
  return(t(results[1:2,which(results[3,] == 1)]))
}

find_intersection <- function(lineA, lineB, nA, nB) {
  dAX = lineA[nA+1,1] - lineA[nA,1]
  dAY = lineA[nA+1,2] - lineA[nA,2]
  
  dBX = lineB[nB+1,1] - lineB[nB,1]
  dBY = lineB[nB+1,2] - lineB[nB,2]
  
  dABX = lineB[nB,1] - lineA[nA, 1]
  dABY = lineB[nB,2] - lineA[nA, 2]
  
  d = dAX * dBY - dAY * dBX
  
  alpha = (dABX * dBY - dABY*dBX)/d
  beta = (dABX * dAY - dABY*dAX)/d
  
  if((alpha*(1-alpha) >= 0) &
     (beta*(1-beta) >= 0)) {   # check whether there is an intersection in between
    intersection = c((1-alpha)*lineA[nA,1] + alpha *lineA[nA+1,1], 
                    (1-alpha)*lineA[nA,2] + alpha *lineA[nA+1,2],1)
    # 1st and 2nd elements: X & Y
    # 3rd element: 1 means found, 0 means not found
  }
  else{
    intersection = c(alpha,beta,0)
  }
  return(intersection)
}

stability_v2 <- function(func_fx, func_fy, gX, k, ss) { 
  delta = 0.001
  delta2 = delta*2
  func_fx_current = func_fx(ss[1], ss[2], gX, k) 
  func_fy_current = func_fy(ss[1], ss[2], gX, k)  
  dfxdx = (func_fx(ss[1]+delta, ss[2], gX, k) - func_fx_current)/delta2
  dfxdy = (func_fx(ss[1], ss[2]+delta, gX, k) - func_fx_current)/delta2
  dfydx = (func_fy(ss[1]+delta, ss[2], gX, k) - func_fy_current)/delta2
  dfydy = (func_fy(ss[1], ss[2]+delta, gX, k) - func_fy_current)/delta2
  
  jacobian = array(c(dfxdx, dfydx, dfxdy, dfydy), c(2,2))
  lambda = eigen(jacobian)$values
  if(class(lambda[1]) == "complex") {
    if(Re(lambda[1]) < 0){
      stability = 4   # stable spiral
    }else{
      stability = 5   # unstable spiral
    }
  }
  else{
    if((lambda[1] < 0) & (lambda[2] <0)){
      stability = 1   # stable 
    }else if((lambda[1] >= 0) & (lambda[2] >= 0)){
      stability = 2   # unstable 
    }else{
      stability = 3   # saddle
    }
  }
  return(stability)
}

# all integrated function
find_steady_states <- function(gX, k, X_range, Y_range) {
  line_1 = null_fx(gX, k, X_range, Y_range)
  line_2 = null_fy(gX, k, X_range, Y_range)
  
  dxdt = fx(line_2[,1], line_2[,2], gX, k)  #dX/dt along the nullcline dY/dt = 0
  dydt = fy(line_1[,1], line_1[,2], gX, k)  #dY/dt along the nullcline dX/dt = 0
  
  ss = find_intersection_all_fast(line_1, line_2, dxdt, dydt)
  
  ss_with_stability =  t(apply(ss, MARGIN = 1, function(ss) {
                          return(c(ss,stability_v2(fx, fy, gX, k, ss)))}))
  return(ss_with_stability)
}

```

# Saddle-node bifurcation

Bifurcation with respect to $g_X$. From the plot below, this is saddle-node bifurcation.

``` {r}
X_range = c(0, 1000)
Y_range = c(0, 1000)

gX = 50
k = 0.1
ss_all = numeric()
for(gX in seq(0, 100, by = 1)){
  ss = find_steady_states(gX, k, X_range, Y_range)
  ss_all = rbind(ss_all, cbind(gX,ss))
}
colnames(ss_all) = c("gX", "X", "Y", "Stability")
plot(ss_all[,1], ss_all[,2], col = ss_all[,4], type = "p", pch = 16, cex = 0.4,
     xlab="gX", ylab="X", xlim=c(0,100), ylim=c(0,1000))
legend("topleft", inset=0.02, legend = c("Stable", "Saddle"), col=c(1,3), lty=1, cex=0.8)
```

# Pitchfork bifurcation

Bifurcation with respect to $k$. This type of bifurcation is called pitchfork bifurcation.

``` {r}
X_range = c(0, 1000)
Y_range = c(0, 1000)

gX = 50
k = 0.1
ss_all = numeric()
for(k in seq(0.01, 0.15, by = 0.001)){
  ss = find_steady_states(gX, k, X_range, Y_range)
  ss_all = rbind(ss_all, cbind(k,ss))
}
colnames(ss_all) = c("k", "X", "Y", "Stability")
plot(ss_all[,1], ss_all[,2], col = ss_all[,4], type = "p", pch = 16, cex = 0.4,
     xlab="gX", ylab="X", xlim=c(0,0.15), ylim=c(0,800))
legend("topleft", inset=0.02, legend = c("Stable", "Saddle"), col=c(1,3), lty=1, cex=0.8)
```

In this example, nullclines can be determined easily. Otherwise, we have to reply on other approaches, such as numerical continuation. We previously discussed the numerical continuation to identify the curve satisfying the condition of an equation with two variables. This can be generalized to the conditions of a set of $n$ equations with $n+1$ variables. In this case, $n=2$. We will leave numerical continuation for the bifurcation analysis for toggle switch circuit as an exercise in Part 3H.
