---
title: "Part 4: Stochastic differential equations"
subtitle: "A. Euler–Maruyama method"
author: "Mingyang Lu"
date: "11/11/2021"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Brownian dynamics 

We previous described that an ODE like $\frac{dX}{dt} = f(X)$ is analogous to an overdamped dynamics of a one-dimensional particle. In physics, random motion of particles in solution can be described by Brownian dynamics, governed by a stochastic differential equation (SDE):

$$\frac{dX}{dt} = f(X) + \sqrt{2D(X)}\eta(t) \tag{1}$$
Compared to the ODE, SDE has an additional term $\eta(t)$, describing the stochastic contribution to the derivatives of $X$. The most common type of noise is called Gaussian white noise, which by definition is a random signal having equal intensity at different frequencies. $D(X)$ represents the level of noise (in physics we know $D$ has the meaning of the diffusion constant). $\eta(t)$ of a Gaussian white noise satisfies

\begin{equation}
\begin{cases} <\eta(t)> = 0  \\
              <\eta(t)\eta(t')> = \delta (t-t')  \tag{2} \end{cases}
\end{equation}

Here, the symbol $<.>$ means the average operation. $<\eta(t)\eta(t')>$ is a correlation function ([see here for more details](https://en.wikipedia.org/wiki/Correlation_function_(statistical_mechanics))), and \delta (t-t') is the [Dirac delta function](https://en.wikipedia.org/wiki/Dirac_delta_function) of $t-t'$. $\delta (t-t')$ is normalized, and it is nonzero only when $t-t'$ approaches to zero. 

SDE is widely applicable to many fields in Science and Engineering, such as stochastic model of stock market, noise in biological phenomena, chemical and metabolic reactions, etc. So, it is important to know how to numerically solve SDE. We will illustrate SDE simulations in the context of Brownian dynamics in current section and gene expression noise of gene circuits in later sections. 

# Euler–Maruyama method

The SDE in Equation (1) can also be expressed as 

$$dX = f(X)dt + \sqrt{2D(X)}dW \tag{3}$$

Here, $W$ a Wiener process (continuous-time stochastic process). 

We can numerically solve the SDE by

$$X_{n+1} = X_n + f(X_n)\Delta t + \sqrt{2D(X_n)} \Delta W_n \tag{4}$$
$\Delta W_n$ represents normal random variables with mean zero and variance $\Delta t$. ($<\Delta W_n^2> = \Delta t$) This is called Euler-Maruyama method, an extension of the Euler method for ODEs to SDEs. Now, we show the implementation in R. 

```{r}
# SDE_Euler_single is a generic function of Euler-Maruyama method for a single SDE, similar to Euler_single from Part 02B
SDE_Euler_single <- function(derivs, noise, para, X0, t.total, dt){
  # derivs: the function of the derivatives 
  # noise: the function of the noise term 
  # X0: initial condition
  # para: pass-in parameters
  # t.total: total simulation time, assuming t starts from 0 at the beginning
  # dt: time step size 
  t_all = seq(0, t.total, by=dt)
  n_all = length(t_all)
  X_all = numeric(n_all)
  X_all[1] = X0
  dW_all = rnorm(n_all, mean=0, sd=sqrt(dt))  # all dW terms, rnorm generate random variables from normal distribution
  for (i in 1:(n_all-1)) {
    t = t_all[i]
    X = X_all[i]
    X_all[i+1] = X + dt * derivs(t, X, para) + dW_all[i] * noise(t, X, para)
  }
  return(cbind(t_all, X_all))   # the output is a matrix of t & X(t) for all time steps
}
```

Note that in the implementation, all random variables $\Delta W$ are generated altogether. This step is more efficient in R, compared to the way to generate them in each step. However, it requires a lot more memory to do so, therefore not ideal for long simulations. The best approach is to use compiled functions with Fortran and C.

A common mistake by a beginner is to add $\sqrt{2D(X)}$ to the $f(X)$ term and integrate with the ODE version of the Euler method. This is wrong, as the variance of the random variables is now $\Delta t^2$ instead of $\Delta t$. 

Now we will apply the Euler-Maruyama method to a few SDEs. We first consider a simple Brownian motion, where $f(X) = 0$.

``` {r}
fx_zero <- function(t, X, para) return(0)
noise <- function(t, X, D) return(sqrt(2*D)) # D is a variable to represent noise levels, 
set.seed(1)

plot(NULL, xlab="t", ylab="X", xlim=c(0,200), ylim=c(-150,150)) 
for(i in seq_len(10)) {
  results = SDE_Euler_single(fx_zero, noise, para = 10, 0, 200, 0.01) # D = 10
  lines(results[,1], results[,2],  col=i)
}
```

We plotted the stochastic time trajectories for ten simulations. They all look very different. In theory, the variance of $X$ equals to $2Dt$. However, numerically testing this relation requires a lot more simulations, so that robust statistics can be reached. 

Let's consider another example: Brownian dynamics under a harmonic potential and constant noise. Here, the Brownian particle is tethered to a spring. According to Hooke's law, $f(X) = - kX$. $k$ is the spring constant, which we set to 1 for simplification. This type of system is also called Ornstein–Uhlenbeck process. 

$$\frac{dX}{dt} = -X + \sqrt{2D}\eta(t) \tag{5}$$

```{r}
fx <- function(t, X, para) return(-X)

set.seed(1)    # Students can try different RNG seed
D_all = c(100, 25, 1)
results1 = SDE_Euler_single(fx, noise, para = D_all[1], 0, 100, 0.01)
results2 = SDE_Euler_single(fx, noise, para = D_all[2], 0, 100, 0.01)
results3 = SDE_Euler_single(fx, noise, para = D_all[3], 0, 100, 0.01)

plot(NULL, xlab="t", ylab="X", xlim=c(0,100), ylim=c(-30,30)) 
lines(results1[,1], results1[,2],  col=2)
lines(results2[,1], results2[,2],  col=3)
lines(results3[,1], results3[,2],  col=4)
legend("bottomright", inset=0.02, title="D", 
       legend = c(paste0(D_all)),
       col=2:4, lty=1, cex=0.8)
```

In the above simulations, we plot the stochastic dynamics for three different noise levels. Unlike the Brownian motion, the dynamics of $X$ here fluctuate around 0 with deviations proportional to the noise level $D$. The larger the $D$, the noisier the $X$ is.

``` {r}
plot(D_all, c(var(results1[,2]), var(results2[,2]), var(results3[,2])), type = "p", xlab = "D", ylab = "Variance")
abline(a=0, b = 1, col = 1)
```

In theory, we know that $<x^2>=\frac{D}{k}$. In the above test, we can find that the simulation results still deviate from the theoretical values. That is because the simulation is not long enough to reach robust statistics. An efficient SDE integrator is required to achieve that. We will leave the detailed evaluation as an exercise. 

# Simulation with package sde

An R package called *sde* can perform SDE simulations fairly easily. Below shows an example of such analysis for the second example (Brownian motion under a harmonic potential).

```{r, message=FALSE}
library(sde)
```

```{r}
d = expression (-x)
s = expression (sqrt(2*100))

results_sde = sde.sim(X0 = 0, N = 10000, delta = 0.01, drift = d, sigma = s, method = "euler")
plot(results_sde)
```

Benchmark is also perform to evaluate the speed of our simulator and sde package. To my surprise, our simple implementation is more than ten times faster. It could be because

(1) Our version didn't use complex data structure in the simulation loop.

(2) The function rnorm is very efficient to generate random numbers from normal distribution.

(3) Generating all random numbers altogether also helps. 

(4) expression() function, although convenient, could be slow.

Certainly, using compiled language will be the most efficient way, and we will need it to compute statistics. We will discuss this more in Part 4C.

```{r, message=FALSE}
library(microbenchmark)  # benchmark
my_benchmark = microbenchmark(ours = SDE_Euler_single(fx, noise, para = 100, 0, 100, 0.01),
               sde = sde.sim(X0 = 0, N = 10000, delta = 0.01, drift = d, sigma = s, method = "euler"),
               times = 10, unit = "s")  
microbenchmark:::autoplot.microbenchmark(my_benchmark)
```

