---
title: "Part 5: Molecular dynamics"
subtitle: "A. Integrators for second order ODEs"
author: "Mingyang Lu"
date: "08/18/2023"
output: 
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will discuss the numerical methods for integrating Newton's equation of motion.

$$ \frac{d^2}{dt^2}x(t) = F(x,t )/m = f(x,t) \tag{1}$$
Here, $x$ represents the position of a particle, and $t$ represents time. $F$ is a function of $x$, representing the force to the particle, and $m$ is the mass. We use $f$ to simplify the expression of Equation (1).

Newton's equation of motion can also be written in the following form:

\begin{equation}
\begin{cases}
  \frac{dv}{dt} = f \tag{2}\\
  \frac{dx}{dt} = v 
\end{cases}
\end{equation}

# Modeling a harmonic oscillator

Harmonic oscillator is a simple model, where

$$ f(x) = - kx$$
```{r}
# force of a harmonic oscillator
f_ho <- function(t, x, k){
  return(-k*x)
}
```

# Euler method

Similar to the Euler method for first-order ODEs, the Euler method can be applied to integrate second-order ODEs iteratively as follows.

$$v_{n+1} = v_n + hf_n  $$
$$x_{n+1} = x_n + hv_n  $$

```{r}
# Euler method
euler_md <- function(f, t0, x0, v0, t.total, dt, ...){
  # f:  2nd derivative function
  # t0: initial time
  # x0: initial position x
  # v0: initial velocity v
  # t.total: total simulation time
  # dt: time step size 
  t_all = seq(t0, t.total, by=dt)
  n_all = length(t_all)
  x_all = numeric(n_all)
  v_all = numeric(n_all)
  x_all[1] = x0
  v_all[1] = v0
  for (i in 1:(n_all-1)) {
    v_all[i+1] = v_all[i] + dt * f(t_all[i], x_all[i],...)
    x_all[i+1] = x_all[i] + dt * v_all[i]
  }
  return(cbind(t_all, x_all, v_all))   # the output is a matrix of t, x, v for all time steps
}

# Run Euler simulation 1, dt = 0.01
dt1 = 0.01; k = 0.1
result_ho_euler = euler_md(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt1, k = k)

# Plotting the outcomes
plot(result_ho_euler[,1], result_ho_euler[,2], type = "l", col=2,
    xlab="t", ylab="values", xlim=c(0,100), ylim=c(-10,10)) 
lines(result_ho_euler[,1], result_ho_euler[,3], col=4)
legend("topleft", inset=0.02, 
       legend = c("x","v"),
       col=c(2,4), lty=1, cex=0.8)
```

We can observe reasonable harmonic oscillations in the simulation when the time step size is small. However, if we choose a larger time step size, now the trajectory drifts away.

```{r}
# Run Euler simulation 2, dt = 0.1
dt2 = 0.1
result_ho_euler2 = euler_md(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt2, k = k)

# Plotting the outcomes
plot(result_ho_euler2[,1], result_ho_euler2[,2], type = "l", col=2,
    xlab="t", ylab="values", xlim=c(0,100), ylim=c(-10,10)) 
lines(result_ho_euler2[,1], result_ho_euler2[,3], col=4)
legend("topleft", inset=0.02, 
       legend = c("x","v"),
       col=c(2,4), lty=1, cex=0.8)
```

We can compute the massic energy (energy per unit mass) $e = \frac{1}{2}kx^2 +  \frac{1}{2}v^2$.

```{r}
# Compute massic energy
energy_ho <- function(result, k) { 
  # result: matrix of simulation outputs, (columns: t, x, v)
  # k: spring constant
  potential_e = 0.5*k*result[,2]**2
  kinetic_e =  0.5*result[,3]**2
  return(potential_e + kinetic_e)
}

# Plotting the energy
plot(result_ho_euler[,1],  energy_ho(result_ho_euler, k), type = "l", col=2, 
     xlab="t", ylab="E", xlim=c(0,100), ylim=c(0,6)) 
lines(result_ho_euler2[,1], energy_ho(result_ho_euler2, k), col=4)
legend("topleft", inset=0.02, 
       legend = c("dt = 0.01","dt = 0.1"),
       col=c(2,4), lty=1, cex=0.8)
```

The massic energy is clearly not conserved for Euler simulations with large time step sizes. For smaller time step sizes, the energy has much less changes, but still not a constant. 

### Symplectic Euler method

There are two important properties for a good MD integrator: perservation of the symplectic form which ensures energy conservation, and time reversibility which ensures to have reverse motions. The Euler method is not symplectic and not time reversible. A very small adjustment to the Euler method would make the method symplectic but not time reversible.

$$v_{n+1} = v_n + hf_n  $$
$$x_{n+1} = x_n + hv_{n+1}  $$
I will leave its implementation and analysis to the readers.

# Leapfrog method

The leapfrog method is a second order integrator for Newton's equation of motion. It requires to compute $v$ at half time steps. 

$$x_{n+1} = x_n + hv_{n+1/2}  $$
$$v_{n+3/2}  = v_{n+1/2}  + hf_{n+1} $$
The leapfrog algorithm requires the initial condition $x_0$ and $v_{1/2}$. The latter can be approximated by

$$ v_{1/2} = v_0 + \frac{1}{2}hf_0$$
```{r}
# Leapfrog method
leapfrog <- function(f, t0, x0, v0, t.total, dt, ...){
  # f:  2nd derivative function
  # t0: initial time
  # x0: initial position x
  # v0: initial velocity v
  # t.total: total simulation time
  # dt: time step size 
  t_all = seq(t0, t.total, by=dt)
  n_all = length(t_all)
  x_all = numeric(n_all)
  v_all = numeric(n_all) # we still save v in the same way, but it should be for n+1/2
  x_all[1] = x0
  v_all[1] = v0 + 0.5*dt*f(t_all[1], x_all[1],...)  # approximation for a half step
  for (i in 1:(n_all-1)) {
    x_all[i+1] = x_all[i] + dt * v_all[i]
    v_all[i+1] = v_all[i] + dt * f(t_all[i+1], x_all[i+1],...)
  }
  return(cbind(t_all, x_all, v_all))   # the output is a matrix of t, x, v for all time steps
}

# Run Leapfrog simulation 1, dt = 0.01
result_ho_leapfrog = leapfrog(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt1, k = k)

# Plotting the outcomes
plot(result_ho_leapfrog[,1], result_ho_leapfrog[,2], type = "l", col=2,
    xlab="t", ylab="values", xlim=c(0,100), ylim=c(-10,10)) 
lines(result_ho_leapfrog[,1] + 0.5*dt1, result_ho_leapfrog[,3], col=4) # note a half time shift for v
legend("topleft", inset=0.02, 
       legend = c("x","v"),
       col=c(2,4), lty=1, cex=0.8)

# Run leapfrog simulation 2, dt = 0.1
result_ho_leapfrog2 = leapfrog(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt2, k = k)

# Plotting the outcomes
plot(result_ho_leapfrog2[,1], result_ho_leapfrog2[,2], type = "l", col=2,
    xlab="t", ylab="values", xlim=c(0,100), ylim=c(-10,10)) 
lines(result_ho_leapfrog2[,1] + 0.5*dt2, result_ho_leapfrog2[,3], col=4) # note a half time shift for v
legend("topleft", inset=0.02, 
       legend = c("x","v"),
       col=c(2,4), lty=1, cex=0.8)

# Plotting the energy
plot(result_ho_leapfrog[,1],  energy_ho(result_ho_leapfrog, k), type = "l", col=2, 
     xlab="t", ylab="E", xlim=c(0,100), ylim=c(0,6)) 
lines(result_ho_leapfrog2[,1], energy_ho(result_ho_leapfrog2, k), col=4)
legend("topleft", inset=0.02, 
       legend = c("dt = 0.01","dt = 0.1"),
       col=c(2,4), lty=1, cex=0.8)
```

In leapfrog simulations, the massic energy does not drift even for large time step sizes. Note there is a small issue in the energy calculation, as the leapfrog method saves $x_n$ and $v_{n+1/2}$, not $x$ and $v$ for the same time points. Because of that, you will find small oscillatory deviation of massic energy when the time step size is large.

# Velocity Verlet method

The leapfrog method does not provide $x$ and $v$ for the same time points. A simple way to address that is to integrate $v$ for two half steps.

$$v_{n+1/2}  = v_{n} + \frac{1}{2}hf_n $$
$$x_{n+1} = x_n + hv_{n+1/2}  $$
$$v_{n+1}  = v_{n+1/2}  + \frac{1}{2}hf_{n+1} $$
```{r}
# Velocity Verlet method
velocity_verlet <- function(f, t0, x0, v0, t.total, dt, ...){
  # f:  2nd derivative function
  # t0: initial time
  # x0: initial position x
  # v0: initial velocity v
  # t.total: total simulation time
  # dt: time step size 
  t_all = seq(t0, t.total, by=dt)
  n_all = length(t_all)
  x_all = numeric(n_all)
  v_all = numeric(n_all)
  x_all[1] = x0
  v_all[1] = v0 
  for (i in 1:(n_all-1)) {
    v_half = v_all[i] + 0.5 * dt * f(t_all[i], x_all[i],...)
    x_all[i+1] = x_all[i] + dt * v_half
    v_all[i+1] = v_half + 0.5 * dt * f(t_all[i+1], x_all[i+1],...)
  }
  return(cbind(t_all, x_all, v_all))   # the output is a matrix of t, x, v for all time steps
}

# Run Velocity Verlet simulation 1, dt = 0.01
result_ho_vv = velocity_verlet(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt1, k = k)

# Plotting the outcomes
plot(result_ho_vv[,1], result_ho_vv[,2], type = "l", col=2,
    xlab="t", ylab="values", xlim=c(0,100), ylim=c(-10,10)) 
lines(result_ho_vv[,1], result_ho_vv[,3], col=4) # note a half time shift for v
legend("topleft", inset=0.02, 
       legend = c("x","v"),
       col=c(2,4), lty=1, cex=0.8)

# Run Velocity Verlet simulation 2, dt = 0.1
result_ho_vv2 = velocity_verlet(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt2, k = k)

# Plotting the outcomes
plot(result_ho_vv2[,1], result_ho_vv2[,2], type = "l", col=2,
    xlab="t", ylab="values", xlim=c(0,100), ylim=c(-10,10)) 
lines(result_ho_vv2[,1], result_ho_vv2[,3], col=4) # note a half time shift for v
legend("topleft", inset=0.02, 
       legend = c("x","v"),
       col=c(2,4), lty=1, cex=0.8)

# Plotting the energy
plot(result_ho_vv[,1],  energy_ho(result_ho_vv, k), type = "l", col=2, 
     xlab="t", ylab="E", xlim=c(0,100), ylim=c(0,6)) 
lines(result_ho_vv2[,1], energy_ho(result_ho_vv2, k), col=4)
legend("topleft", inset=0.02, 
       legend = c("dt = 0.01","dt = 0.1"),
       col=c(2,4), lty=1, cex=0.8)
```

With velocity Verlet, massic energy is conserved regardless of different time step sizes. 

### Position Verlet method

Similar to velocity Verlet method, one can also do position Verlet by integrating $x$ for two half steps. I will leave it for students to think about how to do it.

# Verlet method

If velocity is not needed as an output, we can use the Verlet method, which integrates only $x$.

$$x_{n+2} = 2x_{n+1} - x_{n} + h^2F_{n+1}$$
(Think why this works.)

For the first step, $x_1$ can be computed by:

$$x_1 = x_0 + hv_0 + \frac{1}{2}h^2f_0$$
```{r}
# Verlet method
verlet <- function(f, t0, x0, v0, t.total, dt, ...){
  # f:  2nd derivative function
  # t0: initial time
  # x0: initial position x
  # v0: initial velocity v
  # t.total: total simulation time
  # dt: time step size 
  t_all = seq(t0, t.total, by=dt)
  n_all = length(t_all)
  x_all = numeric(n_all)
  x_all[1] = x0
  x_all[2] = x0 + dt * v0 + 0.5 * dt * dt * f(t_all[1], x_all[1],...)
  for (i in 2:(n_all-1)) {
    x_all[i+1] = 2*x_all[i] - x_all[i-1] + dt * dt * f(t_all[i], x_all[i],...)
  }
  return(cbind(t_all, x_all))   # the output is a matrix of t, x for all time steps
}

# Run Velocity Verlet simulation 1, dt = 0.01
result_ho_v = verlet(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt1, k = k)
# Run Velocity Verlet simulation 2, dt = 0.1
result_ho_v2 = verlet(f = f_ho, t0 = 0, x0 = 1, v0 = 2, t.total = 100, dt = dt2, k = k)

# Plotting the outcomes (only x)
plot(result_ho_v[,1], result_ho_v[,2], type = "l", col=2,
    xlab="t", ylab="x", xlim=c(0,100), ylim=c(-10,10)) 
lines(result_ho_v2[,1], result_ho_v2[,2], col=4)
legend("topleft", inset=0.02, 
       legend = c("dt = 0.01","dt = 0.1"),
       col=c(2,4), lty=1, cex=0.8)
```
