---
title: "Part 6: Stochastic differential equations"
subtitle: "C. SDE integrators"
author: "Mingyang Lu"
date: "08/18/2023"
output: 
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Brownian dynamics revisit

We have simulated the Brownian motion by the model of random walk, where we find $<x>=0$ and $<x^2> \sim t$. These statistical properties allow us to describe the Brownian motion by a continuous time stochastic process, called Wiener process. This mathematical framework can be generalized to describe some other stochastic processes using stochastic differential equations (SDEs):

$$dX = f(X)dt + \sqrt{2D(X)}dW \tag{1}$$

Here, $W$ a Wiener process (continuous-time stochastic process, see the previous note *brownian.Rmd*). $f(x)$ is the derivative function, the same as what we have used in an ODE. $D(X)$ is the state-dependent diffusion coefficient (not a constant anymore) and represents the level of noises.

In physics there is another way to describe a stochastic process by a Langevin equation: 

$$\frac{dX}{dt} = f(X) + \sqrt{2D(X)}\eta(t) \tag{2}$$
,where $\eta(t)$ is a noise term. The most common type of noise is Gaussian white noise, which by definition is a random signal having equal intensity at different frequencies. $\eta(t)$ of a Gaussian white noise (essentially a Gaussian probability distribution) satisfies

\begin{equation}
\begin{cases} <\eta(t)> = 0  \\
              <\eta(t)\eta(t')> = \delta (t-t')  \tag{3} \end{cases}
\end{equation}

Here, the symbol $<.>$ means the average operation. $<\eta(t)\eta(t')>$ is a correlation function ([see here for more details](https://en.wikipedia.org/wiki/Correlation_function_(statistical_mechanics))), and \delta (t-t') is the [Dirac delta function](https://en.wikipedia.org/wiki/Dirac_delta_function) of $t-t'$. $\delta (t-t')$ is normalized, and it is nonzero only when $t-t'$ approaches to zero. $\eta(t)$ of a Gaussian white noise is closely related to the Wiener process.

# Euler–Maruyama method

We can numerically solve the SDE in Equation (1) by

$$X_{n+1} = X_n + f(X_n)\Delta t + \sqrt{2D(X_n)} \Delta W_n \tag{4}$$
$\Delta W_n$ represents normal random variables with mean zero and variance $\Delta t$. ($<\Delta W_n^2> = \Delta t$) This is called Euler-Maruyama method, an extension of the Euler method for ODEs to SDEs. Now, we show the implementation in R. 

```{r}
#Euler-Maruyama method for one-variable SDE
SDE_Euler <- function(derivs, noise, t0, X0, t.total, dt, ...){
  # derivs: the function of the derivatives 
  # noise: the function of the noise term 
  # t0: initial time
  # X0: initial condition
  # t.total: total simulation time
  # dt: time step size 
  t_all = seq(t0, t.total, by=dt)
  n_all = length(t_all)
  X_all = numeric(n_all)
  X_all[1] = X0
  dW_all = rnorm(n_all, mean=0, sd=sqrt(dt))  # all dW terms, rnorm generate random variables from normal distribution
  for (i in 1:(n_all-1)) {
    t = t_all[i]
    X = X_all[i]
    X_all[i+1] = X + dt * derivs(t, X, ...) + dW_all[i] * noise(t, X, ...)
  }
  return(cbind(t_all, X_all))   # the output is a matrix of t & X(t) for all time steps
}
```

Note that in the implementation, all random variables $\Delta W$ are generated altogether. This step is more efficient in R, compared to the way to generate them in each step. However, it requires a lot more memory to do so, therefore not ideal for long simulations. The best approach is to use compiled functions with Fortran and C.

A common mistake by a beginner is to add $\sqrt{2D(X)}$ to the $f(X)$ term and integrate with the ODE version of the Euler method. This is wrong, as the variance of the random variables is now $\Delta t^2$ instead of $\Delta t$. 

Now we will apply the Euler-Maruyama method to a few SDEs. We first consider a simple Brownian motion, where $f(X) = 0$.

``` {r}
f_bm <- function(t, X, D) return(0)
noise_bm <- function(t, X, D) return(sqrt(2*D)) # D is a variable to represent noise levels, 
set.seed(1)

plot(NULL, xlab="t", ylab="X", xlim=c(0,200), ylim=c(-150,150)) 
for(i in seq_len(10)) {
  results = SDE_Euler(derivs = f_bm, noise = noise_bm, t0 = 0, X0 = 0, t.total = 200, dt = 0.01,  D = 10)
  lines(results[,1], results[,2],  col=i)
}
```

The integration of the SDE gives the same results as our previous simulations of the Brownian motion.

Let's consider another example: Brownian dynamics under a harmonic potential and constant noise. Here, the Brownian particle is tethered to a spring. According to Hooke's law, $f(X) = - kX$. $k$ is the spring constant. This type of system is also called Ornstein–Uhlenbeck process. 

$$dX = -kXdt + \sqrt{2D}dW \tag{5}$$

```{r}
f_ou <- function(t, X, k, D) return(-k*X)
noise_ou <- function(t, X, k, D) return(sqrt(2*D)) 

set.seed(1)    # Students can try different RNG seed
D_all = c(100, 25, 1)
results1 = SDE_Euler(derivs = f_ou, noise = noise_ou, t0 = 0, X0 = 0, t.total = 1000, dt = 0.01, k = 1, D = D_all[1])
results2 = SDE_Euler(derivs = f_ou, noise = noise_ou, t0 = 0, X0 = 0, t.total = 1000, dt = 0.01, k = 1, D = D_all[2])
results3 = SDE_Euler(derivs = f_ou, noise = noise_ou, t0 = 0, X0 = 0, t.total = 1000, dt = 0.01, k = 1, D = D_all[3])

plot(NULL, xlab="t", ylab="X", xlim=c(0,100), ylim=c(-30,30)) 
lines(results1[,1], results1[,2],  col=2)
lines(results2[,1], results2[,2],  col=3)
lines(results3[,1], results3[,2],  col=4)
legend("bottomright", inset=0.02, title="D", 
       legend = c(paste0(D_all)),
       col=2:4, lty=1, cex=0.8)
```

In the above simulations, we plot the stochastic dynamics for three different noise levels. Unlike the Brownian motion, the dynamics of $X$ fluctuate around 0 with deviations proportional to the noise level $D$. The larger the $D$, the noisier the $X$ is: $<x^2>=\frac{D}{k}$.

``` {r}
plot(D_all, c(var(results1[,2]), var(results2[,2]), var(results3[,2])), type = "p", xlab = "D", ylab = "Variance")
abline(a=0, b = 1, col = 1)
```

In the above test, we find that the simulation results still slightly deviate from the theoretical values. That is because the simulation is not long enough to reach robust statistics. 

# Milstein method

The Milstein method is a more accurate SDE integrator. Consider an SDE in the following form,

$$dX = f(X)dt + s(X)dW \tag{6}$$
Milstein method integrates SDE by

$$X_{n+1} = X_n + f(X_n)\Delta t + s(X_n) \Delta W_n + \frac{1}{2}s(X_n)s'(X_n)((\Delta W_n)^2-\Delta t) \tag{7}$$

, where $s'$ denotes the derivative of s(X) with respect to $X$. Here, we consider $s(X)=\sqrt{2D(X)}$. The derivation of Equation (1) is by stochastic Taylor expansion. When $D$ is a constant, $s'(X) = 0$. Thus, the Milstein method is equivalent to the Euler-Maruyama method. The Milstein method converges faster than the Euler-Maruyama method for the case in which the noise depends on the state $X$.

``` {r}
# An SDE integrator by the Milstein method
SDE_Milstein <- function(derivs, noise, dnoise, t0, X0, t.total, dt, ...){
  # derivs: the function of the derivatives 
  # noise: the function of the noise term 
  # dnoise: the function of the derivative of the noise with respect to X
  # t0: initial time
  # X0: initial condition
  # t.total: total simulation time
  # dt: time step size 
  t_all = seq(t0, t.total, by=dt)
  n_all = length(t_all)
  X_all = numeric(n_all)
  X_all[1] = X0
  dW_all = rnorm(n_all, mean=0, sd=sqrt(dt))  # all dW terms, rnorm generate random variables from normal distribution
  for (i in 1:(n_all-1)) {
    t = t_all[i]
    X = X_all[i]
    s = noise(t, X, ...)
    dW = dW_all[i]
    X_all[i+1] = X + dt * derivs(t, X, ...) + dW * s  + s*dnoise(t, X, ...)/2*(dW**2-dt)
  }
  return(cbind(t_all, X_all))   # the output is a matrix of t & X(t) for all time steps
}
```

# Gene expression noise in a gene circuit

We consider a circuit of a self-activating gene $X$ with a noise term $s(X) = b\sqrt{X}$.

$dX = [g_0 + g_1 \frac{(X/X_0)^{n_X}}{1+(X/X_0)^{n_X}} - kX]dt + b\sqrt{X}dW \tag{8}$

In Equation (8), the standard deviation of the noise is proportional to $\sqrt{X}$. The noise level becomes larger for larger $X$. But when $X$ approaches to zero, the noise level is close to zero. A similar SDE has been adopted to model the stochastic evolution of interest rates by the Cox-Ingersoll-Ross model in finance. 

```{r}
hill_ex <- function(X,X0,n) {
  a = (X/X0)**n
  return(a/(1+a))
}
f_1g <- function(t, X, b) return(10 + 45 * hill_ex(X, 200, 4) - 0.15*X)  # same in Part 2C
noise_1g <- function(t, X, b) return(b*sqrt(X)) 
dnoise_1g <- function(t, X, b) return(b/sqrt(X)/2) 

b_all = c(5, 2, 0.5)
set.seed(1)
results1 = SDE_Milstein(derivs = f_1g, noise = noise_1g, dnoise = dnoise_1g, 
                        t0 = 0, X0 = 300, t.total = 1000, dt = 0.01, b = b_all[1])
results2 = SDE_Milstein(derivs = f_1g, noise = noise_1g, dnoise = dnoise_1g, 
                        t0 = 0, X0 = 300, t.total = 1000, dt = 0.01, b = b_all[2])
results3 = SDE_Milstein(derivs = f_1g, noise = noise_1g, dnoise = dnoise_1g, 
                        t0 = 0, X0 = 300, t.total = 1000, dt = 0.01, b = b_all[3])

plot(NULL, xlab="t", ylab="X", xlim=c(0,1000), ylim=c(0,700)) 
lines(results1[,1], results1[,2],  col=2)
lines(results2[,1], results2[,2],  col=3)
lines(results3[,1], results3[,2],  col=4)
legend("topright", inset=0.02, title="b", 
       legend = c(paste0(b_all)),
       col=2:4, lty=1, cex=0.8)
```

The circuit of a self-activating gene has two stable steady states: one near $X = 100$ and the other near $X = 300$. The steady states remains in the stochastic dynamics. One interesting feature of stochastic dynamics of multi-stable systems is dynamical transition between steady states. When the noise is low ($b = 0.5$), $X$ only slightly fluctuates around one stable steady state, and the transition to the other stable steady state is unlikely. When the noise becomes larger ($b = 2$), the circuit dynamics start to exhibit transitions between the two states. When the noise becomes very large ($b = 5$), the state transitions become much frequently. 

Lastly, we consider the same gene circuit but a constant noise. In this case, the Euler-Maruyama method gives the same results as the Milstein method.

``` {r}
noise_1g_2 <- function(t, X, b) return(b) 
dnoise_1g_2 <- function(t, X, b) return(0) 

b_all = c(30, 20, 10)
set.seed(1)
results1 = SDE_Milstein(derivs = f_1g, noise = noise_1g_2, dnoise = dnoise_1g_2, 
                        t0 = 0, X0 = 300, t.total = 1000, dt = 0.1, b = b_all[1])
results2 = SDE_Milstein(derivs = f_1g, noise = noise_1g_2, dnoise = dnoise_1g_2, 
                        t0 = 0, X0 = 300, t.total = 1000, dt = 0.1, b = b_all[2])
results3 = SDE_Milstein(derivs = f_1g, noise = noise_1g_2, dnoise = dnoise_1g_2, 
                        t0 = 0, X0 = 300, t.total = 1000, dt = 0.1, b = b_all[3])

plot(NULL, xlab="t", ylab="X", xlim=c(0,1000), ylim=c(0,700)) 
lines(results1[,1], results1[,2],  col=2)
lines(results2[,1], results2[,2],  col=3)
lines(results3[,1], results3[,2],  col=4)
legend("topright", inset=0.02, title="b", 
       legend = c(paste0(b_all)),
       col=2:4, lty=1, cex=0.8)
```

As shown in the above plot, the stochastic dynamics look very different from the previous simulations. (Of course, we are not able to set comparable noise levels for the two cases; thus direct comparison is hard.) However, there are at least two clear differences. 

(1) Compared to the case of the state-dependent noise, the circuit with constant noise has larger $X$ fluctuation around the low expression state and has smaller $X$ fluctuation around the high expression state. 

(2) When the circuit is at the low expression state, high noise would result in $X$ below zero. That is because of the discretization during the SDE integration. Making the time step size $dt$ smaller may help to alleviate this issue. 
