---
title: "Part 7: Partial differential equations"
subtitle: "B. Reaction-diffusion systems"
author: "Mingyang Lu"
date: "2/14/2023"
output: 
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reaction-diffusion equations

If we now use $u(x,t)$ as the state variable and add another term $f(u,t)$ to the diffusion equation, 

$$\frac{\partial u(X,t)}{\partial t} = f(u,t) + D \frac{\partial^2 u(X,t)}{\partial X^2} \tag{1} $$

Equation (1) is called reaction-diffusion equation. $f(u, t)$ represents the reaction term, which we commonly used in a chemical rate equation

$$\frac{du}{dt} = f(u, t) \tag{2}$$

When $f(u, t)$ vanishes, Equation (1) becomes the PDE for a pure diffusion process. 

# Fisher's equation

If we consider a logistic growth as the reaction, Equation (2) for $u(X, t)$ becomes 

$$\frac{\partial u}{\partial t} = ru(1-\frac{u}{B}) + D \frac{\partial^2 u}{\partial X^2} \tag{3} $$

$u$ is the population size, and $B$ is the carrying capacity. This is Fisher's equation, first proposed by Ronald Fisher in 1934. Below is a slightly modified PDE integrator to model reaction-diffusion systems. We use the Dirichlet boundary condition here (zero $u$ at the boundary).

```{r}
## PDE integration with the finite difference method for reaction-diffusion equation 
pde_fd_reaction_diffusion <- function(derivs, ngrid, dX, dt, D, t0, t.total, p0, ...) {
  #derivs: derivative function
  #ngrid: number of grid points
  #dX: X step size
  #dt: time step size
  #D: diffusion constant
  #t0: initial time
  #t.total: total simulation time
  #p0: initial condition: P(X,t = t0)
  
  t_all = seq(t0, t.total, by = dt)
  nt_all = length(t_all)
  factor = D/dX**2*dt
  p = p0
  for (i in seq_len(nt_all-1)){
    p_plus_one = c(p[-1], 0) # a vector, P(X+dX) for all Xs, 0 is added to specify the boundary conditions
    p_minus_one = c(0, p[1:(ngrid-1)])  # a vector, P(X-dX) for all Xs, 0 is added to specify the boundary conditions
    f = sapply(p, function(X) return (derivs(t_all[i],X, ...)))
    p = p + f*dt + factor * (p_plus_one + p_minus_one - 2 * p) # finite difference to update all P(X,t)
  }
  return(p)
}
```

Here, we set a constant $u$ at some center grids as the initial condition of the system. We can observe a bidirectional traveling wave at a constant velocity until the wave reaches to the boundary. 

```{r}
len = 100; dX = 1; dt = 0.01; D = 1
r = 0.64; B = 75
ngrid = as.integer(len/dX)+1  # number of grid points
X_all = c(0:(ngrid-1))*dX  # X values for each grid point

u0 = numeric(ngrid)
u0[(ngrid/2 - 2): (ngrid/2+2)] = 50

fisher <- function(t, u, r, B) return(r*u*(1-u/B))

plot(X_all, u0, type="l", col=1, xlab="X", ylab="u", xlim=c(0,100), ylim=c(0,100)) 

u_1 = u0
for (i in 1:5){
   u_1 = pde_fd_reaction_diffusion(derivs = fisher, ngrid = ngrid, dX = dX, dt = dt, D = D, 
                                t0 = 0, t.total = 5, p0 = u_1, r = r, B = B) 
   lines(X_all,u_1, col=i+1)
}
legend("topright", title = "Iterations", inset=0.02, 
       legend = paste0(1:6),
       col=1:6, lty=1, cex=0.8)
```

For a constant initial condition from the very left end of the system, the traveling wave is towards the right direction.

```{r}
len = 100; dX = 1; dt = 0.01; D = 1
r = 0.64; B = 75
ngrid = as.integer(len/dX)+1  # number of grid points
X_all = c(0:(ngrid-1))*dX  # X values for each grid point

u0 = numeric(ngrid) 
u0[1:5] = 50  # five grids with a constant u

fisher <- function(t, u, r, B) return(r*u*(1-u/B))

plot(X_all, u0, type="l", col=1, xlab="X", ylab="u", xlim=c(0,100), ylim=c(0,100)) 

u_1 = u0
for (i in 1:10){
   u_1 = pde_fd_reaction_diffusion(derivs = fisher, ngrid = ngrid, dX = dX, dt = dt, D = D, 
                                t0 = 0, t.total = 5, p0 = u_1, r = r, B = B) 
   lines(X_all,u_1, col=i+1)
}
legend("topright", title = "Iterations", inset=0.02, 
       legend = paste0(1:11),
       col=1:11, lty=1, cex=0.8)
```

Ronald Fisher used the Fisher's equation to explain the spread of advantage genes in a population. His study has become the foundation of population genetics. In the model of genetics, $u$ (after a variable scaling to remove $B$) represents the frequency of the mutant gene, $x$ is the spatial location, $r$ is the intensity of the selection in favor of the mutant, and $D$ is the rate of diffusion per generation. This model exhibits a traveling wave, which explains the spread of the advantage genes within a population.

# Fokker-Planck equation

A stochastic differential equation (SDE) in the following form

$$\frac{dX}{dt} = f(X,t) + \sqrt{2D(X,t)}\eta(t) \tag{4} $$ 
, where $\eta (t)$ represents a Gaussian white noise, is equivalent to the Fokker-Planck equation:

$$\frac{\partial P}{\partial t} = -\frac{\partial (fP)}{\partial X}  + \frac{\partial^2 (DP)}{\partial X^2} \tag{5} $$
A Fokker-Planck equation is similar to a reaction-diffusion equation, but the variable $P$ represents the probability as the function of $X$ and $t$. The diffusion equation is also a special case of the Fokker-Planck equation when $f=0$.

Similar to the diffusion equation, a Fokker-Planck equation can also be integrated by the finite difference method. From Equations (6-7) in [Part 05A](./05A.html), for any function $F(X,t)$, we have

$$  \frac{F(X + \Delta X, t) - F(X-\Delta X, t)}{2\Delta X} = \frac{\partial F(X,t)}{\partial X} + O(\Delta X^2) \tag{6}$$
This demonstrates that $\frac{\partial F(X,t)}{\partial X}$ can be approximated by  $\frac{F(X + \Delta X, t) - F(X-\Delta X, t)}{2\Delta X}$. Thus, the PDE integration can be achieved by the following formula (for simplification, we assume $D(X,t)$ as a constant):

\begin{equation}
\begin{aligned}
P(X, t+\Delta t) = P(X, t) &- \frac{f(X + \Delta X, t)P(X + \Delta X, t) - f(X-\Delta X, t)P(X-\Delta X, t)}{2\Delta X} \Delta t \\
                           &+ D \frac{P(X + \Delta X, t) + P(X-\Delta X, t) - 2P(X, t)}{\Delta X^2} \Delta t
\end{aligned} \tag{7}
\end{equation}
 
```{r}
## PDE integration with the finite difference method for a generic Fokker-Planck equation (constant D)
pde_fd_fokker_planck <- function(derivs, ngrid, X_all, dX, dt, D, t0, t.total, p0, ...) {
  #derivs: derivative function
  #ngrid: number of grid points
  #X_all: X for all grid points
  #dX: X step size
  #dt: time step size
  #D: diffusion constant
  #t0: initial time
  #t.total: total simulation time
  #p0: initial condition: P(X,t = t0)
  
  t_all = seq(t0, t.total, by = dt)
  nt_all = length(t_all)
  factor = D/dX**2*dt # scaling factor for the diffusion term
  factor_2 = dt/2/dX # sacling factor for the drift term
  p = p0
  for (i in seq_len(nt_all-1)){
    p_plus_one = c(p[-1], 0) # a vector, P(X+dX) for all Xs, 0 is added to specify the boundary conditions
    p_minus_one = c(0, p[1:(ngrid-1)])  # a vector, P(X-dX) for all Xs, 0 is added to specify the boundary conditions
    f = sapply(X_all, function(X) return (derivs(t_all[i],X, ...)))
    fp = f*p
    fp_plus_one = c(fp[-1],0)
    fp_minus_one = c(0, fp[1:(ngrid-1)])
    p = p - factor_2 * (fp_plus_one - fp_minus_one)+ factor * (p_plus_one + p_minus_one - 2 * p)
  }
  return(p)
}
```

We applied the above function on Brownian dynamics under a harmonic potential and constant noise (Ornstein–Uhlenbeck process). 

$$\frac{\partial P}{\partial t} = -\frac{\partial (-kXP)}{\partial X}  + D\frac{\partial^2 P}{\partial X^2} \tag{8}$$

```{r}
len = 100; dX = 1; dt = 0.01; D = 1; k = 0.01
ngrid = as.integer(len/dX)+1  # number of grid points
X_all = (c(1:ngrid) - (ngrid+1)/2)*dX  # X values for each grid point

u0 = numeric(ngrid)
u0[(ngrid/2 - 2): (ngrid/2+2)] = 1
u0 = u0/sum(u0)

spring <- function(t, X, k) return(-k*X)

plot(X_all, u0, type="l", col=1, xlab="X", ylab="u", xlim=c(-50,50), ylim=c(0,0.1)) 
u_1 = u0
for (i in 1:5){
   u_1 = pde_fd_fokker_planck(derivs = spring, ngrid = ngrid, X_all = X_all, dX = dX, dt = dt, D = D, 
                                t0 = 0, t.total = 40, p0 = u_1, k = k) 
   lines(X_all,u_1, col=i+1)
}
points(X_all, sqrt(k/2/pi/D)*exp(-k*X_all**2/2/D), col = 1)
legend("topright", title = "Iterations", inset=0.02, 
       legend = paste0(1:6),
       col=1:6, lty=1, cex=0.8)
```

In the above simulation, the $P$ is localized at the center $X$ initially and converge to a Gaussian distribution. When using a stiffer spring, $P$ converges to a narrower Gaussian. It can be shown the steady state solution of Equation (8) ($\frac{\partial P}{\partial t} =0$) is

$$P_{ss}(X) = \sqrt{\frac{k}{2\pi D}}e^{-\frac{kX^2}{2D}} \tag{9}$$
In the plot, the dots represent the theoretical values of $P_{ss}(X)$.

```{r}
len = 100; dX = 1; dt = 0.01; D = 1; k = 0.03
ngrid = as.integer(len/dX)+1  # number of grid points
X_all = (c(1:ngrid) - (ngrid+1)/2)*dX  # X values for each grid point

u0 = numeric(ngrid)
u0[(ngrid/2 - 2): (ngrid/2+2)] = 1
u0 = u0/sum(u0)

spring <- function(t, X, k) return(-k*X)

plot(X_all, u0, type="l", col=1, xlab="X", ylab="u", xlim=c(-50,50), ylim=c(0,0.1)) 
u_1 = u0
for (i in 1:5){
   u_1 = pde_fd_fokker_planck(derivs = spring, ngrid = ngrid, X_all = X_all, dX = dX, dt = dt, D = D, 
                                t0 = 0, t.total = 20, p0 = u_1, k = k) 
   lines(X_all,u_1, col=i+1)
}
points(X_all, sqrt(k/2/pi/D)*exp(-k*X_all**2/2/D), col = 1)
legend("topright", title = "Iterations", inset=0.02, 
       legend = paste0(1:6),
       col=1:6, lty=1, cex=0.8)
```

$P$ also converges to the same distribution when starting from a different initial condition (uniform distribution).

```{r}
len = 100; dX = 1; dt = 0.01; D = 1; k = 0.03
ngrid = as.integer(len/dX)+1  # number of grid points
X_all = (c(1:ngrid) - (ngrid+1)/2)*dX  # X values for each grid point

u0 = rep(1, ngrid)
u0 = u0/sum(u0)

spring <- function(t, X, k) return(-k*X)

plot(X_all, u0, type="l", col=1, xlab="X", ylab="u", xlim=c(-50,50), ylim=c(0,0.1)) 
u_1 = u0
for (i in 1:5){
   u_1 = pde_fd_fokker_planck(derivs = spring, ngrid = ngrid, X_all = X_all, dX = dX, dt = dt, D = D, 
                                t0 = 0, t.total = 30, p0 = u_1, k = k) 
   lines(X_all,u_1, col=i+1)
}
points(X_all, sqrt(k/2/pi/D)*exp(-k*X_all**2/2/D), col = 1)
legend("topright", title = "Iterations", inset=0.02, 
       legend = paste0(1:6),
       col=1:6, lty=1, cex=0.8)
```

Q: How does $P(X,t)$ change when using Neumann boundary condition? What happens when $D$ is a function of $X$ (*i.e.*, state-dependent)?