---
title: "Part 8: Monte Carlo Simulations"
subtitle: "E. Exercises"
author: "Mingyang Lu"
date: "08/22/2023"
output: 
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Sampling a mixture of bivariate Gaussian distributions

We wish to simulate a set of flow cytometry data for a mixture of cells with two distinct cellular states. Let's consider the following 2D distribution (not normalized) $p(x, y)$, consisting of two bivariate Gaussian distributions. $x$ and $y$ measure the expression levels of two proteins for individual cells.

$$p(x,y) = e^{-(x + 10y - 1)^2 - (x - 0.1y + 2)^2} + e^{-(x - 5y + 2)^2 - (x + 0.2y - 1)^2}. \tag{1}$$

The following R function *p_mixture* computes $p(x,y)$ (the variables $x$ and $y$ are inputted in a vector of size 2).

```{r,fig.width = 6.5, fig.height = 6}
library(ggplot2)
p_mixture <- function(x){
  f1 = x[1]+10*x[2]-1
  f2 = x[1]-0.1*x[2]+2
  f3 = x[1]-5*x[2] + 2
  f4 = x[1]+0.2*x[2] - 1
  return(exp(-f1**2-f2**2) + exp(-f3**2-f4**2))
}

x_all = seq(-5,5, by =0.05)
y_all = seq(-2,2, by =0.01)
data = expand.grid(x_all, y_all)
colnames(data) = c("x", "y")
z_all = apply(data, MARGIN = 1, p_mixture)
data$z = z_all
ggplot(data, aes(x=x, y=y, z = z)) + geom_contour(breaks = 10^(seq(-2,0, by =0.3)), colour = "black")
```

Design a Metropolis-Hastings algorithm to sample the distribution. Show your simulation results by plotting the density map of the sampled data points.

Hint:

(1) Think how to devise an appropriate local move to sample $x$ and $y$.

(2) You can use *geom_bin2d* from the ggplot2 library to plot the density map.

# 2. Stochastic SIR model. 

Consider the following reaction network:

$$S + I {\stackrel{\beta}{\rightarrow}} 2I$$

$$I {\stackrel{\nu}{\rightarrow}} R $$

where $\beta = 0.003$ and $\nu = 1$.

**(a)** We consider the initial condition $S(t=0) = 1000$, $I(t=0) = 1$, and $R(t=0) = 0$. Compute $R_0$. Perform a Gillespie simulation and plot the time dynamics of $S$, $I$ and $R$. Please read the "Gillespie_tutorial.rmd" R Markdown for other modeling examples. You can directly use the functions *gillespie_step* and *ssa*, and you will need to write your own propensity function and stoichiometry function in the following format:

```{r}
my_propensity <- function(Xs) {   # Xs: current state; you can also supply parameters
  
  #output: a vector of propensity for all reactions
}

my_stoichiometry <- function(r_ind) {   # r_ind: the choice of the reaction
  
  # output: the stoichiometry vector for the reaction r_ind
}
```

**(b)** Perform a deterministic simulation of the same model (check the course PPTs for modeling epidemiology). Visually compare the deterministic and stochastic dynamics. For ODE integration, you can use the following RK4 function

```{r}
RK4_generic <- function(derivs, X0, n, t.total, dt, ...){ # Runge-Kutta 4th order, n variables, passing parameters via ellipsis
  # derivs: the function of the derivatives 
  # X0: initial condition, a vector of two elements
  # n: total number of variables 
  # t.total: total simulation time, assuming t starts from 0 at the beginning
  # dt: time step size 
  t_all = seq(0, t.total, by=dt)
  n_all = length(t_all)
  X_all = matrix(0, nrow = n_all, ncol = n)
  X_all[1,] = X0
  for (i in 1:(n_all-1)) {
    t_0= t_all[i]
    t_0.5 = t_0 + 0.5*dt
    t_1 = t_0 + dt
    k1 = dt * derivs(t_0,X_all[i,], ...)
    k2 = dt * derivs(t_0.5,X_all[i,] + k1/2, ...)
    k3 = dt * derivs(t_0.5,X_all[i,] + k2/2, ...)
    k4 = dt * derivs(t_1,X_all[i,] + k3, ...)
    X_all[i+1,] = X_all[i,] + (k1+2*k2+2*k3+k4)/6
  }
  return(cbind(t_all, X_all))   # the output is a matrix of t & X(t) for all time steps
}
```

**(c)** Now we consider another set of parameters $\beta = 0.3$ and $\nu = 1$, and a new initial condition $S(t=0) = 10$, $I(t=0) = 1$, and $R(t=0) = 0$. You will find the same $R_0$ under this condition, thus similar deterministic dynamics. Perform the Gillespie simulation (try stochastic simulations multiple times, each with different random number seeds) and deterministic simulations again. Compare your results. 

If you try a sufficient large number of simulations, you will find a variety of SIR dynamics:

**(1)** dynamics similar to the deterministic dynamics: with a peak in $I(t)$, no $I$ eventually, but with positive $S$.

**(2)** $I(t)$ dies out at the very beginning;

**(3)** $I(t)$ generates a peak, but $I(t)$ dies out earlier and there are a sufficiently large number of $S$ left.

**(4)** $I(t)$ generates a peak, but eventually there is no $S$ and $I$ left. 
