---
title: "10A. Dimensionality reduction"
author: "Mingyang Lu"
old_id: "10A"
---

Biological data are often high-dimensional: a gene-expression experiment measures
thousands of genes across many samples. To visualize and interpret such data we reduce
their dimensionality, mapping each sample to a few coordinates that capture most of the
structure. This chapter builds *principal component analysis* (PCA) from scratch, uses
it on a real gene-expression dataset, and then demonstrates nonlinear embeddings (MDS,
t-SNE, and UMAP).

## 10A.1 Projecting data onto a line

The simplest reduction takes two-dimensional points and describes them by a single
number. Consider 100 points scattered in a narrow, tilted strip.

### R implementation {.impl}

```{r}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
set.seed(1)
x <- runif(100); y <- 0.4 * x + 0.8 + 0.1 * runif(100)
data1 <- data.frame(x = x, y = y)

# fit a line, then project each point onto it
model <- lm(y ~ x, data = data1)
x0 <- c(0, coef(model)[1])                 # a point on the line
dir <- c(1, coef(model)[2]); dir <- dir / sqrt(sum(dir^2))   # unit direction
proj <- as.matrix(data1) %*% dir - sum(x0 * dir)            # 1D coordinate along the line
projected <- t(sapply(proj, function(p) p * dir + x0))      # back to 2D

plot(data1$x, data1$y, xlab = "x", ylab = "y", xlim = c(0, 1), ylim = c(0.5, 1.5), asp = 1)
abline(model, col = 2)
points(projected[, 1], projected[, 2], pch = 16, col = 4)
```

### Python implementation {.impl}

```{python}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
import numpy as np
import matplotlib.pyplot as plt

rng = np.random.default_rng(1)
x = rng.uniform(size=100); y = 0.4 * x + 0.8 + 0.1 * rng.uniform(size=100)
data1 = np.column_stack([x, y])

# fit a line, then project each point onto it
b1, b0 = np.polyfit(x, y, 1)
x0 = np.array([0, b0])
d = np.array([1, b1]); d = d / np.linalg.norm(d)            # unit direction
proj = data1 @ d - x0 @ d                                   # 1D coordinate along the line
projected = np.outer(proj, d) + x0                          # back to 2D

plt.figure(figsize=(5, 5))
plt.scatter(x, y)
xs = np.array([0, 1]); plt.plot(xs, b1 * xs + b0, color="red")
plt.scatter(projected[:, 0], projected[:, 1], color="blue", s=15)
plt.gca().set_aspect("equal"); plt.xlabel("x"); plt.ylabel("y")
plt.xlim(0, 1); plt.ylim(0.5, 1.5); plt.show()
```

Each blue point is the closest position on the line to its original point. The single
coordinate along the line (the projection) keeps most of the spread of the data, since
the points barely deviate from the line. PCA generalizes this idea: it finds the
directions of largest variance automatically, without being told to use a regression
line, and works in any number of dimensions.

## 10A.2 Principal component analysis from scratch

PCA re-expresses the data in a new orthogonal basis, the *principal components*, ordered
by how much variance they capture [@pearson1901lines; @hotelling1933analysis]. The
recipe is: center the data, then find the directions along which it varies most. Those
directions are the eigenvectors of the covariance matrix, and their eigenvalues give the
variance captured. We implement it two equivalent ways: eigen-decomposition of the
covariance matrix, and the *singular value decomposition* (SVD) of the centered data,
which is numerically preferable and much cheaper when there are more features than
samples.

::: {.callout-note title="Recipe 10A.2.1"}
**Objective.** Compute principal components from scratch and confirm two routes agree.

**Model.** The input is a data matrix X with samples in rows and features in columns.

**Method.** Principal component analysis two equivalent ways: the eigen-decomposition of the covariance matrix, and the singular value decomposition of the centered data.

**Test.** The 100x2 "strip" data from 10A.1 (points scattered along a line); centering only, no scaling.

**Verification.** The two routes return identical variances, and on the strip data PC1 carries almost all the variance and points along the strip, recovering the regression direction of 10A.1.
:::

### R implementation {.impl}

```{r}
# PCA from scratch: "eigen" (covariance eigen-decomposition) or "svd" (of centered data)
pca <- function(X, method = "svd") {
  Xc <- scale(X, center = TRUE, scale = FALSE)   # center each feature
  n <- nrow(X)
  if (method == "eigen") {
    e <- eigen(cov(Xc)); values <- e$values; loadings <- e$vectors
  } else {
    s <- svd(Xc); values <- s$d^2 / (n - 1); loadings <- s$v
  }
  list(values = values, loadings = loadings, scores = Xc %*% loadings)
}

# the two routes agree (on the strip data from 10A.1)
X2 <- as.matrix(data1)
round(rbind(eigen = pca(X2, "eigen")$values, svd = pca(X2, "svd")$values), 5)
```

### Python implementation {.impl}

```{python}
# PCA from scratch: "eigen" (covariance eigen-decomposition) or "svd" (of centered data)
def pca(X, method="svd"):
    Xc = X - X.mean(0)                              # center each feature
    n = len(X)
    if method == "eigen":
        vals, vecs = np.linalg.eigh(np.cov(Xc, rowvar=False))
        order = np.argsort(vals)[::-1]
        values, loadings = vals[order], vecs[:, order]
    else:
        U, s, Vt = np.linalg.svd(Xc, full_matrices=False)
        values, loadings = s**2 / (n - 1), Vt.T
    return {"values": values, "loadings": loadings, "scores": Xc @ loadings}

# the two routes agree (on the strip data from 10A.1)
print(np.round(np.vstack([pca(data1, "eigen")["values"], pca(data1, "svd")["values"]]), 5))
```

Both routes return the same variances. For the strip data the first principal component
carries almost all the variance and points along the strip, recovering the direction we
fit by hand in the previous section.

## 10A.3 Interpreting components with loadings

We now apply PCA to a real dataset: expression of 500 genes measured across 26 samples
spanning nine experimental conditions (A through I). The samples are the observations
and the genes are the features, so we transpose the table. The *loadings* (the columns
of the eigenvector matrix) say how much each gene contributes to each component;
inspecting the largest loadings tells us which genes drive the dominant patterns.

### R implementation {.impl}

```{r}
#| fig-width: 7
#| fig-height: 4
#| out-width: "70%"
exp_data <- read.csv("10-high-dim-data/data/exp_data.csv", row.names = 1)
X <- t(as.matrix(exp_data))                        # 26 samples x 500 genes
genes <- colnames(X)

res <- pca(X)                                      # used again in 10A.4
top <- order(abs(res$loadings[, 1]), decreasing = TRUE)[1:10]   # top PC1 genes
barplot(res$loadings[top, 1], names.arg = genes[top], las = 2, ylab = "PC1 loading")
```

```{r}
#| fig-width: 5.5
#| fig-height: 5.5
#| out-width: "60%"
# loadings plot: every gene's PC1 vs PC2 contribution, with the top PC1 genes marked
plot(res$loadings[, 1], res$loadings[, 2], pch = 20, col = "grey70",
     xlab = "PC1 loading", ylab = "PC2 loading")
abline(h = 0, v = 0, lty = 3, col = "grey50")
points(res$loadings[top, 1], res$loadings[top, 2], pch = 19, col = 2)
text(res$loadings[top, 1], res$loadings[top, 2], labels = genes[top], pos = 3, cex = 0.6)
```

### Python implementation {.impl}

```{python}
#| fig-width: 7
#| fig-height: 4
#| out-width: "70%"
import pandas as pd
exp_data = pd.read_csv("10-high-dim-data/data/exp_data.csv", index_col=0)
X = exp_data.values.T                              # 26 samples x 500 genes
genes = exp_data.index.to_numpy()
samples = exp_data.columns.to_numpy()

res = pca(X)                                       # used again in 10A.4
top = np.argsort(np.abs(res["loadings"][:, 0]))[::-1][:10]      # top PC1 genes
plt.figure(figsize=(7, 4))
plt.bar(genes[top], res["loadings"][top, 0])
plt.xticks(rotation=90); plt.ylabel("PC1 loading")
plt.title("Genes contributing most to PC1"); plt.tight_layout(); plt.show()
```

```{python}
#| fig-width: 5.5
#| fig-height: 5.5
#| out-width: "60%"
# loadings plot: every gene's PC1 vs PC2 contribution, with the top PC1 genes marked
plt.figure(figsize=(5.5, 5.5))
plt.scatter(res["loadings"][:, 0], res["loadings"][:, 1], color="grey", s=12)
plt.axhline(0, color="grey", lw=0.7, ls=":"); plt.axvline(0, color="grey", lw=0.7, ls=":")
plt.scatter(res["loadings"][top, 0], res["loadings"][top, 1], color="red", s=25)
for i in top:
    plt.text(res["loadings"][i, 0], res["loadings"][i, 1], genes[i], fontsize=6)
plt.xlabel("PC1 loading"); plt.ylabel("PC2 loading"); plt.title("Gene loadings"); plt.show()
```

Each principal component is a weighted combination of all 500 genes. The handful of
genes with the largest loadings dominate that component, so PC1 can be read as the
coordinated expression pattern of these top genes, the axis along which the samples vary
most. Plotting each gene's PC1 and PC2 loadings together shows most of the 500 genes
clustered near the origin, contributing little to either component, while the ten genes
highlighted above stand out along PC1; several of them also carry a substantial PC2
loading, showing that a gene's influence is not always confined to a single component.

## 10A.4 Visualizing the samples

With the same PCA, we look at the sample side. The *scree plot* shows how much variance
each component explains, guiding how many to keep. Projecting the samples onto the first
two components then reveals their structure.

::: {.callout-note title="Recipe 10A.4.1"}
**Objective.** Project samples onto their leading principal components and summarize the trend with a principal curve.

**Model.** The input is the PCA scores of 26 samples by 500 genes, projected onto the first two components; each sample is labeled by its condition (A through I).

**Method.** A PCA scree plot (variance explained per component) with a 2D PC1-PC2 projection, and a fitted Hastie-Stuetzle principal curve.

**Test.** The gene-expression data (26 samples x 500 genes), colored by condition.

**Verification.** PC1 explains most of the variance, the samples fall in an ordered arc from condition A to I, and the principal curve summarizes that progression as a 1D coordinate per sample.
:::

### R implementation {.impl}

```{r}
#| fig-width: 7
#| fig-height: 4
#| out-width: "70%"
# scree plot: proportion of variance explained
var_prop <- res$values / sum(res$values)
cum_prop <- cumsum(var_prop[1:10])
barplot(var_prop[1:10], names.arg = 1:10, xlab = "Principal component",
        ylab = "Proportion of variance", main = "Scree plot", ylim = c(0, max(cum_prop) * 1.1))
lines(seq_along(var_prop[1:10]) * 1.2 - 0.5, cum_prop, type = "b", col = 2)
legend("topright", c("individual", "cumulative"), col = c("grey30", 2), pch = c(15, 1), lty = c(0, 1))
```

```{r}
#| fig-width: 5.5
#| fig-height: 5.5
#| out-width: "60%"
# project onto PC1-PC2, colored by condition, with a fitted principal curve
library(princurve)
cond <- factor(substr(rownames(X), 1, 1))
pc <- res$scores[, 1:2]
pc_lab <- sprintf("PC%d (%.1f%%)", 1:2, var_prop[1:2] * 100)

fit <- principal_curve(pc)                 # Hastie-Stuetzle principal curve (princurve package)

plot(pc[, 1], pc[, 2], col = cond, pch = 19, xlab = pc_lab[1], ylab = pc_lab[2])
text(pc[, 1], pc[, 2], labels = rownames(X), pos = 3, cex = 0.6)
lines(fit$s[fit$ord, ], col = "red", lwd = 2)
```

### Python implementation {.impl}

```{python}
#| fig-width: 7
#| fig-height: 4
#| out-width: "70%"
# scree plot: proportion of variance explained
var_prop = res["values"] / res["values"].sum()
fig, ax = plt.subplots(figsize=(7, 4))
ax.bar(range(1, 11), var_prop[:10], color="grey")
ax.plot(range(1, 11), np.cumsum(var_prop[:10]), "o-", color="red")
ax.set(xlabel="Principal component", ylabel="Proportion of variance", title="Scree plot")
plt.show()
```

```{python}
#| fig-width: 5.5
#| fig-height: 5.5
#| out-width: "60%"
# project onto PC1-PC2, colored by condition, with a fitted principal curve
from pcurvepy2 import PrincipalCurve
cond = np.array([s[0] for s in samples])
pc = res["scores"][:, :2]
pc_lab = [f"PC{i + 1} ({v * 100:.1f}%)" for i, v in enumerate(var_prop[:2])]

fit = PrincipalCurve()                     # Python port of R's princurve, same Hastie-Stuetzle algorithm
fit.fit(pc, param_s=20)                    # R's princurve auto-tunes its spline smoothing by
                                            # cross-validation; pcurvepy2 needs it supplied directly,
                                            # and the unsmoothed default overfits this small dataset
fit.project_to_curve(pc, points=fit.points, stretch=2.0)   # match R's default curve-end stretch
curve = fit.points_interp[fit.order]

codes = {c: i for i, c in enumerate(sorted(set(cond)))}
plt.figure(figsize=(5.5, 5.5))
plt.scatter(pc[:, 0], pc[:, 1], c=[codes[c] for c in cond], cmap="tab10")
for i, name in enumerate(samples):
    plt.text(pc[i, 0], pc[i, 1], name, fontsize=6)
plt.plot(curve[:, 0], curve[:, 1], color="red", lw=2)
plt.xlabel(pc_lab[0]); plt.ylabel(pc_lab[1]); plt.show()
```

The first component alone explains most of the variance, and the samples fall into an
ordered arc from condition A to I in the PC1-PC2 plane. Fitting a principal curve, a
smooth one-dimensional path that threads through the middle of the points
[@hastie1989principal], summarizes this trajectory: the experiment traces a continuous
progression of gene-expression states, and the position along the curve is a natural
one-dimensional coordinate for each sample. Rather than implementing the fitting
algorithm (project, order, smooth, repeat) from scratch, we use the R `princurve`
package and its Python port `pcurvepy2`, both established implementations of the same
Hastie-Stuetzle procedure.

## 10A.5 Nonlinear embeddings: MDS, t-SNE, and UMAP

PCA is a linear projection. When the data lie on a curved manifold, nonlinear methods
can separate structure that PCA blurs. We demonstrate three, using established libraries
rather than building them from scratch. *Multidimensional scaling* (MDS) places points
in 2D so their pairwise distances match the high-dimensional distances as closely as
possible [@torgerson1952multidimensional]. *t-SNE* [@vandermaaten2008tsne] and *UMAP*
[@mcinnes2018umap] instead preserve local neighborhoods, and are the standard tools for
visualizing high-dimensional biological data, at the cost of being stochastic and
distorting global distances.

::: {.callout-note title="Recipe 10A.5.1"}
**Objective.** Embed high-dimensional samples in 2D three different ways and compare what each preserves.

**Model.** The input is the same 26 samples by 500 genes, colored by condition.

**Method.** Three 2D nonlinear embeddings: classical (Torgerson) multidimensional scaling, t-SNE, and UMAP.

**Test.** The gene-expression data; t-SNE perplexity 5, UMAP 5 neighbors; seed 1.

**Verification.** All three separate the nine conditions; MDS lays them along the A-to-I progression (it preserves global distances), while t-SNE and UMAP emphasize tight local clusters whose between-cluster distances are not meaningful.
:::

### R implementation {.impl}

```{r}
#| fig-width: 9
#| fig-height: 3.4
#| out-width: "95%"
#| warning: false
library(Rtsne); library(uwot)
set.seed(1)
emb_mds  <- cmdscale(dist(X), k = 2)
emb_tsne <- Rtsne(X, dims = 2, perplexity = 5, check_duplicates = FALSE)$Y
emb_umap <- umap(X, n_components = 2, n_neighbors = 5)

par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
plot(emb_mds,  col = cond, pch = 19, xlab = "dim 1", ylab = "dim 2", main = "MDS")
plot(emb_tsne, col = cond, pch = 19, xlab = "dim 1", ylab = "dim 2", main = "t-SNE")
plot(emb_umap, col = cond, pch = 19, xlab = "dim 1", ylab = "dim 2", main = "UMAP")
```

### Python implementation {.impl}

```{python}
#| fig-width: 9
#| fig-height: 3.4
#| out-width: "95%"
#| warning: false
import warnings; warnings.filterwarnings("ignore")
from scipy.spatial.distance import pdist, squareform
from sklearn.decomposition import KernelPCA
from sklearn.manifold import TSNE
import umap

# classical (Torgerson) MDS is eigen-decomposition of the Gower-centered squared-distance
# matrix, equivalent to KernelPCA on that matrix as a precomputed kernel; this matches R's
# cmdscale, unlike sklearn.manifold.MDS, which instead runs the iterative SMACOF algorithm
# and gives a materially different embedding
D2 = squareform(pdist(X)) ** 2
n = D2.shape[0]; J = np.eye(n) - np.ones((n, n)) / n
emb_mds  = KernelPCA(n_components=2, kernel="precomputed").fit_transform(-0.5 * J @ D2 @ J)
emb_tsne = TSNE(n_components=2, perplexity=5, random_state=1).fit_transform(X)
emb_umap = umap.UMAP(n_components=2, n_neighbors=5, random_state=1).fit_transform(X)

col = [codes[c] for c in cond]
fig, axs = plt.subplots(1, 3, figsize=(9, 3.4))
for ax, (name, e) in zip(axs, [("MDS", emb_mds), ("t-SNE", emb_tsne), ("UMAP", emb_umap)]):
    ax.scatter(e[:, 0], e[:, 1], c=col, cmap="tab10")
    ax.set(xlabel="dim 1", ylabel="dim 2", title=name)
plt.tight_layout(); plt.show()
```

All three separate the nine conditions, and MDS, like PCA, lays them out along the A-to-I
progression because it preserves global distances (R's `cmdscale` and the classical MDS
computed above are the same algorithm, so the two panels agree up to rotation and
reflection). t-SNE and UMAP emphasize tight local clusters (the replicates of each
condition group together) but their between-cluster distances should not be read as
meaningful. The R and Python t-SNE/UMAP panels will not overlay exactly: `Rtsne`/`uwot`
and `scikit-learn`/`umap-learn` are independent implementations with different default
optimizers and initializations, so a shared random seed does not make them numerically
identical, only comparable in the cluster structure they recover. For this small, smoothly
varying dataset PCA already captures the essential structure; the nonlinear methods become
more valuable for larger datasets with distinct, non-linear cluster structure.
