---
title: "10A. Dimensionality reduction"
author: "Mingyang Lu"
old_id: "10A"
---

Biological data are often high-dimensional: a gene-expression experiment measures
thousands of genes across many samples. To visualize and interpret such data we reduce
their dimensionality, mapping each sample to a few coordinates that capture most of the
structure. This chapter builds *principal component analysis* (PCA) from scratch, uses
it on a real gene-expression dataset, and then demonstrates nonlinear embeddings (MDS,
t-SNE, and UMAP).

## 10A.1 Projecting data onto a line

The simplest reduction takes two-dimensional points and describes them by a single
number. Consider 100 points scattered in a narrow, tilted strip.

### R implementation {.impl}

```{r}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
set.seed(1)
x <- runif(100); y <- 0.4 * x + 0.8 + 0.1 * runif(100)
data1 <- data.frame(x = x, y = y)

# fit a line, then project each point onto it
model <- lm(y ~ x, data = data1)
x0 <- c(0, coef(model)[1])                 # a point on the line
dir <- c(1, coef(model)[2]); dir <- dir / sqrt(sum(dir^2))   # unit direction
proj <- as.matrix(data1) %*% dir - sum(x0 * dir)            # 1D coordinate along the line
projected <- t(sapply(proj, function(p) p * dir + x0))      # back to 2D

plot(data1$x, data1$y, xlab = "x", ylab = "y", xlim = c(0, 1), ylim = c(0.5, 1.5), asp = 1)
abline(model, col = 2)
points(projected[, 1], projected[, 2], pch = 16, col = 4)
```

### Python implementation {.impl}

```{python}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
import numpy as np
import matplotlib.pyplot as plt

rng = np.random.default_rng(1)
x = rng.uniform(size=100); y = 0.4 * x + 0.8 + 0.1 * rng.uniform(size=100)
data1 = np.column_stack([x, y])

# fit a line, then project each point onto it
b1, b0 = np.polyfit(x, y, 1)
x0 = np.array([0, b0])
d = np.array([1, b1]); d = d / np.linalg.norm(d)            # unit direction
proj = data1 @ d - x0 @ d                                   # 1D coordinate along the line
projected = np.outer(proj, d) + x0                          # back to 2D

plt.figure(figsize=(5, 5))
plt.scatter(x, y)
xs = np.array([0, 1]); plt.plot(xs, b1 * xs + b0, color="red")
plt.scatter(projected[:, 0], projected[:, 1], color="blue", s=15)
plt.gca().set_aspect("equal"); plt.xlabel("x"); plt.ylabel("y")
plt.xlim(0, 1); plt.ylim(0.5, 1.5); plt.show()
```

Each blue point is the closest position on the line to its original point. The single
coordinate along the line (the projection) keeps most of the spread of the data, since
the points barely deviate from the line. PCA generalizes this idea: it finds the
directions of largest variance automatically, without being told to use a regression
line, and works in any number of dimensions.

## 10A.2 Principal component analysis from scratch

PCA re-expresses the data in a new orthogonal basis, the *principal components*, ordered
by how much variance they capture [@pearson1901lines; @hotelling1933analysis]. The
recipe is: center the data, then find the directions along which it varies most. Those
directions are the eigenvectors of the covariance matrix, and their eigenvalues give the
variance captured. We implement it two equivalent ways: eigen-decomposition of the
covariance matrix, and the *singular value decomposition* (SVD) of the centered data,
which is numerically preferable and much cheaper when there are more features than
samples.

### R implementation {.impl}

```{r}
# PCA from scratch: "eigen" (covariance eigen-decomposition) or "svd" (of centered data)
pca <- function(X, method = "svd") {
  Xc <- scale(X, center = TRUE, scale = FALSE)   # center each feature
  n <- nrow(X)
  if (method == "eigen") {
    e <- eigen(cov(Xc)); values <- e$values; loadings <- e$vectors
  } else {
    s <- svd(Xc); values <- s$d^2 / (n - 1); loadings <- s$v
  }
  list(values = values, loadings = loadings, scores = Xc %*% loadings)
}

# the two routes agree (on the strip data from 10A.1)
X2 <- as.matrix(data1)
round(rbind(eigen = pca(X2, "eigen")$values, svd = pca(X2, "svd")$values), 5)
```

### Python implementation {.impl}

```{python}
# PCA from scratch: "eigen" (covariance eigen-decomposition) or "svd" (of centered data)
def pca(X, method="svd"):
    Xc = X - X.mean(0)                              # center each feature
    n = len(X)
    if method == "eigen":
        vals, vecs = np.linalg.eigh(np.cov(Xc, rowvar=False))
        order = np.argsort(vals)[::-1]
        values, loadings = vals[order], vecs[:, order]
    else:
        U, s, Vt = np.linalg.svd(Xc, full_matrices=False)
        values, loadings = s**2 / (n - 1), Vt.T
    return {"values": values, "loadings": loadings, "scores": Xc @ loadings}

# the two routes agree (on the strip data from 10A.1)
print(np.round(np.vstack([pca(data1, "eigen")["values"], pca(data1, "svd")["values"]]), 5))
```

Both routes return the same variances. For the strip data the first principal component
carries almost all the variance and points along the strip, recovering the direction we
fit by hand in the previous section.

## 10A.3 Interpreting components with loadings

We now apply PCA to a real dataset: expression of 500 genes measured across 26 samples
spanning nine experimental conditions (A through I). The samples are the observations
and the genes are the features, so we transpose the table. The *loadings* (the columns
of the eigenvector matrix) say how much each gene contributes to each component;
inspecting the largest loadings tells us which genes drive the dominant patterns.

### R implementation {.impl}

```{r}
#| fig-width: 7
#| fig-height: 4
#| out-width: "70%"
exp_data <- read.csv("10-high-dim-data/data/exp_data.csv", row.names = 1)
X <- t(as.matrix(exp_data))                        # 26 samples x 500 genes
genes <- colnames(X)

res <- pca(X)                                      # used again in 10A.4
top <- order(abs(res$loadings[, 1]), decreasing = TRUE)[1:10]   # top PC1 genes
barplot(res$loadings[top, 1], names.arg = genes[top], las = 2,
        ylab = "PC1 loading", main = "Genes contributing most to PC1")
```

### Python implementation {.impl}

```{python}
#| fig-width: 7
#| fig-height: 4
#| out-width: "70%"
import pandas as pd
exp_data = pd.read_csv("10-high-dim-data/data/exp_data.csv", index_col=0)
X = exp_data.values.T                              # 26 samples x 500 genes
genes = exp_data.index.to_numpy()
samples = exp_data.columns.to_numpy()

res = pca(X)                                       # used again in 10A.4
top = np.argsort(np.abs(res["loadings"][:, 0]))[::-1][:10]      # top PC1 genes
plt.figure(figsize=(7, 4))
plt.bar(genes[top], res["loadings"][top, 0])
plt.xticks(rotation=90); plt.ylabel("PC1 loading")
plt.title("Genes contributing most to PC1"); plt.tight_layout(); plt.show()
```

Each principal component is a weighted combination of all 500 genes. The handful of
genes with the largest loadings dominate that component, so PC1 can be read as the
coordinated expression pattern of these top genes, the axis along which the samples vary
most.

## 10A.4 Visualizing the samples

With the same PCA, we look at the sample side. The *scree plot* shows how much variance
each component explains, guiding how many to keep. Projecting the samples onto the first
two components then reveals their structure.

### R implementation {.impl}

```{r}
#| fig-width: 7
#| fig-height: 4
#| out-width: "70%"
# scree plot: proportion of variance explained
var_prop <- res$values / sum(res$values)
barplot(var_prop[1:10], names.arg = 1:10, xlab = "Principal component",
        ylab = "Proportion of variance", main = "Scree plot")
lines(seq_along(var_prop[1:10]) * 1.2 - 0.5, cumsum(var_prop[1:10]), type = "b", col = 2)
legend("topright", c("individual", "cumulative"), col = c("grey30", 2), pch = c(15, 1), lty = c(0, 1))
```

```{r}
#| fig-width: 5.5
#| fig-height: 5.5
#| out-width: "60%"
# project onto PC1-PC2, colored by condition, with a fitted principal curve
cond <- factor(substr(rownames(X), 1, 1))
pc <- res$scores[, 1:2]

# principal curve from scratch (Hastie-Stuetzle: project, order, smooth; repeat)
principal_curve <- function(P, n_iter = 8, k = 5) {
  n <- nrow(P); lambda <- P[, 1]
  ma <- function(v) { s <- as.numeric(stats::filter(v, rep(1 / k, k), sides = 2)); s[is.na(s)] <- v[is.na(s)]; s }
  for (it in 1:n_iter) {
    ord <- order(lambda); f <- P
    for (d in 1:ncol(P)) f[ord, d] <- ma(P[ord, d])
    curve <- f[ord, ]
    seg <- c(0, cumsum(sqrt(rowSums(diff(curve)^2))))
    for (i in 1:n) lambda[i] <- seg[which.min(rowSums((curve - matrix(P[i, ], n, 2, byrow = TRUE))^2))]
  }
  f[order(lambda), ]
}
curve <- principal_curve(pc)

plot(pc[, 1], pc[, 2], col = cond, pch = 19, xlab = "PC1", ylab = "PC2")
text(pc[, 1], pc[, 2], labels = rownames(X), pos = 3, cex = 0.6)
lines(curve[, 1], curve[, 2], col = "red", lwd = 2)
```

### Python implementation {.impl}

```{python}
#| fig-width: 7
#| fig-height: 4
#| out-width: "70%"
# scree plot: proportion of variance explained
var_prop = res["values"] / res["values"].sum()
fig, ax = plt.subplots(figsize=(7, 4))
ax.bar(range(1, 11), var_prop[:10], color="grey")
ax.plot(range(1, 11), np.cumsum(var_prop[:10]), "o-", color="red")
ax.set(xlabel="Principal component", ylabel="Proportion of variance", title="Scree plot")
plt.show()
```

```{python}
#| fig-width: 5.5
#| fig-height: 5.5
#| out-width: "60%"
# project onto PC1-PC2, colored by condition, with a fitted principal curve
cond = np.array([s[0] for s in samples])
pc = res["scores"][:, :2]

# principal curve from scratch (Hastie-Stuetzle: project, order, smooth; repeat)
def principal_curve(P, n_iter=8, k=5):
    n = len(P); lam = P[:, 0].copy()
    ma = lambda v: np.convolve(v, np.ones(k) / k, mode="same")
    for _ in range(n_iter):
        o = np.argsort(lam); f = P.copy()
        for d in range(P.shape[1]):
            f[o, d] = ma(P[o, d])
        curve = f[o]
        seg = np.concatenate([[0], np.cumsum(np.sqrt((np.diff(curve, axis=0) ** 2).sum(1)))])
        for i in range(n):
            lam[i] = seg[np.argmin(((curve - P[i]) ** 2).sum(1))]
    return f[np.argsort(lam)]
curve = principal_curve(pc)

codes = {c: i for i, c in enumerate(sorted(set(cond)))}
plt.figure(figsize=(5.5, 5.5))
plt.scatter(pc[:, 0], pc[:, 1], c=[codes[c] for c in cond], cmap="tab10")
for i, s in enumerate(samples):
    plt.text(pc[i, 0], pc[i, 1], s, fontsize=6)
plt.plot(curve[:, 0], curve[:, 1], color="red", lw=2)
plt.xlabel("PC1"); plt.ylabel("PC2"); plt.show()
```

The first component alone explains most of the variance, and the samples fall into an
ordered arc from condition A to I in the PC1-PC2 plane. Fitting a principal curve, a
smooth one-dimensional path that threads through the middle of the points
[@hastie1989principal], summarizes this trajectory: the experiment traces a continuous
progression of gene-expression states, and the position along the curve is a natural
one-dimensional coordinate for each sample.

## 10A.5 Nonlinear embeddings: MDS, t-SNE, and UMAP

PCA is a linear projection. When the data lie on a curved manifold, nonlinear methods
can separate structure that PCA blurs. We demonstrate three, using established libraries
rather than building them from scratch. *Multidimensional scaling* (MDS) places points
in 2D so their pairwise distances match the high-dimensional distances as closely as
possible [@torgerson1952multidimensional]. *t-SNE* [@vandermaaten2008tsne] and *UMAP*
[@mcinnes2018umap] instead preserve local neighborhoods, and are the standard tools for
visualizing high-dimensional biological data, at the cost of being stochastic and
distorting global distances.

### R implementation {.impl}

```{r}
#| fig-width: 9
#| fig-height: 3.4
#| out-width: "95%"
#| warning: false
library(Rtsne); library(uwot)
set.seed(1)
emb_mds  <- cmdscale(dist(X), k = 2)
emb_tsne <- Rtsne(X, dims = 2, perplexity = 5, check_duplicates = FALSE)$Y
emb_umap <- umap(X, n_components = 2, n_neighbors = 5)

par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
plot(emb_mds,  col = cond, pch = 19, xlab = "dim 1", ylab = "dim 2", main = "MDS")
plot(emb_tsne, col = cond, pch = 19, xlab = "dim 1", ylab = "dim 2", main = "t-SNE")
plot(emb_umap, col = cond, pch = 19, xlab = "dim 1", ylab = "dim 2", main = "UMAP")
```

### Python implementation {.impl}

```{python}
#| fig-width: 9
#| fig-height: 3.4
#| out-width: "95%"
#| warning: false
import warnings; warnings.filterwarnings("ignore")
from sklearn.manifold import MDS, TSNE
import umap

emb_mds  = MDS(n_components=2, random_state=1, normalized_stress="auto").fit_transform(X)
emb_tsne = TSNE(n_components=2, perplexity=5, random_state=1).fit_transform(X)
emb_umap = umap.UMAP(n_components=2, n_neighbors=5, random_state=1).fit_transform(X)

col = [codes[c] for c in cond]
fig, axs = plt.subplots(1, 3, figsize=(9, 3.4))
for ax, (name, e) in zip(axs, [("MDS", emb_mds), ("t-SNE", emb_tsne), ("UMAP", emb_umap)]):
    ax.scatter(e[:, 0], e[:, 1], c=col, cmap="tab10")
    ax.set(xlabel="dim 1", ylabel="dim 2", title=name)
plt.tight_layout(); plt.show()
```

All three separate the nine conditions, and MDS, like PCA, lays them out along the A-to-I
progression because it preserves global distances. t-SNE and UMAP emphasize tight local
clusters (the replicates of each condition group together) but their between-cluster
distances should not be read as meaningful. For this small, smoothly varying dataset PCA
already captures the essential structure; the nonlinear methods become more valuable for
larger datasets with distinct, non-linear cluster structure.
