---
title: "10B. Clustering"
author: "Mingyang Lu"
old_id: "10B"
---

Clustering groups data points so that points in the same cluster are more similar to each other than to those in other clusters. It is *unsupervised*: unlike the labeled conditions in [Part 10A](10a-dimensionality-reduction.qmd), we look for structure without being told the groups. This chapter builds k-means from scratch, shows how to choose the number of clusters, covers hierarchical clustering, applies clustering to the gene-expression data as a heatmap, and ends with a probabilistic (Gaussian mixture) alternative.

## 10B.1 k-means from scratch

k-means partitions $n$ points into $k$ clusters, each represented by a centroid [@lloyd1982least]. Starting from random centroids, it alternates two steps until convergence: **assign** each point to its nearest centroid, then **update** each centroid to the mean of its assigned points. The result depends on the random start, so we run several restarts and keep the one with the smallest total within-cluster sum of squares (WSS).

::: {.callout-note title="Recipe 10B.1.1"}
**Objective.** Cluster points into groups with k-means implemented from scratch.

**Model.** Three Gaussian blobs in 2D (50 points each, at means 0, 1.5, and 3, standard deviation 0.5); the input is an n x 2 point matrix.

**Method.** k-means clustering (Lloyd's algorithm): random centroids, then alternating nearest-centroid assignment and centroid-mean updates, keeping the lowest within-cluster sum of squares over several restarts.

**Test.** `k = 3`, 10 restarts, up to 100 iterations; seed 123.

**Show.** A scatter of the points colored by cluster, with the three centroids marked.

**Verification.** k-means recovers the three blobs, placing a centroid at the middle of each, and the restarts guard against an unlucky initialization converging to a poor split.
:::

### R implementation {.impl}

```{r}
# one k-means run from a random start
kmeans_once <- function(X, k, max_iter = 100) {
  centroids <- X[sample(nrow(X), k), , drop = FALSE]
  for (iter in 1:max_iter) {
    d <- sapply(1:k, function(i) rowSums(sweep(X, 2, centroids[i, ])^2))   # dist^2 to each centroid
    clusters <- max.col(-d, ties.method = "first")                        # nearest centroid
    new_centroids <- t(sapply(1:k, function(i) colMeans(X[clusters == i, , drop = FALSE])))
    if (sum((new_centroids - centroids)^2) < 1e-10) break
    centroids <- new_centroids
  }
  wss <- sum(sapply(1:k, function(i) sum(rowSums(sweep(X, 2, centroids[i, ])^2)[clusters == i])))
  list(clusters = clusters, centroids = centroids, wss = wss)
}

# k-means with several random restarts; keep the best (lowest WSS)
kmeans_cluster <- function(X, k, n_start = 10, max_iter = 100) {
  best <- NULL
  for (s in 1:n_start) {
    r <- kmeans_once(X, k, max_iter)
    if (is.null(best) || r$wss < best$wss) best <- r
  }
  best
}
```

```{r}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
# three Gaussian blobs
set.seed(123)
X <- rbind(matrix(rnorm(100, 0, 0.5), ncol = 2),
           matrix(rnorm(100, 1.5, 0.5), ncol = 2),
           matrix(rnorm(100, 3, 0.5), ncol = 2))
res <- kmeans_cluster(X, k = 3)
plot(X, col = res$clusters, pch = 19, xlab = "x", ylab = "y", asp = 1)
points(res$centroids, col = 1:3, pch = 8, cex = 2, lwd = 2)
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

# one k-means run from a random start
def kmeans_once(X, k, max_iter=100):
    centroids = X[np.random.choice(len(X), k, replace=False)]
    for _ in range(max_iter):
        d = np.stack([((X - c)**2).sum(1) for c in centroids], 1)   # dist^2 to each centroid
        clusters = d.argmin(1)                                       # nearest centroid
        new = np.stack([X[clusters == i].mean(0) for i in range(k)])
        if ((new - centroids)**2).sum() < 1e-10:
            break
        centroids = new
    wss = sum(((X[clusters == i] - centroids[i])**2).sum() for i in range(k))
    return {"clusters": clusters, "centroids": centroids, "wss": wss}

# k-means with several random restarts; keep the best (lowest WSS)
def kmeans_cluster(X, k, n_start=10, max_iter=100):
    best = None
    for _ in range(n_start):
        r = kmeans_once(X, k, max_iter)
        if best is None or r["wss"] < best["wss"]:
            best = r
    return best
```

```{python}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
# three Gaussian blobs
np.random.seed(123)
X = np.vstack([np.random.normal(m, 0.5, (50, 2)) for m in (0, 1.5, 3)])
res = kmeans_cluster(X, k=3)
plt.figure(figsize=(5, 5))
plt.scatter(X[:, 0], X[:, 1], c=res["clusters"], cmap="brg", s=15)
plt.scatter(res["centroids"][:, 0], res["centroids"][:, 1], c="black", marker="*", s=200)
plt.gca().set_aspect("equal"); plt.xlabel("x"); plt.ylabel("y"); plt.show()
```

k-means recovers the three blobs, placing a centroid at the middle of each. The restarts matter: a single unlucky initialization can converge to a poor split, so keeping the lowest-WSS run guards against that. k-means assumes roughly spherical, equal-sized clusters and requires the number of clusters $k$ in advance.

### Choosing the number of clusters

k-means needs $k$, but we rarely know it. The *elbow method* runs k-means for a range of $k$ and plots the total WSS against $k$. WSS always decreases as $k$ grows, but it drops sharply until the natural number of clusters is reached and then flattens; the "elbow" marks a good choice.

### R implementation {.impl}

```{r}
#| fig-width: 6
#| fig-height: 4
#| out-width: "60%"
set.seed(1)
wss_k <- sapply(1:8, function(k) kmeans_cluster(X, k)$wss)
plot(1:8, wss_k, type = "b", pch = 19, col = 2, xlab = "Number of clusters k",
     ylab = "Total within-cluster SS")
```

### Python implementation {.impl}

```{python}
#| fig-width: 6
#| fig-height: 4
#| out-width: "60%"
np.random.seed(1)
wss_k = [kmeans_cluster(X, k)["wss"] for k in range(1, 9)]
plt.figure(figsize=(6, 4))
plt.plot(range(1, 9), wss_k, "o-", color="red")
plt.xlabel("Number of clusters k"); plt.ylabel("Total within-cluster SS"); plt.show()
```

The curve bends sharply at $k = 3$ and flattens afterward, correctly identifying the three blobs. The elbow is a heuristic, not a rule, but it is a quick and widely used guide.

## 10B.2 Hierarchical clustering

Hierarchical clustering builds a tree (dendrogram) instead of a flat partition, and does not need $k$ up front. *Agglomerative* clustering starts with every point in its own cluster and repeatedly merges the two closest clusters. How "closest" is measured is the *linkage*: **single** (nearest pair), **complete** (farthest pair), **average** (mean pairwise distance), and **Ward** (merge that least increases WSS [@ward1963hierarchical]). Cutting the tree at a chosen height yields a flat clustering. We use the built-in routines (`hclust` in R, `scipy` in Python).

::: {.callout-note title="Recipe 10B.2.1"}
**Objective.** Cluster points by building a dendrogram instead of a flat partition.

**Model.** The input is the same three-blob data, via pairwise Euclidean distances.

**Method.** Agglomerative hierarchical clustering with Ward linkage, cut into a flat partition.

**Test.** The three-blob data, cut into `k = 3` clusters.

**Show.** The dendrogram, and the points colored by the three cut clusters.

**Verification.** The dendrogram shows three clear branches, and cutting into three recovers the blobs, agreeing with k-means.
:::

### R implementation {.impl}

```{r}
#| fig-width: 9
#| fig-height: 4
#| out-width: "95%"
d <- dist(X)                              # pairwise distances
par(mfrow = c(1, 2), mar = c(2, 4, 2, 1))
hc <- hclust(d, method = "ward.D2")
plot(hc, labels = FALSE, main = "Ward linkage", xlab = "", sub = "")
groups <- cutree(hc, k = 3)               # cut into 3 clusters
rect.hclust(hc, k = 3, border = 2:4)
plot(X, col = groups, pch = 19, xlab = "x", ylab = "y", asp = 1, main = "cut into 3 clusters")
```

### Python implementation {.impl}

```{python}
#| fig-width: 9
#| fig-height: 4
#| out-width: "95%"
from scipy.cluster.hierarchy import linkage, dendrogram, fcluster
Z = linkage(X, method="ward")             # pairwise distances + Ward linkage
groups = fcluster(Z, t=3, criterion="maxclust")   # cut into 3 clusters
fig, axs = plt.subplots(1, 2, figsize=(9, 4))
dendrogram(Z, no_labels=True, ax=axs[0]); axs[0].set(title="Ward linkage")
axs[1].scatter(X[:, 0], X[:, 1], c=groups, cmap="brg", s=15)
axs[1].set(xlabel="x", ylabel="y", aspect="equal", title="cut into 3 clusters")
plt.tight_layout(); plt.show()
```

The dendrogram shows three clear branches; cutting it into three groups recovers the blobs, agreeing with k-means. Different linkages can give different trees (single linkage tends to chain points together, Ward produces compact clusters), so the choice matters.

## 10B.3 Clustering the gene-expression data

Applied to the gene-expression data of [Part 10A](10a-dimensionality-reduction.qmd), clustering finds *modules* of genes that vary together across the conditions. We $z$-score each gene (so all genes share a scale), cluster the 500 genes with k-means, and draw a heatmap with the genes ordered by cluster, the canonical way to display expression data.

::: {.callout-note title="Recipe 10B.3.1"}
**Objective.** Cluster genes by expression and display them as a heatmap.

**Model.** The input is the gene-expression data (500 genes x 26 samples), z-scored per gene.

**Method.** k-means clustering of the z-scored genes, shown as a cluster-ordered expression heatmap.

**Test.** `k = 5` clusters; heatmap color scale from -3 to 3; seed 1.

**Show.** A cluster-ordered heatmap of the gene-expression data.

**Verification.** The heatmap splits genes into co-expression modules that rise and fall together across the conditions, resolving the PCA progression of Part 10A into specific gene groups.
:::

### R implementation {.impl}

```{r}
#| fig-width: 6
#| fig-height: 6.5
#| out-width: "62%"
exp_data <- read.csv("10-high-dim-data/data/exp_data.csv", row.names = 1)
Z <- t(scale(t(as.matrix(exp_data))))     # z-score each gene (row)

set.seed(1)
gene_clusters <- kmeans_cluster(Z, k = 5)$clusters
ord <- order(gene_clusters)               # order genes by cluster

pal <- colorRampPalette(c("blue", "white", "red"))(50)
image(1:ncol(Z), 1:nrow(Z), t(Z[ord, ]), col = pal, zlim = c(-3, 3),
      xlab = "sample (A-I)", ylab = "gene (ordered by cluster)", axes = FALSE)
axis(1, at = 1:ncol(Z), labels = colnames(Z), las = 2, cex.axis = 0.5)
abline(h = cumsum(table(gene_clusters)) + 0.5, col = "black")   # cluster boundaries
```

### Python implementation {.impl}

```{python}
#| fig-width: 6
#| fig-height: 6.5
#| out-width: "62%"
import pandas as pd
exp_data = pd.read_csv("10-high-dim-data/data/exp_data.csv", index_col=0)
G = exp_data.values                        # 500 genes x 26 samples
Z = (G - G.mean(1, keepdims=True)) / G.std(1, keepdims=True)   # z-score each gene

np.random.seed(1)
gene_clusters = kmeans_cluster(Z, k=5)["clusters"]
ordr = np.argsort(gene_clusters)           # order genes by cluster

plt.figure(figsize=(6, 6.5))
plt.imshow(Z[ordr], aspect="auto", cmap="bwr", vmin=-3, vmax=3, interpolation="none")
plt.xticks(range(len(exp_data.columns)), exp_data.columns, rotation=90, fontsize=5)
for b in np.cumsum(np.bincount(gene_clusters))[:-1]:
    plt.axhline(b - 0.5, color="black")    # cluster boundaries
plt.xlabel("sample (A-I)"); plt.ylabel("gene (ordered by cluster)"); plt.show()
```

The heatmap splits the genes into co-expression modules: each block of rows is a set of genes that rise and fall together across the A-to-I conditions. Some modules are high in the early conditions and low in the late ones, others the reverse, reflecting the same gene-expression progression seen in the PCA of Part 10A, now resolved into the specific groups of genes that drive it.

## 10B.4 Gaussian mixture models

k-means assigns each point to exactly one cluster (a *hard* assignment) and assumes spherical clusters. A *Gaussian mixture model* (GMM) is a probabilistic alternative: it models the data as a mixture of Gaussian components with their own means, covariances, and weights, fit by the expectation-maximization algorithm [@dempster1977maximum], and gives each point a *probability* of belonging to each cluster (a *soft* assignment). Because each component has its own covariance, a GMM can capture elongated, tilted clusters that k-means cannot, and model selection (via the BIC) can choose the number of components automatically. We use `mclust` [@scrucca2016mclust] in R and scikit-learn in Python.

::: {.callout-note title="Recipe 10B.4.1"}
**Objective.** Cluster points with a Gaussian mixture and pick the number of components automatically.

**Model.** A Gaussian mixture: the data modeled as a mixture of Gaussian components, each with its own mean, covariance, and weight, with soft assignments; the input is the three-blob data.

**Method.** Gaussian mixture models fit by expectation-maximization, with the number of components chosen by the Bayesian information criterion.

**Test.** The three-blob data, choosing among 1 to 6 components; seed 1.

**Show.** A scatter of the points colored by mixture component, and the BIC versus number of components.

**Verification.** The criterion selects three components matching the blobs, and the mixture recovers them with soft probabilities.
:::

### R implementation {.impl}

```{r}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
#| message: false
library(mclust)
gmm <- Mclust(X, verbose = FALSE)         # BIC picks the number of components
plot(gmm, what = "classification")
cat("components chosen by BIC:", gmm$G, "\n")
```

### Python implementation {.impl}

```{python}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
from sklearn.mixture import GaussianMixture
# choose the number of components by BIC
models = [GaussianMixture(n, covariance_type="full", random_state=1).fit(X) for n in range(1, 7)]
bic = [m.bic(X) for m in models]
gmm = models[int(np.argmin(bic))]
labels = gmm.predict(X)
plt.figure(figsize=(5, 5))
plt.scatter(X[:, 0], X[:, 1], c=labels, cmap="brg", s=15)
plt.scatter(gmm.means_[:, 0], gmm.means_[:, 1], c="black", marker="*", s=200)
plt.gca().set_aspect("equal"); plt.xlabel("x"); plt.ylabel("y")
plt.title(f"{gmm.n_components} components (BIC)"); plt.show()
```

The BIC selects three components, matching the blobs, and the GMM recovers them with soft probabilities rather than hard labels. For these well-separated spherical blobs the GMM and k-means agree, but the GMM is the more flexible model: its per-component covariances let it fit clusters of different shapes and sizes, and its probabilistic assignments quantify how confidently each point belongs to its cluster.
