---
title: "10C. Network algorithms"
author: "Mingyang Lu"
old_id: "10C"
---

Many biological systems are naturally described as *networks*: protein-protein
interaction networks, gene regulatory networks, metabolic networks, and cell-signaling
networks all connect components (genes, proteins, metabolites) by their interactions
[@barabasi2004network]. A network, or *graph*, is a set of *vertices* (nodes) joined by
*edges*. This chapter represents networks as adjacency matrices, computes their
structural properties, generates random-network models, detects community structure by
modularity, and models discrete gene-regulatory dynamics with Boolean networks. We
compute the core quantities from scratch and use `igraph` (R) and `networkx` (Python)
for graph layout and the longer standard algorithms.

## 10C.1 Networks and their properties

A network of $n$ vertices is stored as an $n \times n$ *adjacency matrix* $A$, with
$A_{ij} = 1$ when vertices $i$ and $j$ are connected and 0 otherwise (symmetric for an
undirected network). From $A$ we read off the basic structure. The *degree* of a vertex
is its number of neighbors, $k_i = \sum_j A_{ij}$. The *clustering coefficient* measures
how interconnected a vertex's neighbors are: for a vertex of degree $k$ with $e$ edges
among its neighbors, $c = e / \binom{k}{2}$. The *distance* between two vertices is the
length of the shortest path between them, and the *diameter* is the largest such
distance. We illustrate on Zachary's karate club [@zachary1977information], a classic
34-member social network.

### R implementation {.impl}

```{r}
#| fig-width: 6
#| fig-height: 5
#| out-width: "65%"
#| message: false
library(igraph)
g <- make_graph("Zachary"); A <- as.matrix(as_adjacency_matrix(g)); n <- nrow(A)

degree_v <- rowSums(A)                                   # degree of each vertex

# clustering coefficient (averaged over all vertices; degree < 2 contributes 0)
clustering <- function(A) {
  mean(sapply(1:nrow(A), function(u) {
    nb <- which(A[u, ] > 0); k <- length(nb)
    if (k < 2) 0 else sum(A[nb, nb]) / 2 / (k * (k - 1) / 2)
  }))
}

# shortest-path distances via boolean adjacency powers; diameter = largest distance
dist_matrix <- function(A) {
  n <- nrow(A); D <- matrix(Inf, n, n); diag(D) <- 0; reach <- diag(n)
  for (l in 1:(n - 1)) { reach <- (reach %*% A) > 0; D[D == Inf & reach] <- l }
  D
}
D <- dist_matrix(A)
cat("mean degree:", round(mean(degree_v), 2), " clustering:", round(clustering(A), 3),
    " diameter:", max(D[is.finite(D)]), "\n")

bc <- betweenness(g)                                     # betweenness (igraph)
set.seed(1)
plot(g, vertex.size = 4 + degree_v, vertex.color = "gold",
     vertex.label = NA, layout = layout_with_fr)
```

```{r}
#| fig-width: 7
#| fig-height: 3.4
#| out-width: "80%"
#| rmar: false
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
barplot(table(degree_v), xlab = "degree k", ylab = "number of vertices", main = "Degree distribution")
plot(degree_v, bc, xlab = "degree", ylab = "betweenness", main = "Betweenness vs degree", pch = 19)
```

### Python implementation {.impl}

```{python}
#| fig-width: 6
#| fig-height: 5
#| out-width: "65%"
import numpy as np, networkx as nx, matplotlib.pyplot as plt
G = nx.karate_club_graph(); A = nx.to_numpy_array(G, weight=None); n = len(A)

degree_v = A.sum(1).astype(int)                          # degree of each vertex

# clustering coefficient (averaged over all vertices; degree < 2 contributes 0)
def clustering(A):
    n = len(A); c = np.zeros(n)
    for u in range(n):
        nb = np.where(A[u] > 0)[0]; k = len(nb)
        if k >= 2: c[u] = A[np.ix_(nb, nb)].sum() / 2 / (k * (k - 1) / 2)
    return c.mean()

# shortest-path distances via boolean adjacency powers; diameter = largest distance
def dist_matrix(A):
    n = len(A); D = np.full((n, n), np.inf); np.fill_diagonal(D, 0); reach = np.eye(n)
    for l in range(1, n):
        reach = (reach @ A) > 0; D[(D == np.inf) & reach] = l
    return D
D = dist_matrix(A)
print("mean degree:", round(degree_v.mean(), 2), " clustering:", round(clustering(A), 3),
      " diameter:", int(D[np.isfinite(D)].max()))

bc = np.array(list(nx.betweenness_centrality(G).values()))   # betweenness (networkx)
np.random.seed(1); pos = nx.spring_layout(G, seed=1)
plt.figure(figsize=(6, 5))
nx.draw(G, pos, node_size=40 + 20 * degree_v, node_color="gold", edge_color="grey")
plt.show()
```

```{python}
#| fig-width: 7
#| fig-height: 3.4
#| out-width: "80%"
fig, axs = plt.subplots(1, 2, figsize=(7, 3.4))
vals, counts = np.unique(degree_v, return_counts=True)
axs[0].bar(vals, counts); axs[0].set(xlabel="degree k", ylabel="number of vertices", title="Degree distribution")
axs[1].scatter(degree_v, bc * (n - 1) * (n - 2) / 2); axs[1].set(xlabel="degree", ylabel="betweenness", title="Betweenness vs degree")
plt.tight_layout(); plt.show()
```

The karate club has a few high-degree hubs and many low-degree members, a mean degree
near 4.6, a diameter of 5, and a fairly high clustering coefficient (about 0.57):
friends of friends tend to be friends, a hallmark of the "small-world" networks of Watts
and Strogatz [@watts1998collective]. Betweenness, the number of shortest paths passing
through a vertex [@freeman1977set], tracks degree loosely but not exactly; the two
highest-betweenness nodes are the instructor and the club president, the bridges through
which most paths run.

## 10C.2 Random-network models

To judge whether a real network's structure is surprising, we compare it with random
models. The *Erdős-Rényi* model [@erdos1959random] connects each pair of vertices
independently with probability $p$, giving a Poisson-like degree distribution peaked at
the mean. The *Barabási-Albert* model [@barabasi1999emergence] grows a network by
*preferential attachment*: each new vertex attaches to $m$ existing ones with probability
proportional to their degree, so hubs accumulate ever more connections. This produces a
*scale-free* degree distribution with a heavy, power-law tail, closer to what many
biological networks show. Both generators are short to write.

### R implementation {.impl}

```{r}
# Erdos-Renyi: connect each pair with probability p
er_network <- function(n, p) {
  A <- matrix(0, n, n); ut <- upper.tri(A)
  A[ut] <- rbinom(sum(ut), 1, p); A + t(A)
}

# Barabasi-Albert: grow by preferential attachment (m links per new vertex)
ba_network <- function(n, m) {
  A <- matrix(0, n, n); A[1:(m + 1), 1:(m + 1)] <- 1; diag(A) <- 0   # seed clique
  for (v in (m + 2):n) {
    deg <- rowSums(A)[1:(v - 1)]
    targets <- sample(1:(v - 1), m, prob = deg)                      # attach prop. to degree
    A[v, targets] <- 1; A[targets, v] <- 1
  }
  A
}
```

```{r}
#| fig-width: 7
#| fig-height: 3.4
#| out-width: "80%"
#| rmar: false
set.seed(1)
er <- er_network(1000, 0.006); ba <- ba_network(1000, 3)
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
hist(rowSums(er), breaks = 20, main = "Erdos-Renyi", xlab = "degree", col = "grey")
plot(table(rowSums(ba)), log = "xy", main = "Barabasi-Albert (log-log)",
     xlab = "degree", ylab = "count")
```

### Python implementation {.impl}

```{python}
# Erdos-Renyi: connect each pair with probability p
def er_network(n, p):
    A = (np.triu(np.random.rand(n, n), 1) < p).astype(int); return A + A.T

# Barabasi-Albert: grow by preferential attachment (m links per new vertex)
def ba_network(n, m):
    A = np.zeros((n, n), int); A[:m + 1, :m + 1] = 1; np.fill_diagonal(A, 0)   # seed clique
    for v in range(m + 1, n):
        deg = A[:v].sum(1)
        targets = np.random.choice(v, m, replace=False, p=deg / deg.sum())     # attach prop. to degree
        A[v, targets] = 1; A[targets, v] = 1
    return A
```

```{python}
#| fig-width: 7
#| fig-height: 3.4
#| out-width: "80%"
np.random.seed(1)
er = er_network(1000, 0.006); ba = ba_network(1000, 3)
fig, axs = plt.subplots(1, 2, figsize=(7, 3.4))
axs[0].hist(er.sum(1), bins=20, color="grey"); axs[0].set(title="Erdos-Renyi", xlabel="degree")
v, c = np.unique(ba.sum(1), return_counts=True)
axs[1].loglog(v, c, "o"); axs[1].set(title="Barabasi-Albert (log-log)", xlabel="degree", ylabel="count")
plt.tight_layout(); plt.show()
```

The Erdős-Rényi degrees cluster tightly around the mean, so a highly connected hub is
essentially impossible. The Barabási-Albert degrees instead fall off as a straight line
on log-log axes, the signature of a power law: most vertices have few links but a few
hubs have very many. Biological networks such as protein interaction networks resemble
the scale-free model far more than the random one, which is why hubs (often essential
genes) are a recurring theme.

## 10C.3 Community detection by modularity

Networks often divide into *communities*, groups of vertices more densely connected
among themselves than to the rest. Newman's *modularity* $Q$ measures how good a division
is, comparing the fraction of edges inside groups with the fraction expected if edges
were placed at random while preserving degrees [@newman2006modularity],

$$ Q = \frac{1}{2m} \sum_{i,j} \left( A_{ij} - \frac{k_i k_j}{2m} \right) \delta(c_i, c_j) , \tag{1} $$

where $m$ is the number of edges, $k_i$ the degree of vertex $i$, and $\delta(c_i, c_j)$
is 1 when $i$ and $j$ are in the same community. Newman's *spectral method* splits a
network in two using the *modularity matrix* $B_{ij} = A_{ij} - k_i k_j / 2m$: the signs
of the leading eigenvector of $B$ assign vertices to the two groups, the division that
best increases $Q$.

### R implementation {.impl}

```{r}
#| fig-width: 6
#| fig-height: 5
#| out-width: "65%"
# modularity of a given community assignment (Eq. 1)
modularity_Q <- function(A, membership) {
  m <- sum(A) / 2; k <- rowSums(A)
  B <- A - outer(k, k) / (2 * m)
  sum(B * outer(membership, membership, "==")) / (2 * m)
}

# Newman spectral bisection: sign of the leading eigenvector of the modularity matrix
spectral_split <- function(A) {
  m <- sum(A) / 2; k <- rowSums(A)
  B <- A - outer(k, k) / (2 * m)
  ifelse(eigen(B, symmetric = TRUE)$vectors[, 1] > 0, 1, 2)
}

comm <- spectral_split(A)
cat("community sizes:", table(comm), " modularity Q:", round(modularity_Q(A, comm), 3), "\n")
set.seed(1)
plot(g, vertex.color = c("tomato", "skyblue")[comm], vertex.size = 8,
     vertex.label = NA, layout = layout_with_fr)
```

### Python implementation {.impl}

```{python}
#| fig-width: 6
#| fig-height: 5
#| out-width: "65%"
# modularity of a given community assignment (Eq. 1)
def modularity_Q(A, membership):
    m = A.sum() / 2; k = A.sum(1); B = A - np.outer(k, k) / (2 * m)
    same = membership[:, None] == membership[None, :]
    return (B * same).sum() / (2 * m)

# Newman spectral bisection: sign of the leading eigenvector of the modularity matrix
def spectral_split(A):
    m = A.sum() / 2; k = A.sum(1); B = A - np.outer(k, k) / (2 * m)
    return (np.linalg.eigh(B)[1][:, -1] > 0).astype(int)

comm = spectral_split(A)
print("community sizes:", np.bincount(comm), " modularity Q:", round(modularity_Q(A, comm), 3))
plt.figure(figsize=(6, 5))
nx.draw(G, pos, node_color=["tomato" if c else "skyblue" for c in comm],
        node_size=120, edge_color="grey")
plt.show()
```

The spectral split cleanly separates the karate club into two communities of sizes 18
and 16, with modularity $Q \approx 0.37$, and this partition closely matches the actual
split of the club into two factions after a dispute, the outcome Zachary recorded. For
more than two communities one applies the bisection recursively or uses a fast optimizer
such as the Louvain method [@blondel2008fast] (`cluster_louvain` in igraph,
`greedy_modularity_communities` in networkx), which we would reach for on larger networks.

## 10C.4 Boolean network models

Gene regulatory networks can be modeled without kinetic parameters by treating each gene
as either on (1) or off (0) and updating its state by a logical rule of its regulators'
states [@kauffman1969metabolic]. The system state is a binary vector; in *synchronous*
updating, every gene is updated together at each discrete time step, $x_i(t+1) =
f_i(\mathbf{x}(t))$. Iterating from any start eventually falls into an *attractor*: a
fixed point (a steady state, $\mathbf{x}(t+1) = \mathbf{x}(t)$) or a cycle (an
oscillation). We enumerate the attractors by running every possible initial state to its
repeat.

### R implementation {.impl}

```{r}
# enumerate attractors: run every start state to the first repeat; the cycle is the attractor
boolean_attractors <- function(update, n) {
  states <- as.matrix(expand.grid(rep(list(0:1), n))); found <- list()
  for (r in 1:nrow(states)) {
    s <- states[r, ]; seen <- character(0); traj <- list()
    repeat {
      key <- paste(s, collapse = "")
      if (key %in% seen) {
        cyc <- traj[which(seen == key):length(seen)]
        canon <- paste(sort(sapply(cyc, paste, collapse = "")), collapse = "|")
        if (is.null(found[[canon]])) found[[canon]] <- cyc
        break
      }
      seen <- c(seen, key); traj[[length(traj) + 1]] <- s; s <- update(s)
    }
  }
  found
}
show_att <- function(a) sapply(a, function(cyc) paste0("{", paste(sapply(cyc, paste, collapse = ""), collapse = " -> "), "}"))

# toggle switch: two mutually repressing genes, A = NOT B, B = NOT A
toggle <- function(s) c(1 - s[2], 1 - s[1])
cat("toggle attractors:  ", show_att(boolean_attractors(toggle, 2)), "\n")

# Boolean repressilator: A -| B -| C -| A
repressilator <- function(s) c(1 - s[3], 1 - s[1], 1 - s[2])
cat("repressilator:      ", show_att(boolean_attractors(repressilator, 3)), "\n")
```

```{r}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
# state-transition graph of the toggle switch (each of 4 states -> its successor)
states <- as.matrix(expand.grid(A = 0:1, B = 0:1))
labels <- apply(states, 1, paste, collapse = "")
edges <- unlist(lapply(1:4, function(r) {
  nxt <- toggle(states[r, ]); c(labels[r], paste(nxt, collapse = ""))
}))
gt <- graph(edges, directed = TRUE)
set.seed(3)
plot(gt, vertex.size = 34, vertex.color = "lightgreen", edge.arrow.size = 0.5,
     main = "Toggle-switch state transitions")
```

### Python implementation {.impl}

```{python}
# enumerate attractors: run every start state to the first repeat; the cycle is the attractor
from itertools import product
def boolean_attractors(update, n):
    found = {}
    for s0 in product([0, 1], repeat=n):
        s = list(s0); seen = []
        while True:
            key = tuple(s)
            if key in seen:
                cyc = seen[seen.index(key):]; found[tuple(sorted(cyc))] = cyc; break
            seen.append(key); s = update(s)
    return list(found.values())
def show_att(a):
    return [" {" + " -> ".join("".join(map(str, s)) for s in cyc) + "}" for cyc in a]

# toggle switch: two mutually repressing genes, A = NOT B, B = NOT A
toggle = lambda s: [1 - s[1], 1 - s[0]]
print("toggle attractors: ", show_att(boolean_attractors(toggle, 2)))

# Boolean repressilator: A -| B -| C -| A
repressilator = lambda s: [1 - s[2], 1 - s[0], 1 - s[1]]
print("repressilator:     ", show_att(boolean_attractors(repressilator, 3)))
```

```{python}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
# state-transition graph of the toggle switch (each of 4 states -> its successor)
states = list(product([0, 1], repeat=2))
GT = nx.DiGraph()
for s in states:
    GT.add_edge("".join(map(str, s)), "".join(map(str, toggle(list(s)))))
plt.figure(figsize=(5, 5))
nx.draw(GT, nx.spring_layout(GT, seed=3), with_labels=True, node_size=1200,
        node_color="lightgreen", arrowsize=15)
plt.title("Toggle-switch state transitions"); plt.show()
```

The toggle switch has two fixed-point attractors, `10` and `01` (the two mutually
exclusive on-states of a bistable switch), plus a two-state oscillation `00 ↔ 11`; the
state-transition graph makes the two stable states and the cycle visible at a glance. The
Boolean repressilator has two cyclic attractors: a short `000 -> 111` oscillation and a
six-state cycle that is the discrete analog of the limit cycle we found for its continuous
version in earlier chapters. Despite their
extreme simplicity, Boolean networks reproduce the qualitative dynamics of real
regulatory circuits: Li et al. showed that the Boolean network of the yeast cell cycle
has a single dominant attractor corresponding to the biological cell-cycle sequence
[@li2004yeast], evidence that the network's wiring, not finely tuned parameters, sets its
behavior.
