---
title: "1A. Getting started with R and Python"
author: "Mingyang Lu"
old_id: "01A"
---

## 1A.1 Setting up your environment

This book runs R and Python side by side, with [Quarto](https://quarto.org)
rendering both into one document and [VS Code](https://code.visualstudio.com)
as the editor. Install, in this order:

1. **R**, from [CRAN](https://cran.r-project.org).
2. **Python** (3.x), from [python.org](https://www.python.org/downloads/) or
   your system's package manager.
3. **Quarto**, from [quarto.org/docs/get-started](https://quarto.org/docs/get-started/).
4. **VS Code**, from [code.visualstudio.com](https://code.visualstudio.com),
   then add three extensions from the Extensions panel: **R** (by
   REditorSupport), **Python** (by Microsoft), and **Quarto**.
5. **Claude Code**, a command-line AI assistant that reads, edits, and runs
   code inside your own project rather than a generic chat window. Install it
   with

   ```bash
   npm install -g @anthropic-ai/claude-code
   ```

   which requires [Node.js](https://nodejs.org). Anthropic also publishes a
   VS Code extension for it, so a Claude Code panel can sit next to your
   editor; check Anthropic's current documentation for the exact setup, since
   install methods change over time.

Once these are installed, open a project folder in VS Code and run `claude`
from the integrated terminal to start a session there. The rest of this
chapter is about what to do with that session once it's open.

## 1A.2 Learning numerical methods with AI

An AI assistant can now turn a short description into a working script, so
producing code is no longer the hard part of programming. What an assistant
cannot do reliably is tell you whether the result is right. That gap matters
most in numerical work, where a wrong method rarely crashes: it returns a
number that still plots as a smooth curve. The skills worth building are the
ones the assistant lacks, understanding a method well enough to know what it
should produce, and checking that it did.

This is where the code in this book earns its place. Every method is given as a
correct, tested implementation, already checked against a known answer. That
makes it a fixed reference you can trust, while code an assistant generates is
only plausible until you prove otherwise. You will use the provided code four
ways: as a worked example to read, as a target to rebuild, as the reference to
check your own or an assistant's version against, and as a starting point to
modify once you want to go beyond the chapter.

A good way to work through each method is:

1. **Understand the method before looking at any code.** From the concept and
   the equations, decide what the result should be and how you will know it is
   right, for instance a closed-form solution or a quantity that must stay
   constant.
2. **Read the provided implementation** to see one correct version, and ask the
   assistant to explain any line you do not recognize.
3. **Rebuild it yourself**, letting the assistant fill in parts you can describe
   precisely. Rewriting the method is what fixes it in your mind; accepting
   generated code without this step does not.
4. **Verify against the reference and the check from step 1**, not against how
   the output looks. Run your version and the book's on the same input and
   compare. Where they disagree is where the learning is.
5. **Extend it past the chapter**: change the model, add a variable, alter a
   boundary. The assistant makes this quick, and the understanding from the
   earlier steps keeps the extension honest. This is also where your work stops
   resembling everyone else's.

To support this, each method in the chapters that follow opens with a short box
giving its objective, the method, a test case, and how to verify the result.
Read it, then pause and attempt the code before reading the implementation.
Even a failed attempt makes the reference far more useful when you turn to it.

When you do ask the assistant, ask well. Name the actual code and the actual
confusion: paste the function, then ask what a specific operator, argument, or
line does and why. A vague request such as "explain this file" returns a vague
answer. Verify the explanation the same way you verify code: change an input,
predict the result, run it, and compare. An explanation you have not checked is
a plausible guess, and assistants produce those freely.

The rest of this chapter demonstrates the reading-and-checking part of this loop
on three small examples of increasing depth: a script you can read at a glance,
three R and Python quirks that reward a careful read, and one complete
simulation. Every session shown was tested with Claude Opus 4.8; a different
model or version will phrase things differently, but the paste-ask-verify
technique is the same.

## 1A.3 A minimal example

Here is the smallest script that still has something worth asking about: a
comment, a variable, a function, and a loop with a condition inside it.

### R implementation {.impl}

```{r}
# classify each number in a list as even or odd
numbers <- c(4, 7, 10, 13)
classify <- function(x) {
  if (x %% 2 == 0) "even" else "odd"
}
for (n in numbers) {
  cat(n, "is", classify(n), "\n")
}
```

### Python implementation {.impl}

```{python}
# classify each number in a list as even or odd
numbers = [4, 7, 10, 13]
def classify(x):
    return "even" if x % 2 == 0 else "odd"
for n in numbers:
    print(n, "is", classify(n))
```

To run this yourself, create a new file `classify.R` (or `classify.py`) in
VS Code, paste in the code above, and use the Run button the R or Python
extension adds to the editor, or run `Rscript classify.R` /
`python3 classify.py` from the integrated terminal.

Then open a Claude Code session in the same folder: split the editor so the
file you are reading stays visible on one side while Claude Code runs on the
other, rather than switching back and forth between two overlapping windows.

::: {.content-visible when-format="html:js"}
<table class="vscode-mockup">
<tr><th>01a-getting-started.qmd</th><th>TERMINAL - claude</th></tr>
<tr>
<td><pre>## 1A.3 A minimal example

numbers &lt;- c(4, 7, 10, 13)
classify &lt;- function(x) {
  if (x %% 2 == 0) "even" else "odd"
}
for (n in numbers) {
  cat(n, "is", classify(n), "\n")
}</pre></td>
<td><pre>&gt; Explain the classify function
&gt; below, line by line.

In R, a function returns the value of
the last expression it evaluates, and
an if/else block is itself an
expression...</pre></td>
</tr>
</table>
:::

::: {.content-visible unless-format="html:js"}
```
┌─ VS Code ────────────────────────────────┬────────────────────────────────┐
│ 01a-getting-started.qmd                  │ TERMINAL - claude              │
│ ───────────────────────────────────────  │ ─────────────────────────────  │
│ ## 1A.3 A minimal example                 │ > Explain the classify         │
│                                            │ > function below, line by      │
│ numbers <- c(4, 7, 10, 13)               │ > line.                        │
│ classify <- function(x) {                │                                 │
│   if (x %% 2 == 0) "even" else "odd"     │ In R, a function returns the   │
│ }                                         │ value of the last expression   │
│ for (n in numbers) {                     │ it evaluates, and an if/else   │
│   cat(n, "is", classify(n), "\n")        │ block is itself an             │
│ }                                         │ expression...                  │
└────────────────────────────────────────────┴────────────────────────────────┘
```
:::

Ask about the one part that is not obvious on a first read: the inline
`if (...) "even" else "odd"` in R, or the equivalent conditional expression
in Python. A session might go like this:

```
> What does `if (x %% 2 == 0) "even" else "odd"` return, and why is there no
> explicit `return`?

In R, a function returns the value of the last expression it evaluates, and
an if/else block is itself an expression: it evaluates to whichever branch
ran. So this line evaluates to the string "even" or "odd", and because it is
the only expression in `classify`, that string becomes the function's return
value. `%%` is R's modulo operator, so `x %% 2 == 0` is TRUE exactly when x
is even.
```

Verify it: call `classify(11)` and confirm you get `"odd"`, then try
`classify(0)` and check the answer still makes sense.

## 1A.4 A second example: three R and Python quirks

Some R and Python behavior looks fine until it silently is not. These three
come up constantly in scientific code and reward a deliberate AI-assisted
read rather than a guess.

**Ellipsis and `**kwargs`.** R's `...` and Python's `**kwargs` let a function
forward whatever arguments it received on to another function whose full
signature it does not need to know. This pattern shows up in nearly every
integrator in this book, starting with `Euler()` in
[Part 2B](../02-odes/02b-numerical-integration.qmd).

### R implementation {.impl}

```{r}
# a function whose second argument is itself a function, forwarded extra
# arguments via ...
func_main <- function(a, func_2nd, ...) {
  return(a + func_2nd(...))
}
func1 <- function(d, e) d + e
func_main(a = 1, func_2nd = func1, d = 2.4, e = 0.6)
```

### Python implementation {.impl}

```{python}
# a function whose second argument is itself a function, forwarded extra
# arguments via **kwargs
def func_main(a, func_2nd, **kwargs):
    return a + func_2nd(**kwargs)
def func1(d, e):
    return d + e
func_main(a=1, func_2nd=func1, d=2.4, e=0.6)
```

Paste `func_main` into Claude Code and ask "what does `...`/`**kwargs` do
here, and why does `func1` need to accept exactly `d` and `e`?" A good answer
explains that the ellipsis collects any named arguments beyond `a` and
`func_2nd`, then re-expands them at `func_2nd(...)`. If `func1`'s parameter
names do not match what was forwarded, the call fails. Verify it: rename
`func1`'s parameters to `f` and `g` without changing the call, and confirm
you get an error, since neither `d` nor `e` is among `func1`'s parameter
names anymore.

**The factor-vs-string default.** R's `data.frame()` used to convert every
character column into a factor by default. R 4.0, released in 2020, flipped
that default so strings stay strings. Code written for the older default, or
a tutorial that assumes it, can behave differently on a modern R install with
no error at all, just a different column type. This is R-specific: Python's
`pandas.DataFrame` never auto-converts strings this way, so there is no
equivalent quirk to show on the Python side.

### R implementation {.impl}

```{r}
my_data <- data.frame(
  atom = c("N", "CA", "CB", "C", "O"),
  mass = c(14, 12, 12, 12, 16)
)
class(my_data$atom)   # character, on R 4.0 and later

my_data_old_default <- data.frame(
  atom = c("N", "CA", "CB", "C", "O"),
  mass = c(14, 12, 12, 12, 16),
  stringsAsFactors = TRUE
)
class(my_data_old_default$atom)   # factor, matching R's pre-4.0 default
```

Ask Claude Code "why would the same `data.frame()` call produce a character
column on one R installation and a factor on another?" A good answer names
the R 4.0 default change directly, not just "it depends on your settings."
Verify it: run `R.version.string` and check which behavior your own
installation gives by default.

**Data frame vs. matrix.** A matrix requires every element to be the same
type and supports linear algebra directly. A data frame allows a different
type per column and is what most statistics and plotting functions in R
expect. This is also R-specific enough to leave as a single example: Python's
analogous distinction (a NumPy array vs. a pandas DataFrame) is less prone to
this particular mix-up, since the two rarely get passed to the same function
interchangeably.

### R implementation {.impl}

```{r}
my_data <- data.frame(
  mass = c(14, 12, 12, 12, 16),
  size = c(1.2, 1.4, 1.4, 1.4, 1.1)
)
model <- lm(size ~ mass, data = my_data)
coef(model)
```

Ask Claude Code "why does `lm()` take a data frame here instead of a matrix,
and what would break if `my_data` had a character column mixed in with `mass`
and `size`?" Verify it: add an `atom` character column to `my_data` and
confirm `lm(size ~ mass, data = my_data)` still runs unaffected, since a
column not named in the formula just comes along for the ride.

## 1A.5 A worked example: Conway's Game of Life

Conway's Game of Life is a grid of cells, each alive or dead, updated
together at every step by three rules applied to each cell's eight
neighbors [@gardner1970game]:

- A live cell with 2 or 3 live neighbors survives.
- A dead cell with exactly 3 live neighbors becomes alive.
- Every other cell dies or stays dead.

### R implementation {.impl}

The code below is shown only in R; the point of this example is the
AI-assisted reading technique, not the simulation itself, and the technique
applies the same way in Python.

```r
n <- 10
grid <- matrix(0, n, n)

# a glider: a five-cell pattern that translates diagonally across the grid
# one cell every four generations, without changing shape
glider <- rbind(c(1, 2), c(2, 3), c(3, 1), c(3, 2), c(3, 3))
for (i in seq_len(nrow(glider))) grid[glider[i, 1], glider[i, 2]] <- 1

# count each cell's live neighbors
neighbor_sum <- function(g) {
  n <- nrow(g)
  idx_r <- row(g); idx_c <- col(g)
  total <- matrix(0, n, n)
  for (di in -1:1) for (dj in -1:1) {
    if (di == 0 && dj == 0) next
    r  <- ((idx_r - 1 + di) %% n) + 1
    cc <- ((idx_c - 1 + dj) %% n) + 1
    total <- total + matrix(g[cbind(as.vector(r), as.vector(cc))], n, n)
  }
  total
}

# apply the three rules to every cell at once
step <- function(g) {
  nb <- neighbor_sum(g)
  ((nb == 3) | (g == 1 & nb == 2)) * 1
}

library(ggplot2); library(gganimate); library(gifski)

n_gen <- 40
frames <- vector("list", n_gen)
g <- grid
for (t in seq_len(n_gen)) {
  frames[[t]] <- data.frame(row = as.vector(row(g)), col = as.vector(col(g)),
                             state = as.vector(g), generation = t)
  g <- step(g)
}
df <- do.call(rbind, frames)
df$state <- factor(df$state, levels = c(0, 1))

p <- ggplot(df, aes(x = col, y = row, fill = state)) +
  geom_tile(color = "grey80") +
  scale_fill_manual(values = c("0" = "white", "1" = "red"), guide = "none") +
  scale_y_reverse() + coord_fixed() +
  labs(title = "Generation: {current_frame}", x = NULL, y = NULL) +
  theme_void() + theme(plot.title = element_text(hjust = 0.5))

anim <- animate(p + transition_manual(generation), fps = 5, nframes = n_gen,
                 width = 400, height = 400, renderer = gifski_renderer())
anim_save("game-of-life.gif", animation = anim)
```

::: {.content-visible when-format="html:js"}
![Forty generations of a glider on a 10x10 grid. Red cells are alive, white
cells are dead.](images/01A/game-of-life.gif){width=40% fig-align="center"}
:::

Watch the animation closely around generation 35 to 40: part of the glider
appears to jump to the opposite side of the grid rather than sliding off it.
The `neighbor_sum` function above never says anything about what happens at
the edges of the 10x10 grid, so paste it into Claude Code and ask directly:

```
> What is `((idx_r - 1 + di) %% n) + 1` doing, and why does the glider
> reappear on the opposite edge instead of just disappearing off the grid?

The %% (modulo) here wraps an index back into the range 1..n whenever di or
dj would push it past an edge: a neighbor "above row 1" becomes row n, and a
neighbor "past column n" becomes column 1. That makes the grid a torus
rather than a flat sheet with edges, so a cell near row 1 is a genuine
neighbor of a cell in row n. This is called a periodic boundary condition:
instead of an edge where nothing happens or a wall, the far edge of the grid
is treated as touching the near edge. That is exactly why the glider
reappears on the opposite side rather than vanishing.
```

Verify it: pick a cell in the top row and compute `((0 - 1 + -1) %% 10) + 1`
by hand to confirm it does map to row 10, not to a negative or out-of-range
index.

Periodic boundaries are not just a trick for a small grid of cells.
[Part 5C](../05-molecular-dynamics/05c-box-of-particles.qmd) applies the same
idea to a continuous-space molecular dynamics simulation: a particle leaving
one side of a simulation box re-enters from the opposite side, using a
function (`bc_periodic`) built on the same wraparound rule shown here.

::: {.callout-tip title="Try it yourself"}
Pick any function elsewhere in this book that you have not already read in
depth, a later chapter's ODE integrator, or `neighbor_sum` above with a
different neighborhood shape. Paste it into Claude Code, ask one specific
question about the one part you do not understand, and check the answer by
changing an input and predicting the output before you run it.
:::
