---
title: "1A. Getting started with R and Python"
author: "Mingyang Lu"
old_id: "01A"
---

## 1A.1 Setting up your environment

This book runs R and Python side by side, with [Quarto](https://quarto.org)
rendering both into one document and [VS Code](https://code.visualstudio.com)
as the editor. Install, in this order:

1. **R**, from [CRAN](https://cran.r-project.org).
2. **Python** (3.x), from [python.org](https://www.python.org/downloads/) or
   your system's package manager.
3. **Quarto**, from [quarto.org/docs/get-started](https://quarto.org/docs/get-started/).
4. **VS Code**, from [code.visualstudio.com](https://code.visualstudio.com),
   then add three extensions from the Extensions panel: **R** (by
   REditorSupport), **Python** (by Microsoft), and **Quarto**.
5. **Claude Code**, a command-line AI assistant that reads, edits, and runs
   code inside your own project rather than a generic chat window. Install it
   with

   ```bash
   npm install -g @anthropic-ai/claude-code
   ```

   which requires [Node.js](https://nodejs.org). Anthropic also publishes a
   VS Code extension for it, so a Claude Code panel can sit next to your
   editor; check Anthropic's current documentation for the exact setup, since
   install methods change over time.

Once these are installed, open a project folder in VS Code and run `claude`
from the integrated terminal to start a session there. The rest of this
chapter is about what to do with that session once it's open.

## 1A.2 Learning numerical methods with AI

An AI assistant can now turn a short description into a working script, so
producing code is no longer the hard part of programming. What an assistant
cannot do reliably is tell you whether the result is right. That gap matters
most in numerical work, where a wrong method rarely crashes: it returns a
number that still plots as a smooth curve. The skills worth building are the
ones the assistant lacks, understanding a method well enough to know what it
should produce, and checking that it did.

This is where the code in this book earns its place. Every method is given as a
correct, tested implementation, already checked against a known answer. That
makes it a fixed reference you can trust, while code an assistant generates is
only plausible until you prove otherwise. You will use the provided code four
ways: as a worked example to read, as a target to rebuild, as the reference to
check your own or an assistant's version against, and as a starting point to
modify once you want to go beyond the chapter.

A good way to work through each method is:

1. **Understand the method before looking at any code.** From the concept and
   the equations, decide what the result should be and how you will know it is
   right, for instance a closed-form solution or a quantity that must stay
   constant.
2. **Read the provided implementation** to see one correct version, and ask the
   assistant to explain any line you do not recognize.
3. **Rebuild it yourself**, letting the assistant fill in parts you can describe
   precisely. Rewriting the method is what fixes it in your mind; accepting
   generated code without this step does not.
4. **Verify against the reference and the check from step 1**, not against how
   the output looks. Run your version and the book's on the same input and
   compare. Where they disagree is where the learning is.
5. **Extend it past the chapter**: change the model, add a variable, alter a
   boundary. The assistant makes this quick, and the understanding from the
   earlier steps keeps the extension honest. This is also where your work stops
   resembling everyone else's.

To make this concrete, each worked method in this book is framed as a **recipe**, a
short box with six fields: **Objective** (what the code should accomplish),
**Model** (the governing equations and kinetics, stated fully enough to reproduce the
system), **Method** (the numerical approach), **Test** (a concrete case to run, with
every parameter, input, and initial condition), **Show** (the output to produce, such
as a plot or printed values), and **Verification** (how to tell the result is right,
which is a check on the output, not the output itself). The Model and Test fields are
what make a recipe self-contained: the equations plus the parameters are enough to
rebuild the system without seeing the code. A recipe is something to read and attempt
before you look at the implementation. Even a failed attempt makes the reference code
far more useful when you turn to it. (The Model field is dropped for the occasional
example that has no governing equations.)

Recipe boxes appear for the major methods and applications, not for every code block.
Each was first drafted by an AI assistant (Claude Opus 4.8) and then checked and
corrected by hand. Producing them this way doubles as a test of the assistant itself:
a method it cannot compress into a correct objective, test, and verification is one it
has not really understood. Every box is collected in the
[recipe index](appendix-recipe-index.qmd).

A recipe also converts into a **structured prompt**, which is how you get help on
the parts you cannot yet write without giving up the understanding. Instead of a
vague request, you fill a fixed template with the recipe's fields:

> In [R or Python], [Objective]. The model is [Model]. Use [Method]. Test it on
> [Test]. Produce [Show], and as a separate check, confirm that [Verification].
> Implement the method explicitly with short comments rather than calling a routine
> that does it in one step, and explain in a sentence why that check confirms the
> result.

Write it in plain text, typing any equation the way you would in an email; an
assistant reads `dX/dt = g - kX` as readily as the typeset form. Asking for the output
and the check separately matters: the thing to produce (the Show field) is usually not
the same as the quantity you verify against, so naming both keeps the assistant from
returning only the check. The last two requests are what make this a learning prompt
rather than one that just vends code: writing the method out so you can see it, and
saying why the check confirms correctness. In the HTML edition, each recipe box carries
a **Copy prompt** button that fills this template from the box for you. Whatever the
assistant returns, run the verification yourself before you trust it.

When you instead ask about code that already exists, ask well. Name the actual code
and the actual confusion: paste the function, then ask what a specific operator,
argument, or line does and why. A vague request such as "explain this file" returns
a vague answer. Verify the explanation the same way you verify code: change an
input, predict the result, run it, and compare. An explanation you have not checked
is a plausible guess, and assistants produce those freely.

The next section works this recipe-and-prompt loop through a small example. The two
after it introduce the R and Python features you will meet most often, and a
complete simulation to read closely. Every session shown was tested with Claude Opus
4.8; a different model or version will phrase things differently, but the technique
is the same.

## 1A.3 A minimal example

Here is the smallest script that still has something worth asking about: a
comment, a variable, a function, and a loop with a condition inside it. Here is
its recipe:

::: {.callout-note title="Recipe 1A.3.1"}
**Objective.** Label each number in a list as "even" or "odd".

**Method.** Loop over the numbers, test each with the modulo (remainder) operator,
and print its label.

**Test.** The list 4, 7, 10, 13.

**Show.** The printed even/odd label for each number in the list.

**Verification.** 4 and 10 print as even; 7 and 13 print as odd.
:::

Try the recipe first: the **Copy prompt** button on the box turns it into a prompt
you can hand to an assistant, then compare what comes back with the version below.

### R implementation {.impl}

```{r}
# classify each number in a list as even or odd
numbers <- c(4, 7, 10, 13)
classify <- function(x) {
  if (x %% 2 == 0) "even" else "odd"
}
for (n in numbers) {
  cat(n, "is", classify(n), "\n")
}
```

### Python implementation {.impl}

```{python}
# classify each number in a list as even or odd
numbers = [4, 7, 10, 13]
def classify(x):
    return "even" if x % 2 == 0 else "odd"
for n in numbers:
    print(n, "is", classify(n))
```

To run this yourself, create a new file `classify.R` (or `classify.py`) in
VS Code, paste in the code above, and use the Run button the R or Python
extension adds to the editor, or run `Rscript classify.R` /
`python3 classify.py` from the integrated terminal.

Then open a Claude Code session in the same folder: split the editor so the
file you are reading stays visible on one side while Claude Code runs on the
other, rather than switching back and forth between two overlapping windows.

::: {.content-visible when-format="html:js"}
<table class="vscode-mockup">
<tr><th>01a-getting-started.qmd</th><th>TERMINAL - claude</th></tr>
<tr>
<td><pre>## 1A.3 A minimal example

numbers &lt;- c(4, 7, 10, 13)
classify &lt;- function(x) {
  if (x %% 2 == 0) "even" else "odd"
}
for (n in numbers) {
  cat(n, "is", classify(n), "\n")
}</pre></td>
<td><pre>&gt; Explain the classify function
&gt; below, line by line.

In R, a function returns the value of
the last expression it evaluates, and
an if/else block is itself an
expression...</pre></td>
</tr>
</table>
:::

::: {.content-visible unless-format="html:js"}
```
┌─ VS Code ────────────────────────────────┬────────────────────────────────┐
│ 01a-getting-started.qmd                  │ TERMINAL - claude              │
│ ───────────────────────────────────────  │ ─────────────────────────────  │
│ ## 1A.3 A minimal example                 │ > Explain the classify         │
│                                            │ > function below, line by      │
│ numbers <- c(4, 7, 10, 13)               │ > line.                        │
│ classify <- function(x) {                │                                 │
│   if (x %% 2 == 0) "even" else "odd"     │ In R, a function returns the   │
│ }                                         │ value of the last expression   │
│ for (n in numbers) {                     │ it evaluates, and an if/else   │
│   cat(n, "is", classify(n), "\n")        │ block is itself an             │
│ }                                         │ expression...                  │
└────────────────────────────────────────────┴────────────────────────────────┘
```
:::

Ask about the one part that is not obvious on a first read: the inline
`if (...) "even" else "odd"` in R, or the equivalent conditional expression
in Python. A session might go like this:

```
> What does `if (x %% 2 == 0) "even" else "odd"` return, and why is there no
> explicit `return`?

In R, a function returns the value of the last expression it evaluates, and
an if/else block is itself an expression: it evaluates to whichever branch
ran. So this line evaluates to the string "even" or "odd", and because it is
the only expression in `classify`, that string becomes the function's return
value. `%%` is R's modulo operator, so `x %% 2 == 0` is TRUE exactly when x
is even.
```

Verify it: call `classify(11)` and confirm you get `"odd"`, then try
`classify(0)` and check the answer still makes sense.

The R and Python versions do the same thing with different spellings, and that
contrast is worth asking about too. Paste both blocks and ask the assistant to list
the differences line by line. A good answer points out that R builds the list with
`c(...)` while Python uses `[...]`; R's remainder operator is `%%` and Python's is
`%`; R returns the value of its last expression, the bare `if/else`, whereas Python
needs an explicit `return` with the conditional expression `A if cond else B`; and R
prints with `cat`, Python with `print`. Seeing one idea in two spellings is the
fastest way to stop treating either language's punctuation as magic.

A few other ways to use an assistant on a script you are reading:

- **Predict, then run.** Before running, tell the assistant what you expect each
  line to print and ask whether you are right. Being wrong out loud is where the
  learning happens.
- **Translate and diff.** Ask it to rewrite the R version in Python (or the
  reverse), run both, and confirm the outputs match. A translation that agrees line
  for line means you understood the original.
- **Change one thing.** Ask what happens if you replace `== 0` with `== 1`, then
  check the prediction by editing and re-running.

Each of these keeps you doing the thinking and uses the assistant to check it, which
is the habit the rest of the book is built around.

## 1A.4 A little R and Python

This book teaches R and Python by using them rather than through a full language
course. For a proper grounding, these are good starting points.

For R:

- [An Introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html), the official manual from CRAN.
- [R & Bioconductor Manual](http://manuals.bioinformatics.ucr.edu/home/R_BioCondManual) by Thomas Girke, aimed at biological data.
- [Efficient R Programming](https://csgillespie.github.io/efficientR/) by Colin Gillespie and Robin Lovelace.

For Python:

- [The Python Tutorial](https://docs.python.org/3/tutorial/), the official introduction from python.org.
- [Scientific Python Lectures](https://lectures.scientific-python.org), covering NumPy, SciPy, and Matplotlib.
- The reference documentation for [NumPy](https://numpy.org/doc/stable/) and [Matplotlib](https://matplotlib.org/stable/).

Below are two features new readers most often stumble on, each shown in both
languages: forwarding arguments with an ellipsis, and the difference between a
matrix and a data frame. When a later chapter's code looks unfamiliar, that is the
moment to paste it into an assistant and ask.

**Ellipsis and `**kwargs`.** R's `...` and Python's `**kwargs` let a function
forward whatever arguments it received on to another function whose full
signature it does not need to know. This pattern shows up in nearly every
integrator in this book, starting with `Euler()` in
[Part 2B](../02-odes/02b-numerical-integration.qmd).

### R implementation {.impl}

```{r}
# a function whose second argument is itself a function, forwarded extra
# arguments via ...
func_main <- function(a, func_2nd, ...) {
  return(a + func_2nd(...))
}
func1 <- function(d, e) d + e
func_main(a = 1, func_2nd = func1, d = 2.4, e = 0.6)
```

### Python implementation {.impl}

```{python}
# a function whose second argument is itself a function, forwarded extra
# arguments via **kwargs
def func_main(a, func_2nd, **kwargs):
    return a + func_2nd(**kwargs)
def func1(d, e):
    return d + e
func_main(a=1, func_2nd=func1, d=2.4, e=0.6)
```

Paste `func_main` into Claude Code and ask "what does `...`/`**kwargs` do
here, and why does `func1` need to accept exactly `d` and `e`?" A good answer
explains that the ellipsis collects any named arguments beyond `a` and
`func_2nd`, then re-expands them at `func_2nd(...)`. If `func1`'s parameter
names do not match what was forwarded, the call fails. Verify it: rename
`func1`'s parameters to `f` and `g` without changing the call, and confirm
you get an error, since neither `d` nor `e` is among `func1`'s parameter
names anymore.

**Matrix vs. data frame.** A **matrix** holds elements of a single type and
supports linear algebra directly; a **data frame** allows a different type per
column and is what most statistics and plotting functions expect. Confusing the two
is a common source of trouble, and both languages draw the same distinction. In R,
`matrix()` builds the first and `data.frame()` the second.

### R implementation {.impl}

```{r}
# a matrix holds one type and does linear algebra
m <- matrix(c(1, 2, 3, 4), nrow = 2)
m %*% m                                   # matrix multiplication

# a data frame holds a column per variable, each with its own type
df <- data.frame(atom = c("N", "CA", "C", "O"),   # character column
                 mass = c(14, 12, 12, 16))          # numeric column
str(df)                                   # each column keeps its own type
```

### Python implementation {.impl}

Python draws the same line between a NumPy array (one type, for linear algebra) and
a pandas data frame (a column per variable, mixed types).

```{python}
import numpy as np
import pandas as pd

# a NumPy array holds one type and does linear algebra
m = np.array([[1, 2], [3, 4]])
print(m @ m)                              # matrix multiplication

# a pandas DataFrame holds a column per variable, each with its own type
df = pd.DataFrame({"atom": ["N", "CA", "C", "O"],   # string column
                   "mass": [14, 12, 12, 16]})         # numeric column
print(df.dtypes)                          # each column keeps its own type
```

Ask the assistant why a matrix cannot hold a text column alongside numbers, and when
you would reach for each. Verify it: try to place a string into the numeric matrix
(`matrix(c(1, "a", 3, 4), 2)` in R, `np.array([1, "a", 3, 4])` in Python) and watch
every element turn into text, since a matrix or array must share one type.

## 1A.5 A worked example: Conway's Game of Life

Conway's Game of Life is a grid of cells, each alive or dead, updated
together at every step by three rules applied to each cell's eight
neighbors [@gardner1970game]:

- A live cell with 2 or 3 live neighbors survives.
- A dead cell with exactly 3 live neighbors becomes alive.
- Every other cell dies or stays dead.

::: {.callout-note title="Recipe 1A.5.1"}
**Objective.** Simulate Conway's Game of Life and follow a glider across the grid.

**Model.** A grid of cells, each alive (1) or dead (0), updated synchronously by three rules applied to each cell's eight neighbors: a live cell with 2 or 3 live neighbors survives; a dead cell with exactly 3 live neighbors becomes alive; every other cell is dead next generation. The grid edges wrap around (a periodic boundary), so it is a torus.

**Method.** Count each cell's live neighbors with wraparound, then apply the three rules to all cells at once.

**Test.** A 10x10 grid seeded with a glider at cells (1,2), (2,3), (3,1), (3,2), (3,3); run for several generations.

**Show.** The grid at several successive generations, with the glider visible.

**Verification.** The glider keeps its five-cell shape and drifts one step diagonally every four generations; on reaching an edge it reappears on the opposite side.
:::

### R implementation {.impl}

```{r}
#| rmar: false
n <- 10
grid <- matrix(0, n, n)
# a glider: a five-cell pattern that translates diagonally without changing shape
glider <- rbind(c(1, 2), c(2, 3), c(3, 1), c(3, 2), c(3, 3))
for (i in seq_len(nrow(glider))) grid[glider[i, 1], glider[i, 2]] <- 1

# count each cell's eight live neighbors, with wraparound (periodic) edges
neighbor_sum <- function(g) {
  n <- nrow(g)
  idx_r <- row(g); idx_c <- col(g)
  total <- matrix(0, n, n)
  for (di in -1:1) for (dj in -1:1) {
    if (di == 0 && dj == 0) next
    r  <- ((idx_r - 1 + di) %% n) + 1
    cc <- ((idx_c - 1 + dj) %% n) + 1
    total <- total + matrix(g[cbind(as.vector(r), as.vector(cc))], n, n)
  }
  total
}

# one generation: apply the three rules to every cell at once
step <- function(g) {
  nb <- neighbor_sum(g)
  ((nb == 3) | (g == 1 & nb == 2)) * 1
}

# run four generations (one full glider cycle) and show the start and end
g_end <- grid
for (t in 1:4) g_end <- step(g_end)
par(mfrow = c(1, 2), mar = c(1, 1, 2, 1), pty = "s")
image(1:n, 1:n, t(grid[n:1, ]),  col = c("white", "red"), axes = FALSE,
      xlab = "", ylab = "", main = "generation 1")
image(1:n, 1:n, t(g_end[n:1, ]), col = c("white", "red"), axes = FALSE,
      xlab = "", ylab = "", main = "generation 5")
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

n = 10
grid = np.zeros((n, n), dtype=int)
# a glider: a five-cell pattern that translates diagonally without changing shape
glider = [(0, 1), (1, 2), (2, 0), (2, 1), (2, 2)]
for i, j in glider:
    grid[i, j] = 1

# count each cell's eight live neighbors, with wraparound (periodic) edges
def neighbor_sum(g):
    total = np.zeros_like(g)
    for di in (-1, 0, 1):
        for dj in (-1, 0, 1):
            if di == 0 and dj == 0:
                continue
            total += np.roll(np.roll(g, di, axis=0), dj, axis=1)  # roll wraps around
    return total

# one generation: apply the three rules to every cell at once
def step(g):
    nb = neighbor_sum(g)
    return ((nb == 3) | ((g == 1) & (nb == 2))).astype(int)

# run four generations (one full glider cycle) and show the start and end
g_end = grid.copy()
for _ in range(4):
    g_end = step(g_end)
fig, ax = plt.subplots(1, 2, figsize=(6, 3))
# assign each call to _ so its matplotlib return object is not echoed as output
for a, g, title in zip(ax, [grid, g_end], ["generation 1", "generation 5"]):
    _ = a.imshow(g, cmap="Reds", vmin=0, vmax=1)
    _ = a.set_title(title)
    _ = a.axis("off")
plt.show()
```

The two snapshots show the glider one full cycle later: the same five-cell shape,
shifted one step diagonally. A pair of frames only hints at the motion, so for a
clearer view we render the whole run as an animation. The code below (R's `gganimate`
here) is shown as a listing you can run; the resulting GIF is pre-generated and
embedded, so the book does not rebuild it on every render.

```r
library(ggplot2); library(gganimate); library(gifski)

n_gen <- 40
frames <- vector("list", n_gen)
g <- grid
for (t in seq_len(n_gen)) {
  frames[[t]] <- data.frame(row = as.vector(row(g)), col = as.vector(col(g)),
                             state = as.vector(g), generation = t)
  g <- step(g)
}
df <- do.call(rbind, frames)
df$state <- factor(df$state, levels = c(0, 1))

p <- ggplot(df, aes(x = col, y = row, fill = state)) +
  geom_tile(color = "grey80") +
  scale_fill_manual(values = c("0" = "white", "1" = "red"), guide = "none") +
  scale_y_reverse() + coord_fixed() +
  labs(title = "Generation: {current_frame}", x = NULL, y = NULL) +
  theme_void() + theme(plot.title = element_text(hjust = 0.5))

anim <- animate(p + transition_manual(generation), fps = 5, nframes = n_gen,
                 width = 400, height = 400, renderer = gifski_renderer())
anim_save("game-of-life.gif", animation = anim)
```

::: {.content-visible when-format="html:js"}
![Forty generations of a glider on a 10x10 grid. Red cells are alive, white
cells are dead.](images/01A/game-of-life.gif){width=40% fig-align="center"}
:::

Watch the animation closely around generation 35 to 40: part of the glider
appears to jump to the opposite side of the grid rather than sliding off it.
The `neighbor_sum` function above never says anything about what happens at
the edges of the 10x10 grid, so paste it into Claude Code and ask directly:

```
> What is `((idx_r - 1 + di) %% n) + 1` doing, and why does the glider
> reappear on the opposite edge instead of just disappearing off the grid?

The %% (modulo) here wraps an index back into the range 1..n whenever di or
dj would push it past an edge: a neighbor "above row 1" becomes row n, and a
neighbor "past column n" becomes column 1. That makes the grid a torus
rather than a flat sheet with edges, so a cell near row 1 is a genuine
neighbor of a cell in row n. This is called a periodic boundary condition:
instead of an edge where nothing happens or a wall, the far edge of the grid
is treated as touching the near edge. That is exactly why the glider
reappears on the opposite side rather than vanishing.
```

Verify it: pick a cell in the top row and compute `((0 - 1 + -1) %% 10) + 1`
by hand to confirm it does map to row 10, not to a negative or out-of-range
index.

Periodic boundaries are not just a trick for a small grid of cells.
[Part 5C](../05-molecular-dynamics/05c-box-of-particles.qmd) applies the same
idea to a continuous-space molecular dynamics simulation: a particle leaving
one side of a simulation box re-enters from the opposite side, using a
function (`bc_periodic`) built on the same wraparound rule shown here.

::: {.callout-tip title="Try it yourself"}
Compare the two `neighbor_sum` functions above: the R version wraps indices with the
modulo operator (`%%`), while the Python version shifts the whole grid with
`np.roll`. Ask the assistant why both produce the same periodic boundary. Then pick
any function you have not read closely, this one or a later chapter's ODE integrator,
paste one version into Claude Code, ask one specific question about the part you do
not understand, and check the answer by changing an input and predicting the output
before you run it.
:::
