---
title: "1B. Efficient programming"
author: "Mingyang Lu"
old_id: "01B"
---

Our goal is not just to learn how to write R or Python code, but learn how to write code that runs efficiently. Efficient programming is especially important for numerical methods, as they often involve a simulation that reruns thousands of times, or a loop over every particle in a system, which would become extremely slow if not optimized. Both R and Python are interpreted languages, so their execution performance can't compete with a compiled language like C or Fortran on raw speed. Yet, each language has its own tricks to improve code efficiency and close the gap with compiled languages.

## 1B.1 Avoid growing vectors

The following is slow for large `n`, because the script copies the entire vector every time it grows by one element.

### R implementation {.impl}

```{r}
n <- 10
v <- c()
for (i in seq_len(n)) {
  v <- c(v, i^2)
}
v
```

A better approach allocates the final length up front and fills it in place.

```{r}
n <- 10
v <- numeric(n)
for (i in seq_len(n)) {
  v[i] <- i^2
}
v
```

### Python implementation {.impl}

```{python}
n = 10
v = []
for i in range(1, n + 1):
    v.append(i**2)
print(v)
```

```{python}
import numpy as np

n = 10
v = np.zeros(n)
for i in range(1, n + 1):
    v[i - 1] = i**2
print(v)
```

## 1B.2 Vectorize your code

An even better approach replaces the loop with a single vector operation.

### R implementation {.impl}

```{r}
n <- 10
v <- seq_len(n)
v <- v * v
v
```

### Python implementation {.impl}

```{python}
n = 10
v = np.arange(1, n + 1)
v = v**2
print(v)
```

Loops are especially costly when they run thousands of times inside a larger calculation. Consider computing the mean and standard deviation of a series of numbers,

$$\sigma(\mathbf{x})=\sqrt{\frac{\sum_{j=1}^{n}(\mathbf{x_j}-\mathbf{\overline{x}})^2}{n}} ,$$

which is algebraically equal to, and cheaper to compute as,

$$\sigma(\mathbf{x})=\sqrt{\frac{\sum_{j=1}^{n}\mathbf{x_j}^2}{n} - \left(\frac{\sum_{j=1}^{n}\mathbf{x_j}}{n}\right)^2} .$$

### R implementation {.impl}

```{r}
# a running-sum loop, the way this would be written in C or Fortran
num <- 100
my_sum <- 0
my_sum2 <- 0
for (i in seq_len(num)) {
  my_sum <- my_sum + i
  my_sum2 <- my_sum2 + i^2
}
my_mean <- my_sum / num
my_sd <- sqrt(my_sum2 / num - my_mean^2)
c(my_mean, my_sd)
```

```{r}
# vectorized: mean() and mean(v^2) replace both running sums
num <- 100
v <- seq_len(num)
my_mean <- mean(v)
my_sd <- sqrt(mean(v^2) - my_mean^2)
c(my_mean, my_sd)
```

`apply` performs the same calculation for every column of a matrix at once.

```{r}
mat <- matrix(rnorm(16), nrow = 4)
means <- apply(mat, 2, mean)
sd <- sqrt(apply(mat^2, 2, mean) - means^2)
sd
```

R's own `sd()` divides by $n-1$ rather than $n$, the usual statistical convention for an unbiased estimate from a sample:

```{r}
apply(mat, 2, sd)
```

### Python implementation {.impl}

```{python}
import math

# a running-sum loop, the way this would be written in C or Fortran
num = 100
my_sum = 0
my_sum2 = 0
for i in range(1, num + 1):
    my_sum += i
    my_sum2 += i**2
my_mean = my_sum / num
my_sd = math.sqrt(my_sum2 / num - my_mean**2)
print(my_mean, my_sd)
```

```{python}
# vectorized: np.mean(v) and np.mean(v**2) replace both running sums
num = 100
v = np.arange(1, num + 1)
my_mean = np.mean(v)
my_sd = np.sqrt(np.mean(v**2) - my_mean**2)
print(my_mean, my_sd)
```

`apply` on a `pandas` DataFrame performs the same calculation for every column of a matrix at once.

```{python}
import pandas as pd

mat = np.random.normal(size=(4, 4))
df = pd.DataFrame(mat)
means = df.apply(np.mean, axis=0)
sd = np.sqrt(df.apply(lambda col: np.mean(col**2), axis=0) - means**2)
print(sd)
```

## 1B.3 Byte-compiling and JIT compilation

An interpreted language works out what a function does every time that function is called, and the repeated bookkeeping is a large part of why an R or Python loop is slower than the same loop in C. Compiling the function once into a lower-level form removes the repeated work. R and Python differ in whether you have to ask for it.

In R, byte compiler has been on by default since R 3.4 (2017): a function is compiled to byte code the first time it is called, and package code is byte-compiled when the package is installed. The `compiler` package still provides `cmpfun` for compiling a function by hand, and it still works, but on a current R no performance change would be noticed.

CPython makes no such promise. A Python loop is interpreted on every pass, however many times it runs, so compilation has to be requested explicitly. The usual tool is [Numba](https://numba.pydata.org), whose `@njit` decorator compiles a function to machine code the first time it is called with a given argument type. Later calls skip the interpreter entirely and run at a speed comparable to compiled C.

### Python implementation {.impl}

```{python}
from numba import njit
import time

def my_sd_loop(x):
    n = len(x)
    s = 0.0; s2 = 0.0
    for i in range(n):
        s += x[i]
        s2 += x[i]**2
    m = s / n
    return math.sqrt(s2 / n - m**2)

@njit
def my_sd_loop_jit(x):
    n = len(x)
    s = 0.0; s2 = 0.0
    for i in range(n):
        s += x[i]
        s2 += x[i]**2
    m = s / n
    return math.sqrt(s2 / n - m**2)

v = np.random.normal(size=1000)
_ = my_sd_loop_jit(v)   # first call triggers compilation; excluded from the timing below

t0 = time.time()
for _ in range(1000):
    _ = my_sd_loop(v)   # assigned to `_` so reticulate doesn't echo all 1000 return values
t1 = time.time()
for _ in range(1000):
    _ = my_sd_loop_jit(v)
t2 = time.time()
print("interpreted:", t1 - t0, " compiled:", t2 - t1)
```

The first call is left out of the timing because it pays for the compilation itself. After that, the JIT-compiled loop runs many times faster than the interpreted one. That gap is what R now closes on its own, and what Python still asks you to close by hand.

## 1B.4 Calling compiled Fortran code

Both languages can call a function written in a compiled language directly. Here is a Fortran subroutine computing the same standard deviation, in `01-intro/src/sub_cal_sd.f90`:

```fortran
subroutine cal_sd(n,x,sd)
  implicit none
  integer,intent(in) :: n
  double precision, intent(in) :: x(n)
  double precision, intent(out) :: sd
  integer :: i
  double precision :: mean, mean2
  mean = 0.d0
  mean2 = 0.d0
  do i = 1, n
    mean = mean + x(i)
    mean2 = mean2 + x(i)**2
  end do
  mean = mean/n
  mean2 = mean2/n
  sd = sqrt(mean2 - mean**2)
end subroutine cal_sd
```

Only integer, double precision, and logical arguments are allowed across the Fortran-R or Fortran-Python interface. Compile it to a shared library with `gfortran`:

```bash
gfortran -fpic -shared 01-intro/src/sub_cal_sd.f90 -o 01-intro/src/sub_cal_sd.so
```

### R implementation {.impl}

```{r}
dyn.load("01-intro/src/sub_cal_sd.so")

my_sd_R <- function(x) {
  if (!is.numeric(x)) stop("argument x must be numeric")
  my_sd <- 0.0
  results <- .Fortran("cal_sd", as.integer(length(x)), as.double(x), my_sd)
  return(results[[3]])   # the third argument, sd, is the one returned by the subroutine
}
my_sd_R(seq_len(100))
```

### Python implementation {.impl}

The same compiled shared library can be called from Python through `ctypes`. As in [Part 2B](../02-odes/02b-numerical-integration.qmd), gfortran appends a trailing underscore to the subroutine name, and arguments are passed by reference:

```{python}
import ctypes

_lib = ctypes.CDLL("01-intro/src/sub_cal_sd.so")

def my_sd_fortran(x):
    x = np.asarray(x, dtype=np.float64)
    n = ctypes.c_int(len(x))
    sd = ctypes.c_double(0.0)   # written into by the Fortran call, then read back
    _lib.cal_sd_(ctypes.byref(n), x.ctypes.data_as(ctypes.POINTER(ctypes.c_double)),
                 ctypes.byref(sd))
    return sd.value

my_sd_fortran(np.arange(1, 101))
```

## 1B.5 Parallelization

If a calculation is repeated many times independently, splitting the repetitions across CPU cores can help, provided each repetition does enough work to outweigh the cost of coordinating across processes.

### R implementation {.impl}

```{r}
library(parallel)
numCores <- detectCores()
numCores

tests <- seq(1, 150)
my_test <- function(test_id) {
  v <- rnorm(1000000)
  my_sd_R(v)
}

t0 <- Sys.time()
results_serial <- lapply(tests, my_test)
t1 <- Sys.time()
results_parallel <- mclapply(tests, my_test, mc.cores = numCores)
t2 <- Sys.time()
cat("serial:", as.numeric(t1 - t0), "s  parallel:", as.numeric(t2 - t1), "s\n")
```

### Python implementation {.impl}

```{python}
import multiprocessing as mp

# "fork" copies this process, so my_test is already defined in each worker;
# the macOS default "spawn" re-imports the calling script, which this is not
ctx = mp.get_context("fork")
numCores = ctx.cpu_count()
print(numCores)

def my_test(test_id):
    rng = np.random.default_rng(test_id)
    v = rng.normal(size=1_000_000)
    return my_sd_fortran(v)

tests = list(range(150))

t0 = time.time()
results_serial = [my_test(t) for t in tests]
t1 = time.time()
with ctx.Pool(numCores) as pool:
    results_parallel = pool.map(my_test, tests)
t2 = time.time()
print("serial:", t1 - t0, "s  parallel:", t2 - t1, "s")
```

## 1B.6 Using AI to speed up your code

The techniques above are exactly the kind of thing worth asking an AI assistant about, both to get a first pass and to sanity-check your own optimization. Paste the growing-vector loop from 1B.1 into Claude Code and ask something concrete: "this loop rebuilds `v` on every iteration; how would you rewrite it to avoid that, and why is the rewrite faster?" A useful answer does not just hand you the vectorized version. It names the actual mechanism, that `c(v, i^2)` allocates an entirely new, longer vector and copies the old contents into it every single iteration, so the loop's cost grows with the square of `n` rather than linearly.

Verify it the same way you would verify any other AI-assisted explanation: time both versions yourself.

```{r}
n <- 20000
t0 <- Sys.time()
v <- c(); for (i in seq_len(n)) v <- c(v, i^2)
t1 <- Sys.time()
v2 <- numeric(n); for (i in seq_len(n)) v2[i] <- i^2
t2 <- Sys.time()
cat("growing:", as.numeric(t1 - t0), "s  preallocated:", as.numeric(t2 - t1), "s\n")
```

The gap you measure here would allow you to confirm whether the fix actually worked or not. This same paste-explain-verify loop applies to any inefficiency you run into later in this book.