---
title: "1C. Numerical methods"
author: "Mingyang Lu"
old_id: "01C"
---

## 1C.1 What is a numerical method?

Most quantities of interest in this book, an equilibrium concentration, a
trajectory over time, the root of an equation, cannot be written down in closed
form. A numerical method is the computational recipe that gets you an answer
anyway: it starts from a mathematical model (an equation, a system of equations,
a curve to be reconstructed from data), replaces whatever is continuous or exact
in that model with something a computer can represent and manipulate, and
repeats a well-defined update until the result is close enough to trust.

Two properties decide whether a numerical method is any good. **Accuracy**
depends on where error enters: the model itself may only approximate the real
system (modeling error), replacing a continuous quantity with a finite
representation introduces further error (discretization error), and stopping an
iterative process early leaves a residual error (iteration error). **Efficiency**
is the computational cost of getting that accuracy, in both runtime and memory.
Every method in this book, from the Babylonian method below to the ODE
integrators of [Part 2B](../02-odes/02b-numerical-integration.qmd), is a particular
trade-off between these two.

## 1C.2 Machine precision

A computer represents a real number with a fixed number of bits, so most real
numbers are stored as the nearest representable approximation rather than
exactly. R and Python both use double precision (64 bits) for their default
numeric type, giving roughly 16 significant decimal digits.

### R implementation {.impl}

```{r}
0.1 + 0.2 == 0.3
.Machine$double.eps
```

### Python implementation {.impl}

```{python}
0.1 + 0.2 == 0.3
import sys
sys.float_info.epsilon
```

`0.1` and `0.2` cannot be represented exactly in binary floating point, so their
sum lands a tiny bit off from the double closest to `0.3`, and the direct
comparison is `FALSE`/`False` in both languages. `.Machine$double.eps` /
`sys.float_info.epsilon` (about $2.2\times 10^{-16}$) is machine epsilon, the
smallest relative gap between two distinct representable numbers; it sets a
hard floor on how precisely any double-precision calculation can agree with the
exact answer, a limit that reappears in 1C.3 below.

::: {.callout-note title="Practice Q1"}
Write a function `real_equal(x, y, tol)` that reports whether two real numbers
should be considered equal, given that direct `==` comparison is unreliable.
(Hint: compare `abs(x - y)` against a small tolerance rather than testing for
exact equality.) Test it on `0.1 + 0.2` and `0.3`.
:::

## 1C.3 The Babylonian method for square roots

Long before iterative numerical methods had that name, they were already in
use: a Babylonian clay tablet catalogued as YBC 7289 records a base-60
approximation of $\sqrt{2}$ accurate to about six decimal digits
[@fowler1998square]. The method it reflects, now called the Babylonian method,
finds $\sqrt{a}$ for $a > 0$ by repeatedly averaging a guess with $a$ divided by
that guess:

$$x_{n+1} = \frac{1}{2} \left(x_n + \frac{a}{x_{n}}\right)$$

starting from any $x_1 > 0$, until

$$ |x_{n+1}^2 - a| < \epsilon$$

for a chosen tolerance $\epsilon$. (This is in fact a special case of Newton's
method for finding the root of $f(x) = x^2 - a$; [Part 2F](../02-odes/02f-bifurcation-curves.qmd)
implements Newton's method for a general $f$.)

::: {.callout-note title="Recipe 1C.3.1"}
**Objective.** Compute the square root of a positive number by iteration, without a built-in square-root function.

**Model.** The square root of `a > 0` is the positive solution of `x^2 = a`.

**Method.** The Babylonian iteration `x_next = (x + a/x) / 2`, from any positive guess, repeated until `|x_next^2 - a| < epsilon`.

**Test.** `a = 2`; starting guesses 1, 100, 200; tolerance `epsilon = 1e-6`.

**Verification.** Every guess converges to about 1.41421; the result squared is within the tolerance of 2, and a far starting guess just needs a few more iterations.
:::

### R implementation {.impl}

```{r}
# The Babylonian method for finding square root
# a: a positive real number; we will solve the square root of a.
# x1: the initial guess of the solution. x1 should be positive too.
# epsilon: a small positive number for the tolerance.
babylonian <- function(a, x1, epsilon = 1e-6) {
  if (a <= 0 || x1 <= 0) {
    print("a and x1 should be both positive!")
    return(0)
  } else {
    x <- x1
    i <- 0
    while (abs(x * x - a) > epsilon) {
      i <- i + 1
      x <- (x + a / x) / 2
    }
    print(paste0("Number of iterations: ", i))
    return(x)
  }
}

babylonian(2, 1)
babylonian(2, 100)
babylonian(2, 200)
```

### Python implementation {.impl}

```{python}
# The Babylonian method for finding square root
# a: a positive real number; we will solve the square root of a.
# x1: the initial guess of the solution. x1 should be positive too.
# epsilon: a small positive number for the tolerance.
def babylonian(a, x1, epsilon=1e-6):
    if a <= 0 or x1 <= 0:
        print("a and x1 should be both positive!")
        return 0
    else:
        x = x1
        i = 0
        while abs(x * x - a) > epsilon:
            i += 1
            x = (x + a / x) / 2
        print(f"Number of iterations: {i}")
        return x

babylonian(2, 1)
babylonian(2, 100)
babylonian(2, 200)
```

The method converges fast: each iteration roughly doubles the number of
correct digits. Tracking the digits of accuracy against $\sqrt{2}$ directly
confirms this, and also shows machine precision (1C.2) in action:

```{r}
x <- 1
for (i in 1:7) {
  x <- (x + 2 / x) / 2
  err <- abs(x - sqrt(2))
  digits <- if (err > 0) -log10(err) else Inf
  cat("iteration", i, "- digits of accuracy:", round(digits, 1), "\n")
}
```

The digit count roughly doubles for the first several iterations (1.1, 2.6,
5.7, 11.8), then flattens out around 15-16 digits by iteration 5: that plateau
is machine epsilon's limit on double precision. Reaching genuinely dozens of
correct digits, as historical demonstrations of this method sometimes show,
requires arbitrary-precision arithmetic (R's `Rmpfr`, Python's `decimal` or
`mpmath`) rather than R's or Python's default double.

::: {.callout-note title="Practice Q2"}
Modify `babylonian()` to also print the relative error, $|x_{n+1} - x_n| /
x_n$, at every iteration, and confirm it shrinks roughly quadratically (matches
the doubling-digits pattern above) until it hits the machine-epsilon floor.
:::

## 1C.4 Algorithmic complexity (Big-O notation)

Big-O notation describes how a method's cost, whether time or memory, grows as
the problem size $n$ grows, ignoring constant factors and lower-order terms.
It answers "if I double $n$, roughly how much more expensive does this get?"
rather than giving an exact operation count. A few complexity classes recur
throughout this book:

| Complexity | Meaning | Example |
|:-----------|:--------|:--------|
| $O(1)$ | constant, independent of $n$ | checking whether an integer is even or odd |
| $O(\log n)$ | logarithmic | binary search over a sorted list |
| $O(n)$ | linear | finding the maximum of $n$ unsorted numbers |
| $O(n \log n)$ | linearithmic | quicksort (average case) |
| $O(n^2)$ | quadratic | selection sort |
| $O(2^n)$, $O(n!)$ | exponential, factorial | enumerating all subsets, or all permutations, of $n$ items |

[Part 9B](../09-optimization/09b-dynamic-programming.qmd) leans on this
vocabulary directly, contrasting an $O(n^2 2^n)$ dynamic-programming solution
against a brute-force $O(n!)$ one.

::: {.callout-note title="Practice Q3"}
The Babylonian method's iteration count grows with $-\log_{10}\epsilon$ (since
each step roughly doubles the number of correct digits, the number of
iterations needed grows only with the logarithm of the number of digits
requested). What is its time complexity in terms of the tolerance $\epsilon$,
and how does that compare to the complexity classes in the table above?
:::

## 1C.5 Curve interpolation

Representing a curve on a computer usually means storing it as a finite set of
$(x, y)$ points and filling in the gaps between them. The simplest choice,
linear interpolation, connects consecutive points with straight-line segments;
it is cheap but has a visible kink at every data point. A smoother alternative,
cubic spline interpolation [@deboor2001splines], fits a separate cubic
polynomial between each pair of consecutive points and chooses their
coefficients so that value, first derivative, and second derivative all agree
at every shared point, so the whole curve looks smooth rather than piecewise.

For $n+1$ points $(x_0, y_0), \dots, (x_n, y_n)$, the cubic piece between
$(x_i, y_i)$ and $(x_{i+1}, y_{i+1})$ is

$$y_i(x) = a_i x^3 + b_i x^2 + c_i x + d_i , \tag{1}$$

and matching value, first derivative, and second derivative at each interior
point $x_i$ gives

$$ y_i(x_i) = y_{i+1}(x_i), \quad \frac{dy_i}{dx}(x_i) = \frac{dy_{i+1}}{dx}(x_i), \quad \frac{d^2y_i}{dx^2}(x_i) = \frac{d^2y_{i+1}}{dx^2}(x_i) . \tag{2}$$

Two more conditions are still needed to pin down the remaining coefficients;
the natural choice sets the second derivative to zero at the two endpoints,

$$ \frac{d^2y_0}{dx^2}(x_0) = \frac{d^2y_{n-1}}{dx^2}(x_n) = 0 . \tag{3}$$

For a small example ($n = 3$ points, so two cubic pieces), Equations (1)-(3)
reduce to a system of 8 linear equations in the 8 unknown coefficients, solved
directly. Here is a 3-point data set of average temperatures at different
latitudes:

```{r}
x <- c(5, 15, 25)
y <- c(-3.02, -3.07, -3.17)
plot(x, y, xlab = "Latitude", ylab = "Average temperature", type = "b", col = "red", lwd = 0.5)
```

### R implementation {.impl}

```{r}
vec_c <- c(y[1], y[2], y[2], y[3], 0, 0, 0, 0)  # right-hand side of the linear system

a <- matrix(c(x[1]**3, x[1]**2, x[1], 1, 0, 0, 0, 0,
              x[2]**3, x[2]**2, x[2], 1, 0, 0, 0, 0,
              0, 0, 0, 0, x[2]**3, x[2]**2, x[2], 1,
              0, 0, 0, 0, x[3]**3, x[3]**2, x[3], 1,
              3 * x[2]**2, 2 * x[2], 1, 0, -3 * x[2]**2, -2 * x[2], -1, 0,
              6 * x[2], 2, 0, 0, -6 * x[2], -2, 0, 0,
              6 * x[1], 2, 0, 0, 0, 0, 0, 0,
              0, 0, 0, 0, 6 * x[3], 2, 0, 0), nrow = 8, byrow = TRUE)

vec_p <- solve(a, vec_c)  # spline coefficients: a0,b0,c0,d0, a1,b1,c1,d1

lin <- approx(x, y, n = 100)  # linear interpolation, for comparison

plot(x, y, xlab = "Latitude", ylab = "Average temperature", type = "b", col = "red", lwd = 0.5)
lines(lin$x, lin$y, lwd = 1, col = "grey50", lty = 2)
curve(vec_p[1] * x**3 + vec_p[2] * x**2 + vec_p[3] * x + vec_p[4],
      from = x[1], to = x[2], add = TRUE, lwd = 1.5, col = "black")
curve(vec_p[5] * x**3 + vec_p[6] * x**2 + vec_p[7] * x + vec_p[8],
      from = x[2], to = x[3], add = TRUE, lwd = 1.5, col = "blue")
legend("bottomleft", inset = .02,
       legend = c("Data", "Linear", "Cubic spline 1", "Cubic spline 2"),
       col = c("red", "grey50", "black", "blue"), lty = c(1, 2, 1, 1), cex = 0.8)
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

x = np.array([5, 15, 25])
y = np.array([-3.02, -3.07, -3.17])

vec_c = np.array([y[0], y[1], y[1], y[2], 0, 0, 0, 0])  # right-hand side of the linear system

a = np.array([[x[0]**3, x[0]**2, x[0], 1, 0, 0, 0, 0],
              [x[1]**3, x[1]**2, x[1], 1, 0, 0, 0, 0],
              [0, 0, 0, 0, x[1]**3, x[1]**2, x[1], 1],
              [0, 0, 0, 0, x[2]**3, x[2]**2, x[2], 1],
              [3 * x[1]**2, 2 * x[1], 1, 0, -3 * x[1]**2, -2 * x[1], -1, 0],
              [6 * x[1], 2, 0, 0, -6 * x[1], -2, 0, 0],
              [6 * x[0], 2, 0, 0, 0, 0, 0, 0],
              [0, 0, 0, 0, 6 * x[2], 2, 0, 0]], dtype=float)

vec_p = np.linalg.solve(a, vec_c)  # spline coefficients: a0,b0,c0,d0, a1,b1,c1,d1

x_lin = np.linspace(x[0], x[-1], 100)  # linear interpolation, for comparison
y_lin = np.interp(x_lin, x, y)

x_vals_1 = np.linspace(x[0], x[1], 100)
x_vals_2 = np.linspace(x[1], x[2], 100)

_ = plt.plot(x, y, "ro-", label="Data", linewidth=0.5)
_ = plt.plot(x_lin, y_lin, "--", color="grey", linewidth=1, label="Linear")
_ = plt.plot(x_vals_1, vec_p[0] * x_vals_1**3 + vec_p[1] * x_vals_1**2 +
             vec_p[2] * x_vals_1 + vec_p[3], label="Cubic spline 1", linewidth=1.5)
_ = plt.plot(x_vals_2, vec_p[4] * x_vals_2**3 + vec_p[5] * x_vals_2**2 +
             vec_p[6] * x_vals_2 + vec_p[7], label="Cubic spline 2", linewidth=1.5)
plt.xlabel("Latitude")
plt.ylabel("Average temperature")
_ = plt.legend(loc="lower left")
plt.show()
```

The two cubic pieces meet smoothly at the middle point, while the linear
interpolation (dashed) has a visible kink there, the trade-off this section
opened with: linear interpolation is simpler and cheaper, but cubic spline
interpolation better matches how a physical quantity like temperature actually
varies continuously with latitude.
