---
title: "1D. Exercises"
author: "Mingyang Lu"
old_id: "01D"
---

## 1D.1 Radius of gyration

In this exercise, we compute the radius of gyration of a protein structure.

The radius of gyration $R_g$ is defined as

$$R_g=\sqrt{\frac{\sum_{j=1}^{N}m_j \Delta \mathbf{r_j} \cdot \Delta \mathbf{r_j}}{\sum_{j=1}^{N}m_j}} ,$$

where $\mathbf{r_j}$ is the coordinate vector of atom $j$ (its x, y, and z values),
$m_j$ is the atomic mass of atom $j$, and $N$ is the total number of atoms. The center
of mass $\mathbf{\overline{r}}$ is

$$\mathbf{\overline{r}}=\frac{\sum_{j=1}^{N}m_j\mathbf{r_j}}{\sum_{j=1}^{N}m_j} ,$$

and the displacement vector $\Delta \mathbf{r_j}$ is $\mathbf{r_j} - \mathbf{\overline{r}}$.

### R implementation {.impl}

The `bio3d` package downloads and parses structures from the Protein Data Bank
directly.

```{r}
#| warning: false
if (!"bio3d" %in% rownames(installed.packages())) {
  install.packages("bio3d")
}
library(bio3d)

mystructure <- read.pdb("7l3u")   # load online data
coord <- mystructure$atom
class(coord)
head(coord, 3)
```

`coord` is a data frame with one row per atom; the atom type needed for its mass is
in the `elesy` column.

### Python implementation {.impl}

`Bio.PDB` (from `biopython`) plays the same role. This helper downloads a structure
and collects its atomic coordinates into a `pandas` DataFrame:

```{python}
#| warning: false
import tempfile
from Bio import PDB
import pandas as pd

def pdb_to_dataframe(pdb_id):
    pdbl = PDB.PDBList()
    pdb_file_path = pdbl.retrieve_pdb_file(pdb_id, pdir=tempfile.gettempdir(), file_format="pdb")
    pdb_parser = PDB.PDBParser(QUIET=True)
    structure = pdb_parser.get_structure(pdb_id, pdb_file_path)
    atom_data = []
    for model in structure:
        for chain in model:
            for residue in chain:
                for atom in residue:
                    atom_data.append({
                        "atom_name": atom.get_name(),
                        "residue_name": residue.resname,
                        "chain_id": chain.id,
                        "residue_number": residue.id[1],
                        "x": atom.coord[0],
                        "y": atom.coord[1],
                        "z": atom.coord[2],
                        "b": atom.bfactor,
                    })
    return pd.DataFrame(atom_data)

coord = pdb_to_dataframe("7l3u")
coord.shape
coord.head(3)
```

**(a)** Starting from `coord`, compute $R_g$ for structure `7l3u`. Watch the dot
product in the definition of $R_g$: it is $\Delta \mathbf{r_j} \cdot \Delta
\mathbf{r_j}$, not $|\Delta \mathbf{r_j}|$. Atomic masses can be looked up from the
element symbol; `bio3d` provides `atom2mass()` for this (`?atom2mass`), or use any
periodic-table lookup you like in Python.

**(b)** Ask an AI assistant to explain any part of the provided code you find
unfamiliar (the nested loop over `model`/`chain`/`residue`/`atom` in the Python
version, or what `read.pdb()` returns in R), following the same read-then-verify
approach from [Part 1A](01a-getting-started.qmd).

## 1D.2 Gene expression analysis

You are given a gene expression dataset (derived from GEO accession GSE17708)
recording expression levels for 500 genes across 26 samples. Your task is to
compute summary statistics, filter out the least variable genes, and look at
correlations between the remaining ones.

### R implementation {.impl}

```{r}
exp_data <- read.csv("01-intro/data/exp_data.csv", row.names = 1)
class(exp_data)
dim(exp_data)   # 500 genes, 26 conditions
```

### Python implementation {.impl}

```{python}
import pandas as pd

exp_data = pd.read_csv("01-intro/data/exp_data.csv", index_col=0)
type(exp_data)
exp_data.shape   # 500 genes, 26 conditions
```

Each row of `exp_data` is a gene, and each column is a sample.

**(a) Summary statistics.** For each sample, compute the mean, median, and standard
deviation of expression across genes, and make a bar plot showing the mean and
standard deviation per sample.

**(b) Filtering.** Remove genes whose standard deviation (across samples) is below
0.5. Report how many genes remain (you should get fewer than 50).

**(c) Gene expression correlations.** Within the filtered set, compute the
Pearson correlation between every pair of genes' expression profiles, and report
the 10 most correlated pairs.

## 1D.3 Stirling's approximation

The naive way to compute $\ln n!$ is to take the log of the factorial directly:

### R implementation {.impl}

```{r}
ln_fact <- function(n) {
  return(log(factorial(n)))
}
ln_fact(10)
ln_fact(170)
ln_fact(171)
```

### Python implementation {.impl}

```{python}
import math

def ln_fact(n):
    return math.log(math.factorial(n))

ln_fact(10)
ln_fact(170)
ln_fact(1000)
```

R's `factorial(171)` returns `Inf`, so `ln_fact` breaks there: R represents
`factorial(n)` as a double, and doubles overflow past about $1.8 \times 10^{308}$,
which $171!$ already exceeds. Python's `math.factorial` instead returns an
arbitrary-precision integer, so `ln_fact(1000)` still gives a finite answer, but
try `ln_fact(100000)` and time it: multiplying out a thousands-of-digits integer,
one factor at a time, gets slow well before it overflows anything.

Both languages hit a wall for large $n$, just for different reasons and at
different points (think about why, for each). A more scalable approach avoids
forming $n!$ at all and instead sums logs directly:

$$f(n) = \ln n!=\sum_{j=1}^{n}\ln j .$$

**(a) Benchmarking.** Implement $f(n)$ for large $n$ using each of the following,
in both R and Python:

  (1) a loop that accumulates the sum term by term;
  (2) a vectorized sum (R: `sum(log(1:n))`; Python: `numpy`);
  (3) an apply-family construct (R: `sapply`/`vapply`; Python: `map`, or a list
      comprehension);
  (4) a compiled C or Fortran routine, called the way [Part 1B](01b-efficient-programming.qmd)
      calls compiled code.

  Benchmark each method's runtime (R: `microbenchmark`; Python: `timeit`) for a
  moderately large $n$ (say $10^5$ or $10^6$), consistent with the previous
  chapter.

**(b) Deviation of Stirling's approximation.** Stirling's approximation
[@stirling1730methodus] states

$$\ln n! \approx n\ln n - n .$$

Using any method from (a) to compute the left-hand side, plot the relative
deviation

$$1-\frac{n\ln n - n}{\ln n!}$$

as a function of $n$, and describe how it behaves as $n$ grows.

## 1D.4 Prime numbers

A prime number is a positive integer greater than 1 with no divisors other than 1
and itself.

**(a)** Write a function that takes an integer and returns whether it is prime,
by testing divisibility by every integer from 2 up to its square root.

**(b)** Apply your function to every integer up to 100 and print the primes you
find.

## 1D.5 Explaining an unfamiliar one-liner

R and Python both let you fold a sequence down to a single value without writing
an explicit loop: R's `Reduce()`, and Python's `functools.reduce()` paired with a
generator expression. Here is one such line in each language, computing something
about a list of numbers without saying what:

```r
x <- c(3, 1, 4, 1, 5, 9, 2, 6)
Reduce(function(acc, v) if (v > acc) v else acc, x)
```

```python
from functools import reduce
x = [3, 1, 4, 1, 5, 9, 2, 6]
reduce(lambda acc, v: v if v > acc else acc, x)
```

**(a)** Before running either snippet, ask an AI assistant to explain what it
computes and how the accumulator argument (`acc`) changes on each step.

**(b)** Run the code and confirm the AI's explanation against the actual output.

**(c)** Hand-trace the R version for a shorter list, say `c(3, 1, 4)`, writing out
the value of `acc` after each step, and check it matches what the AI described.

## 1D.6 Finding an injected bug

The function below is meant to return the *second-largest* distinct value in a
numeric vector/list (for example, the second-largest of `c(5, 1, 5, 3)` is 3, since
the two 5's count as one distinct value). It has a bug.

```r
second_largest <- function(x) {
  u <- sort(unique(x), decreasing = TRUE)
  return(u[1])
}
```

```python
def second_largest(x):
    u = sorted(set(x), reverse=True)
    return u[0]
```

**(a)** Without running the code, predict what `second_largest(c(5, 1, 5, 3))` (or
the Python equivalent) returns, and check whether it matches the description
above.

**(b)** Paste the (buggy) function into an AI assistant and ask it to find and fix
the bug. Compare the fix it proposes to your own diagnosis in (a).

**(c)** In your own words, explain why the fix is correct, and test it on an edge
case the original description doesn't cover: a vector with fewer than two distinct
values.

## 1D.7 Porting and critiquing a translation

Pick one function you wrote for [Part 1B](01b-efficient-programming.qmd) (for
example the growing-vector fix from 1B.1, or one of the `ln_fact` variants from
1D.3) in whichever of R or Python you find more natural. Ask an AI assistant to
translate it into the other language.

**(a)** Run the translation and confirm it produces the same output as the
original on a few test inputs.

**(b)** Critique the translation: does it use the idioms a fluent programmer in
the target language would actually use, or does it read like a line-by-line
transliteration (for example, an explicit index-based loop in Python where a
vectorized `numpy` expression would be natural, or vice versa)? Note any subtle
difference in edge-case behavior between the two versions.
