---
title: "7A. Modeling diffusion"
author: "Mingyang Lu"
old_id: "07A"
---

In [Part 6](../06-stochastic/06b-brownian-motion.qmd) we modeled Brownian motion
with an SDE. For a purely diffusive one-variable system,

$$ \frac{dX}{dt} = \sqrt{2D}\,\eta(t) , \tag{1} $$

with $\eta(t)$ the Gaussian white noise, we followed a single random trajectory
$X(t)$ and recovered its statistics ($\langle X \rangle$, $\langle X^2 \rangle$)
from many runs. Here we take the complementary view and follow the probability
distribution $P(X, t)$ directly. It obeys the *diffusion equation*
[@gardiner2009stochastic],

$$ \frac{\partial P(X, t)}{\partial t} = D \frac{\partial^2 P(X, t)}{\partial X^2} , \tag{2} $$

a partial differential equation (PDE): $P$ depends on two variables, and $\partial
P / \partial t$ is a *partial derivative*, taken while holding $X$ fixed,

$$ \frac{\partial P(X, t)}{\partial t} = \lim_{\Delta t \to 0} \frac{P(X, t + \Delta t) - P(X, t)}{\Delta t} . \tag{3} $$

## 7A.1 The finite difference method

The most direct way to solve a PDE like Equation (2) is the *finite difference
method* [@leveque2007finite], which replaces each derivative by a difference over a
grid. Write $P_i^n$ for the value at grid point $X_i$ and time
$t_n = t_0 + n\,\Delta t$. A Taylor expansion in time gives a forward difference for
the time derivative, accurate to $O(\Delta t)$,

$$ \frac{\partial P}{\partial t} \approx \frac{P_i^{n+1} - P_i^n}{\Delta t} . \tag{4} $$

Adding the Taylor expansions of $P_{i+1}^n$ and $P_{i-1}^n$ cancels the odd-order
terms and gives a centered difference for the second space derivative, accurate to
$O(\Delta X^2)$,

$$ \frac{\partial^2 P}{\partial X^2} \approx \frac{P_{i+1}^n + P_{i-1}^n - 2 P_i^n}{\Delta X^2} . \tag{5} $$

Substituting both into Equation (2) gives an explicit update, the diffusion analog
of the Euler method for ODEs,

$$ P_i^{n+1} = P_i^n + D\, \frac{\Delta t}{\Delta X^2} \left( P_{i+1}^n + P_{i-1}^n - 2 P_i^n \right) . \tag{6} $$

The prefactor $D\,\Delta t / \Delta X^2$ must stay small for stability, so the time
step cannot be large relative to $\Delta X^2$. We also need boundary and initial
conditions. Here we impose a *Dirichlet* condition, $P = 0$ at the two ends
$X = \pm l/2$, by padding the shifted arrays with zeros.

::: {.callout-note title="Recipe 7A.1.1"}
**Objective.** Build the explicit finite-difference scheme for the 1D diffusion equation.

**Model.** The 1D diffusion equation `dP/dt = D*d2P/dX2` for a distribution `P(X, t)`, with no reaction term.

**Method.** The explicit forward-time centered-space finite-difference method, `P_i_next = P_i + D*(dt/dX^2)*(P_{i+1} + P_{i-1} - 2*P_i)`, with Dirichlet boundaries (P = 0 at both ends).

**Test.** The generic solver `pde_fd_diffusion(ngrid, X_all, dX, dt, D, ...)`; the stability factor `D*dt/dX^2` must stay small.

**Show.** The distribution P advanced several finite-difference steps (exercised in 7A.2).

**Verification.** The scheme is the diffusion analog of Euler integration for ODEs, advancing P one time step from its neighbors.
:::

### R implementation {.impl}

```{r}
# finite-difference step for the diffusion equation, Dirichlet (P = 0) boundaries
pde_fd_diffusion <- function(ngrid, X_all, dX, dt, D, t0, t.total, p0) {
  t_all <- seq(t0, t.total, by = dt); nt_all <- length(t_all)
  factor <- D / dX^2 * dt
  p <- p0
  for (i in seq_len(nt_all - 1)) {
    p_plus  <- c(p[-1], 0)              # P(X + dX), 0 at the right boundary
    p_minus <- c(0, p[1:(ngrid-1)])     # P(X - dX), 0 at the left boundary
    p <- p + factor * (p_plus + p_minus - 2 * p)
  }
  return(p)
}
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

# finite-difference step for the diffusion equation, Dirichlet (P = 0) boundaries
def pde_fd_diffusion(ngrid, X_all, dX, dt, D, t0, t_total, p0):
    t_all = np.arange(t0, t_total + dt, dt); nt_all = len(t_all)
    factor = D / dX**2 * dt
    p = p0.copy()
    for i in range(nt_all - 1):
        p_plus  = np.concatenate((p[1:], [0]))     # P(X + dX), 0 at the right boundary
        p_minus = np.concatenate(([0], p[:-1]))    # P(X - dX), 0 at the left boundary
        p = p + factor * (p_plus + p_minus - 2 * p)
    return p
```

## 7A.2 Diffusion from the center

We start all the probability at the origin, $P(X = 0, t = 0) = 1$, and watch it
spread. On a box of size $l = 100$ with $\Delta X = 1$, $\Delta t = 0.01$, and
$D = 1$, we plot the distribution after each of five successive runs.

::: {.callout-note title="Recipe 7A.2.1"}
**Objective.** Diffuse an initially concentrated distribution and compare its spread to theory.

**Model.** The 1D diffusion equation `dP/dt = D*d2P/dX2`.

**Method.** The explicit finite-difference diffusion integrator (from 7A.1), tracking the mean, variance, and total probability.

**Test.** Domain length 100, `dX = 1`, `dt = 0.01`, `D = 1`; all probability initially at X = 0; compared with the theoretical variance `sigma^2 = 2*D*t`.

**Show.** A plot of P(X) spreading over time, and its variance versus t against `2*D*t`.

**Verification.** The point distribution spreads into a widening Gaussian whose variance grows as `2*D*t` at early times, while the total probability slowly leaks as the Dirichlet ends absorb it.
:::

### R implementation {.impl}

```{r}
l <- 100; dX <- 1; dt <- 0.01; D <- 1
ngrid <- as.integer(l / dX) + 1
X_all <- (1:ngrid - (ngrid + 1) / 2) * dX     # grid, centered on 0

p0 <- numeric(ngrid)
p0[which.min(abs(X_all))] <- 1                # all probability at X = 0

plot(X_all, p0, type = "l", col = 1, xlab = "X", ylab = "P", xlim = c(-50, 50), ylim = c(0, 0.05))
p_now <- p0
for (i in 1:5) {
  p_now <- pde_fd_diffusion(ngrid, X_all, dX, dt, D, t0 = 0, t.total = 50, p0 = p_now)
  lines(X_all, p_now, col = i + 1)
}
legend("topright", inset = 0.02, title = "Iterations", legend = 0:5, col = 1:6, lty = 1, cex = 0.8)
```

Each run broadens the distribution into a wider Gaussian, exactly the spreading we
saw for the random walk. To quantify it we track the mean and variance of $X$,

$$
\begin{aligned}
\langle X \rangle &= \frac{\sum_i P(X_i, t)\, X_i}{\sum_i P(X_i, t)} , \\[1mm]
\langle X^2 \rangle &= \frac{\sum_i P(X_i, t)\, X_i^2}{\sum_i P(X_i, t)} ,
\end{aligned} \tag{7}
$$

with $\sigma^2(X) = \langle X^2 \rangle - \langle X \rangle^2$.

```{r}
#| fig-width: 9
#| fig-height: 3.2
#| out-width: "92%"
meanX <- function(p, X) as.numeric(p %*% X / sum(p))
sdX   <- function(p, X) sqrt(as.numeric(p %*% X^2 / sum(p)) - meanX(p, X)^2)

num_sim <- 100
meanX_all <- sdX_all <- p_tot_all <- numeric(num_sim)
p_now <- p0
for (i in 1:num_sim) {
  p_now <- pde_fd_diffusion(ngrid, X_all, dX, dt, D, t0 = 0, t.total = 10, p0 = p_now)
  meanX_all[i] <- meanX(p_now, X_all); sdX_all[i] <- sdX(p_now, X_all); p_tot_all[i] <- sum(p_now)
}
t_all <- 10 * (1:num_sim)

par(mfrow = c(1, 3), mar = c(4, 4, 1, 1))
plot(t_all, p_tot_all, xlab = "t", ylab = "Total P", ylim = c(0, 1))
plot(t_all, meanX_all, xlab = "t", ylab = "<X>", ylim = c(-5, 5))
plot(t_all, sdX_all^2, xlab = "t", ylab = expression(sigma(X)^2), ylim = c(0, 500))
curve(2 * D * x, add = TRUE)                  # theory: sigma^2 = 2 D t
```

### Python implementation {.impl}

```{python}
l = 100; dX = 1; dt = 0.01; D = 1
ngrid = int(l / dX) + 1
X_all = (np.arange(1, ngrid + 1) - (ngrid + 1) / 2) * dX     # grid, centered on 0

p0 = np.zeros(ngrid)
p0[np.argmin(np.abs(X_all))] = 1                             # all probability at X = 0

plt.plot(X_all, p0, color="black", label="0")
p_now = p0.copy()
for i in range(5):
    p_now = pde_fd_diffusion(ngrid, X_all, dX, dt, D, t0=0, t_total=50, p0=p_now)
    plt.plot(X_all, p_now, label=f"{i + 1}")
plt.xlabel("X"); plt.ylabel("P"); plt.xlim(-50, 50); plt.ylim(0, 0.05)
plt.legend(title="Iterations", loc="upper right"); plt.show()
```

```{python}
#| fig-width: 9
#| fig-height: 3.2
#| out-width: "92%"
def meanX(p, X): return np.dot(p, X) / np.sum(p)
def sdX(p, X):   return np.sqrt(np.dot(p, X**2) / np.sum(p) - meanX(p, X)**2)

num_sim = 100
meanX_all = np.zeros(num_sim); sdX_all = np.zeros(num_sim); p_tot_all = np.zeros(num_sim)
p_now = p0.copy()
for i in range(num_sim):
    p_now = pde_fd_diffusion(ngrid, X_all, dX, dt, D, t0=0, t_total=10, p0=p_now)
    meanX_all[i] = meanX(p_now, X_all); sdX_all[i] = sdX(p_now, X_all); p_tot_all[i] = np.sum(p_now)
t_all = 10 * np.arange(1, num_sim + 1)

fig, axs = plt.subplots(1, 3, figsize=(9, 3.2))
axs[0].plot(t_all, p_tot_all, "o", ms=3); axs[0].set(xlabel="t", ylabel="Total P", ylim=(0, 1))
axs[1].plot(t_all, meanX_all, "o", ms=3); axs[1].set(xlabel="t", ylabel="<X>", ylim=(-5, 5))
axs[2].plot(t_all, sdX_all**2, "o", ms=3); axs[2].set(xlabel="t", ylabel=r"$\sigma(X)^2$", ylim=(0, 500))
axs[2].plot(t_all, 2 * D * t_all, color="black")            # theory: sigma^2 = 2 D t
plt.tight_layout(); plt.show()
```

The variance grows as $\sigma^2 = 2Dt$ at early times, matching the random-walk
result. The total probability, which should be conserved, slowly leaks away
because the Dirichlet boundaries absorb whatever reaches the edges; once $P$ is
appreciable near the ends, the variance also grows more slowly than $2Dt$.

## 7A.3 Diffusion from the left

A different initial condition places all the probability on the left half,
$P(X < 0, t = 0) = 1$ (normalized). The distribution then spreads rightward,
filling the box.

### R implementation {.impl}

```{r}
p0 <- numeric(ngrid)
p0[X_all < 0] <- 1                            # all probability on the left half
p0 <- p0 / sum(p0)

plot(X_all, p0, type = "l", col = 1, xlab = "X", ylab = "P", xlim = c(-50, 50), ylim = c(0, 0.02))
p_now <- p0
for (i in 1:5) {
  p_now <- pde_fd_diffusion(ngrid, X_all, dX, dt, D, t0 = 0, t.total = 50, p0 = p_now)
  lines(X_all, p_now, col = i + 1)
}
legend("topright", inset = 0.02, title = "Iterations", legend = 0:5, col = 1:6, lty = 1, cex = 0.8)
```

### Python implementation {.impl}

```{python}
p0 = np.zeros(ngrid)
p0[X_all < 0] = 1                             # all probability on the left half
p0 = p0 / np.sum(p0)

plt.plot(X_all, p0, color="black", label="0")
p_now = p0.copy()
for i in range(5):
    p_now = pde_fd_diffusion(ngrid, X_all, dX, dt, D, t0=0, t_total=50, p0=p_now)
    plt.plot(X_all, p_now, label=f"{i + 1}")
plt.xlabel("X"); plt.ylabel("P"); plt.xlim(-50, 50); plt.ylim(0, 0.02)
plt.legend(title="Iterations", loc="upper right"); plt.show()
```

The step spreads and flattens, diffusing from the filled left side toward the
empty right, until it would eventually even out (were it not for the absorbing
ends).

## 7A.4 Other boundary conditions

The Dirichlet condition ($P = 0$ at the ends) is only one choice, and different
boundaries give different dynamics. A **Neumann** condition instead fixes the flux
by setting $\partial P / \partial X = 0$ at each end, which prevents probability
from leaking out; numerically, one copies each boundary point from its inward
neighbor. Implementing it and comparing the dynamics is left as an
[exercise](07f-exercises.qmd). A
third choice, the **periodic** boundary, wraps the grid so that $P(X_0) = P(X_n)$
and $P(X_{n+1}) = P(X_1)$; we use it for the pattern-forming systems of
[Part 7C](07c-turing-instability.qmd).
