---
title: "7E. Two-dimensional reaction-diffusion"
author: "Mingyang Lu"
old_id: "07E"
---

[Part 7C](07c-turing-instability.qmd) followed two-component Turing systems on a
line. The same diffusion-driven instability acts in two spatial dimensions, where it
produces the spots, stripes, and labyrinths seen on animal coats and many other
biological surfaces [@turing1952chemical]. Here we extend the Gierer-Meinhardt models
of Part 7C [@gierer1972theory] from a one-dimensional grid to a two-dimensional
lattice, keeping the same reaction terms but now letting each species diffuse in both
$X$ and $Y$.

For a two-component system with fields $u(X, Y, t)$ and $v(X, Y, t)$, the
reaction-diffusion equations are

$$
\begin{aligned}
\frac{\partial u}{\partial t} &= f(u, v) + D_u \left( \frac{\partial^2 u}{\partial X^2} + \frac{\partial^2 u}{\partial Y^2} \right) , \\
\frac{\partial v}{\partial t} &= g(u, v) + D_v \left( \frac{\partial^2 v}{\partial X^2} + \frac{\partial^2 v}{\partial Y^2} \right) ,
\end{aligned} \tag{1}
$$

with reaction terms $f$, $g$ and diffusion constants $D_u$, $D_v$. We discretize the
two-dimensional Laplacian with the five-point stencil of
[Part 7D](07d-pattern-formation-dictyostelium.qmd), giving the explicit update

$$
\begin{aligned}
u_{i,j}^{n+1} = u_{i,j}^n &+ f\,\Delta t \\
&+ D_u \frac{\Delta t}{\Delta X^2}\left( u_{i+1,j}^n + u_{i-1,j}^n + u_{i,j+1}^n + u_{i,j-1}^n - 4 u_{i,j}^n \right) ,
\end{aligned} \tag{2}
$$

and the same for $v$ with $D_v$. As in Part 7C we use a *periodic* boundary, wrapping
the lattice in both directions so the pattern tiles the plane without an edge.

## 7E.1 A two-dimensional integrator

The one-dimensional integrator of Part 7C called the reaction term at each grid point
in turn. In two dimensions that would be slow, so we exploit the fact that the
reaction is *local* (the value of $f$ and $g$ at a point depends only on $u$ and $v$
at that point) and evaluate it on the whole grid at once. The diffusion term is the
periodic five-point Laplacian, computed by shifting the array in each direction and
wrapping the edges.

One caveat specific to two dimensions: each point now has four neighbors instead of
two, so the explicit scheme is stable only when $D\,\Delta t / \Delta X^2 < 1/4$. We
therefore use a smaller time step, $\Delta t = 0.005$, than the 1D runs of Part 7C.

### R implementation {.impl}

```{r}
# periodic 2D Laplacian, five-point stencil (wraps in both directions)
lap2d <- function(M, n)
  rbind(M[-1, ], M[1, ]) + rbind(M[n, ], M[-n, ]) +
  cbind(M[, -1], M[, 1]) + cbind(M[, n], M[, -n]) - 4 * M

# one explicit finite-difference solve of a 2-component reaction-diffusion system in 2D
pde_fd_rd_2d <- function(reaction, D, u, v, n, dX, dt, t.total, ...) {
  nt <- as.integer(t.total / dt)
  fu <- D[1] / dX^2 * dt; fv <- D[2] / dX^2 * dt
  for (step in seq_len(nt)) {
    r <- reaction(u, v, ...)                        # reaction on the whole grid at once
    u_new <- u + r[[1]] * dt + fu * lap2d(u, n)
    v_new <- v + r[[2]] * dt + fv * lap2d(v, n)
    u <- u_new; v <- v_new
  }
  list(u = u, v = v)
}
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

# periodic 2D Laplacian, five-point stencil (wraps in both directions)
def lap2d(M):
    return np.roll(M, 1, 0) + np.roll(M, -1, 0) + np.roll(M, 1, 1) + np.roll(M, -1, 1) - 4 * M

# one explicit finite-difference solve of a 2-component reaction-diffusion system in 2D
def pde_fd_rd_2d(reaction, D, u, v, n, dX, dt, t_total, **kw):
    nt = int(t_total / dt); fu, fv = D[0] / dX**2 * dt, D[1] / dX**2 * dt
    for _ in range(nt):
        ru, rv = reaction(u, v, **kw)               # reaction on the whole grid at once
        u, v = u + ru * dt + fu * lap2d(u), v + rv * dt + fv * lap2d(v)
    return u, v
```

To watch a pattern develop, we run the integrator in successive blocks of time,
starting from a nearly uniform state $u = v \approx 1$ with a small random
perturbation, and display the $u$ field after each block. The helper below wraps this
loop and draws the six snapshots on a shared grid.

### R implementation {.impl}

```{r}
run_rd_2d <- function(reaction, D, mu, n, dX, dt, block, nplot, seed = 10) {
  set.seed(seed)
  u <- matrix(1 + runif(n * n, -0.1, 0.1), n, n)
  v <- matrix(1 + runif(n * n, -0.1, 0.1), n, n)
  vir <- hcl.colors(20, "viridis")
  par(mfrow = c(2, 3), pty = "s", mar = c(1, 1, 2, 1))
  image(1:n, 1:n, u, col = vir, axes = FALSE, main = "t = 0"); box()
  for (i in 1:(nplot - 1)) {
    r <- pde_fd_rd_2d(reaction, D, u, v, n, dX, dt, block, mu = mu)
    u <- r$u; v <- r$v
    image(1:n, 1:n, u, col = vir, axes = FALSE, main = paste("t =", i * block)); box()
  }
}
```

### Python implementation {.impl}

```{python}
def run_rd_2d(reaction, D, mu, n, dX, dt, block, nplot, seed=10):
    rng = np.random.default_rng(seed)
    u = 1 + rng.uniform(-0.1, 0.1, (n, n)); v = 1 + rng.uniform(-0.1, 0.1, (n, n))
    fig, axs = plt.subplots(2, 3, figsize=(9, 6.2))
    for k, ax in enumerate(axs.flat):
        if k > 0:
            u, v = pde_fd_rd_2d(reaction, D, u, v, n, dX, dt, block, mu=mu)
        ax.imshow(u.T, origin="lower", cmap="viridis", aspect="equal")
        ax.set_title(f"t = {k * block}"); ax.axis("off")
    plt.tight_layout(); plt.show()
```

## 7E.2 A substrate-depletion model

The substrate-depletion model of Part 7C keeps $D_u = d$, $D_v = 1$, and

$$
\begin{aligned}
f(u, v) &= u^2 v - u , \\
g(u, v) &= \mu (1 - u^2 v) ,
\end{aligned} \tag{3}
$$

with $v$ a fast-diffusing substrate that activates $u$ and is depleted by it. On the
line this gave a periodic array of peaks. On the plane, with $d = 0.1$ and
$\mu = 1.5$, the same instability grows into a labyrinth of stripes.

### R implementation {.impl}

```{r}
#| fig-width: 9
#| fig-height: 6.2
#| out-width: "95%"
#| rmar: false
asdm <- function(u, v, mu) { u2v <- u * u * v; list(u2v - u, mu * (1 - u2v)) }

n <- 101; dX <- 0.2; dt <- 0.005; D <- c(0.1, 1); mu <- 1.5
run_rd_2d(asdm, D, mu, n, dX, dt, block = 20, nplot = 6)
```

### Python implementation {.impl}

```{python}
#| fig-width: 9
#| fig-height: 6.2
#| out-width: "95%"
def asdm(u, v, mu):
    u2v = u * u * v
    return u2v - u, mu * (1 - u2v)

n = 101; dX = 0.2; dt = 0.005; D = np.array([0.1, 1.0]); mu = 1.5
run_rd_2d(asdm, D, mu, n, dX, dt, block=20, nplot=6)
```

The nearly uniform initial field first develops fine ripples, which coarsen and link
up into winding stripes. The stripe width is set by the same fastest-growing mode
that fixed the peak spacing in one dimension. Each color scale is set per snapshot,
so the early panels show structure that is still very small in amplitude.

## 7E.3 An activator-inhibitor model

Switching the reaction terms to the activator-inhibitor model, again with
$D_u = d$, $D_v = 1$ and

$$
\begin{aligned}
f(u, v) &= u^2 / v - u , \\
g(u, v) &= \mu (u^2 - v) ,
\end{aligned} \tag{4}
$$

gives a different pattern from the same initial conditions and the same parameters.

### R implementation {.impl}

```{r}
#| fig-width: 9
#| fig-height: 6.2
#| out-width: "95%"
#| rmar: false
aim <- function(u, v, mu) list(u^2 / v - u, mu * u^2 - v)

run_rd_2d(aim, D, mu, n, dX, dt, block = 20, nplot = 6)
```

### Python implementation {.impl}

```{python}
#| fig-width: 9
#| fig-height: 6.2
#| out-width: "95%"
def aim(u, v, mu):
    return u**2 / v - u, mu * u**2 - v

run_rd_2d(aim, D, mu, n, dX, dt, block=20, nplot=6)
```

Instead of stripes, the activator-inhibitor model settles into a regular array of
spots, with high-$u$ peaks on a low-$u$ background. The two Gierer-Meinhardt variants
thus select two of the classic two-dimensional Turing morphologies, stripes and
spots, from the same random start. Which one appears is a property of the reaction
kinetics, not of the initial condition: as in the earlier chapters, the R and Python
runs use independent random perturbations and so produce different specific
realizations of the same pattern type.
