---
title: "9A. MCMC optimization methods"
author: "Mingyang Lu"
old_id: "09A"
---

The Metropolis-Hastings algorithm of [Part 8B](../08-monte-carlo/08b-metropolis-algorithm.qmd)
samples a probability distribution efficiently, but it struggles as a *global
optimizer*. Treating a cost function $f$ as an energy and sampling the Boltzmann
distribution $P \propto e^{-f/T}$, the acceptance probability

$$ a = \min\left(1, e^{-(f_{x'} - f_x)/T}\right) \tag{1} $$

becomes tiny at low temperature $T$. When $f$ has several local minima, a low-$T$ chain
gets trapped in one basin and rarely crosses the barriers to the others, while a high-$T$
chain crosses barriers freely but never settles into any minimum. This chapter develops
MCMC methods that overcome this trade-off: simulated annealing, parallel tempering, and
simulated tempering.

## 9A.1 The Himmelblau test function

We use Himmelblau's function [@himmelblau1972applied], a standard optimization
benchmark,

$$ f(x, y) = (x^2 + y - 11)^2 + (x + y^2 - 7)^2 . \tag{2} $$

It has four well-separated local minima, all with $f = 0$, which makes it a good stage
for methods that must find or sample multiple basins.

### R implementation {.impl}

```{r}
#| fig-width: 5
#| fig-height: 5
#| out-width: "60%"
library(ggplot2)
f_Himmelblau <- function(x) (x[1]^2 + x[2] - 11)^2 + (x[1] + x[2]^2 - 7)^2

# contour of the landscape, reused as a background throughout the chapter
grid_df <- expand.grid(x = seq(-6, 6, 0.1), y = seq(-6, 6, 0.1))
grid_df$z <- apply(grid_df, 1, f_Himmelblau)
base_contour <- function()
  list(geom_contour(data = grid_df, aes(x, y, z = z), breaks = 10^seq(-2, 2.5, 0.5),
                    colour = "grey60"),
       scale_colour_gradientn(colours = c("blue", "yellow", "red"), name = "step"),
       coord_fixed(), theme_bw())

ggplot() + base_contour()
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt
from matplotlib.colors import LinearSegmentedColormap

def f_Himmelblau(x):
    return (x[0]**2 + x[1] - 11)**2 + (x[0] + x[1]**2 - 7)**2

# contour of the landscape, reused as a background throughout the chapter
gx = np.arange(-6, 6, 0.1)
XX, YY = np.meshgrid(gx, gx); ZZ = f_Himmelblau([XX, YY])
BYR = LinearSegmentedColormap.from_list("byr", ["blue", "yellow", "red"])

def plot_traj(ax, x, title=""):
    ax.contour(XX, YY, ZZ, levels=10**np.arange(-2, 3, 0.5), colors="grey")
    ax.scatter(x[:, 0], x[:, 1], c=np.arange(len(x)), cmap=BYR, s=1)
    ax.set(title=title, xlim=(-6, 6), ylim=(-6, 6), aspect="equal")

fig, ax = plt.subplots(figsize=(5, 5))
ax.contour(XX, YY, ZZ, levels=10**np.arange(-2, 3, 0.5), colors="black")
ax.set_aspect("equal"); plt.show()
```

The four basins sit roughly at the corners of the plotted region.

## 9A.2 Metropolis-Hastings sampling

We first apply plain Metropolis-Hastings at several fixed temperatures. Each step
proposes a small uniform displacement and accepts it with probability given by
Equation (1). The `mh` function also returns its acceptance rate, which we reuse in the
parallel-tempering method below.

::: {.callout-note title="Recipe 9A.2.1"}
**Objective.** Explore a multi-minimum landscape with Metropolis-Hastings at fixed temperature.

**Model.** Himmelblau's function `f(x, y) = (x^2 + y - 11)^2 + (x + y^2 - 7)^2`, which has four equal minima at f = 0.

**Method.** The Metropolis-Hastings algorithm at fixed temperature, with a uniform proposal displacement and acceptance `a = min(1, exp(-de/T))`.

**Test.** Start at (0, 0), `1e4` steps, step size 0.5, temperatures `T = 1, 10, 30, 50`; seed 1 per temperature.

**Verification.** No single temperature both explores widely and settles: T = 1 traps in one basin, while T = 50 roams but never pins a minimum.
:::

### R implementation {.impl}

```{r}
# Metropolis-Hastings sampler for a cost function func
mh <- function(n, func, x0, nstep, dx_max, temp, ifprint = FALSE) {
  x <- matrix(0, nstep + 1, n); e <- numeric(nstep + 1)
  x[1, ] <- x0; e[1] <- func(x0); num_accept <- 0
  for (i in 1:nstep) {
    xnew <- x[i, ] + runif(n, -dx_max, dx_max)
    de <- func(xnew) - e[i]
    if (de < 0 || runif(1) < exp(-de / temp)) {
      x[i + 1, ] <- xnew; e[i + 1] <- e[i] + de; num_accept <- num_accept + 1
    } else { x[i + 1, ] <- x[i, ]; e[i + 1] <- e[i] }
  }
  rate <- num_accept / nstep
  if (ifprint) cat("acceptance rate", rate, "\n")
  list(x = x, e = e, rate = rate)
}
```

```{r}
#| fig-width: 8
#| fig-height: 8
#| out-width: "85%"
# sample at four temperatures, each starting from the origin
nstep <- 10^4
traj_all <- do.call(rbind, lapply(c(1, 10, 30, 50), function(temp) {
  set.seed(1)
  r <- mh(2, f_Himmelblau, c(0, 0), nstep, dx_max = 0.5, temp = temp)
  data.frame(x = r$x[, 1], y = r$x[, 2], step = 0:nstep, temp = paste0("T = ", temp))
}))
traj_all$temp <- factor(traj_all$temp, levels = paste0("T = ", c(1, 10, 30, 50)))
ggplot() + base_contour() +
  geom_path(data = traj_all, aes(x, y, colour = step), linewidth = 0.2) +
  facet_wrap(~temp)
```

### Python implementation {.impl}

```{python}
# Metropolis-Hastings sampler for a cost function func
def mh(n, func, x0, nstep, dx_max, temp, ifprint=False):
    x = np.zeros((nstep + 1, n)); e = np.zeros(nstep + 1)
    x[0] = x0; e[0] = func(x0); num_accept = 0
    for i in range(nstep):
        xnew = x[i] + np.random.uniform(-dx_max, dx_max, n)
        de = func(xnew) - e[i]
        if de < 0 or np.random.rand() < np.exp(-de / temp):
            x[i + 1] = xnew; e[i + 1] = e[i] + de; num_accept += 1
        else:
            x[i + 1] = x[i]; e[i + 1] = e[i]
    rate = num_accept / nstep
    if ifprint:
        print("acceptance rate", rate)
    return {"x": x, "e": e, "rate": rate}
```

```{python}
#| fig-width: 8
#| fig-height: 8
#| out-width: "85%"
# sample at four temperatures, each starting from the origin
nstep = 10**4
fig, axs = plt.subplots(2, 2, figsize=(8, 8))
for ax, temp in zip(axs.flat, (1, 10, 30, 50)):
    np.random.seed(1)
    r = mh(2, f_Himmelblau, np.zeros(2), nstep, dx_max=0.5, temp=temp)
    plot_traj(ax, r["x"], f"T = {temp}")
plt.tight_layout(); plt.show()
```

At $T = 1$ the chain is trapped in the single basin it first reaches. As $T$ rises it
explores more of the landscape, sampling several basins by $T = 50$, but then it no
longer settles cleanly into any minimum. No single fixed temperature both explores
widely and pins down a minimum.

## 9A.3 Simulated annealing

Simulated annealing (SA) resolves the trade-off by *lowering* the temperature during a
single run [@kirkpatrick1983optimization]. It starts hot enough to roam across all
basins and cools toward zero to settle into a minimum. Two common cooling schedules are
a linear decrease and a geometric decrease (multiplying $T$ by a factor just below 1
each step); the geometric schedule spends more steps at low temperature and usually
converges faster. We fold both into one function.

::: {.callout-note title="Recipe 9A.3.1"}
**Objective.** Find a global minimum by cooling a Metropolis chain.

**Model.** Himmelblau's function `f(x, y) = (x^2 + y - 11)^2 + (x + y^2 - 7)^2`, which has four equal minima at f = 0.

**Method.** Simulated annealing, lowering the temperature during a single run on a geometric (or linear) cooling schedule with Metropolis acceptance.

**Test.** Start at (0, 0), `1e4` steps, step size 0.5, maximum temperature 50, geometric schedule (scaling 0.999); four runs with seeds 1 to 4.

**Verification.** Each run cools from a broad search into a single minimum, but which minimum depends on the random path, so different seeds land in different basins.
:::

### R implementation {.impl}

```{r}
# simulated annealing with a linear or geometric cooling schedule
sa <- function(n, func, nstep, dx_max, temp_max, schedule = "geometric", scaling = 0.999) {
  t <- if (schedule == "linear") seq(temp_max, 0, length.out = nstep + 1)
       else temp_max * scaling^(0:nstep)
  x <- matrix(0, nstep + 1, n); e <- numeric(nstep + 1); e[1] <- func(x[1, ])
  for (i in 1:nstep) {
    xnew <- x[i, ] + runif(n, -dx_max, dx_max)
    de <- func(xnew) - e[i]
    if (de < 0 || runif(1) < exp(-de / t[i])) {
      x[i + 1, ] <- xnew; e[i + 1] <- e[i] + de
    } else { x[i + 1, ] <- x[i, ]; e[i + 1] <- e[i] }
  }
  list(x = x, e = e, t = t)
}
```

```{r}
#| fig-width: 7
#| fig-height: 3.2
#| out-width: "80%"
#| rmar: false
# the two cooling schedules
nstep <- 10^4
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(seq(50, 0, length.out = nstep + 1), type = "l", col = 2, xlab = "Step",
     ylab = "T", main = "linear")
plot(50 * 0.999^(0:nstep), type = "l", col = 4, xlab = "Step", ylab = "T", main = "geometric")
```

```{r}
#| fig-width: 8
#| fig-height: 8
#| out-width: "85%"
# four independent geometric-cooling runs; different seeds land in different minima
traj_all <- do.call(rbind, lapply(1:4, function(seed) {
  set.seed(seed)
  r <- sa(2, f_Himmelblau, nstep, dx_max = 0.5, temp_max = 50, scaling = 0.999)
  data.frame(x = r$x[, 1], y = r$x[, 2], step = 0:nstep, seed = paste0("seed ", seed))
}))
ggplot() + base_contour() +
  geom_path(data = traj_all, aes(x, y, colour = step), linewidth = 0.2) +
  facet_wrap(~seed)
```

### Python implementation {.impl}

```{python}
# simulated annealing with a linear or geometric cooling schedule
def sa(n, func, nstep, dx_max, temp_max, schedule="geometric", scaling=0.999):
    if schedule == "linear":
        t = np.linspace(temp_max, 0, nstep + 1)
    else:
        t = temp_max * scaling**np.arange(nstep + 1)
    x = np.zeros((nstep + 1, n)); e = np.zeros(nstep + 1); e[0] = func(x[0])
    for i in range(nstep):
        xnew = x[i] + np.random.uniform(-dx_max, dx_max, n)
        de = func(xnew) - e[i]
        if de < 0 or np.random.rand() < np.exp(-de / t[i]):
            x[i + 1] = xnew; e[i + 1] = e[i] + de
        else:
            x[i + 1] = x[i]; e[i + 1] = e[i]
    return {"x": x, "e": e, "t": t}
```

```{python}
#| fig-width: 7
#| fig-height: 3.2
#| out-width: "80%"
# the two cooling schedules
nstep = 10**4
fig, axs = plt.subplots(1, 2, figsize=(7, 3.2))
axs[0].plot(np.linspace(50, 0, nstep + 1), color="red"); axs[0].set(xlabel="Step", ylabel="T", title="linear")
axs[1].plot(50 * 0.999**np.arange(nstep + 1), color="blue"); axs[1].set(xlabel="Step", ylabel="T", title="geometric")
plt.tight_layout(); plt.show()
```

```{python}
#| fig-width: 8
#| fig-height: 8
#| out-width: "85%"
# four independent geometric-cooling runs; different seeds land in different minima
fig, axs = plt.subplots(2, 2, figsize=(8, 8))
for ax, seed in zip(axs.flat, (1, 2, 3, 4)):
    np.random.seed(seed)
    r = sa(2, f_Himmelblau, nstep, dx_max=0.5, temp_max=50, scaling=0.999)
    plot_traj(ax, r["x"], f"seed {seed}")
plt.tight_layout(); plt.show()
```

Each run cools from a broad high-temperature search (blue) into one minimum (red), but
which minimum depends on the random path, so different seeds find different basins.
Because SA does not guarantee the global optimum, one typically runs it many times and
keeps the best result.

## 9A.4 Parallel tempering

Parallel tempering (PT), or replica exchange [@hukushima1996exchange], runs several
chains at once, each at its own fixed temperature, and periodically proposes to *swap*
the coordinates of two replicas. Hot replicas cross barriers and feed configurations
down to cold replicas, which refine them in the minima. A swap between replicas $i$ and
$j$ is accepted with

$$ a = \min\left(1, e^{(f_i - f_j)(1/T_i - 1/T_j)}\right) , \tag{3} $$

which preserves detailed balance so every replica still samples the Boltzmann
distribution at its own temperature. We generalize to any number of replicas, running
the `mh` sampler on each between swaps and only swapping adjacent temperatures (in
alternating even/odd rounds for ergodicity).

::: {.callout-note title="Recipe 9A.4.1"}
**Objective.** Sample every basin in one run by exchanging replicas across temperatures.

**Model.** Himmelblau's function `f(x, y) = (x^2 + y - 11)^2 + (x + y^2 - 7)^2`, which has four equal minima at f = 0.

**Method.** Parallel tempering (replica exchange): run a Metropolis chain at each of several temperatures and swap adjacent-temperature replicas with acceptance `a = min(1, exp((f_i - f_j)*(1/T_i - 1/T_j)))`.

**Test.** 6 replicas at `T = 2.3, 5, 10, 20, 40, 80` with step sizes 0.3 to 2.5, all started at the origin, 100 swap rounds of 100 steps; seed 1.

**Verification.** The coldest replica stays inside minima yet still hops between them via configurations passed down from hotter replicas, while the hottest roams the whole landscape.
:::

### R implementation {.impl}

```{r}
# parallel tempering over nrep replicas at increasing temperatures
pt_multi <- function(n, func, x0, nrep, niter, nstep, dx_max, temp, ifprint = FALSE) {
  x <- lapply(1:nrep, function(j) matrix(0, niter * (nstep + 1), n))
  e <- matrix(0, niter * (nstep + 1), nrep)
  for (j in 1:nrep) x[[j]][1, ] <- x0[j, ]
  num_accept <- num_proposed <- numeric(nrep - 1)
  for (i in 1:niter) {
    start <- (i - 1) * (nstep + 1) + 1; end <- i * (nstep + 1)
    res <- lapply(1:nrep, function(j) mh(n, func, x[[j]][start, ], nstep, dx_max[j], temp[j]))
    for (j in 1:nrep) { x[[j]][start:end, ] <- res[[j]]$x; e[start:end, j] <- res[[j]]$e }
    if (i == niter) break
    for (j in 1:nrep) { x[[j]][end + 1, ] <- x[[j]][end, ]; e[end + 1, j] <- e[end, j] }
    j_seq <- if (i %% 2 == 1) seq(3, nrep, 2) else seq(2, nrep, 2)   # alternate swap rounds
    for (j in j_seq) {
      num_proposed[j - 1] <- num_proposed[j - 1] + 1
      a <- exp((e[end, j - 1] - e[end, j]) * (1 / temp[j - 1] - 1 / temp[j]))
      if (runif(1) < a) {
        num_accept[j - 1] <- num_accept[j - 1] + 1
        x[[j - 1]][end + 1, ] <- x[[j]][end, ]; x[[j]][end + 1, ] <- x[[j - 1]][end, ]
        e[end + 1, j - 1] <- e[end, j]; e[end + 1, j] <- e[end, j - 1]
      }
    }
  }
  if (ifprint) cat("swap acceptance:", num_accept / num_proposed, "\n")
  list(x = x, e = e)
}
```

```{r}
#| fig-width: 9
#| fig-height: 6
#| out-width: "92%"
set.seed(1)
nrep <- 6; niter <- 100; nstep <- 100
temp <- c(2.3, 5, 10, 20, 40, 80); dx_max <- c(0.3, 0.5, 0.7, 1, 1.5, 2.5)
res_pt <- pt_multi(2, f_Himmelblau, matrix(0, nrep, 2), nrep, niter, nstep, dx_max, temp, ifprint = TRUE)

ntot <- niter * (nstep + 1)
traj_all <- do.call(rbind, lapply(1:nrep, function(j) {
  data.frame(x = res_pt$x[[j]][, 1], y = res_pt$x[[j]][, 2], step = 1:ntot,
             rep = factor(paste0("T = ", temp[j]), levels = paste0("T = ", temp)))
}))
ggplot() + base_contour() +
  geom_path(data = traj_all, aes(x, y, colour = step), linewidth = 0.2) +
  facet_wrap(~rep)
```

### Python implementation {.impl}

```{python}
# parallel tempering over nrep replicas at increasing temperatures
def pt_multi(n, func, x0, nrep, niter, nstep, dx_max, temp, ifprint=False):
    x = [np.zeros((niter * (nstep + 1), n)) for _ in range(nrep)]
    e = np.zeros((niter * (nstep + 1), nrep))
    for j in range(nrep):
        x[j][0] = x0[j]
    num_accept = np.zeros(nrep - 1); num_proposed = np.zeros(nrep - 1)
    for i in range(1, niter + 1):
        start = (i - 1) * (nstep + 1); end = i * (nstep + 1)
        res = [mh(n, func, x[j][start], nstep, dx_max[j], temp[j]) for j in range(nrep)]
        for j in range(nrep):
            x[j][start:end] = res[j]["x"]; e[start:end, j] = res[j]["e"]
        if i == niter:
            break
        for j in range(nrep):
            x[j][end] = x[j][end - 1]; e[end, j] = e[end - 1, j]
        j_seq = range(2, nrep, 2) if i % 2 == 1 else range(1, nrep, 2)   # alternate swap rounds
        for j in j_seq:
            num_proposed[j - 1] += 1
            a = np.exp((e[end, j - 1] - e[end, j]) * (1 / temp[j - 1] - 1 / temp[j]))
            if np.random.rand() < a:
                num_accept[j - 1] += 1
                x[j - 1][end] = x[j][end - 1]; x[j][end] = x[j - 1][end - 1]
                e[end, j - 1] = e[end - 1, j]; e[end, j] = e[end - 1, j - 1]
    if ifprint:
        print("swap acceptance:", num_accept / num_proposed)
    return {"x": x, "e": e}
```

```{python}
#| fig-width: 9
#| fig-height: 6
#| out-width: "92%"
np.random.seed(1)
nrep = 6; niter = 100; nstep = 100
temp = [2.3, 5, 10, 20, 40, 80]; dx_max = [0.3, 0.5, 0.7, 1, 1.5, 2.5]
res_pt = pt_multi(2, f_Himmelblau, np.zeros((nrep, 2)), nrep, niter, nstep, dx_max, temp, ifprint=True)

fig, axs = plt.subplots(2, 3, figsize=(9, 6))
for ax, j in zip(axs.flat, range(nrep)):
    plot_traj(ax, res_pt["x"][j], f"T = {temp[j]}")
plt.tight_layout(); plt.show()
```

The coldest replica ($T = 2.3$) stays tightly inside minima yet still hops between them,
thanks to configurations passed down from the hotter replicas, while the hottest
($T = 80$) roams over the whole landscape. Monitoring the swap acceptance rate helps
choose the temperature spacing: if adjacent temperatures are too far apart, swaps are
rarely accepted and more replicas are needed.

## 9A.5 Simulated tempering

Simulated tempering (ST) achieves a similar effect with a *single* chain whose
temperature itself performs a random walk over a discrete ladder $T_1 < \dots <
T_{n_\text{temp}}$ [@marinari1992simulated]. A mixing parameter $\alpha$ sets the
fraction of steps spent moving in space (ordinary MH at the current temperature) versus
moving in temperature. A proposed temperature change to $T'$ is accepted with

$$ a = \min\left(1, \frac{c'}{c}\, e^{-f(1/T' - 1/T)}\right) , \tag{4} $$

where the weights $c$ bias how often each temperature is visited (larger weights at low
temperature emphasize refining minima). The weights would be the partition functions to
sample all temperatures equally, but since those are unknown we choose them by hand.

::: {.callout-note title="Recipe 9A.5.1"}
**Objective.** Cross barriers and settle into minima within one chain by letting the temperature wander.

**Model.** Himmelblau's function `f(x, y) = (x^2 + y - 11)^2 + (x + y^2 - 7)^2`, which has four equal minima at f = 0.

**Method.** Simulated tempering: a single chain whose temperature random-walks a discrete ladder, with temperature moves accepted as `a = min(1, (c'/c)*exp(-f*(1/T' - 1/T)))`.

**Test.** Start at (0, 0), `1e4` steps, mixing fraction 0.9, step size 1, a 16-rung ladder `T = 5, 10, ..., 80` with weights decreasing linearly from 5 to 1; seed 1.

**Verification.** Because the temperature wanders up and down, one trajectory both crosses barriers when hot and settles into minima when cold, visiting all four basins.
:::

### R implementation {.impl}

```{r}
# simulated tempering: one chain, temperature random-walks a discrete ladder
st <- function(n, func, x0, nstep, alpha, dx_max, temp_all, weight_all, ifprint = FALSE) {
  ntemp <- length(temp_all)
  x <- matrix(0, nstep + 1, n); ind_t <- integer(nstep + 1); e <- numeric(nstep + 1)
  x[1, ] <- x0; ind_t[1] <- sample(ntemp, 1); e[1] <- func(x0)
  for (i in 1:nstep) {
    temp <- temp_all[ind_t[i]]
    if (runif(1) <= alpha) {                       # spatial MH move
      ind_t[i + 1] <- ind_t[i]
      xnew <- x[i, ] + runif(n, -dx_max, dx_max); de <- func(xnew) - e[i]
      if (de < 0 || runif(1) < exp(-de / temp)) { x[i + 1, ] <- xnew; e[i + 1] <- e[i] + de }
      else { x[i + 1, ] <- x[i, ]; e[i + 1] <- e[i] }
    } else {                                        # temperature move
      ind_new <- if (sample(2, 1) == 1) max(1, ind_t[i] - 1) else min(ind_t[i] + 1, ntemp)
      temp_new <- temp_all[ind_new]
      a <- weight_all[ind_new] / weight_all[ind_t[i]] * exp(-e[i] * (1 / temp_new - 1 / temp))
      ind_t[i + 1] <- if (runif(1) < a) ind_new else ind_t[i]
      x[i + 1, ] <- x[i, ]; e[i + 1] <- e[i]
    }
  }
  list(x = x, e = e, ind_t = ind_t)
}
```

```{r}
#| fig-width: 9
#| fig-height: 4
#| out-width: "92%"
set.seed(1)
nstep <- 10^4
temp_all <- seq(5, 80, 5); weight_all <- seq(5, 1, length.out = length(temp_all))
res_st <- st(2, f_Himmelblau, numeric(2), nstep, alpha = 0.9, dx_max = 1,
             temp_all = temp_all, weight_all = weight_all)

library(patchwork)
g_traj <- ggplot() + base_contour() +
  geom_path(data = data.frame(x = res_st$x[, 1], y = res_st$x[, 2], step = 0:nstep),
            aes(x, y, colour = step), linewidth = 0.2) + ggtitle("trajectory")
g_temp <- ggplot(data.frame(step = 0:nstep, temp = temp_all[res_st$ind_t]), aes(step, temp)) +
  geom_line(colour = "red") + labs(x = "Step", y = "Temperature") + theme_bw()
g_traj + g_temp                                # side-by-side via patchwork
```

### Python implementation {.impl}

```{python}
# simulated tempering: one chain, temperature random-walks a discrete ladder
def st(n, func, x0, nstep, alpha, dx_max, temp_all, weight_all, ifprint=False):
    ntemp = len(temp_all)
    x = np.zeros((nstep + 1, n)); ind_t = np.zeros(nstep + 1, dtype=int); e = np.zeros(nstep + 1)
    x[0] = x0; ind_t[0] = np.random.randint(ntemp); e[0] = func(x0)
    for i in range(nstep):
        temp = temp_all[ind_t[i]]
        if np.random.rand() <= alpha:                  # spatial MH move
            ind_t[i + 1] = ind_t[i]
            xnew = x[i] + np.random.uniform(-dx_max, dx_max, n); de = func(xnew) - e[i]
            if de < 0 or np.random.rand() < np.exp(-de / temp):
                x[i + 1] = xnew; e[i + 1] = e[i] + de
            else:
                x[i + 1] = x[i]; e[i + 1] = e[i]
        else:                                          # temperature move
            ind_new = max(0, ind_t[i] - 1) if np.random.randint(2) == 0 else min(ind_t[i] + 1, ntemp - 1)
            temp_new = temp_all[ind_new]
            a = weight_all[ind_new] / weight_all[ind_t[i]] * np.exp(-e[i] * (1 / temp_new - 1 / temp))
            ind_t[i + 1] = ind_new if np.random.rand() < a else ind_t[i]
            x[i + 1] = x[i]; e[i + 1] = e[i]
    return {"x": x, "e": e, "ind_t": ind_t}
```

```{python}
#| fig-width: 9
#| fig-height: 4
#| out-width: "92%"
np.random.seed(1)
nstep = 10**4
temp_all = np.arange(5, 81, 5); weight_all = np.linspace(5, 1, len(temp_all))
res_st = st(2, f_Himmelblau, np.zeros(2), nstep, alpha=0.9, dx_max=1,
            temp_all=temp_all, weight_all=weight_all)

fig, axs = plt.subplots(1, 2, figsize=(9, 4))
plot_traj(axs[0], res_st["x"], "trajectory")
axs[1].plot(range(nstep + 1), temp_all[res_st["ind_t"]], color="red")
axs[1].set(xlabel="Step", ylabel="Temperature")
plt.tight_layout(); plt.show()
```

Because the temperature wanders up and down, a single ST trajectory both crosses
barriers (when it is hot) and settles into minima (when it is cold), visiting all four
basins in one run. Unlike simulated annealing, which ends cold and stops, simulated
tempering keeps sampling, and unlike parallel tempering it needs only one chain.
