---
title: "9C. Genetic algorithm"
author: "Mingyang Lu"
old_id: "09C"
---

A *genetic algorithm* (GA) is a stochastic global optimizer inspired by natural
selection [@holland1975adaptation]. It maintains a *population* of candidate solutions
and improves it over successive *generations* by three operators borrowed from
evolution: *crossover* (combining two solutions), *mutation* (randomly perturbing one),
and *selection* (keeping the fittest). It needs only a genetic representation of a
solution (usually a vector) and a fitness function to optimize.

## 9C.1 Genetic algorithm for continuous variables

We first optimize a continuous function, the Rastrigin function [@torn1989global],

$$ f(x_1, x_2) = 20 + \sum_{i=1}^{2} \left( x_i^2 - 10\cos(2\pi x_i) \right) , $$

a standard benchmark that is strongly *multimodal*: a regular lattice of local minima
sits on a broad bowl, with the global minimum $f = 0$ at the origin. Escaping the many
local minima is exactly the kind of global search a genetic algorithm is built for. The
function is also *separable* (each coordinate contributes independently), which suits the
crossover below: it recombines whole coordinates from two parents, so a good $x_1$ from
one parent can join a good $x_2$ from another. The genetic representation is the vector
$(x_1, x_2)$ and the fitness is $f$ (to be minimized). Crossover swaps variables between
two parents to make two offspring; mutation nudges one randomly chosen variable by a
small uniform step. Each generation we make `np` offspring from `np` parents, then keep
the best `np` of the combined pool.

::: {.callout-note title="Recipe 9C.1.1"}
**Objective.** Minimize a rugged multimodal function with a genetic algorithm.

**Model.** The Rastrigin function `f(x1, x2) = 20 + sum(x_i^2 - 10*cos(2*pi*x_i))`, whose global minimum f = 0 sits at the origin amid many local minima.

**Method.** A genetic algorithm for continuous variables: crossover swaps whole variables between parents, mutation nudges one variable, and the fittest members are kept each generation (elitist selection).

**Test.** Two variables on `[-5.12, 5.12]^2`, population 50, crossover rate 0.7, 100 generations, mutation step 10% of the range; seed 1.

**Verification.** The best score drops toward zero, converging on the global minimum at the origin despite the many local minima.
:::

### R implementation {.impl}

```{r}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
library(ggplot2)
f_rastrigin <- function(x) 20 + sum(x^2 - 10 * cos(2 * pi * x))

grid_df <- expand.grid(x1 = seq(-5.12, 5.12, 0.05), x2 = seq(-5.12, 5.12, 0.05))
grid_df$z <- apply(grid_df, 1, f_rastrigin)
ggplot(grid_df, aes(x1, x2, z = z)) +
  geom_contour(breaks = c(1, 3, 6, 12, 22, 38, 60), colour = "grey50") + coord_fixed() + theme_bw()
```

```{r}
# swap variables between two parents; nudge one variable of a parent
crossover_opt <- function(n, x1, x2) {
  w <- sample(0:1, n, replace = TRUE)
  rbind(x1 * w + x2 * (1 - w), x2 * w + x1 * (1 - w))
}
mutation_opt <- function(n, x, xrange, dx_max) {
  i <- sample(n, 1); x[i] <- x[i] + runif(1, -dx_max[i], dx_max[i])
  x[i] <- min(max(x[i], xrange[i, 1]), xrange[i, 2]); x
}

# genetic algorithm for a continuous function (minimized)
ga_opt <- function(n, func, np, rate_x, ngen, xrange, dx_ratio) {
  np <- ceiling(np / 2) * 2
  dx_max <- (xrange[, 2] - xrange[, 1]) * dx_ratio
  x_par <- sapply(1:n, function(i) runif(np, xrange[i, 1], xrange[i, 2]))
  s_par <- apply(x_par, 1, func)
  x_off <- matrix(0, np, n); s_keep <- matrix(0, ngen, np)
  for (g in 1:ngen) {
    ir <- sample(np)
    for (j in 1:(np / 2)) {
      x1 <- x_par[ir[2 * j - 1], ]; x2 <- x_par[ir[2 * j], ]
      if (runif(1) < rate_x) x_off[(2 * j - 1):(2 * j), ] <- crossover_opt(n, x1, x2)
      else { x_off[2 * j - 1, ] <- mutation_opt(n, x1, xrange, dx_max)
             x_off[2 * j, ] <- mutation_opt(n, x2, xrange, dx_max) }
    }
    s_off <- apply(x_off, 1, func)
    x_all <- rbind(x_par, x_off); s_all <- c(s_par, s_off)
    keep <- order(s_all)[1:np]                     # keep the fittest np
    x_par <- x_all[keep, ]; s_par <- s_all[keep]; s_keep[g, ] <- s_par
  }
  list(x = x_par[1, ], s = s_par[1], s_keep = s_keep)
}

set.seed(1)
xrange <- matrix(c(-5.12, 5.12, -5.12, 5.12), 2, byrow = TRUE)
res <- ga_opt(2, f_rastrigin, np = 50, rate_x = 0.7, ngen = 100, xrange, dx_ratio = 0.1)
cat("best x:", round(res$x, 3), " best score:", signif(res$s, 3), "\n")
matplot(1:100, cbind(apply(res$s_keep, 1, min), rowMeans(res$s_keep), apply(res$s_keep, 1, max)),
        type = "l", lty = 1, col = c(2, 1, 4), xlab = "Generation", ylab = "Score", log = "y")
legend("topright", c("min", "mean", "max"), col = c(2, 1, 4), lty = 1, cex = 0.8)
```

### Python implementation {.impl}

```{python}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
import numpy as np
import matplotlib.pyplot as plt

def f_rastrigin(x):
    x = np.asarray(x)
    return 20 + np.sum(x**2 - 10 * np.cos(2 * np.pi * x), axis=0)

gx = np.arange(-5.12, 5.12, 0.05)
X1, X2 = np.meshgrid(gx, gx); Z = f_rastrigin([X1, X2])
plt.figure(figsize=(5, 5))
plt.contour(X1, X2, Z, levels=[1, 3, 6, 12, 22, 38, 60], colors="grey")
plt.gca().set_aspect("equal"); plt.xlabel("x1"); plt.ylabel("x2"); plt.show()
```

```{python}
# swap variables between two parents; nudge one variable of a parent
def crossover_opt(n, x1, x2):
    w = np.random.randint(0, 2, n)
    return np.vstack((x1 * w + x2 * (1 - w), x2 * w + x1 * (1 - w)))
def mutation_opt(n, x, xrange, dx_max):
    i = np.random.randint(n); x = x.copy()
    x[i] = np.clip(x[i] + np.random.uniform(-dx_max[i], dx_max[i]), xrange[i, 0], xrange[i, 1])
    return x

# genetic algorithm for a continuous function (minimized)
def ga_opt(n, func, np_, rate_x, ngen, xrange, dx_ratio):
    np_ += np_ % 2
    dx_max = (xrange[:, 1] - xrange[:, 0]) * dx_ratio
    x_par = np.random.uniform(xrange[:, 0], xrange[:, 1], size=(np_, n))
    s_par = np.apply_along_axis(func, 1, x_par)
    x_off = np.zeros((np_, n)); s_keep = np.zeros((ngen, np_))
    for g in range(ngen):
        ir = np.random.permutation(np_)
        for j in range(np_ // 2):
            x1, x2 = x_par[ir[2 * j]], x_par[ir[2 * j + 1]]
            if np.random.rand() < rate_x:
                x_off[2 * j:2 * j + 2] = crossover_opt(n, x1, x2)
            else:
                x_off[2 * j] = mutation_opt(n, x1, xrange, dx_max)
                x_off[2 * j + 1] = mutation_opt(n, x2, xrange, dx_max)
        s_off = np.apply_along_axis(func, 1, x_off)
        x_all = np.vstack((x_par, x_off)); s_all = np.concatenate((s_par, s_off))
        keep = np.argsort(s_all)[:np_]                 # keep the fittest np
        x_par, s_par = x_all[keep], s_all[keep]; s_keep[g] = s_par
    return {"x": x_par[0], "s": s_par[0], "s_keep": s_keep}

np.random.seed(1)
xrange = np.array([[-5.12, 5.12], [-5.12, 5.12]])
res = ga_opt(2, f_rastrigin, 50, 0.7, 100, xrange, 0.1)
print("best x:", np.round(res["x"], 3), " best score:", float(f"{res['s']:.3g}"))
s = res["s_keep"]
plt.plot(range(100), s.min(1), "r", label="min")
plt.plot(range(100), s.mean(1), "k", label="mean")
plt.plot(range(100), s.max(1), "b", label="max")
plt.yscale("log"); plt.xlabel("Generation"); plt.ylabel("Score"); plt.legend(); plt.show()
```

The best score drops toward zero, converging on the global minimum at the origin
despite the many local minima that would trap a purely local method. The min/mean/max
curves show the whole population improving, with the spread narrowing as it converges.
Performance depends on the crossover rate, population size, and number of generations.

## 9C.2 Genetic algorithm for the TSP

For the traveling salesman problem the genetic representation is a *permutation* of the
city indices, and the fitness is the total tour length. Permutations need
permutation-preserving operators. For mutation we use either a *point* swap (exchange
two cities) or an *inversion* (reverse the order of a segment).

::: {.callout-note title="Recipe 9C.2.1"}
**Objective.** Approximate the shortest tour of many cities with a genetic algorithm.

**Model.** The traveling salesman problem with the tour-length objective (the sum of consecutive city distances plus the return leg); a candidate solution is a permutation of the city indices.

**Method.** A genetic algorithm on permutations, using permutation-preserving mutation (swap two cities, or reverse a segment) and order-preserving crossover.

**Test.** A 10-city instance, population 10, crossover rate 0.8, 50 generations.

**Verification.** The genetic algorithm recovers the exact shortest tour found by dynamic programming on the small instance, and scales to larger instances where exact methods cannot.
:::

### R implementation {.impl}

```{r}
# total tour length for a city order z
cal_s <- function(n, d, z) {
  dis <- 0; for (i in 1:(n - 1)) dis <- dis + d[z[i], z[i + 1]]
  dis + d[z[1], z[n]]
}
# point mutation: swap two cities
mut_point <- function(n, z) { s <- sample(n, 2); z[s] <- z[rev(s)]; z }
# inversion mutation: reverse a segment
mut_inverse <- function(n, z) { s <- sample(n, 2); z[s[1]:s[2]] <- z[s[2]:s[1]]; z }
```

### Python implementation {.impl}

```{python}
# total tour length for a city order z
def cal_s(n, d, z):
    dis = sum(d[z[i], z[i + 1]] for i in range(n - 1))
    return dis + d[z[0], z[n - 1]]
# point mutation: swap two cities
def mut_point(n, z):
    s = np.random.choice(n, 2, replace=False); z = z.copy()
    z[s[0]], z[s[1]] = z[s[1]], z[s[0]]; return z
# inversion mutation: reverse a segment
def mut_inverse(n, z):
    s = sorted(np.random.choice(n, 2, replace=False)); z = z.copy()
    z[s[0]:s[1] + 1] = z[s[0]:s[1] + 1][::-1]; return z
```

## 9C.3 Crossover operators for permutations

Ordinary crossover would produce invalid tours with repeated or missing cities, so
permutations use special operators [@goldberg1989genetic]. We implement three classic
ones. *Partially mapped crossover* (PMX) copies a segment from one parent and repairs
the rest through the position mapping between parents. *Order crossover* (OX) keeps a
segment from one parent and fills the rest in the order they appear in the other.
*Cycle crossover* (CX) builds offspring from cycles of positions shared by the parents.

### R implementation {.impl}

```{r}
# PMX: partially mapped crossover
pmx <- function(n, z1, z2) {
  s <- sort(sample(n, 2)); if (s[1] == 1 && s[2] == n) return(rbind(z1, z2))
  seg <- s[1]:s[2]; rest <- setdiff(1:n, seg)
  c1 <- z1[seg]; c2 <- z2[seg]; z1n <- z2n <- numeric(n)
  z1n[seg] <- c2
  for (i in rest) { cand <- z1[i]; while (cand %in% c2) cand <- c1[c2 == cand]; z1n[i] <- cand }
  z2n[seg] <- c1
  for (i in rest) { cand <- z2[i]; while (cand %in% c1) cand <- c2[c1 == cand]; z2n[i] <- cand }
  rbind(z1n, z2n)
}

# OX: order crossover
ox <- function(n, z1, z2) {
  s <- sort(sample(n, 2)); if (s[1] == 1 && s[2] == n) return(rbind(z1, z2))
  fill_from <- function(keep, other) {
    ordered <- c(tail(other, -s[2]), head(other, s[2]))
    c(keep, setdiff(ordered, keep))
  }
  z2n <- fill_from(z1[s[1]:s[2]], z2); z1n <- fill_from(z2[s[1]:s[2]], z1)
  if (s[1] > 1) { sh <- s[1] - 1
    z2n <- c(tail(z2n, sh), head(z2n, -sh)); z1n <- c(tail(z1n, sh), head(z1n, -sh)) }
  rbind(z1n, z2n)
}

# CX: cycle crossover
cx <- function(n, z1, z2) {
  z1n <- numeric(n); z1n[1] <- z1[1]; ind <- which(z1 == z2[1])
  while (z1n[ind] == 0) { z1n[ind] <- z1[ind]; ind <- which(z1 == z2[ind]) }
  rest <- which(z1n == 0); z1n[rest] <- z2[rest]
  rbind(z1n, z1 + z2 - z1n)
}
```

```{r}
# demonstrate the operators on two example tours
set.seed(1)
n <- 8; z1 <- c(3, 4, 8, 2, 7, 1, 6, 5); z2 <- c(4, 2, 5, 1, 6, 8, 3, 7)
rbind(point = mut_point(n, z1), inverse = mut_inverse(n, z1))
pmx(n, z1, z2); ox(n, z1, z2); cx(n, z1, z2)
```

### Python implementation {.impl}

```{python}
# PMX: partially mapped crossover
def pmx(n, z1, z2):
    s = sorted(np.random.choice(n, 2, replace=False))
    if s[0] == 0 and s[1] == n - 1: return np.vstack((z1, z2))
    seg = np.arange(s[0], s[1] + 1); rest = np.setdiff1d(np.arange(n), seg)
    c1, c2 = z1[seg], z2[seg]; z1n = np.zeros(n, int); z2n = np.zeros(n, int)
    z1n[seg] = c2
    for i in rest:
        cand = z1[i]
        while cand in c2: cand = c1[np.where(c2 == cand)[0][0]]
        z1n[i] = cand
    z2n[seg] = c1
    for i in rest:
        cand = z2[i]
        while cand in c1: cand = c2[np.where(c1 == cand)[0][0]]
        z2n[i] = cand
    return np.vstack((z1n, z2n))

# OX: order crossover
def ox(n, z1, z2):
    s = sorted(np.random.choice(n, 2, replace=False))
    if s[0] == 0 and s[1] == n - 1: return np.vstack((z1, z2))
    def fill_from(keep, other):
        ordered = np.concatenate((other[s[1] + 1:], other[:s[1] + 1]))
        return np.concatenate((keep, np.array([c for c in ordered if c not in keep])))
    z2n = fill_from(z1[s[0]:s[1] + 1], z2); z1n = fill_from(z2[s[0]:s[1] + 1], z1)
    if s[0] > 0:
        z2n = np.concatenate((z2n[-s[0]:], z2n[:-s[0]])); z1n = np.concatenate((z1n[-s[0]:], z1n[:-s[0]]))
    return np.vstack((z1n, z2n))

# CX: cycle crossover
def cx(n, z1, z2):
    z1n = np.zeros(n, int); mask = np.zeros(n, bool)
    z1n[0] = z1[0]; mask[0] = True; ind = np.where(z1 == z2[0])[0][0]
    while not mask[ind]:
        z1n[ind] = z1[ind]; mask[ind] = True; ind = np.where(z1 == z2[ind])[0][0]
    z1n[~mask] = z2[~mask]
    return np.vstack((z1n, z1 + z2 - z1n))
```

```{python}
# demonstrate the operators on two example tours
np.random.seed(1)
n = 8; z1 = np.array([3, 4, 8, 2, 7, 1, 6, 5]); z2 = np.array([4, 2, 5, 1, 6, 8, 3, 7])
print("point:  ", mut_point(n, z1)); print("inverse:", mut_inverse(n, z1))
print("pmx:\n", pmx(n, z1, z2)); print("ox:\n", ox(n, z1, z2)); print("cx:\n", cx(n, z1, z2))
```

## 9C.4 Testing on the TSP

The GA for the TSP follows the same loop as the continuous version, now with the
permutation operators. We start from random tours and, each generation, pair parents to
produce offspring by crossover or mutation, then keep the fittest.

### R implementation {.impl}

```{r}
# genetic algorithm for the TSP
tsp_ga <- function(n, d, np, rate_x, ngen, func_x, func_m) {
  np <- ceiling(np / 2) * 2
  z_par <- t(replicate(np, sample(n))); s_par <- apply(z_par, 1, function(z) cal_s(n, d, z))
  z_off <- matrix(0, np, n); s_keep <- matrix(0, ngen, np)
  for (g in 1:ngen) {
    ir <- sample(np)
    for (j in 1:(np / 2)) {
      z1 <- z_par[ir[2 * j - 1], ]; z2 <- z_par[ir[2 * j], ]
      if (runif(1) < rate_x) z_off[(2 * j - 1):(2 * j), ] <- func_x(n, z1, z2)
      else { z_off[2 * j - 1, ] <- func_m(n, z1); z_off[2 * j, ] <- func_m(n, z2) }
    }
    s_off <- apply(z_off, 1, function(z) cal_s(n, d, z))
    z_all <- rbind(z_par, z_off); s_all <- c(s_par, s_off)
    keep <- order(s_all)[1:np]
    z_par <- z_all[keep, ]; s_par <- s_all[keep]; s_keep[g, ] <- s_par
  }
  list(z = z_par[1, ], s = s_par[1], s_keep = s_keep)
}

# helper to plot the tour and the score convergence side by side
plot_tsp <- function(cities, res, ngen, title) {
  p <- c(res$z, res$z[1]); s <- res$s_keep
  par(mfrow = c(1, 2), mar = c(4, 4, 2, 1), pty = "s")
  plot(cities$x, cities$y, pch = 19, cex = 0.6, xlab = "x", ylab = "y", main = title, asp = 1)
  lines(cities$x[p], cities$y[p], col = 2)
  par(pty = "m")
  matplot(1:ngen, cbind(apply(s, 1, min), rowMeans(s), apply(s, 1, max)),
          type = "l", lty = 1, col = c(2, 1, 4), xlab = "Generation", ylab = "Score")
}
```

```{r}
#| fig-width: 9
#| fig-height: 4.5
#| out-width: "92%"
#| rmar: false
# ten-city instance: the GA finds the exact optimum
c10 <- data.frame(
  x = c(0, -28.87, -14.66, -29.06, -36.04, -50.48, -50.59, -0.14, -21.50, -43.07),
  y = c(0,   0.00,  43.39,  43.22,  21.61,  -7.37,  21.59, 28.73,  -7.32, -14.55))
d10 <- as.matrix(dist(c10))
set.seed(1)
res <- tsp_ga(10, d10, np = 10, rate_x = 0.8, ngen = 50, func_x = pmx, func_m = mut_point)
plot_tsp(c10, res, 50, paste("10 cities, length", round(res$s, 1)))
```

```{r}
#| fig-width: 9
#| fig-height: 4.5
#| out-width: "92%"
#| rmar: false
# the 52-city berlin52 instance (TSPLIB): a much harder problem
c52 <- read.table("09-optimization/data/berlin52.tsp", col.names = c("ind", "x", "y"))
d52 <- as.matrix(dist(c52[, c("x", "y")]))     # distances from x, y only
set.seed(1)
res_pp <- tsp_ga(52, d52, np = 40, rate_x = 0.7, ngen = 2000, func_x = pmx, func_m = mut_point)
plot_tsp(c52, res_pp, 2000, paste("PMX + point swap, length", round(res_pp$s)))
```

```{r}
#| fig-width: 9
#| fig-height: 4.5
#| out-width: "92%"
#| rmar: false
# order crossover with inversion mutation does much better
set.seed(1)
res_oi <- tsp_ga(52, d52, np = 40, rate_x = 0.7, ngen = 2000, func_x = ox, func_m = mut_inverse)
plot_tsp(c52, res_oi, 2000, paste("OX + inversion, length", round(res_oi$s)))
```

### Python implementation {.impl}

```{python}
# genetic algorithm for the TSP
def tsp_ga(n, d, np_, rate_x, ngen, func_x, func_m):
    np_ += np_ % 2
    z_par = np.array([np.random.permutation(n) for _ in range(np_)])
    s_par = np.array([cal_s(n, d, z) for z in z_par])
    z_off = np.zeros((np_, n), int); s_keep = np.zeros((ngen, np_))
    for g in range(ngen):
        ir = np.random.permutation(np_)
        for j in range(np_ // 2):
            z1, z2 = z_par[ir[2 * j]], z_par[ir[2 * j + 1]]
            if np.random.rand() < rate_x:
                z_off[2 * j:2 * j + 2] = func_x(n, z1, z2)
            else:
                z_off[2 * j] = func_m(n, z1); z_off[2 * j + 1] = func_m(n, z2)
        s_off = np.array([cal_s(n, d, z) for z in z_off])
        z_all = np.vstack((z_par, z_off)); s_all = np.concatenate((s_par, s_off))
        keep = np.argsort(s_all)[:np_]
        z_par, s_par = z_all[keep], s_all[keep]; s_keep[g] = s_par
    return {"z": z_par[0], "s": s_par[0], "s_keep": s_keep}

# helper to plot the tour and the score convergence side by side
def plot_tsp(cities, res, ngen, title):
    p = list(res["z"]) + [res["z"][0]]; s = res["s_keep"]
    fig, ax = plt.subplots(1, 2, figsize=(9, 4.5))
    ax[0].plot(cities[:, 0], cities[:, 1], "o", color="black", ms=3)
    ax[0].plot(cities[p, 0], cities[p, 1], color="red")
    ax[0].set(xlabel="x", ylabel="y", title=title, aspect="equal")
    ax[1].plot(range(ngen), s.min(1), "r"); ax[1].plot(range(ngen), s.mean(1), "k")
    ax[1].plot(range(ngen), s.max(1), "b")
    ax[1].set(xlabel="Generation", ylabel="Score")
    plt.tight_layout(); plt.show()
```

```{python}
#| fig-width: 9
#| fig-height: 4.5
#| out-width: "92%"
# ten-city instance: the GA finds the exact optimum
c10 = np.array([[0, 0], [-28.87, 0], [-14.66, 43.39], [-29.06, 43.22], [-36.04, 21.61],
                [-50.48, -7.37], [-50.59, 21.59], [-0.14, 28.73], [-21.50, -7.32], [-43.07, -14.55]])
d10 = np.linalg.norm(c10[:, None, :] - c10[None, :, :], axis=-1)
np.random.seed(1)
res = tsp_ga(10, d10, 10, 0.8, 50, pmx, mut_point)
plot_tsp(c10, res, 50, f"10 cities, length {res['s']:.1f}")
```

```{python}
#| fig-width: 9
#| fig-height: 4.5
#| out-width: "92%"
# the 52-city berlin52 instance (TSPLIB): a much harder problem
c52 = np.loadtxt("09-optimization/data/berlin52.tsp")[:, 1:]    # x, y columns
d52 = np.linalg.norm(c52[:, None, :] - c52[None, :, :], axis=-1)
np.random.seed(1)
res_pp = tsp_ga(52, d52, 40, 0.7, 2000, pmx, mut_point)
plot_tsp(c52, res_pp, 2000, f"PMX + point swap, length {res_pp['s']:.0f}")
```

```{python}
#| fig-width: 9
#| fig-height: 4.5
#| out-width: "92%"
# order crossover with inversion mutation does much better
np.random.seed(1)
res_oi = tsp_ga(52, d52, 40, 0.7, 2000, ox, mut_inverse)
plot_tsp(c52, res_oi, 2000, f"OX + inversion, length {res_oi['s']:.0f}")
```

On the ten-city instance the GA recovers the exact optimum found by dynamic programming
in [Part 9B](09b-dynamic-programming.qmd). That Held-Karp solution is guaranteed to be
optimal, but its cost grows exponentially ($O(n^2 2^n)$) and becomes hopeless well
before 52 cities; the GA reproduces the known answer here and, unlike the exact method,
still runs on the larger instance. The 52-city berlin52 instance [@reinelt1991tsplib] is
much harder (its known optimal tour length is about 7542). Partially mapped crossover
with point-swap mutation converges to a poor, self-crossing tour, whereas order crossover
with inversion mutation, which better preserves useful sub-tours, reaches a considerably
shorter one, though still above the optimum. The choice of operators, not just the
parameters, drives the performance.

The convergence plots also show the population losing diversity and plateauing well
above 7542: the greedy "keep the best `np`" selection converges prematurely. A basic GA
like this can be pushed much closer to the optimum with a less greedy selection that
preserves diversity (for example tournament selection), or by hybridizing it with a
local search such as 2-opt that refines each tour (a memetic algorithm). Even so, unlike
the exact Held-Karp algorithm, the GA gives only a heuristic solution with no guarantee
of optimality.
