---
title: "2C. Modeling bacterial growth"
author: "Mingyang Lu"
old_id: "02C"
code-tools: true
---

This chapter uses ODEs to model bacterial growth (exponential growth and logistic
growth) and reuses the numerical integrators developed in
[Part 2B](02b-numerical-integration.qmd). Each section works through the model and
then poses a short exercise (in a highlighted box) for you to complete by editing
and re-running the code. The growth models follow the standard treatment in
Sontag's lecture notes on mathematical systems biology [@sontag_systemsbio].

## 2C.1 Exponential growth

The simplest model assumes a constant per-capita growth rate $r$, so the
bacterial count $N$ obeys

\begin{equation}
\frac{dN}{dt} = rN , \qquad N(t=0) = N_0 . \tag{1}
\end{equation}

This has the closed-form solution

\begin{equation}
N(t) = N_0 e^{rt} . \tag{2}
\end{equation}

(Interpret $N$ as a continuous population, not necessarily an integer.) We first
evaluate and plot the exact solution, on a linear axis and on a log-scaled axis
where the exponential appears as a straight line. We then reproduce it
numerically with the Euler and RK4 integrators from
[Part 2B](02b-numerical-integration.qmd) (repeated here so this page runs on its
own); zooming in at large $t$, RK4 tracks the exact solution far more closely than
Euler at the same step size. Finally, each language's built-in ODE solver
reproduces the same result: `deSolve` in R (here a fixed-step Euler) and SciPy's
`solve_ivp` in Python (adaptive RK45).

::: {.callout-note title="Recipe 2C.1.1"}
**Objective.** Model exponential bacterial growth and reproduce the exact solution numerically.

**Model.** A population growing at a constant per-capita rate r: `dN/dt = r*N`, with exact solution `N(t) = N0*exp(r*t)`.

**Method.** Explicit integration (Euler and RK4) and a general ODE solver.

**Test.** r = 0.1, N0 = 1; t = 0 to 100; step dt = 0.1.

**Show.** A log-scaled plot of N(t) for Euler, RK4, and the exact solution on one set of axes.

**Verification.** On a log-scaled y axis the exact solution is a straight line; at large t, RK4 tracks it far more closely than Euler; the final value matches `N(100) = exp(10)`, about 22026.
:::

### R implementation {.impl}

```{r}
# evaluate and plot the exact solution N(t) = N0 e^{rt}
N_exact <- function(t, r, N0) N0 * exp(r * t)   # exact solution, Equation (2)

r <- 0.1; N0 <- 1
t <- seq(0, 100, 1)
N <- N_exact(t, r, N0)

plot(t, N, xlab = "t", ylab = "N(t)", xlim = c(0, 100), ylim = c(0, 1000))
```

```{r}
# the same solution on a log-scaled y axis (a straight line)
plot(t, N, log = "y", xlab = "t", ylab = "N(t) (log scale)", xlim = c(0, 100))
```

```{r}
# integrate with our own Euler and RK4; zoom in where Euler's error shows
dN <- function(t, N, r) r * N            # derivative, Equation (1)

Euler <- function(derivs, X0, t.total, dt, ...) {
  t_all <- seq(0, t.total, by = dt); n <- length(t_all)
  X <- numeric(n); X[1] <- X0
  for (i in 1:(n - 1)) X[i + 1] <- X[i] + dt * derivs(t_all[i], X[i], ...)
  cbind(t_all, X)
}

RK4 <- function(derivs, X0, t.total, dt, ...) {
  t_all <- seq(0, t.total, by = dt); n <- length(t_all)
  X <- numeric(n); X[1] <- X0
  for (i in 1:(n - 1)) {
    h <- dt; t0 <- t_all[i]
    k1 <- h * derivs(t0,       X[i],          ...)
    k2 <- h * derivs(t0 + h/2, X[i] + k1 / 2, ...)
    k3 <- h * derivs(t0 + h/2, X[i] + k2 / 2, ...)
    k4 <- h * derivs(t0 + h,   X[i] + k3,     ...)
    X[i + 1] <- X[i] + (k1 + 2 * k2 + 2 * k3 + k4) / 6
  }
  cbind(t_all, X)
}

sol_euler <- Euler(dN, N0, 100, 0.1, r = r)
sol_rk4   <- RK4(dN, N0, 100, 0.1, r = r)

# zoom in on large t, where the Euler error is most visible
plot(t, N, xlab = "t", ylab = "N", xlim = c(66, 70), ylim = c(700, 1000), col = 4)
lines(sol_euler[, 1], sol_euler[, 2], col = 2)
lines(sol_rk4[, 1],   sol_rk4[, 2],   col = 3)
legend("topleft", inset = 0.02, legend = c("Euler dt=0.1", "RK4 dt=0.1", "Exact"),
       col = c(2, 3, 4), lty = c(1, 1, NA), pch = c(NA, NA, 1), cex = 0.8)
```

```{r}
#| message: false
# integrate with the deSolve library (fixed-step Euler)
library(deSolve)
dN_deSolve <- function(t, N, parms) list(parms["r"] * N)
sol_de <- ode(y = N0, times = seq(0, 100, 0.1), func = dN_deSolve,
              parms = c(r = 0.1), method = "euler")
tail(sol_de, 3)   # final points match the exact value N(100) = e^{10}
```

### Python implementation {.impl}

```{python}
# evaluate and plot the exact solution N(t) = N0 e^{rt}
import numpy as np
import matplotlib.pyplot as plt

def N_exact(t, r, N0):                # exact solution, Equation (2)
    return N0 * np.exp(r * t)

r, N0 = 0.1, 1
t = np.arange(0, 101, 1)
N = N_exact(t, r, N0)

plt.plot(t, N, "o", markersize=3)
plt.xlabel("t"); plt.ylabel("N(t)"); plt.xlim(0, 100); plt.ylim(0, 1000)
plt.show()
```

```{python}
# the same solution on a log-scaled y axis (a straight line)
plt.plot(t, N, "o", markersize=3)
plt.yscale("log")
plt.xlabel("t"); plt.ylabel("N(t) (log scale)"); plt.xlim(0, 100)
plt.show()
```

```{python}
# integrate with our own Euler and RK4; zoom in where Euler's error shows
def dN(t, N, r):                      # derivative, Equation (1)
    return r * N

def Euler(derivs, X0, t_total, dt, **kw):
    t_all = np.arange(0, t_total + dt, dt); n = len(t_all)
    X = np.zeros(n); X[0] = X0
    for i in range(1, n):
        X[i] = X[i - 1] + dt * derivs(t_all[i - 1], X[i - 1], **kw)
    return np.column_stack((t_all, X))

def RK4(derivs, X0, t_total, dt, **kw):
    t_all = np.arange(0, t_total + dt, dt); n = len(t_all)
    X = np.zeros(n); X[0] = X0
    for i in range(0, n - 1):
        h = dt; t0 = t_all[i]
        k1 = h * derivs(t0,         X[i],          **kw)
        k2 = h * derivs(t0 + h / 2, X[i] + k1 / 2, **kw)
        k3 = h * derivs(t0 + h / 2, X[i] + k2 / 2, **kw)
        k4 = h * derivs(t0 + h,     X[i] + k3,     **kw)
        X[i + 1] = X[i] + (k1 + 2 * k2 + 2 * k3 + k4) / 6
    return np.column_stack((t_all, X))

sol_euler = Euler(dN, N0, 100, 0.1, r=r)
sol_rk4   = RK4(dN, N0, 100, 0.1, r=r)

plt.plot(sol_euler[:, 0], sol_euler[:, 1], color="red",   label="Euler dt=0.1")
plt.plot(sol_rk4[:, 0],   sol_rk4[:, 1],   color="green", label="RK4 dt=0.1")
plt.plot(t, N, "o", color="blue", markersize=3, label="Exact")
plt.xlim(66, 70); plt.ylim(700, 1000)
plt.xlabel("t"); plt.ylabel("N")
plt.legend(loc="upper left", fontsize=8)
plt.show()
```

```{python}
# integrate with SciPy's solve_ivp (adaptive RK45)
from scipy.integrate import solve_ivp
sol = solve_ivp(lambda t, N: r * N, [0, 100], [N0], t_eval=np.arange(0, 101, 1))
sol.y[0][-3:]                          # final points ~ N(100) = e^{10}
```

::: {.callout-note title="Practice Q1"}
Re-run the exponential-growth simulation with other integration methods and step
sizes, for example `deSolve`'s `ode45`/`ode23` in R, or different `method=`
options of SciPy's `solve_ivp` in Python. For each, compare against the exact
solution and assess the accuracy. (Hint: `?deSolve` in R, `help(solve_ivp)` in
Python.) Edit this page's source (see the tip above) to add your code.
:::

## 2C.2 Logistic growth

The logistic model adds a limit to growth (the carrying capacity $B$):

\begin{equation}
\frac{dN}{dt} = rN\left(1 - \frac{N}{B}\right) , \qquad N(t=0) = N_0 , \tag{3}
\end{equation}

with closed-form solution [@sontag_systemsbio]

\begin{equation}
N(t) = \frac{N_0 B}{N_0 + (B - N_0)e^{-rt}} . \tag{4}
\end{equation}

The exact solution is plotted below (for $r = 0.1$, $B = 100$, $N_0 = 1$) to give
you a target to check your simulations against.

::: {.callout-note title="Recipe 2C.2.1"}
**Objective.** Plot the exact logistic-growth curve as a target to check simulations against.

**Model.** Logistic growth at rate r with carrying capacity B: `dN/dt = r*N*(1 - N/B)`, exact solution `N(t) = N0*B / (N0 + (B - N0)*exp(-r*t))`.

**Method.** Direct evaluation of the exact solution over time.

**Test.** r = 0.1, B = 100, N0 = 1; t = 0 to 100.

**Show.** A plot of the logistic curve N(t) rising to the carrying capacity B.

**Verification.** The curve rises sigmoidally from N0 and levels off at the carrying capacity `B = 100`.
:::

### R implementation {.impl}

```{r}
logistic_exact <- function(t, r, B, N0) N0 * B / (N0 + (B - N0) * exp(-r * t))

N0 <- 1; r <- 0.1; B <- 100
t <- seq(0, 100, 1)
N_log <- logistic_exact(t, r, B, N0)

plot(t, N_log, xlab = "t", ylab = "N(t)", xlim = c(0, 100), ylim = c(0, 110))
```

### Python implementation {.impl}

```{python}
def logistic_exact(t, r, B, N0):
    return N0 * B / (N0 + (B - N0) * np.exp(-r * t))

N0, r, B = 1, 0.1, 100
t = np.arange(0, 101, 1)
N_log = logistic_exact(t, r, B, N0)

plt.plot(t, N_log, "o", markersize=3)
plt.xlabel("t"); plt.ylabel("N(t)"); plt.xlim(0, 100); plt.ylim(0, 110)
plt.show()
```

::: {.callout-note title="Practice Q2"}
For the logistic model with $B = 100$, $N_0 = 1$, $r = 0.1$: simulate Equation
(3) with your `Euler`/`RK4` functions (or `deSolve`/`solve_ivp`). Adapt the
exponential-growth example. The `dN` derivative, `Euler`, `RK4`, and the library
solvers all carry over. Edit this page's source (see the tip above) to add your
code.

(a) Compare the simulation with the exact solution above. At what times $t$ are
the errors largest, and why?

(b) Now take the initial condition $N_0 = 200$ (above the carrying capacity).
Simulate $N(t)$ and describe what happens.
:::
