---
title: "2D. Stability and effective potential"
author: "Mingyang Lu"
old_id: "02D"
---

Having learned how to integrate ODEs, we now ask a more qualitative question:
given a gene circuit, what are its steady states, are they stable, and how many
are there? We answer this graphically and analytically, and then package the
answer into an *effective potential* — a landscape whose valleys are the stable
states.

## 2D.1 Stability of steady states

A steady state $X_s$ of $\frac{dX}{dt} = f(X)$ satisfies $f(X_s) = 0$. It is
**stable** if a small perturbation $X \to X + \Delta X$ decays back to $X_s$, and
**unstable** otherwise. Linearizing $f$ about $X_s$ gives the criterion

\begin{equation}
X_s \text{ is }
  \begin{cases}
    \text{stable}   & \dfrac{df(X)}{dX}\Big|_{X_s} < 0 \\[2mm]
    \text{unstable} & \dfrac{df(X)}{dX}\Big|_{X_s} \ge 0
  \end{cases} \tag{1}
\end{equation}

For the constitutively expressed gene from [Part 2A](02a-modeling-gene-circuits.qmd),
$f(X) = g - kX$, there is a single steady state $X_s = g/k$. Plotting the two
terms $g$ and $kX$ separately, their intersection is the steady state. A small
increase in $X$ raises $kX$ above $g$, so $\frac{dX}{dt} < 0$ and $X$ falls back —
the state is stable (consistent with $\frac{df}{dX} = -k < 0$).

The self-inhibiting gene, $f(X) = g_0 + g_1 H^{inh}(X) - kX$, likewise has a
single, stable steady state: $g(X)$ decreases monotonically, so it crosses the
line $kX$ exactly once, and $\frac{df}{dX} < 0$ there. We confirm this by
computing $\frac{df}{dX}$ numerically with a central finite difference.

### R implementation {.impl}

```{r}
# Constitutive gene: dX/dt = g - kX
g <- 50; k <- 0.1
X_all <- seq(400, 600, 1)
plot(X_all, rep(g, length(X_all)), type = "l", col = 1,
     xlab = "X (nM)", ylab = "Rates", xlim = c(400, 600), ylim = c(40, 60))
lines(X_all, k * X_all, col = 2)
legend("bottomright", inset = 0.02, legend = c("g", "kX"), col = 1:2, lty = 1, cex = 0.8)
```

```{r}
# Self-inhibiting gene: g(X) = g0 + g1 * H_inh(X)
hill_inh <- function(X, X_th, n) {   # inhibitory Hill function
  a <- (X / X_th)^n
  1 / (1 + a)
}

g0 <- 10; g1 <- 60; k <- 0.1; X_th <- 200; n <- 4
X_all <- seq(0, 600, 1)
plot(X_all, g0 + g1 * hill_inh(X_all, X_th, n), type = "l", col = 1,
     xlab = "X (nM)", ylab = "Rates", xlim = c(0, 600), ylim = c(0, 70))
lines(X_all, k * X_all, col = 2)
legend("top", inset = 0.02, legend = c("g(X)", "kX"), col = 1:2, lty = 1, cex = 0.8)
```

```{r}
# df/dX by a central finite difference (vectorized)
f_1g_self_inh <- function(X, g0, g1, X_th, n, k) g0 + g1 * hill_inh(X, X_th, n) - k * X

cal_dfdx <- function(X_all, dt, derivs, ...) {
  f_all   <- derivs(X_all, ...)
  f_plus  <- c(f_all[-1], NA)                # f(X + dX)
  f_minus <- c(NA, f_all[-length(f_all)])    # f(X - dX)
  (f_plus - f_minus) / (2 * dt)              # (f(X+dX) - f(X-dX)) / (2 dX)
}

X_all <- seq(0, 600, 0.1)
dfdX <- cal_dfdx(X_all, dt = 0.1, derivs = f_1g_self_inh,
                 g0 = 10, g1 = 60, X_th = 200, n = 4, k = 0.1)
plot(X_all, dfdX, type = "l", col = 1,
     xlab = "X (nM)", ylab = "df/dX", xlim = c(0, 600), ylim = c(-0.5, 0.1))
abline(h = 0, lty = 2, col = 2)
```

At the steady state (around 250 nM) $\frac{df}{dX} < 0$, so it is stable.

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

# Constitutive gene: dX/dt = g - kX
g, k = 50, 0.1
X_all = np.arange(400, 601, 1)
plt.plot(X_all, np.full_like(X_all, g, dtype=float), color="black", label="g")
plt.plot(X_all, k * X_all, color="red", label="kX")
plt.xlabel("X (nM)"); plt.ylabel("Rates"); plt.xlim(400, 600); plt.ylim(40, 60)
plt.legend(loc="lower right", fontsize=8); plt.show()
```

```{python}
def hill_inh(X, X_th, n):        # inhibitory Hill function
    a = (X / X_th)**n
    return 1 / (1 + a)

g0, g1, k, X_th, n = 10, 60, 0.1, 200, 4
X_all = np.arange(0, 601, 1)
plt.plot(X_all, g0 + g1 * hill_inh(X_all, X_th, n), color="black", label="g(X)")
plt.plot(X_all, k * X_all, color="red", label="kX")
plt.xlabel("X (nM)"); plt.ylabel("Rates"); plt.xlim(0, 600); plt.ylim(0, 70)
plt.legend(loc="upper right", fontsize=8); plt.show()
```

```{python}
def f_1g_self_inh(X, g0, g1, X_th, n, k):
    return g0 + g1 * hill_inh(X, X_th, n) - k * X

def cal_dfdx(X_all, dt, derivs, **kwargs):
    f_all   = derivs(X_all, **kwargs)
    f_plus  = np.concatenate([f_all[1:], [np.nan]])   # f(X + dX)
    f_minus = np.concatenate([[np.nan], f_all[:-1]])  # f(X - dX)
    return (f_plus - f_minus) / (2 * dt)

X_all = np.arange(0, 600.1, 0.1)
dfdX = cal_dfdx(X_all, 0.1, f_1g_self_inh, g0=10, g1=60, X_th=200, n=4, k=0.1)
plt.plot(X_all, dfdX, color="black", label="df/dX")
plt.axhline(0, linestyle="--", color="red")
plt.xlabel("X (nM)"); plt.ylabel("df/dX"); plt.xlim(0, 600); plt.ylim(-0.5, 0.1)
plt.show()
```

## 2D.2 Multi-stability

The self-activating gene, $f(X) = g_0 + g_1 H^{ex}(X) - kX$, behaves very
differently. Plotting $g(X)$ and $kX$ while varying $k$, the line $kX$ can cross
the sigmoidal $g(X)$ either once or **three** times. With $k = 0.15$ there are
three steady states: the outer two are stable and the middle one is unstable.
A system with more than one stable steady state is called **multistable** (here,
bistable) [@alon2019systems].

Intuitively: when $X$ is low there is too little protein to activate
transcription, so $X$ stays low; when $X$ is high it activates its own promoter,
so $X$ stays high. Simulating the ODE (with RK4) from different initial
conditions shows the trajectories converging to the two stable states, with the
unstable state acting as the watershed between their basins.

### R implementation {.impl}

```{r}
hill_ex <- function(X, X_th, n) {   # excitatory Hill function
  a <- (X / X_th)^n
  a / (1 + a)
}

g0 <- 10; g1 <- 45; X_th <- 200; n <- 4
X_all <- seq(0, 600, 1)
plot(X_all, g0 + g1 * hill_ex(X_all, X_th, n), type = "l", col = 1,
     xlab = "X (nM)", ylab = "Rates", xlim = c(0, 600), ylim = c(0, 60))
lines(X_all, 0.10 * X_all, col = 2)
lines(X_all, 0.15 * X_all, col = 3)
lines(X_all, 0.20 * X_all, col = 4)
legend("topleft", inset = 0.02,
       legend = c("g(X)", "kX (k=0.1)", "kX (k=0.15)", "kX (k=0.2)"),
       col = 1:4, lty = 1, cex = 0.8)
```

```{r}
#| message: false
library(deSolve)

f_1g_self_act <- function(X, g0, g1, X_th, n, k) g0 + g1 * hill_ex(X, X_th, n) - k * X

dy_self_act <- function(t, y, parms) {
  with(as.list(c(y, parms)), list(g0 + g1 * hill_ex(y, X_th, n) - k * y))
}

parms <- c(g0 = 10, g1 = 45, X_th = 200, n = 4, k = 0.15)
t_all <- seq(0, 80, 0.1)
X0_all <- seq(50, 400, 50)

plot(NULL, xlab = "t (Minute)", ylab = "X (nM)", xlim = c(0, 80), ylim = c(0, 500))
for (i in seq_along(X0_all)) {
  res <- ode(y = X0_all[i], times = t_all, func = dy_self_act, parms = parms, method = "rk4")
  lines(res[, 1], res[, 2], col = i + 1)
}
legend("bottomright", inset = 0.02, title = "Initial conditions",
       legend = paste0("X0 = ", X0_all), col = seq_along(X0_all) + 1, lty = 1, cex = 0.8)
```

```{r}
X_all <- seq(0, 500, 0.1)
dfdX <- cal_dfdx(X_all, dt = 0.1, derivs = f_1g_self_act,
                 g0 = 10, g1 = 45, X_th = 200, n = 4, k = 0.15)
plot(X_all, dfdX, type = "l", col = 1,
     xlab = "X (nM)", ylab = "df/dX", xlim = c(0, 500), ylim = c(-0.2, 0.2))
abline(h = 0, lty = 2, col = 2)
```

The $\frac{df}{dX}$ curve crosses zero three times; it is negative at the two
outer (stable) states and positive at the middle (unstable) one.

### Python implementation {.impl}

```{python}
def hill_ex(X, X_th, n):        # excitatory Hill function
    a = (X / X_th)**n
    return a / (1 + a)

g0, g1, X_th, n = 10, 45, 200, 4
X_all = np.arange(0, 601, 1)
plt.plot(X_all, g0 + g1 * hill_ex(X_all, X_th, n), color="black", label="g(X)")
plt.plot(X_all, 0.10 * X_all, color="red",   label="kX (k=0.1)")
plt.plot(X_all, 0.15 * X_all, color="green", label="kX (k=0.15)")
plt.plot(X_all, 0.20 * X_all, color="blue",  label="kX (k=0.2)")
plt.xlabel("X (nM)"); plt.ylabel("Rates"); plt.xlim(0, 600); plt.ylim(0, 60)
plt.legend(loc="upper left", fontsize=8); plt.show()
```

```{python}
from scipy.integrate import solve_ivp

def f_1g_self_act(X, g0, g1, X_th, n, k):
    return g0 + g1 * hill_ex(X, X_th, n) - k * X

parms = dict(g0=10, g1=45, X_th=200, n=4, k=0.15)
t_all = np.arange(0, 80.1, 0.1)
X0_all = np.arange(50, 401, 50)

for X0 in X0_all:
    res = solve_ivp(lambda t, y: [f_1g_self_act(y[0], **parms)], [0, 80], [X0],
                    t_eval=t_all, method="RK45", rtol=1e-6, atol=1e-8)
    plt.plot(res.t, res.y[0], label=f"X0 = {X0}")
plt.xlabel("t (Minute)"); plt.ylabel("X (nM)"); plt.xlim(0, 80); plt.ylim(0, 500)
plt.legend(title="Initial conditions", loc="lower right", fontsize=8); plt.show()
```

```{python}
X_all = np.arange(0, 500.1, 0.1)
dfdX = cal_dfdx(X_all, 0.1, f_1g_self_act, g0=10, g1=45, X_th=200, n=4, k=0.15)
plt.plot(X_all, dfdX, color="black", label="df/dX")
plt.axhline(0, linestyle="--", color="red")
plt.xlabel("X (nM)"); plt.ylabel("df/dX"); plt.xlim(0, 500); plt.ylim(-0.2, 0.2)
plt.show()
```

## 2D.3 Effective potential

A gene circuit is a non-equilibrium system, so energy is not strictly defined.
Still, for a one-variable system we can make an analogy with an overdamped
particle whose position is $X$ and whose velocity $\frac{dX}{dt} = f(X)$ is
proportional to the force $-\,U'(X)$. This defines an **effective potential**

$$U(X) = -\int_{X_0}^{X} f(x)\,dx + U(X_0) ,$$

a one-dimensional version of Waddington's epigenetic landscape
[@waddington1957strategy], with stable states at the valleys and unstable states
at the peaks. We set $U(0) = 0$ and integrate with the trapezoidal rule
[@press2007numerical]:

$$U(X + \Delta x) = U(X) - \frac{f(X) + f(X + \Delta x)}{2}\,\Delta x .$$

We implement this first with an explicit loop, then vectorized with a cumulative
sum, and benchmark the two. Finally we show how the landscape changes with $k$:
at $k = 0.15$ there are two basins (bistability); the lower-lying basin is the
more stable state, and the barrier height indicates how hard a noise-driven
transition between states would be (see Parts 4 and 6). At $k = 0.1$ or $k = 0.2$
a single basin remains (monostability).

### R implementation {.impl}

```{r}
# Trapezoidal integration with an explicit loop
cal_int <- function(Xmin, Xmax, dX, f, ...) {
  X_all <- seq(Xmin, Xmax, dX)
  nX <- length(X_all)
  U_all <- numeric(nX)
  for (i in 1:(nX - 1)) {
    X <- X_all[i]
    U_all[i + 1] <- U_all[i] - (f(X, ...) + f(X + dX, ...)) / 2 * dX
  }
  cbind(X_all, U_all)
}

res1 <- cal_int(Xmin = 0, Xmax = 500, dX = 0.1, f = f_1g_self_act,
                g0 = 10, g1 = 45, X_th = 200, n = 4, k = 0.15)
plot(res1[, 1], res1[, 2], type = "l", col = 1,
     xlab = "X (nM)", ylab = "Effective potential", xlim = c(0, 500))
```

```{r}
# Same integration, vectorized with a cumulative sum
cal_int_vector <- function(Xmin, Xmax, dX, f, ...) {
  X_all <- seq(Xmin, Xmax, dX)
  f_all <- f(X_all, ...)
  f_plus <- c(f_all[-1], NA)                 # f(X + dX)
  trap <- (f_all + f_plus) * dX / 2          # trapezoid areas
  U_all <- -cumsum(trap)
  cbind(X_all, U_all)
}

res2 <- cal_int_vector(Xmin = 0, Xmax = 500, dX = 0.1, f = f_1g_self_act,
                       g0 = 10, g1 = 45, X_th = 200, n = 4, k = 0.15)
plot(res2[, 1], res2[, 2], type = "l", col = 1,
     xlab = "X (nM)", ylab = "Effective potential", xlim = c(0, 500))
```

```{r}
#| message: false
#| warning: false
library(microbenchmark)
library(ggplot2)
mb <- microbenchmark(
  loop   = cal_int(0, 500, 0.1, f_1g_self_act, g0 = 10, g1 = 45, X_th = 200, n = 4, k = 0.15),
  vector = cal_int_vector(0, 500, 0.1, f_1g_self_act, g0 = 10, g1 = 45, X_th = 200, n = 4, k = 0.15),
  times = 10, unit = "ms")
autoplot(mb)
```

```{r}
# Landscapes for k = 0.2 (monostable) and k = 0.1 (monostable)
res3 <- cal_int_vector(0, 500, 0.1, f_1g_self_act, g0 = 10, g1 = 45, X_th = 200, n = 4, k = 0.2)
res4 <- cal_int_vector(0, 500, 0.1, f_1g_self_act, g0 = 10, g1 = 45, X_th = 200, n = 4, k = 0.1)
plot(res3[, 1], res3[, 2], type = "l", col = 2,
     xlab = "X (nM)", ylab = "Effective potential", xlim = c(0, 500),
     ylim = range(res3[, 2], res4[, 2], na.rm = TRUE))   # fit both curves
lines(res4[, 1], res4[, 2], col = 4)
legend("topright", inset = 0.02, legend = c("k = 0.2", "k = 0.1"), col = c(2, 4), lty = 1, cex = 0.8)
```

The vectorized version is both shorter and faster.

### Python implementation {.impl}

```{python}
def cal_int(Xmin, Xmax, dX, f, **kwargs):
    X_all = np.arange(Xmin, Xmax + dX, dX)
    U_all = np.zeros(len(X_all))
    for i in range(len(X_all) - 1):
        X = X_all[i]
        U_all[i + 1] = U_all[i] - (f(X, **kwargs) + f(X + dX, **kwargs)) / 2 * dX
    return np.column_stack((X_all, U_all))

res1 = cal_int(0, 500, 0.1, f_1g_self_act, g0=10, g1=45, X_th=200, n=4, k=0.15)
plt.plot(res1[:, 0], res1[:, 1], color="black")
plt.xlabel("X (nM)"); plt.ylabel("Effective potential"); plt.xlim(0, 500); plt.show()
```

```{python}
def cal_int_vector(Xmin, Xmax, dX, f, **kwargs):
    X_all = np.arange(Xmin, Xmax + dX, dX)
    f_all = f(X_all, **kwargs)
    f_plus = np.concatenate([f_all[1:], [np.nan]])   # f(X + dX)
    trap = (f_all + f_plus) * dX / 2
    U_all = -np.cumsum(trap)
    return np.column_stack((X_all, U_all))

res2 = cal_int_vector(0, 500, 0.1, f_1g_self_act, g0=10, g1=45, X_th=200, n=4, k=0.15)
plt.plot(res2[:, 0], res2[:, 1], color="black")
plt.xlabel("X (nM)"); plt.ylabel("Effective potential"); plt.xlim(0, 500); plt.show()
```

```{python}
import timeit
kw = dict(g0=10, g1=45, X_th=200, n=4, k=0.15)
n_rep = 10
t_loop = timeit.timeit(lambda: cal_int(0, 500, 0.1, f_1g_self_act, **kw), number=n_rep) / n_rep
t_vec  = timeit.timeit(lambda: cal_int_vector(0, 500, 0.1, f_1g_self_act, **kw), number=n_rep) / n_rep
print(f"loop      : {t_loop*1e3:8.2f} ms")
print(f"vectorized: {t_vec*1e3:8.2f} ms")
```

```{python}
res3 = cal_int_vector(0, 500, 0.1, f_1g_self_act, g0=10, g1=45, X_th=200, n=4, k=0.2)
res4 = cal_int_vector(0, 500, 0.1, f_1g_self_act, g0=10, g1=45, X_th=200, n=4, k=0.1)
plt.plot(res3[:, 0], res3[:, 1], color="red",  label="k = 0.2")
plt.plot(res4[:, 0], res4[:, 1], color="blue", label="k = 0.1")
plt.xlabel("X (nM)"); plt.ylabel("Effective potential"); plt.xlim(0, 500)
plt.legend(); plt.show()
```
