---
title: "2E. Bifurcation"
author: "Mingyang Lu"
old_id: "02E"
---

We continue with the single self-activating gene from
[Part 2A](02a-modeling-gene-circuits.qmd) (Equations (7) and (8)). As we saw in
[Part 2D](02d-effective-potential.qmd), varying the degradation rate $k$ changes
the number of steady states: the circuit is monostable for some $k$ and bistable
(two stable states plus one unstable state) for others. To see the whole picture
at once, we treat $k$ as a *control parameter* and plot the steady-state level
$X_s$ against $k$. This curve is a one-dimensional **bifurcation diagram**. We use
the rate equation

\begin{equation}
\frac{dX}{dt} = f(X,k) = 10 + 45\frac{X^4}{X^4 + 200^4} - kX , \tag{1}
\end{equation}

with state variable $X$ and control parameter $k$. Every point on the diagram is
a steady state, so it satisfies $f(X, k) = 0$. This chapter works through several
numerical strategies for tracing that curve, from a special-case shortcut to
general-purpose continuation. Some are robust; others are homemade and work well
only for this saddle-node example, as noted where they appear.

## 2E.1 Separation of variables

For this particular $f$, the steady-state condition $f(X, k) = 0$ can be solved
for $k$ explicitly as a function of $X$:

$$ k = \frac{10}{X} + 45\frac{X^3}{X^4 + 200^4} . $$

Sweeping $X$ over a range (avoiding $X = 0$) and evaluating $k$ traces the curve
directly, with no root-finding needed.

### R implementation {.impl}

```{r}
func_k <- function(X) 10 / X + 45 * X^3 / (X^4 + 200^4)

X_all <- seq(50, 600, by = 5)   # X grid; avoid X = 0
k_all <- func_k(X_all)

plot(k_all, X_all, xlab = "k", ylab = "X", type = "p", col = 2,
     xlim = c(0.09, 0.21), ylim = c(0, 600))
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

def func_k(X):
    return 10 / X + 45 * X**3 / (X**4 + 200**4)

X_all = np.arange(50, 605, 5)   # X grid; avoid X = 0
k_all = func_k(X_all)

plt.scatter(k_all, X_all, facecolors="none", edgecolors="red", s=20)
plt.xlabel("k"); plt.ylabel("X")
plt.xlim(0.09, 0.21); plt.ylim(0, 600)
plt.show()
```

The resulting curve is S-shaped and shows how the circuit's behavior changes with
$k$. It has two turning points, near $k = 0.125$ and $k = 0.170$, where a stable
steady state and the unstable steady state meet and annihilate; these are the
**bifurcation points**. For $k$ below the first point the system is monostable at
a relatively high $X$; between the two points it is bistable, with a high and a
low stable state (and an unstable one in between); above the second point it is
monostable again at a low $X$. Because the branches meet and vanish at the turning
points, this is a *saddle-node* bifurcation [@strogatz2015nonlinear].

This method is exact and cheap, but it depends on being able to solve
$f(X, k) = 0$ for $k$ in closed form. The remaining methods do not, so they also
apply to circuits where no such shortcut exists.

## 2E.2 Direct method: find all roots

The most direct general approach samples $k$ uniformly and, at each $k$, finds
every root $X_s$ of $f(X, k) = 0$. We classify each root from the sign of
$\frac{df}{dX}$ ([Part 2D](02d-effective-potential.qmd)), stable when it is
negative and unstable otherwise,
estimated here with a small finite step $dx$. This is the most robust method in
the chapter and works well for our example, though for a general nonlinear $f$
there is no guarantee of finding every root, and it is slower than the methods
that follow.

The roots come out as a scatter of $(k, X_s)$ points that are not ordered along
the curve, so joining them in the order found would give a tangled line. We
reorder them with a nearest-neighbor walk: start at the leftmost point, then
repeatedly step to the closest unused point until all points are used. (In the
inverse-distance matrix used below, the nearest point is the one with the largest
score.)

### R implementation {.impl}

`rootSolve::uniroot.all` finds all roots of a function on an interval in one call
[@soetaert2009rootsolve].

```{r}
#| message: false
# For each sampled k, find all roots of f(X, k) = 0 and label each by stability
library(rootSolve)

# Model derivative from Equation (1); reused by the methods below
derivs_k <- function(X, k) 10 + 45 * X^4 / (X^4 + 200^4) - k * X

dx <- 0.1                          # small step to estimate df/dX for stability
k_all <- seq(0.1, 0.2, by = 0.001)
results <- matrix(0, nrow = length(k_all) * 3, ncol = 3)  # columns: k, Xs, stability

ind <- 0
for (k in k_all) {
  roots <- uniroot.all(derivs_k, k = k, interval = c(0, 800))
  for (x in roots) {
    ind <- ind + 1
    stability <- if (sign(derivs_k(X = x + dx, k = k)) < 0) 1 else 2  # 1 stable, 2 unstable
    results[ind, ] <- c(k, x, stability)
  }
}
results <- results[1:ind, ]

plot(results[, 1], results[, 2], pch = 1, cex = 0.3, col = results[, 3],
     xlab = expression(k ~ (Minute^{-1})), ylab = "Xs (nM)",
     xlim = c(0.09, 0.21), ylim = c(0, 600))
legend("topright", inset = 0.02, legend = c("Stable", "Unstable"),
       col = 1:2, pch = 1, cex = 0.8)
```

```{r}
# joining the points in the order found: not yet ordered along the curve
plot(results[, 1], results[, 2], type = "l", col = 2,
     xlab = expression(k ~ (Minute^{-1})), ylab = "Xs (nM)",
     xlim = c(0.09, 0.21), ylim = c(0, 600))
```

```{r}
#| warning: false
# reorder the scattered roots into a connected curve (nearest-neighbor walk)
d <- as.matrix(dist(results[, 1:2]))   # pairwise distances between points
m <- 1 / d                             # inverse distance; nearest point scores highest
num_total <- nrow(results)
sequence <- integer(num_total)
if_unused <- rep(1, num_total)         # 1 unused, 0 used

point_current <- which.min(results[, 1])   # start at the smallest-k point
sequence[1] <- point_current
if_unused[point_current] <- 0
for (i in 2:num_total) {
  point_current <- which.max(m[, point_current] * if_unused)  # closest unused point
  sequence[i] <- point_current
  if_unused[point_current] <- 0
}

plot(results[sequence, 1], results[sequence, 2], type = "l", col = 2,
     xlab = expression(k ~ (Minute^{-1})), ylab = "Xs (nM)",
     xlim = c(0.09, 0.21), ylim = c(0, 600))
```

### Python implementation {.impl}

`scipy.optimize.fsolve` returns one root per initial guess, so we pass a vector of
guesses and then discard spurious roots (where $|f|$ is not near zero) and any
outside the range. Several guesses can converge to the same root, so we also drop
duplicates; `uniroot.all` in R does this in one step.

```{python}
#| warning: false
# For each sampled k, find all roots of f(X, k) = 0 and label each by stability
import warnings
warnings.filterwarnings("ignore")   # suppress fsolve/numpy runtime warnings
from scipy.optimize import fsolve

def derivs_k(X, k):
    return 10 + 45 * X**4 / (X**4 + 200**4) - k * X

dx = 0.1
k_all = np.arange(0.1, 0.201, 0.001)
results = np.zeros((len(k_all) * 10, 3))    # columns: k, Xs, stability

ind = 0
for k in k_all:
    roots = fsolve(derivs_k, x0=np.arange(50, 500, 50), args=(k,))
    for x in roots:
        if abs(derivs_k(x, k)) > 0.01:      # fsolve can return non-roots; keep true roots only
            continue
        if 0 <= x <= 800:
            stability = 1 if np.sign(derivs_k(x + dx, k)) < 0 else 2
            results[ind] = [k, x, stability]
            ind += 1
results = results[:ind]

# several initial guesses may land on the same root; keep unique (k, Xs)
_, uniq = np.unique(np.round(results[:, :2], 2), axis=0, return_index=True)
results = results[np.sort(uniq)]

stable = results[results[:, 2] == 1]
unstable = results[results[:, 2] == 2]
plt.scatter(stable[:, 0], stable[:, 1], s=10, color="black", label="Stable")
plt.scatter(unstable[:, 0], unstable[:, 1], s=10, color="red", label="Unstable")
plt.xlabel(r"$k$ (Minute$^{-1}$)"); plt.ylabel(r"$X_s$ (nM)")
plt.xlim(0.09, 0.21); plt.ylim(0, 600)
plt.legend(loc="upper right", fontsize=8); plt.show()
```

```{python}
# joining the points in the order found: not yet ordered along the curve
plt.plot(results[:, 0], results[:, 1], color="red", linewidth=1)
plt.xlabel(r"$k$ (Minute$^{-1}$)"); plt.ylabel(r"$X_s$ (nM)")
plt.xlim(0.09, 0.21); plt.ylim(0, 600); plt.show()
```

```{python}
#| warning: false
# reorder the scattered roots into a connected curve (nearest-neighbor walk)
from scipy.spatial.distance import cdist

d = cdist(results[:, :2], results[:, :2])   # pairwise distances
m = 1 / d                                    # inverse distance; nearest point scores highest
np.fill_diagonal(m, 0)                        # avoid divide-by-zero on the diagonal

num_total = len(results)
sequence = np.zeros(num_total, dtype=int)
if_unused = np.ones(num_total, dtype=bool)

point_current = int(np.argmin(results[:, 0]))   # start at the smallest-k point
sequence[0] = point_current
if_unused[point_current] = False
for i in range(1, num_total):
    point_current = int(np.argmax(m[:, point_current] * if_unused))
    sequence[i] = point_current
    if_unused[point_current] = False

plt.plot(results[sequence, 0], results[sequence, 1], color="red", linewidth=1)
plt.xlabel(r"$k$ (Minute$^{-1}$)"); plt.ylabel(r"$X_s$ (nM)")
plt.xlim(0.09, 0.21); plt.ylim(0, 600); plt.show()
```

The scatter shows both stable branches (high and low $X$) and the unstable middle
branch, colored by stability. After reordering, the points connect into the same
smooth S-curve found by separation of variables, now obtained without an explicit
formula for $k(X)$.
