---
title: "2E. Bifurcation"
author: "Mingyang Lu"
old_id: "02E"
---

We continue with the single self-activating gene from
[Part 2A](02a-modeling-gene-circuits.qmd) (Equations (7) and (8)). As we saw in
[Part 2D](02d-effective-potential.qmd), varying the degradation rate $k$ changes
the number of steady states: the circuit is monostable for some $k$ and bistable
(two stable states plus one unstable state) for others. To see the whole picture
at once, we treat $k$ as a *control parameter* and plot the steady-state level
$X_s$ against $k$. This curve is a one-dimensional **bifurcation diagram**. We use
the rate equation

\begin{equation}
\frac{dX}{dt} = f(X,k) = 10 + 45\frac{X^4}{X^4 + 200^4} - kX , \tag{1}
\end{equation}

with state variable $X$ and control parameter $k$. Every point on the diagram is
a steady state, so it satisfies $f(X, k) = 0$. This chapter works through several
numerical strategies for tracing that curve, from a special-case shortcut to
general-purpose continuation. Some are robust; others are homemade and work well
only for this saddle-node example, as noted where they appear.

## 2E.1 Separation of variables

For this particular $f$, the steady-state condition $f(X, k) = 0$ can be solved
for $k$ explicitly as a function of $X$:

$$ k = \frac{10}{X} + 45\frac{X^3}{X^4 + 200^4} . $$

Sweeping $X$ over a range (avoiding $X = 0$) and evaluating $k$ traces the curve
directly, with no root-finding needed.

### R implementation {.impl}

```{r}
func_k <- function(X) 10 / X + 45 * X^3 / (X^4 + 200^4)

X_all <- seq(50, 600, by = 5)   # X grid; avoid X = 0
k_all <- func_k(X_all)

plot(k_all, X_all, xlab = "k", ylab = "X", type = "p", col = 2,
     xlim = c(0.09, 0.21), ylim = c(0, 600))
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

def func_k(X):
    return 10 / X + 45 * X**3 / (X**4 + 200**4)

X_all = np.arange(50, 605, 5)   # X grid; avoid X = 0
k_all = func_k(X_all)

plt.scatter(k_all, X_all, facecolors="none", edgecolors="red", s=20)
plt.xlabel("k"); plt.ylabel("X")
plt.xlim(0.09, 0.21); plt.ylim(0, 600)
plt.show()
```

The resulting curve is S-shaped and shows how the circuit's behavior changes with
$k$. It has two turning points, near $k = 0.125$ and $k = 0.170$, where a stable
steady state and the unstable steady state meet and annihilate; these are the
**bifurcation points**. For $k$ below the first point the system is monostable at
a relatively high $X$; between the two points it is bistable, with a high and a
low stable state (and an unstable one in between); above the second point it is
monostable again at a low $X$. Because the branches meet and vanish at the turning
points, this is a *saddle-node* bifurcation [@strogatz2015nonlinear].

This method is exact and cheap, but it depends on being able to solve
$f(X, k) = 0$ for $k$ in closed form. The remaining methods do not, so they also
apply to circuits where no such shortcut exists.

## 2E.2 Direct method: find all roots

The most direct general approach samples $k$ uniformly and, at each $k$, finds
every root $X_s$ of $f(X, k) = 0$. We classify each root from the sign of
$\frac{df}{dX}$ ([Part 2D](02d-effective-potential.qmd)), stable when it is
negative and unstable otherwise,
estimated here with a small finite step $dx$. This is the most robust method in
the chapter and works well for our example, though for a general nonlinear $f$
there is no guarantee of finding every root, and it is slower than the methods
that follow.

The roots come out as a scatter of $(k, X_s)$ points that are not ordered along
the curve, so joining them in the order found would give a tangled line. We
reorder them with a nearest-neighbor walk: start at the leftmost point, then
repeatedly step to the closest unused point until all points are used. (In the
inverse-distance matrix used below, the nearest point is the one with the largest
score.)

### R implementation {.impl}

`rootSolve::uniroot.all` finds all roots of a function on an interval in one call
[@soetaert2009rootsolve].

```{r}
#| message: false
# For each sampled k, find all roots of f(X, k) = 0 and label each by stability
library(rootSolve)

# Model derivative from Equation (1); reused by the methods below
derivs_k <- function(X, k) 10 + 45 * X^4 / (X^4 + 200^4) - k * X

dx <- 0.1                          # small step to estimate df/dX for stability
k_all <- seq(0.1, 0.2, by = 0.001)
results <- matrix(0, nrow = length(k_all) * 3, ncol = 3)  # columns: k, Xs, stability

ind <- 0
for (k in k_all) {
  roots <- uniroot.all(derivs_k, k = k, interval = c(0, 800))
  for (x in roots) {
    ind <- ind + 1
    stability <- if (sign(derivs_k(X = x + dx, k = k)) < 0) 1 else 2  # 1 stable, 2 unstable
    results[ind, ] <- c(k, x, stability)
  }
}
results <- results[1:ind, ]

plot(results[, 1], results[, 2], pch = 1, cex = 0.3, col = results[, 3],
     xlab = expression(k ~ (Minute^{-1})), ylab = "Xs (nM)",
     xlim = c(0.09, 0.21), ylim = c(0, 600))
legend("topright", inset = 0.02, legend = c("Stable", "Unstable"),
       col = 1:2, pch = 1, cex = 0.8)
```

```{r}
# joining the points in the order found: not yet ordered along the curve
plot(results[, 1], results[, 2], type = "l", col = 2,
     xlab = expression(k ~ (Minute^{-1})), ylab = "Xs (nM)",
     xlim = c(0.09, 0.21), ylim = c(0, 600))
```

```{r}
#| warning: false
# reorder the scattered roots into a connected curve (nearest-neighbor walk)
d <- as.matrix(dist(results[, 1:2]))   # pairwise distances between points
m <- 1 / d                             # inverse distance; nearest point scores highest
num_total <- nrow(results)
sequence <- integer(num_total)
if_unused <- rep(1, num_total)         # 1 unused, 0 used

point_current <- which.min(results[, 1])   # start at the smallest-k point
sequence[1] <- point_current
if_unused[point_current] <- 0
for (i in 2:num_total) {
  point_current <- which.max(m[, point_current] * if_unused)  # closest unused point
  sequence[i] <- point_current
  if_unused[point_current] <- 0
}

plot(results[sequence, 1], results[sequence, 2], type = "l", col = 2,
     xlab = expression(k ~ (Minute^{-1})), ylab = "Xs (nM)",
     xlim = c(0.09, 0.21), ylim = c(0, 600))
```

### Python implementation {.impl}

`scipy.optimize.fsolve` returns one root per initial guess, so we pass a vector of
guesses and then discard spurious roots (where $|f|$ is not near zero) and any
outside the range. Several guesses can converge to the same root, so we also drop
duplicates; `uniroot.all` in R does this in one step.

```{python}
#| warning: false
# For each sampled k, find all roots of f(X, k) = 0 and label each by stability
import warnings
warnings.filterwarnings("ignore")   # suppress fsolve/numpy runtime warnings
from scipy.optimize import fsolve

def derivs_k(X, k):
    return 10 + 45 * X**4 / (X**4 + 200**4) - k * X

dx = 0.1
k_all = np.arange(0.1, 0.201, 0.001)
results = np.zeros((len(k_all) * 10, 3))    # columns: k, Xs, stability

ind = 0
for k in k_all:
    roots = fsolve(derivs_k, x0=np.arange(50, 500, 50), args=(k,))
    for x in roots:
        if abs(derivs_k(x, k)) > 0.01:      # fsolve can return non-roots; keep true roots only
            continue
        if 0 <= x <= 800:
            stability = 1 if np.sign(derivs_k(x + dx, k)) < 0 else 2
            results[ind] = [k, x, stability]
            ind += 1
results = results[:ind]

# several initial guesses may land on the same root; keep unique (k, Xs)
_, uniq = np.unique(np.round(results[:, :2], 2), axis=0, return_index=True)
results = results[np.sort(uniq)]

stable = results[results[:, 2] == 1]
unstable = results[results[:, 2] == 2]
plt.scatter(stable[:, 0], stable[:, 1], s=10, color="black", label="Stable")
plt.scatter(unstable[:, 0], unstable[:, 1], s=10, color="red", label="Unstable")
plt.xlabel(r"$k$ (Minute$^{-1}$)"); plt.ylabel(r"$X_s$ (nM)")
plt.xlim(0.09, 0.21); plt.ylim(0, 600)
plt.legend(loc="upper right", fontsize=8); plt.show()
```

```{python}
# joining the points in the order found: not yet ordered along the curve
plt.plot(results[:, 0], results[:, 1], color="red", linewidth=1)
plt.xlabel(r"$k$ (Minute$^{-1}$)"); plt.ylabel(r"$X_s$ (nM)")
plt.xlim(0.09, 0.21); plt.ylim(0, 600); plt.show()
```

```{python}
#| warning: false
# reorder the scattered roots into a connected curve (nearest-neighbor walk)
from scipy.spatial.distance import cdist

d = cdist(results[:, :2], results[:, :2])   # pairwise distances
m = 1 / d                                    # inverse distance; nearest point scores highest
np.fill_diagonal(m, 0)                        # avoid divide-by-zero on the diagonal

num_total = len(results)
sequence = np.zeros(num_total, dtype=int)
if_unused = np.ones(num_total, dtype=bool)

point_current = int(np.argmin(results[:, 0]))   # start at the smallest-k point
sequence[0] = point_current
if_unused[point_current] = False
for i in range(1, num_total):
    point_current = int(np.argmax(m[:, point_current] * if_unused))
    sequence[i] = point_current
    if_unused[point_current] = False

plt.plot(results[sequence, 0], results[sequence, 1], color="red", linewidth=1)
plt.xlabel(r"$k$ (Minute$^{-1}$)"); plt.ylabel(r"$X_s$ (nM)")
plt.xlim(0.09, 0.21); plt.ylim(0, 600); plt.show()
```

The scatter shows both stable branches (high and low $X$) and the unstable middle
branch, colored by stability. After reordering, the points connect into the same
smooth S-curve found by separation of variables, now obtained without an explicit
formula for $k(X)$.

## 2E.3 Bisection method

The library root-finder can be replaced by a homemade one. The **bisection
method** [@press2007numerical] brackets a root: if $f$ has opposite signs at the
two ends of an interval, a root must lie between them, and repeatedly halving the
interval (keeping the half across which the sign still changes) converges to it.
Bisection is simple and robust, but it has two limitations. It needs a sign
change, so it misses roots where $f$ only touches zero, and it returns nothing
when the two endpoints share a sign even if roots exist between them. It also
finds only one root per interval. Both effects show up when we test it on the
whole range $[0, 600]$: at $k = 0.12$ and $k = 0.2$ (monostable) it returns the
single root, while at $k = 0.15$ (bistable) it returns just one of the three.

To find every root, we bracket: scan the range in many small windows and apply
bisection to each window that contains a sign change (the `all_roots_bracketing`
driver, with `resolution` setting the window width). Applied across $k$, this
reproduces the same scattered $(k, X_s)$ points as in 2E.2. Reordering them into
the curve is the identical nearest-neighbor walk from 2E.2 and is omitted here.
Both implementations reuse the model derivative `derivs_k` defined in 2E.2.

### R implementation {.impl}

```{r}
# bisection: find one root of f where it changes sign across the interval
bisection <- function(f, interval, epsilon = 1e-6, ...) {
  x_min <- interval[1]; f1 <- f(x_min, ...)
  x_max <- interval[2]; f2 <- f(x_max, ...)
  if (abs(f1) < epsilon) return(x_min)
  if (abs(f2) < epsilon) return(x_max)
  if (f1 * f2 > 0) return(NA)                     # same sign: no bracketed root
  while (f1 * f2 < 0) {
    x_mid <- (x_min + x_max) / 2
    f3 <- f(x_mid, ...)
    if (abs(f3) < epsilon) return(x_mid)
    if (f1 * f3 > 0) { f1 <- f3; x_min <- x_mid }  # sign change is in the upper half
    else             { f2 <- f3; x_max <- x_mid }  # sign change is in the lower half
  }
}

for (k in c(0.12, 0.15, 0.2)) {
  root <- bisection(function(X) derivs_k(X, k), interval = c(0, 600))
  cat(sprintf("k = %.2f: root = %.2f\n", k, root))
}
```

```{r}
# all roots by bracketing: scan the range in small windows and bisect each
all_roots_bracketing <- function(f, method, interval, resolution = 1e-2, ...) {
  x_all <- seq(interval[1], interval[2], by = resolution * (interval[2] - interval[1]))
  roots <- numeric(0)
  for (i in seq_len(length(x_all) - 1)) {
    root <- method(f, c(x_all[i], x_all[i + 1]), ...)
    if (!is.na(root)) roots <- c(roots, root)
  }
  if (length(roots) == 0) NA else unique(roots)
}
```

```{r}
# trace the curve: at each k, find every root by bracketed bisection
k_all <- seq(0.1, 0.2, by = 0.001)
results <- matrix(0, ncol = 2, nrow = length(k_all) * 3)   # columns: k, Xs
ind <- 1
for (k in k_all) {
  roots <- all_roots_bracketing(function(X) derivs_k(X, k), bisection,
                                interval = c(0, 600), resolution = 1e-2)
  num_X <- length(roots)
  results[ind:(ind + num_X - 1), 1] <- k
  results[ind:(ind + num_X - 1), 2] <- roots
  ind <- ind + num_X
}
results <- results[1:(ind - 1), ]

plot(results[, 1], results[, 2], pch = 1, cex = 0.3, col = 2,
     xlab = expression(k ~ (Minute^{-1})), ylab = "Xs (nM)",
     xlim = c(0.09, 0.21), ylim = c(0, 600))
```

### Python implementation {.impl}

```{python}
# bisection: find one root of f where it changes sign across the interval
def bisection(f, interval, epsilon=1e-6, **kwargs):
    x_min, x_max = interval
    f1 = f(x_min, **kwargs); f2 = f(x_max, **kwargs)
    if abs(f1) < epsilon: return x_min
    if abs(f2) < epsilon: return x_max
    if f1 * f2 > 0: return None                    # same sign: no bracketed root
    while f1 * f2 < 0:
        x_mid = (x_min + x_max) / 2
        f3 = f(x_mid, **kwargs)
        if abs(f3) < epsilon: return x_mid
        if f1 * f3 > 0: f1, x_min = f3, x_mid       # sign change is in the upper half
        else:           f2, x_max = f3, x_mid       # sign change is in the lower half
    return None

for k in (0.12, 0.15, 0.2):
    root = bisection(lambda X: derivs_k(X, k), (0, 600))
    print(f"k = {k}: root = {root:.2f}")
```

```{python}
# all roots by bracketing: scan the range in small windows and bisect each
def all_roots_bracketing(f, method, interval, resolution=1e-2, **kwargs):
    x_all = np.linspace(interval[0], interval[1], int(1 / resolution) + 1)
    roots = []
    for i in range(len(x_all) - 1):
        root = method(f, (x_all[i], x_all[i + 1]), **kwargs)
        if root is not None:
            roots.append(root)
    return np.unique(roots).tolist() if roots else []
```

```{python}
# trace the curve: at each k, find every root by bracketed bisection
k_all = np.arange(0.1, 0.201, 0.001)
results = np.zeros((len(k_all) * 3, 2))    # columns: k, Xs
ind = 0
for k in k_all:
    roots = all_roots_bracketing(lambda X: derivs_k(X, k), bisection,
                                 (0, 600), resolution=1e-2)
    n = len(roots)
    results[ind:ind + n, 0] = k
    results[ind:ind + n, 1] = roots
    ind += n
results = results[:ind]

plt.scatter(results[:, 0], results[:, 1], color="red", s=3)
plt.xlabel(r"$k$ (Minute$^{-1}$)"); plt.ylabel(r"$X_s$ (nM)")
plt.xlim(0.09, 0.21); plt.ylim(0, 600); plt.show()
```

The three-value test confirms the single-root limitation: at $k = 0.15$ bisection
on $[0, 600]$ returns only one of the three steady states. The bracketed driver
recovers all of them, giving the same scatter (and, after the 2E.2 reordering,
the same S-curve).
