---
title: "2F. Bifurcation: finding curves"
author: "Mingyang Lu"
old_id: "02E"
---

In [Part 2E](02e-bifurcation.qmd) we built the bifurcation diagram by *sampling*:
at each value of $k$ we located the steady states (the roots of $f(X, k) = 0$, or
the states an ODE relaxes onto), producing a cloud of points that we then
connected into a curve. That works, but it treats the diagram as a set of
independent points and depends on a post-hoc reordering step to make a curve out
of them.

This chapter takes the complementary view: the bifurcation set
$\{(k, X) : f(X, k) = 0\}$ is a *curve* in the $(k, X)$ plane, and we trace it
directly. Two families of methods do this. The **contour method** treats
$f(k, X)$ as a surface and extracts its zero-level contour, interpolating within a
grid to place the curve accurately. **Numerical continuation** starts from one
point on the curve and walks along it with a predictor–corrector step. These are
the general-purpose tools of bifurcation analysis, and they return the curve
already ordered.

## 2F.1 The model derivative

We use the same self-activating-gene model as in [Part 2E](02e-bifurcation.qmd),

\begin{equation}
\frac{dX}{dt} = f(X, k) = 10 + 45\frac{X^4}{X^4 + 200^4} - kX . \tag{1}
\end{equation}

Every method below evaluates $f(X, k)$, so we define it once in each language.
Because each chapter renders as an independent session, this repeats the
`derivs_k` used in [Part 2E](02e-bifurcation.qmd) rather than relying on it.

### R implementation {.impl}

```{r}
derivs_k <- function(X, k) 10 + 45 * X^4 / (X^4 + 200^4) - k * X
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

def derivs_k(X, k):
    return 10 + 45 * X**4 / (X**4 + 200**4) - k * X
```

## 2F.2 Contour method

Write $z = f(k, X)$, a surface over the $(k, X)$ plane. The steady states are
exactly the points where this surface crosses zero, so the bifurcation curve is
the **zero-level contour** of $z$. Standard contour algorithms take $f$ on a grid
of $(k, X)$ and trace the level set $z = 0$; both R's `contour()` and Matplotlib's
`contour()` do this. The method is powerful and stable, and, unlike the sampling
methods of Part 2E, it returns the curve as connected, ordered line segments
rather than a point cloud. Its one weakness is that it can miss a branch if the
solution set is not connected. Here $f(X, k) = 0$ is a single connected S-shaped
curve, so one contour captures all of it.

### R implementation {.impl}

```{r}
# evaluate f on a grid of (k, X), then draw its zero-level contour
k_all <- seq(0.1, 0.2, by = 0.001)
X_all <- seq(0, 600, by = 5)
z <- outer(k_all, X_all, function(k, X) derivs_k(X, k))   # z[i, j] = f(X_all[j], k_all[i])

contour(k_all, X_all, z, levels = 0, col = 2, drawlabels = FALSE,
        xlab = "k", ylab = "X")
```

### Python implementation {.impl}

```{python}
# evaluate f on a grid of (k, X), then draw its zero-level contour
k_all = np.arange(0.1, 0.2 + 0.001, 0.001)
X_all = np.arange(0, 600 + 5, 5)
K, Xg = np.meshgrid(k_all, X_all)          # grid of k and X
z = derivs_k(Xg, K)                         # f(X, k) at every grid point

plt.contour(k_all, X_all, z, levels=[0], colors="red")
plt.xlabel("k"); plt.ylabel("X")
plt.show()
```

The contour traces the whole S-curve as connected segments, already ordered by
the algorithm, with no separate reordering step. This ordered output is the main
practical advantage over the point-sampling methods, and it is what the
interpolation and continuation methods below build on.
