---
title: "2F. Bifurcation: finding curves"
author: "Mingyang Lu"
old_id: "02E"
---

In [Part 2E](02e-bifurcation.qmd) we built the bifurcation diagram by *sampling*:
at each value of $k$ we located the steady states (the roots of $f(X, k) = 0$, or
the states an ODE relaxes onto), producing a cloud of points that we then
connected into a curve. That works, but it treats the diagram as a set of
independent points and depends on a post-hoc reordering step to make a curve out
of them.

This chapter takes the complementary view: the bifurcation set
$\{(k, X) : f(X, k) = 0\}$ is a *curve* in the $(k, X)$ plane, and we trace it
directly. Two families of methods do this. The **contour method** treats
$f(k, X)$ as a surface and extracts its zero-level contour, interpolating within a
grid to place the curve accurately. **Numerical continuation** starts from one
point on the curve and walks along it with a predictor–corrector step. These are
the general-purpose tools of bifurcation analysis, and they return the curve
already ordered.

## 2F.1 The model derivative

We use the same self-activating-gene model as in [Part 2E](02e-bifurcation.qmd),

\begin{equation}
\frac{dX}{dt} = f(X, k) = 10 + 45\frac{X^4}{X^4 + 200^4} - kX . \tag{1}
\end{equation}

Every method below evaluates $f(X, k)$, so we define it once in each language.
Because each chapter renders as an independent session, this repeats the
`derivs_k` used in [Part 2E](02e-bifurcation.qmd) rather than relying on it.

### R implementation {.impl}

```{r}
derivs_k <- function(X, k) 10 + 45 * X^4 / (X^4 + 200^4) - k * X
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

def derivs_k(X, k):
    return 10 + 45 * X**4 / (X**4 + 200**4) - k * X
```

## 2F.2 Contour method

Write $z = f(k, X)$, a surface over the $(k, X)$ plane. The steady states are
exactly the points where this surface crosses zero, so the bifurcation curve is
the **zero-level contour** of $z$. Standard contour algorithms take $f$ on a grid
of $(k, X)$ and trace the level set $z = 0$; both R's `contour()` and Matplotlib's
`contour()` do this. The method is powerful and stable, and, unlike the sampling
methods of Part 2E, it returns the curve as connected, ordered line segments
rather than a point cloud. Its one weakness is that it can miss a branch if the
solution set is not connected. Here $f(X, k) = 0$ is a single connected S-shaped
curve, so one contour captures all of it.

### R implementation {.impl}

```{r}
# evaluate f on a grid of (k, X), then draw its zero-level contour
k_all <- seq(0.1, 0.2, by = 0.001)
X_all <- seq(0, 600, by = 5)
z <- outer(k_all, X_all, function(k, X) derivs_k(X, k))   # z[i, j] = f(X_all[j], k_all[i])

contour(k_all, X_all, z, levels = 0, col = 2, drawlabels = FALSE,
        xlab = "k", ylab = "X")
```

### Python implementation {.impl}

```{python}
# evaluate f on a grid of (k, X), then draw its zero-level contour
k_all = np.arange(0.1, 0.2 + 0.001, 0.001)
X_all = np.arange(0, 600 + 5, 5)
K, Xg = np.meshgrid(k_all, X_all)          # grid of k and X
z = derivs_k(Xg, K)                         # f(X, k) at every grid point

plt.contour(k_all, X_all, z, levels=[0], colors="red")
plt.xlabel("k"); plt.ylabel("X")
plt.show()
```

The contour traces the whole S-curve as connected segments, already ordered by
the algorithm, with no separate reordering step. This ordered output is the main
practical advantage over the point-sampling methods, and it is what the
interpolation and continuation methods below build on.

## 2F.3 Bilinear interpolation

The built-in contour routines are black boxes. To see how contour extraction
works, we build a simple one from **bilinear interpolation**, the two-dimensional
analogue of linear interpolation. Inside a grid square with corners
$(x_1, y_1), (x_2, y_1), (x_1, y_2), (x_2, y_2)$ and corner values
$f_{ij} = f(x_i, y_j)$, a solution of $f(x, y) = 0$ exists whenever the four
corners are not all the same sign. We first interpolate linearly in $x$ along the
bottom and top edges,

\begin{equation}
\begin{cases}
f(x, y_1) = \dfrac{x_2 - x}{x_2 - x_1} f_{11} + \dfrac{x - x_1}{x_2 - x_1} f_{21} \\[2mm]
f(x, y_2) = \dfrac{x_2 - x}{x_2 - x_1} f_{12} + \dfrac{x - x_1}{x_2 - x_1} f_{22}
\end{cases} \tag{2}
\end{equation}

and then interpolate those linearly in $y$, giving the full bilinear surface

\begin{equation}
f(x, y) = \frac{1}{(x_2 - x_1)(y_2 - y_1)}
\begin{pmatrix} x_2 - x & x - x_1 \end{pmatrix}
\begin{pmatrix} f_{11} & f_{12} \\ f_{21} & f_{22} \end{pmatrix}
\begin{pmatrix} y_2 - y \\ y - y_1 \end{pmatrix} .
\end{equation}

Setting $f(x, y) = 0$ and solving for $y$ gives

\begin{equation}
y = \frac{f(x, y_1)\, y_2 - f(x, y_2)\, y_1}{f(x, y_1) - f(x, y_2)} , \tag{3}
\end{equation}

which has the same form as the false-position step of
[Part 2E](02e-bifurcation.qmd). To trace the contour we visit every grid square
whose corners differ in sign, sample $x$ across it, and read off the matching $y$
from Equations (2) and (3), keeping only points that land inside the square.

### R implementation {.impl}

```{r}
# bilinear contour in one grid square: sample x, solve f = 0 for y (Eqs 2-3)
# returns flattened x,y pairs where the z = 0 curve crosses the square, or NULL
contour_segment_bilinear <- function(x_all, y_all, z, ind_x, ind_y, npoints = 20) {
  f11 <- z[ind_x, ind_y];     f21 <- z[ind_x + 1, ind_y]
  f12 <- z[ind_x, ind_y + 1]; f22 <- z[ind_x + 1, ind_y + 1]
  if (length(unique(sign(c(f11, f21, f12, f22)))) == 1) return(NULL)  # all same sign: no crossing

  x1 <- x_all[ind_x]; x2 <- x_all[ind_x + 1]
  y1 <- y_all[ind_y]; y2 <- y_all[ind_y + 1]
  x_points <- seq(x1, x2, length.out = npoints)
  f_x_y1 <- ((x2 - x_points) * f11 + (x_points - x1) * f21) / (x2 - x1)   # Eq (2)
  f_x_y2 <- ((x2 - x_points) * f12 + (x_points - x1) * f22) / (x2 - x1)   # Eq (2)
  y_points <- (f_x_y1 * y2 - f_x_y2 * y1) / (f_x_y1 - f_x_y2)             # Eq (3)

  keep <- which((y_points - y1) * (y2 - y_points) > 0)   # keep y inside (y1, y2)
  if (length(keep) == 0) return(NULL)
  c(t(cbind(x_points[keep], y_points[keep])))
}
```

```{r}
# build f on the grid, then extract the zero contour square by square
k_all <- seq(0.1, 0.2, by = 0.001)
X_all <- seq(0, 600, by = 5)
z <- outer(k_all, X_all, function(k, X) derivs_k(X, k))   # z[i, j] = f(X_all[j], k_all[i])

segments <- list()
for (i in seq_len(length(k_all) - 1)) {
  for (j in seq_len(length(X_all) - 1)) {
    seg <- contour_segment_bilinear(k_all, X_all, z, i, j)
    if (length(seg) > 0) segments[[length(segments) + 1]] <- seg
  }
}
null_contour <- matrix(unlist(segments), ncol = 2, byrow = TRUE)

plot(null_contour[, 1], null_contour[, 2], pch = 20, cex = 0.3, col = 2,
     xlab = expression(k ~ (Minute^{-1})), ylab = "Xs (nM)",
     xlim = c(0.09, 0.21), ylim = c(0, 600))
```

### Python implementation {.impl}

```{python}
# bilinear contour in one grid square: sample x, solve f = 0 for y (Eqs 2-3)
# returns x,y pairs where the z = 0 curve crosses the square, or None
def contour_segment_bilinear(x_all, y_all, z, ind_x, ind_y, npoints=20):
    f11 = z[ind_x, ind_y];     f21 = z[ind_x + 1, ind_y]
    f12 = z[ind_x, ind_y + 1]; f22 = z[ind_x + 1, ind_y + 1]
    if len(np.unique(np.sign([f11, f21, f12, f22]))) == 1:   # all same sign: no crossing
        return None
    x1, x2 = x_all[ind_x], x_all[ind_x + 1]
    y1, y2 = y_all[ind_y], y_all[ind_y + 1]
    x_points = np.linspace(x1, x2, npoints)
    f_x_y1 = ((x2 - x_points) * f11 + (x_points - x1) * f21) / (x2 - x1)   # Eq (2)
    f_x_y2 = ((x2 - x_points) * f12 + (x_points - x1) * f22) / (x2 - x1)   # Eq (2)
    y_points = (f_x_y1 * y2 - f_x_y2 * y1) / (f_x_y1 - f_x_y2)             # Eq (3)

    keep = (y_points - y1) * (y2 - y_points) > 0                          # keep y inside (y1, y2)
    pts = np.column_stack((x_points[keep], y_points[keep]))
    return pts if pts.size else None
```

```{python}
# build f on the grid, then extract the zero contour square by square
k_all = np.arange(0.1, 0.2 + 0.001, 0.001)
X_all = np.arange(0, 600 + 5, 5)
K, Xg = np.meshgrid(k_all, X_all, indexing="ij")   # z[i, j] = f(X_all[j], k_all[i])
z = derivs_k(Xg, K)

segments = []
for i in range(len(k_all) - 1):
    for j in range(len(X_all) - 1):
        seg = contour_segment_bilinear(k_all, X_all, z, i, j)
        if seg is not None:
            segments.append(seg)
null_contour = np.vstack(segments)

plt.scatter(null_contour[:, 0], null_contour[:, 1], color="red", s=2)
plt.xlabel(r"$k$ (Minute$^{-1}$)"); plt.ylabel(r"$X_s$ (nM)")
plt.xlim(0.09, 0.21); plt.ylim(0, 600); plt.show()
```

The hand-built extractor recovers the same S-curve as the library contour
routine. Bilinear interpolation is cheap and widely used (in image processing,
for instance), but it treats $f$ as flat-sided within each square, so a coarse
grid shows visible facets. A smoother fit needs a higher-order scheme.
