---
title: "8C. Particles in a box: MCMC sampling"
author: "Mingyang Lu"
old_id: "08C"
---

In [Part 5C](../05-molecular-dynamics/05c-box-of-particles.qmd) we studied a box of
particles interacting through the Lennard-Jones potential by *molecular dynamics*,
integrating Newton's equations to follow the trajectory in time. Here we study the
same box with the *Metropolis-Hastings algorithm* of
[Part 8B](08b-metropolis-algorithm.qmd), sampling configurations directly from the
Boltzmann distribution instead of propagating dynamics.

The two methods target the same equilibrium properties but from opposite angles.
Molecular dynamics tracks positions and velocities through time; Monte Carlo needs
only positions. The reason is that the kinetic energy in the Boltzmann distribution
depends only on the velocities and factors out of the partition function (a result
from statistical mechanics we take on faith here). So the sampling depends solely on
the particle positions and the potential energy, and we can drop velocities and
kinetic energy entirely.

## 8C.1 The Lennard-Jones potential

As in Part 5C, particles interact pairwise through the Lennard-Jones force

$$ F(r) = \frac{1}{r^{13}} - \frac{1}{r^{7}} , \tag{1} $$

whose potential energy [@lennardjones1924] is

$$ U(r) = \int -F(r)\, dr = \frac{1}{12 r^{12}} - \frac{1}{6 r^{6}} , \tag{2} $$

with a minimum at $r = 1$. We write the potential as a function of $r^2$ rather than
$r$, which lets the pairwise routines skip a square root. Distances use the periodic
box and the minimum image convention introduced in
[Part 5C](../05-molecular-dynamics/05c-box-of-particles.qmd), so each particle
interacts with the nearest image of every other particle.

### R implementation {.impl}

```{r}
# Lennard-Jones potential, as a function of r squared (avoids a sqrt)
potential <- function(r2) {
  r6 <- r2^3
  1 / (12 * r6 * r6) - 1 / (6 * r6)
}

r_all <- seq(0.85, 4, by = 0.01)
plot(r_all, potential(r_all^2), type = "l", xlab = "r", ylab = "U",
     xlim = c(0, 4), ylim = c(-0.2, 0.3))
abline(h = 0, lty = 3, col = 2)
```

```{r}
# pairwise potential between particles i and j, minimum image convention
potential_ij <- function(xi, xj, a) {
  dx <- xi - xj
  dx <- dx - round(dx / a) * a            # nearest periodic image
  potential(sum(dx * dx))
}

# total potential energy over all pairs
potential_tot <- function(x_all, a) {
  ntot <- length(x_all) / 2; p_tot <- 0
  for (i in 1:(ntot - 1)) {
    xi <- x_all[(2 * i - 1):(2 * i)]
    for (j in (i + 1):ntot) {
      xj <- x_all[(2 * j - 1):(2 * j)]
      p_tot <- p_tot + potential_ij(xi, xj, a)
    }
  }
  p_tot
}
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

# Lennard-Jones potential, as a function of r squared (avoids a sqrt)
def potential(r2):
    r6 = r2**3
    return 1 / (12 * r6 * r6) - 1 / (6 * r6)

r_all = np.arange(0.85, 4, 0.01)
plt.plot(r_all, potential(r_all**2), color="blue")
plt.axhline(0, color="red", linestyle=":")
plt.xlabel("r"); plt.ylabel("U"); plt.xlim(0, 4); plt.ylim(-0.2, 0.3); plt.show()
```

```{python}
# pairwise potential between particles i and j, minimum image convention
def potential_ij(xi, xj, a):
    dx = xi - xj
    dx = dx - np.round(dx / a) * a          # nearest periodic image
    return potential(np.dot(dx, dx))

# total potential energy over all pairs
def potential_tot(x_all, a):
    ntot = len(x_all) // 2; p_tot = 0.0
    for i in range(ntot - 1):
        xi = x_all[2 * i:2 * i + 2]
        for j in range(i + 1, ntot):
            xj = x_all[2 * j:2 * j + 2]
            p_tot += potential_ij(xi, xj, a)
    return p_tot
```

The potential has the familiar Lennard-Jones shape: a steep repulsive wall at short
range and a shallow attractive well with its minimum at $r = 1$. A configuration is
stored as a flat vector `x_all` of length $2 N$, holding $(x_1, y_1, x_2, y_2, \dots)$.

## 8C.2 Local moves and incremental energy

Metropolis-Hastings needs the *change* in energy for each proposed move, not the full
energy. We propose a local move, displacing one randomly chosen particle by a small
amount. Because only that particle's interactions change, we recompute just the terms
involving it, which costs $O(N)$ instead of the $O(N^2)$ of a full recomputation.
This is the same idea as the incremental energy update for the Ising model in
[Part 8B](08b-metropolis-algorithm.qmd).

### R implementation {.impl}

```{r}
# energy change from moving particle i to x_new (only its terms change)
dp <- function(x_all, a, x_new, i) {
  xi <- x_all[(2 * i - 1):(2 * i)]
  ntot <- length(x_all) / 2; p_old <- 0; p_new <- 0
  for (j in 1:ntot) {
    if (j != i) {
      xj <- x_all[(2 * j - 1):(2 * j)]
      p_old <- p_old + potential_ij(xi, xj, a)
      p_new <- p_new + potential_ij(x_new, xj, a)
    }
  }
  p_new - p_old
}
```

### Python implementation {.impl}

```{python}
# energy change from moving particle i to x_new (only its terms change)
def dp(x_all, a, x_new, i):
    xi = x_all[2 * i:2 * i + 2]
    ntot = len(x_all) // 2; p_old = 0.0; p_new = 0.0
    for j in range(ntot):
        if j != i:
            xj = x_all[2 * j:2 * j + 2]
            p_old += potential_ij(xi, xj, a)
            p_new += potential_ij(x_new, xj, a)
    return p_new - p_old
```

## 8C.3 Metropolis-Hastings sampling

We start from 25 particles on an evenly spaced grid. The box size $a = 6.25 = 5
\times 1.25$ is chosen so the grid tiles periodically: every nearest-neighbor
spacing, including across the boundary, equals 1.25. The sampler picks a random
particle, proposes a small uniform displacement, wraps it back into the box, and
accepts the move with the Metropolis rule using the incremental energy `dp`. As with
the Ising model, a move that lowers the energy is always accepted.

::: {.callout-note title="Recipe 8C.3.1"}
**Objective.** Sample equilibrium configurations of a Lennard-Jones particle box with Metropolis Monte Carlo.

**Model.** A 2D box of N particles interacting through the Lennard-Jones potential `U(r) = 1/(12*r^12) - 1/(6*r^6)` with minimum-image periodic boundaries, sampled from the Boltzmann distribution `proportional to exp(-U/T)` over positions.

**Method.** The Metropolis-Hastings algorithm with local single-particle moves and an incremental energy update.

**Test.** 25 particles on a 5x5 grid at spacing 1.25 (box side `a = 6.25`), move size `dxmax = 0.25`, temperature `T = 0.05`, 5000 equilibration then 10000 sampling steps; seed 10.

**Verification.** During equilibration the energy falls as particles condense from the open grid into a denser cluster, then fluctuates around a low plateau.
:::

### R implementation {.impl}

```{r}
#| fig-width: 4.2
#| fig-height: 4.2
#| out-width: "50%"
# initial configuration: 25 particles on a grid
set.seed(10)
ntot <- 25; a <- 6.25
grid_pts <- seq(0, 5, by = 1.25)
mat_xy <- as.matrix(expand.grid(grid_pts, grid_pts))
x0 <- c(t(mat_xy))                                  # flat (x1, y1, x2, y2, ...)
xy0 <- matrix(x0, ncol = 2, byrow = TRUE)
par(pty = "s")
plot(xy0[, 1], xy0[, 2], pch = 1, xlab = "X", ylab = "Y", xlim = c(0, a), ylim = c(0, a))
potential_tot(x0, a)                                # initial energy
```

```{r}
# Metropolis-Hastings sampler with local moves
metropolis_particle <- function(ntot, cal_e, cal_de, x0, dxmax, nstep, temp, a) {
  x <- matrix(0, nstep + 1, 2 * ntot); e <- numeric(nstep + 1)
  x[1, ] <- x0; e[1] <- cal_e(x0, a); num_accept <- 0
  for (i in 1:nstep) {
    k <- sample(ntot, 1)                             # random particle
    x_new <- x[i, (2 * k - 1):(2 * k)] + runif(2, -dxmax, dxmax)
    x_new <- x_new - floor(x_new / a) * a            # periodic boundary
    de <- cal_de(x[i, ], a, x_new, k)
    if (de < 0 || runif(1) < exp(-de / temp)) {
      x[i + 1, ] <- x[i, ]; x[i + 1, (2 * k - 1):(2 * k)] <- x_new
      e[i + 1] <- e[i] + de; num_accept <- num_accept + 1
    } else { x[i + 1, ] <- x[i, ]; e[i + 1] <- e[i] }
  }
  cat("acceptance rate =", num_accept / nstep, "\n")
  list(x = x, e = e)
}
```

The first run is an equilibration phase: the grid starts at a higher energy than the
low temperature $T = 0.05$ favors, so the energy drops as the particles rearrange. We
tuned `dxmax` to keep the acceptance rate near 40-50%.

```{r}
set.seed(10)
dxmax <- 0.25; temp <- 0.05
res_init <- metropolis_particle(ntot, potential_tot, dp, x0, dxmax, nstep = 5000, temp, a)
plot(0:5000, res_init$e, type = "l", col = 2, xlab = "Step", ylab = "Total energy")
x1 <- res_init$x[5001, ]
```

```{r}
#| fig-width: 4.2
#| fig-height: 4.2
#| out-width: "50%"
# configuration after equilibration
xy1 <- matrix(x1, ncol = 2, byrow = TRUE)
par(pty = "s")
plot(xy1[, 1], xy1[, 2], pch = 1, xlab = "X", ylab = "Y", xlim = c(0, a), ylim = c(0, a))
```

```{r}
# second phase: sampling for the radial distribution function
res_final <- metropolis_particle(ntot, potential_tot, dp, x1, dxmax, nstep = 10000, temp, a)
plot(0:10000, res_final$e, type = "l", col = 2, xlab = "Step", ylab = "Total energy")
```

### Python implementation {.impl}

```{python}
#| fig-width: 4.2
#| fig-height: 4.2
#| out-width: "50%"
# initial configuration: 25 particles on a grid
np.random.seed(10)
ntot = 25; a = 6.25
grid_pts = np.arange(0, 6.25, 1.25)
xx, yy = np.meshgrid(grid_pts, grid_pts)
xy0 = np.column_stack([xx.ravel(), yy.ravel()])
x0 = xy0.reshape(-1)                                 # flat (x1, y1, x2, y2, ...)
plt.figure(figsize=(4.2, 4.2))
plt.scatter(xy0[:, 0], xy0[:, 1], facecolors="none", edgecolors="C0")
plt.gca().set_aspect("equal")
plt.xlabel("X"); plt.ylabel("Y"); plt.xlim(0, a); plt.ylim(0, a); plt.show()
print(potential_tot(x0, a))                          # initial energy
```

```{python}
# Metropolis-Hastings sampler with local moves
def metropolis_particle(ntot, cal_e, cal_de, x0, dxmax, nstep, temp, a):
    x = np.zeros((nstep + 1, 2 * ntot)); e = np.zeros(nstep + 1)
    x[0] = x0; e[0] = cal_e(x0, a); num_accept = 0
    for i in range(nstep):
        k = np.random.randint(ntot)                  # random particle
        x_new = x[i, 2 * k:2 * k + 2] + np.random.uniform(-dxmax, dxmax, size=2)
        x_new = x_new - np.floor(x_new / a) * a       # periodic boundary
        de = cal_de(x[i], a, x_new, k)
        if de < 0 or np.random.uniform() < np.exp(-de / temp):
            x[i + 1] = x[i]; x[i + 1, 2 * k:2 * k + 2] = x_new
            e[i + 1] = e[i] + de; num_accept += 1
        else:
            x[i + 1] = x[i]; e[i + 1] = e[i]
    print("acceptance rate =", num_accept / nstep)
    return {"x": x, "e": e}
```

```{python}
np.random.seed(10)
dxmax = 0.25; temp = 0.05
res_init = metropolis_particle(ntot, potential_tot, dp, x0, dxmax, 5000, temp, a)
plt.plot(range(5001), res_init["e"], "r-")
plt.xlabel("Step"); plt.ylabel("Total energy"); plt.show()
x1 = res_init["x"][5000]
```

```{python}
#| fig-width: 4.2
#| fig-height: 4.2
#| out-width: "50%"
# configuration after equilibration
xy1 = x1.reshape(-1, 2)
plt.figure(figsize=(4.2, 4.2))
plt.scatter(xy1[:, 0], xy1[:, 1], facecolors="none", edgecolors="C0")
plt.gca().set_aspect("equal")
plt.xlabel("X"); plt.ylabel("Y"); plt.xlim(0, a); plt.ylim(0, a); plt.show()
```

```{python}
# second phase: sampling for the radial distribution function
res_final = metropolis_particle(ntot, potential_tot, dp, x1, dxmax, 10000, temp, a)
plt.plot(range(10001), res_final["e"], "r-")
plt.xlabel("Step"); plt.ylabel("Total energy"); plt.show()
```

During equilibration the energy falls steadily as the particles condense from the
open grid into a denser, lower-energy arrangement; the post-equilibration snapshot
shows them clustered rather than spread on the original lattice. In the second phase
the energy fluctuates around a low plateau, which is the set of near-equilibrium
configurations we sample for the analysis below.

## 8C.4 The radial distribution function

The radial distribution function (RDF) measures how particle density varies with
distance from a reference particle, exactly as computed from the molecular-dynamics
trajectory in [Part 5C](../05-molecular-dynamics/05c-box-of-particles.qmd). Here we
average it over the second-phase configurations. Successive MCMC steps move only one
particle and are highly correlated, so we thin the trajectory, using every 20th
configuration as a roughly independent sample.

### R implementation {.impl}

```{r}
# radial distribution function, averaged over sampled configurations
cal_rdf <- function(ntot, results, a, dr) {
  nt_all <- nrow(results)
  r_all <- seq(dr, a / 2, by = dr); nr <- length(r_all)
  counts <- numeric(nr)
  for (nt in 1:nt_all) {
    xy <- matrix(results[nt, ], ncol = 2, byrow = TRUE)
    dx <- outer(xy[, 1], xy[, 1], "-"); dx <- dx - round(dx / a) * a   # minimum image
    dy <- outer(xy[, 2], xy[, 2], "-"); dy <- dy - round(dy / a) * a
    rij <- sqrt(dx^2 + dy^2); diag(rij) <- NA           # exclude self-pairs
    ind <- ceiling(rij / dr)
    keep <- !is.na(ind) & ind <= nr
    counts <- counts + tabulate(ind[keep], nbins = nr)
  }
  r_mean <- r_all - 0.5 * dr
  rho <- counts / nt_all / ntot / (2 * pi * dr * r_mean)
  list(r = r_mean, rdf = rho / (ntot / a^2))            # normalize by bulk density
}

res_rdf <- cal_rdf(ntot, res_final$x[seq(1, 10001, by = 20), ], a, dr = 0.05)
plot(res_rdf$r, res_rdf$rdf, type = "l", col = 2, xlab = "r", ylab = "g(r)")
```

### Python implementation {.impl}

```{python}
# radial distribution function, averaged over sampled configurations
def cal_rdf(ntot, results, a, dr):
    nt_all = results.shape[0]
    r_all = np.arange(dr, a / 2, dr); nr = len(r_all)
    counts = np.zeros(nr)
    for nt in range(nt_all):
        xy = results[nt].reshape(-1, 2)
        dx = np.subtract.outer(xy[:, 0], xy[:, 0]); dx -= np.round(dx / a) * a   # minimum image
        dy = np.subtract.outer(xy[:, 1], xy[:, 1]); dy -= np.round(dy / a) * a
        rij = np.sqrt(dx**2 + dy**2); np.fill_diagonal(rij, np.nan)  # exclude self-pairs
        ind = np.ceil(rij / dr)
        keep = ~np.isnan(ind) & (ind <= nr)
        counts += np.bincount(ind[keep].astype(int) - 1, minlength=nr)
    r_mean = r_all - 0.5 * dr
    rho = counts / nt_all / ntot / (2 * np.pi * dr * r_mean)
    return r_mean, rho / (ntot / a**2)                  # normalize by bulk density

r_mean, rdf = cal_rdf(ntot, res_final["x"][::20], a, dr=0.05)
plt.plot(r_mean, rdf, "r-")
plt.xlabel("r"); plt.ylabel("g(r)"); plt.show()
```

The RDF rises to a sharp first peak near $r = 1$, the equilibrium separation of the
Lennard-Jones potential, and shows further peaks at larger distances, the signature
of short-range order in the condensed cluster. This matches the picture from the
molecular-dynamics simulation of Part 5C: two very different sampling methods,
Newtonian dynamics and Metropolis Monte Carlo, recover the same equilibrium
structure.

::: {.callout-note title="Practice Q1"}
Compare this Monte Carlo calculation with the molecular-dynamics simulation of
Part 5C. Which reaches equilibrium in fewer operations for this system? Then consider
efficiency of the sampler itself: the local move recomputes only $O(N)$ interaction
terms per step. Could you propose a move that displaces several particles at once, and
would it improve or hurt the acceptance rate?
:::
