---
title: "8D. Gillespie algorithm"
author: "Mingyang Lu"
old_id: "08D"
---

The stochastic differential equations of [Part 6](../06-stochastic/06a-random-number-generators.qmd)
add continuous noise to a deterministic rate equation. For a chemical reaction
network with small molecule numbers, a different and more fundamental description is
often needed: the molecules are discrete and react one event at a time. The
*Gillespie algorithm* [@gillespie1976general; @gillespie1977exact], also called the
stochastic simulation algorithm (SSA), generates exact sample trajectories of such a
network from its chemical master equation.

We describe the system by a state vector $\mathbf{S}$ holding the number of molecules
of each species. Each reaction $j$ converts reactants into products,

$$ \sum_i \alpha_{ij} S_i \rightarrow \sum_i \beta_{ij} S_i , $$

and changes the state by its *stoichiometry vector* $\boldsymbol{\gamma}_j$, the
$j$-th column of the stoichiometry matrix with entries
$\Gamma_{ij} = \beta_{ij} - \alpha_{ij}$. Each reaction has a *propensity* $R_j$, the
probability per unit time that it fires, given by the law of mass action times a
combinatorial degeneracy factor [@gillespie1992rigorous],

$$ R_j = k_j \prod_{i} \binom{S_i}{\alpha_{ij}} \approx k_j \prod_i S_i^{\alpha_{ij}} , $$

where the binomial factor counts the distinct ways of choosing the reactant
molecules (for example, $S(S-1)/2$ ways to pick two of $S$ identical molecules).

## 8D.1 The algorithm

Given the propensities, each Gillespie step advances the system by one reaction event:

1. compute all propensities $R_j(\mathbf{S})$;
2. draw the waiting time $\tau$ to the next event from an exponential distribution
   with rate $R_\text{tot} = \sum_j R_j$;
3. pick which reaction fires, choosing reaction $j$ with probability
   $R_j / R_\text{tot}$;
4. advance time $t \rightarrow t + \tau$ and update the state
   $\mathbf{S} \rightarrow \mathbf{S} + \boldsymbol{\gamma}_j$.

Repeating until a stopping time gives one exact realization. The generic driver below
takes user-supplied `propensity` and `stoichiometry` functions, so the same code runs
any reaction network.

### R implementation {.impl}

```{r}
# one Gillespie step: advance by a single reaction event
gillespie_step <- function(t, Xs, propensity, stoichiometry, ...) {
  p_all <- propensity(Xs, ...); p_sum <- sum(p_all)
  if (abs(p_sum) < 1e-9) return(list(t = t, Xs = Xs, if_stop = TRUE))  # no reaction possible
  dt <- rexp(1, rate = p_sum)                       # waiting time to next event
  event <- sample(length(p_all), 1, prob = p_all)   # which reaction fires
  list(t = t + dt, Xs = Xs + stoichiometry(event, ...), if_stop = FALSE)
}

# full stochastic simulation up to tmax or iter_max events
ssa <- function(propensity, stoichiometry, X0, tmax, iter_max, ...) {
  Xs <- X0; n <- length(Xs)
  t_all <- numeric(iter_max); X_all <- matrix(0, iter_max, n)
  t_all[1] <- 0; X_all[1, ] <- Xs
  t <- 0; iter <- 1
  while (t < tmax && iter < iter_max) {
    out <- gillespie_step(t, Xs, propensity, stoichiometry, ...)
    if (out$if_stop) break
    iter <- iter + 1; t <- out$t; Xs <- out$Xs
    t_all[iter] <- t; X_all[iter, ] <- Xs
  }
  cbind(t_all[1:iter], X_all[1:iter, , drop = FALSE])
}
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

# one Gillespie step: advance by a single reaction event
def gillespie_step(t, Xs, propensity, stoichiometry, **kw):
    p_all = propensity(Xs, **kw); p_sum = np.sum(p_all)
    if abs(p_sum) < 1e-9:
        return {"t": t, "Xs": Xs, "if_stop": True}          # no reaction possible
    dt = np.random.exponential(1 / p_sum)                    # waiting time to next event
    event = np.random.choice(len(p_all), p=p_all / p_sum)    # which reaction fires
    return {"t": t + dt, "Xs": Xs + stoichiometry(event, **kw), "if_stop": False}

# full stochastic simulation up to tmax or iter_max events
def ssa(propensity, stoichiometry, X0, tmax, iter_max, **kw):
    Xs = np.array(X0); n = len(Xs)
    t_all = np.zeros(iter_max); X_all = np.zeros((iter_max, n))
    t_all[0] = 0; X_all[0] = Xs
    t = 0; it = 0
    while t < tmax and it < iter_max - 1:
        out = gillespie_step(t, Xs, propensity, stoichiometry, **kw)
        if out["if_stop"]:
            break
        it += 1; t = out["t"]; Xs = out["Xs"]
        t_all[it] = t; X_all[it] = Xs
    return np.column_stack((t_all[:it + 1], X_all[:it + 1]))
```

## 8D.2 A constitutively transcribed gene

The simplest example is a gene produced at a constant rate $g$ and degraded at rate
$k$, a birth-death process,

$$ 0 \xrightarrow{g} x \xrightarrow{k} 0 , $$

whose deterministic rate equation $dx/dt = g - kx$ has steady state
$\bar{x} = g/k$. We specify the network with two functions: the propensities
$(g,\, kx)$ and the stoichiometry ($+1$ for production, $-1$ for degradation).

### R implementation {.impl}

```{r}
p_func1 <- function(Xs, g, k) c(g, k * Xs[1])           # propensities: birth, death
s_func1 <- function(r_ind, g, k) cbind(c(1), c(-1))[, r_ind]   # +1 or -1
```

```{r}
#| fig-width: 9
#| fig-height: 3.2
#| out-width: "95%"
#| rmar: false
# stochastic dynamics started at the deterministic steady state, three copy numbers
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
k <- 0.1
for (g in c(100, 10, 1)) {
  set.seed(101)
  xbar <- g / k
  res <- ssa(p_func1, s_func1, c(xbar), tmax = 100, iter_max = 20000, g = g, k = k)
  plot(res[, 1], res[, 2], type = "s", col = 2, xlab = "Time", ylab = "x",
       main = bquote(bar(x) == .(xbar)), ylim = c(xbar * 0.8, xbar * 1.2))
  abline(h = xbar, col = 3)
}
```

### Python implementation {.impl}

```{python}
def p_func1(Xs, g, k):
    return np.array([g, k * Xs[0]])                    # propensities: birth, death
def s_func1(r_ind, g, k):
    return np.array([[1], [-1]])[r_ind]                # +1 or -1
```

```{python}
#| fig-width: 9
#| fig-height: 3.2
#| out-width: "95%"
# stochastic dynamics started at the deterministic steady state, three copy numbers
fig, axs = plt.subplots(1, 3, figsize=(9, 3.2))
k = 0.1
for ax, g in zip(axs, (100, 10, 1)):
    np.random.seed(101)
    xbar = g / k
    res = ssa(p_func1, s_func1, [xbar], tmax=100, iter_max=20000, g=g, k=k)
    ax.step(res[:, 0], res[:, 1], color="red")
    ax.axhline(xbar, color="green")
    ax.set(xlabel="Time", ylabel="x", title=f"x̄ = {xbar:.0f}", ylim=(xbar * 0.8, xbar * 1.2))
plt.tight_layout(); plt.show()
```

The fluctuations around the steady state grow in absolute size but shrink in relative
size as the copy number increases: a striking feature of intrinsic noise. To quantify
it we compute the time-weighted mean and standard deviation, weighting each state by
how long the system dwells there (the waiting times $\Delta t_i$ are not constant, so
plain averaging would be wrong):

$$
\langle x \rangle = \frac{\sum_i x_i\, \Delta t_i}{\sum_i \Delta t_i} , \qquad
\text{SD}(x) = \sqrt{\langle x^2 \rangle - \langle x \rangle^2} .
$$

### R implementation {.impl}

```{r}
# time-weighted mean, SD, and relative error of each species
cal_stat <- function(mat) {
  mat <- na.omit(mat); nt <- nrow(mat); nx <- ncol(mat)
  dt <- mat[-1, 1] - mat[-nt, 1]; t_tot <- sum(dt)
  mean  <- t(mat[-nt, 2:nx, drop = FALSE])    %*% dt / t_tot
  mean2 <- t(mat[-nt, 2:nx, drop = FALSE]^2)  %*% dt / t_tot
  sd <- sqrt(mean2 - mean^2)
  list(mean = mean, sd = sd, error = sd / mean)
}
```

```{r}
#| fig-width: 9
#| fig-height: 4
#| out-width: "95%"
#| rmar: false
k <- 0.1; ss_all <- c(1000, 100, 10)
stats <- lapply(c(100, 10, 1), function(g) {
  set.seed(101)
  cal_stat(ssa(p_func1, s_func1, c(g / k), 100, 20000, g = g, k = k))
})
sd_all    <- sapply(stats, function(s) s$sd)
error_all <- sapply(stats, function(s) s$error)

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(ss_all, sd_all, type = "b", col = 2, xlab = "Steady-state x", ylab = "SD",
     xlim = c(0, 1000), ylim = c(0, 35))
curve(sqrt(x), add = TRUE, col = 3)
legend("topleft", c("Simulation", "sqrt(x)"), col = c(2, 3), lty = 1, cex = 0.8)
plot(ss_all, error_all, type = "b", col = 2, xlab = "Steady-state x",
     ylab = "SD / Mean", xlim = c(0, 1000), ylim = c(0, 0.4))
```

### Python implementation {.impl}

```{python}
# time-weighted mean, SD, and relative error of each species
def cal_stat(mat):
    mat = mat[~np.isnan(mat).any(axis=1)]; nt = mat.shape[0]
    dt = mat[1:, 0] - mat[:-1, 0]; t_tot = np.sum(dt)
    mean  = mat[:-1, 1:].T  @ dt / t_tot
    mean2 = (mat[:-1, 1:]**2).T @ dt / t_tot
    sd = np.sqrt(mean2 - mean**2)
    return {"mean": mean, "sd": sd, "error": sd / mean}
```

```{python}
#| fig-width: 9
#| fig-height: 4
#| out-width: "95%"
k = 0.1; ss_all = np.array([1000, 100, 10])
stats = []
for g in (100, 10, 1):
    np.random.seed(101)
    stats.append(cal_stat(ssa(p_func1, s_func1, [g / k], 100, 20000, g=g, k=k)))
sd_all    = np.array([s["sd"][0] for s in stats])
error_all = np.array([s["error"][0] for s in stats])

fig, axs = plt.subplots(1, 2, figsize=(9, 4))
xx = np.arange(1, 1001)
axs[0].plot(ss_all, sd_all, "o-", color="red")
axs[0].plot(xx, np.sqrt(xx), color="green", label="sqrt(x)")
axs[0].set(xlabel="Steady-state x", ylabel="SD", xlim=(0, 1000), ylim=(0, 35)); axs[0].legend()
axs[1].plot(ss_all, error_all, "o-", color="red")
axs[1].set(xlabel="Steady-state x", ylabel="SD / Mean", xlim=(0, 1000), ylim=(0, 0.4))
plt.tight_layout(); plt.show()
```

The standard deviation follows $\text{SD}(x) = \sqrt{\bar{x}}$, the Poisson result for
a birth-death process, so the *relative* noise $\text{SD}/\text{mean} = 1/\sqrt{\bar{x}}$
falls as the copy number rises. This is why low-abundance species are intrinsically
noisier. Starting instead from $x = 0$, the trajectories climb toward the steady state
along the deterministic curve $\bar{x}(1 - e^{-kt})$, with fluctuations that again
shrink in relative terms at higher copy number.

### R implementation {.impl}

```{r}
#| fig-width: 9
#| fig-height: 3.2
#| out-width: "95%"
#| rmar: false
# growth from zero toward the steady state
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
k <- 0.1
for (g in c(100, 10, 1)) {
  set.seed(101)
  res <- ssa(p_func1, s_func1, c(0), tmax = 100, iter_max = 20000, g = g, k = k)
  plot(res[, 1], res[, 2], type = "s", col = 2, xlab = "Time", ylab = "x",
       main = bquote(bar(x) == .(g / k)), ylim = c(0, 1.2 * g / k))
  lines(res[, 1], g / k * (1 - exp(-k * res[, 1])), col = 3)
}
```

### Python implementation {.impl}

```{python}
#| fig-width: 9
#| fig-height: 3.2
#| out-width: "95%"
# growth from zero toward the steady state
fig, axs = plt.subplots(1, 3, figsize=(9, 3.2))
k = 0.1
for ax, g in zip(axs, (100, 10, 1)):
    np.random.seed(101)
    res = ssa(p_func1, s_func1, [0], tmax=100, iter_max=20000, g=g, k=k)
    ax.step(res[:, 0], res[:, 1], color="red")
    ax.plot(res[:, 0], g / k * (1 - np.exp(-k * res[:, 0])), color="green")
    ax.set(xlabel="Time", ylabel="x", title=f"x̄ = {g / k:.0f}", ylim=(0, 1.2 * g / k))
plt.tight_layout(); plt.show()
```

## 8D.3 Bursting transcription

Genes often produce proteins in *bursts*: a promoter fires occasionally but releases
several molecules each time. We model this by making each production event add $n$
molecules at rate $g/n$, so the mean production rate $g$ (and the deterministic steady
state $\bar{x} = g/k$) stay fixed as $n$ varies,

$$ 0 \xrightarrow{g/n} n\,x , \qquad x \xrightarrow{k} 0 . $$

### R implementation {.impl}

```{r}
p_func2 <- function(Xs, g, k, n) c(g / n, k * Xs[1])          # burst production, degradation
s_func2 <- function(r_ind, g, k, n) cbind(c(n), c(-1))[, r_ind]   # +n or -1
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "92%"
#| rmar: false
# larger bursts (bigger n) at the same mean level
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
g <- 16; k <- 0.1
res_burst <- list()
for (n in c(1, 2, 4, 8)) {
  set.seed(91)
  res <- ssa(p_func2, s_func2, c(160), tmax = 2000, iter_max = 500000, g = g, k = k, n = n)
  res_burst[[as.character(n)]] <- res
  plot(res[, 1], res[, 2], type = "s", col = 2, xlab = "Time", ylab = "x",
       main = paste("n =", n), ylim = c(100, 220))
  abline(h = 160, col = 3)
}
```

```{r}
#| fig-width: 9
#| fig-height: 4
#| out-width: "95%"
#| rmar: false
# mean stays fixed, noise grows with burst size
n_all <- c(1, 2, 4, 8)
stats_b <- lapply(res_burst, cal_stat)
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(n_all, sapply(stats_b, function(s) s$mean), type = "b", col = 2,
     xlab = "n", ylab = "Mean", xlim = c(0, 8), ylim = c(0, 200))
plot(n_all, sapply(stats_b, function(s) s$sd), type = "b", col = 2,
     xlab = "n", ylab = "SD", xlim = c(0, 8), ylim = c(0, 35))
```

### Python implementation {.impl}

```{python}
def p_func2(Xs, g, k, n):
    return np.array([g / n, k * Xs[0]])                      # burst production, degradation
def s_func2(r_ind, g, k, n):
    return np.array([[n], [-1]])[r_ind]                      # +n or -1
```

```{python}
#| fig-width: 8
#| fig-height: 6
#| out-width: "92%"
# larger bursts (bigger n) at the same mean level
fig, axs = plt.subplots(2, 2, figsize=(8, 6))
g = 16; k = 0.1
res_burst = {}
for ax, n in zip(axs.flat, (1, 2, 4, 8)):
    np.random.seed(91)
    res = ssa(p_func2, s_func2, [160], tmax=2000, iter_max=500000, g=g, k=k, n=n)
    res_burst[n] = res
    ax.step(res[:, 0], res[:, 1], color="red")
    ax.axhline(160, color="green")
    ax.set(xlabel="Time", ylabel="x", title=f"n = {n}", ylim=(100, 220))
plt.tight_layout(); plt.show()
```

```{python}
#| fig-width: 9
#| fig-height: 4
#| out-width: "95%"
# mean stays fixed, noise grows with burst size
n_all = [1, 2, 4, 8]
stats_b = [cal_stat(res_burst[n]) for n in n_all]
fig, axs = plt.subplots(1, 2, figsize=(9, 4))
axs[0].plot(n_all, [s["mean"][0] for s in stats_b], "o-", color="red")
axs[0].set(xlabel="n", ylabel="Mean", xlim=(0, 8), ylim=(0, 200))
axs[1].plot(n_all, [s["sd"][0] for s in stats_b], "o-", color="red")
axs[1].set(xlabel="n", ylabel="SD", xlim=(0, 8), ylim=(0, 35))
plt.tight_layout(); plt.show()
```

The mean stays at 160 regardless of $n$, but the standard deviation grows with the
burst size: larger, rarer bursts inject more noise than frequent single-molecule
events, even at the same average production rate. Transcriptional bursting is thus a
source of gene-expression noise beyond the Poisson floor of the simple birth-death
process.

## 8D.4 A self-inhibiting gene

Finally we model a gene that represses itself. The promoter has an unbound state $D$
and a bound state $C$; two molecules of $M$ can bind the promoter (a dimer),
switching it from $D$ to $C$. Production is faster from the unbound promoter
($g_0 > g_1$), so accumulating $M$ shuts down its own production. The five reactions
are production from each promoter state, degradation of $M$, and binding/unbinding:

$$
\begin{aligned}
D &\xrightarrow{g_0} D + M , & M &\xrightarrow{k} 0 , & 2M + D &\xrightarrow{k_\text{on}} C , \\
C &\xrightarrow{k_\text{off}} 2M + D , & C &\xrightarrow{g_1} C + M . &&
\end{aligned}
$$

The binding propensity carries the degeneracy factor $M(M-1)/2$ for choosing two of
the $M$ molecules, and the effective dissociation constant is
$K = \sqrt{2 k_\text{off} / k_\text{on}}$. We omit the promoter-occupancy traces and
focus on the $M$ dynamics.

### R implementation {.impl}

```{r}
# state Xs = (D, C, M): D unbound promoter, C bound promoter, M molecules
p_func3 <- function(Xs, g0, g1, k, kon, koff) {
  d <- Xs[1]; c <- Xs[2]; m <- Xs[3]
  c(g0 * d, k * m, kon * d * m * (m - 1) / 2, koff * c, g1 * c)
}
s_func3 <- function(r_ind, g0, g1, k, kon, koff) {
  cbind(c(0, 0, 1),    # production from D: M + 1
        c(0, 0, -1),   # degradation: M - 1
        c(-1, 1, -2),  # binding: D - 1, C + 1, M - 2
        c(1, -1, 2),   # unbinding: D + 1, C - 1, M + 2
        c(0, 0, 1))[, r_ind]   # production from C: M + 1
}
```

```{r}
#| fig-width: 9
#| fig-height: 3.2
#| out-width: "95%"
#| rmar: false
# three regimes: self-inhibition (high copy), no feedback, self-inhibition (low copy)
runs <- list(
  list(main = "self-inhibition",  g0 = 55,  g1 = 5,   kon = 0.002, koff = 90, X0 = c(1, 0, 300), tmax = 400, ymax = 600),
  list(main = "no feedback",      g0 = 30,  g1 = 30,  kon = 0,     koff = 0,  X0 = c(1, 0, 300), tmax = 300, ymax = 600),
  list(main = "low copy number",  g0 = 5.5, g1 = 0.5, kon = 0.2,   koff = 90, X0 = c(1, 0, 30),  tmax = 300, ymax = 60)
)
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
stats_si <- list()
for (r in runs) {
  set.seed(101)
  res <- ssa(p_func3, s_func3, r$X0, r$tmax, 50000, g0 = r$g0, g1 = r$g1, k = 0.1, kon = r$kon, koff = r$koff)
  stats_si[[r$main]] <- cal_stat(res)
  plot(res[, 1], res[, 4], type = "s", col = 2, xlab = "Time", ylab = "M",
       main = r$main, xlim = c(0, r$tmax), ylim = c(0, r$ymax))
}
# mean and SD of M in each regime
sapply(stats_si, function(s) c(mean_M = s$mean[3], sd_M = s$sd[3]))
```

### Python implementation {.impl}

```{python}
# state Xs = (D, C, M): D unbound promoter, C bound promoter, M molecules
def p_func3(Xs, g0, g1, k, kon, koff):
    d, c, m = Xs
    return np.array([g0 * d, k * m, kon * d * m * (m - 1) / 2, koff * c, g1 * c])
def s_func3(r_ind, g0, g1, k, kon, koff):
    S = np.array([[0, 0, 1],    # production from D: M + 1
                  [0, 0, -1],   # degradation: M - 1
                  [-1, 1, -2],  # binding: D - 1, C + 1, M - 2
                  [1, -1, 2],   # unbinding: D + 1, C - 1, M + 2
                  [0, 0, 1]])   # production from C: M + 1
    return S[r_ind]
```

```{python}
#| fig-width: 9
#| fig-height: 3.2
#| out-width: "95%"
# three regimes: self-inhibition (high copy), no feedback, self-inhibition (low copy)
runs = [
    dict(main="self-inhibition", g0=55,  g1=5,   kon=0.002, koff=90, X0=[1, 0, 300], tmax=400, ymax=600),
    dict(main="no feedback",     g0=30,  g1=30,  kon=0,     koff=0,  X0=[1, 0, 300], tmax=300, ymax=600),
    dict(main="low copy number", g0=5.5, g1=0.5, kon=0.2,   koff=90, X0=[1, 0, 30],  tmax=300, ymax=60),
]
fig, axs = plt.subplots(1, 3, figsize=(9, 3.2))
stats_si = {}
for ax, r in zip(axs, runs):
    np.random.seed(101)
    res = ssa(p_func3, s_func3, r["X0"], r["tmax"], 50000,
              g0=r["g0"], g1=r["g1"], k=0.1, kon=r["kon"], koff=r["koff"])
    stats_si[r["main"]] = cal_stat(res)
    ax.step(res[:, 0], res[:, 3], color="red")
    ax.set(xlabel="Time", ylabel="M", title=r["main"], xlim=(0, r["tmax"]), ylim=(0, r["ymax"]))
plt.tight_layout(); plt.show()
# mean and SD of M in each regime
for name, s in stats_si.items():
    print(name, "mean_M =", round(s["mean"][2], 1), "sd_M =", round(s["sd"][2], 1))
```

The middle panel turns the feedback off (setting $k_\text{on} = k_\text{off} = 0$ and
$g_0 = g_1$ so the mean $M$ is unchanged), reducing the circuit to a constitutively
expressed gene. Comparing it with the self-inhibiting case on the left, the
self-inhibiting gene has a lower standard deviation: negative autoregulation
suppresses gene-expression noise, a result established both theoretically and
experimentally [@becskei2000engineering]. At low copy number (right panel), the
binding and unbinding of just a few molecules make the dynamics strongly bursty,
resembling noisy oscillations.
