---
title: "4B. Modeling systems with time delays"
author: "Mingyang Lu"
old_id: "04B"
---

We now apply DDEs to three familiar models and watch how a time delay changes their behavior. Throughout we integrate with the second-order Heun scheme from [Part 4A](04a-delayed-differential-equations.qmd), redefined at the start of the first example so the chapter is self-contained.

## 4B.1 Delay logistic growth

Adding a delay to the logistic model gives the **Hutchinson equation** [@hutchinson1948circular], in which the growth of $N$ is limited by the population one delay $\tau$ in the past,

$$ \frac{dN}{dt} = r\, N(t)\left[1 - \frac{N(t - \tau)}{B}\right] , \tag{1} $$

with carrying capacity $B$. We fix $B = 100$ and $\tau = 1$ and vary the growth rate $r$. The delay drives a Hopf bifurcation at $r\tau = \pi/2$: below it the population settles at $B$; above it, it oscillates forever. Four values of $r$ trace the transition: $r = 0.3$ (monotonic growth to $B$), $r = 1.5$ (just below $\pi/2$, a damped oscillation settling at $B$), $r = \pi/2 \approx 1.57$ (the Hopf point, an oscillation that decays extremely slowly), and $r = 1.7$ (above $\pi/2$, a sustained oscillation, a stable limit cycle).

::: {.callout-note title="Recipe 4B.1.1"}
**Objective.** Simulate delay-driven logistic growth and locate the onset of sustained oscillation.

**Model.** The Hutchinson delay-logistic equation, logistic growth whose crowding term uses the population one delay in the past, `dN/dt = r*N(t)*(1 - N(t-tau)/B)`, with carrying capacity B.

**Method.** The second-order Heun integrator for delay differential equations.

**Test.** Constant history `N(t) = 1`; `tau = 1`, `B = 100`, `dt = 0.01`; growth rates r = 0.3, 1.5, pi/2 (about 1.57), and 1.7.

**Show.** A plot of N(t) for each r, from monotonic growth to a sustained oscillation.

**Verification.** Crossing the Hopf point `r*tau = pi/2`: r = 0.3 grows monotonically to B, r = 1.5 is a damped oscillation settling at B, r = pi/2 decays extremely slowly, and r = 1.7 becomes a sustained oscillation (a stable limit cycle).
:::

### R implementation {.impl}

```{r}
# Heun (2nd-order) integrator for a DDE with one constant delay (same as Part 4A)
DDE_Heun <- function(derivs, t0, X0, t.total, dt, tau, ...) {
  t_all <- seq(t0 - tau, t.total, by = dt)
  n_all <- length(t_all)
  X_all <- numeric(n_all)
  n_delay <- as.integer(tau / dt)
  X_all[1:(n_delay + 1)] <- X0
  for (i in (n_delay + 1):(n_all - 1)) {
    k1 <- dt * derivs(t_all[i], X_all[i], X_all[i - n_delay], ...)
    X_all[i + 1] <- X_all[i] + k1
    k2 <- dt * derivs(t_all[i + 1], X_all[i + 1], X_all[i + 1 - n_delay], ...)
    X_all[i + 1] <- X_all[i] + (k1 + k2) / 2
  }
  cbind(t_all, X_all)
}
```

```{r}
dN_logistic_delays <- function(t, N, N_old, r, B) r * N * (1 - N_old / B)

# (r, total time, x-max, y-max) for the four regimes
cases <- list(c(0.3, 40, 40, 120), c(1.5, 200, 200, 220),
              c(pi / 2, 200, 200, 220), c(1.7, 80, 80, 240))
labs <- c("r = 0.3", "r = 1.5", "r = pi/2", "r = 1.7")

for (j in seq_along(cases)) {
  cs <- cases[[j]]
  res <- DDE_Heun(dN_logistic_delays, 0, 1, cs[2], 0.01, 1, r = cs[1], B = 100)
  plot(res[, 1], res[, 2], type = "l", col = 2, xlab = "t", ylab = "N",
       xlim = c(-1, cs[3]), ylim = c(0, cs[4]))
  legend("topright", inset = 0.02, legend = labs[j], col = 2, lty = 1, cex = 0.8)
}
```

### Python implementation {.impl}

```{python}
import numpy as np
import matplotlib.pyplot as plt

# Heun (2nd-order) integrator for a DDE with one constant delay (same as Part 4A)
def DDE_Heun(derivs, t0, X0, t_total, dt, tau, **kw):
    t_all = np.arange(t0 - tau, t_total + dt, dt)
    n_all = len(t_all)
    X_all = np.zeros(n_all)
    n_delay = int(tau / dt)
    X_all[0:n_delay + 1] = X0
    for i in range(n_delay, n_all - 1):
        k1 = dt * derivs(t_all[i], X_all[i], X_all[i - n_delay], **kw)
        X_all[i + 1] = X_all[i] + k1
        k2 = dt * derivs(t_all[i + 1], X_all[i + 1], X_all[i + 1 - n_delay], **kw)
        X_all[i + 1] = X_all[i] + (k1 + k2) / 2
    return np.column_stack((t_all, X_all))
```

```{python}
def dN_logistic_delays(t, N, N_old, r, B):
    return r * N * (1 - N_old / B)

# (r, total time, x-max, y-max, legend label) for the four regimes
cases = [(0.3, 40, 40, 120, "r = 0.3"), (1.5, 200, 200, 220, "r = 1.5"),
         (np.pi / 2, 200, 200, 220, "r = pi/2"), (1.7, 80, 80, 240, "r = 1.7")]

for r, tmax, xl, yl, lab in cases:
    res = DDE_Heun(dN_logistic_delays, 0, 1, tmax, 0.01, 1, r=r, B=100)
    plt.figure()
    plt.plot(res[:, 0], res[:, 1], color="red")
    plt.xlabel("t"); plt.ylabel("N"); plt.xlim(-1, xl); plt.ylim(0, yl)
    plt.legend([lab], loc="upper right"); plt.show()
```

## 4B.2 A self-inhibiting gene with delay

A gene whose product represses its own transcription forms a negative-feedback loop. Because transcription and translation take time, the repression acts on the protein level a delay $\tau$ earlier,

$$ \frac{dX}{dt} = g_0 + g_1\frac{1}{1 + (X(t - \tau)/X_{th})^n} - k X(t) , \tag{2} $$

where $g_0$ is the leakage rate, $g_0 + g_1$ the maximum rate (reached when $X$ is low and the inhibitory Hill term approaches $1$), $X_{th}$ the Hill threshold, $n$ the Hill coefficient, and $k$ the degradation rate. With $g_0 = 10$, $g_1 = 60$, $X_{th} = 200$, $n = 4$, $k = 0.1$, the delay decides the dynamics: a short delay ($\tau = 5$) settles at the steady state near $250$, but as $\tau$ grows the response turns into a damped and then a sustained oscillation. Delayed negative feedback is a generic route to biological oscillations.

::: {.callout-note title="Recipe 4B.2.1"}
**Objective.** Simulate a self-repressing gene whose feedback acts after a delay, and watch oscillations emerge as the delay grows.

**Model.** Delayed negative autoregulation: basal plus repressive Hill production driven by the protein level a delay earlier, minus linear degradation, `dX/dt = g0 + g1/(1 + (X(t-tau)/Xth)^n) - k*X(t)`.

**Method.** The second-order Heun integrator for delay differential equations.

**Test.** Constant history `X(t) = 1`; `g0 = 10, g1 = 60, Xth = 200, n = 4, k = 0.1`, `dt = 0.01`, to t = 200; delay swept `tau = 5, 10, 15, 20`.

**Show.** A plot of X(t) for each delay tau, from a steady state to a sustained oscillation.

**Verification.** A short delay (tau = 5) settles at the steady state near 250; larger delays turn the response into a damped and then a sustained oscillation.
:::

### R implementation {.impl}

```{r}
hill_inh <- function(X, X_th, n) 1 / (1 + (X / X_th)^n)

# delayed self-inhibition: repression depends on X at the earlier time (X_old)
f_1g_self_inh_delays <- function(t, X, X_old, g0, g1, X_th, n, k) {
  g0 + g1 * hill_inh(X_old, X_th, n) - k * X
}

tau_all <- seq(5, 20, 5)
plot(NULL, xlab = "t", ylab = "X", xlim = c(-20, 200), ylim = c(150, 450))
for (j in seq_along(tau_all)) {
  res <- DDE_Heun(f_1g_self_inh_delays, 0, 200, 200, 0.01, tau_all[j],
                  g0 = 10, g1 = 60, X_th = 200, n = 4, k = 0.1)
  lines(res[, 1], res[, 2], col = j)
}
legend("topleft", inset = 0.02, legend = paste0("tau = ", tau_all),
       col = seq_along(tau_all), lty = 1, cex = 0.8)
```

### Python implementation {.impl}

```{python}
def hill_inh(X, X_th, n):
    return 1 / (1 + (X / X_th)**n)

# delayed self-inhibition: repression depends on X at the earlier time (X_old)
def f_1g_self_inh_delays(t, X, X_old, g0, g1, X_th, n, k):
    return g0 + g1 * hill_inh(X_old, X_th, n) - k * X

for tau in range(5, 21, 5):
    res = DDE_Heun(f_1g_self_inh_delays, 0, 200, 200, 0.01, tau,
                   g0=10, g1=60, X_th=200, n=4, k=0.1)
    plt.plot(res[:, 0], res[:, 1], label=f"tau = {tau}")
plt.xlabel("t"); plt.ylabel("X"); plt.xlim(-20, 200); plt.ylim(150, 450)
plt.legend(loc="upper left"); plt.show()
```

## 4B.3 Delayed Lotka-Volterra

Finally, a two-variable delay: in the Lotka-Volterra predator-prey model each species reacts to the other after a lag,

\begin{equation}
\begin{cases}
\dfrac{dN}{dt} = N(t)\,[a - b\,P(t - \tau_1)] \\[2mm]
\dfrac{dP}{dt} = P(t)\,[c\,N(t - \tau_2) - d]
\end{cases} \tag{3}
\end{equation}

For simplicity we set $\tau_1 = \tau_2 = \tau$, which needs an integrator that carries a *vector* history. The undelayed model ($\tau = 0$) has neutrally stable closed orbits (the center of [Part 3D](../03-phase-plane/03d-predator-prey.qmd)). The delay breaks that neutral balance: even a one-step delay ($\tau = 0.01$) makes the orbit spiral outward, and larger delays deviate faster.

::: {.callout-note title="Recipe 4B.3.1"}
**Objective.** Simulate a predator-prey model in which each species responds to the other after a lag.

**Model.** The Lotka-Volterra system with delayed cross-terms, `dN/dt = N(t)*(a - b*P(t-tau))`, `dP/dt = P(t)*(c*N(t-tau) - d)`.

**Method.** The generic multi-variable Heun integrator for delay differential equations, carrying a vector state and vector history.

**Test.** `a = 1, b = 0.03, c = 0.02, d = 1`; initial (N, P) = (30, 10); `dt = 0.01`, to t = 50; delay `tau = 0` and `tau = 0.01`.

**Show.** A phase-plane plot of the predator-prey orbit for tau = 0 and tau = 0.01.

**Verification.** Without delay (tau = 0) the orbit is a neutrally stable closed loop; even a one-step delay (tau = 0.01) makes it spiral outward.
:::

### R implementation {.impl}

```{r}
# Lotka-Volterra with delays; Xs is (N, P) now, Xs_old is (N, P) a delay ago
derivs_LV_delays <- function(t, Xs, Xs_old, a, b, c, d) {
  c(Xs[1] * (a - b * Xs_old[2]), Xs[2] * (c * Xs_old[1] - d))
}

# Heun DDE integrator for a generic multi-variable system (vector state + history)
DDE_Heun_generic <- function(derivs, t0, X0, t.total, dt, tau, ...) {
  t_all <- seq(t0 - tau, t.total, by = dt)
  nt <- length(t_all)
  X_all <- matrix(0, nt, length(X0))
  n_delay <- as.integer(tau / dt)
  for (i in 1:(n_delay + 1)) X_all[i, ] <- X0
  for (i in (n_delay + 1):(nt - 1)) {
    k1 <- dt * derivs(t_all[i], X_all[i, ], X_all[i - n_delay, ], ...)
    X_all[i + 1, ] <- X_all[i, ] + k1
    k2 <- dt * derivs(t_all[i + 1], X_all[i + 1, ], X_all[i + 1 - n_delay, ], ...)
    X_all[i + 1, ] <- X_all[i, ] + (k1 + k2) / 2
  }
  cbind(t_all, X_all)
}
```

```{r}
a <- 1; b <- 0.03; c <- 0.02; d <- 1; X0 <- c(30, 10)
res_nodelay <- DDE_Heun_generic(derivs_LV_delays, 0, X0, 50, 0.01, 0,    a, b, c, d)
res_delay   <- DDE_Heun_generic(derivs_LV_delays, 0, X0, 50, 0.01, 0.01, a, b, c, d)

# time courses without delay: the usual oscillation
plot(res_nodelay[, 1], res_nodelay[, 2], type = "l", col = 2, xlab = "t",
     ylab = "Population", xlim = c(-1, 50), ylim = c(0, 160))
lines(res_nodelay[, 1], res_nodelay[, 3], col = 3)
legend("topright", inset = 0.02, legend = c("N", "P"), col = c(2, 3), lty = 1, cex = 0.8)
```

```{r}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
# phase plane: the delay turns the closed orbit into an outward spiral
par(pty = "s")   # square plot region
plot(res_nodelay[, 2], res_nodelay[, 3], type = "l", col = 2, xlab = "N", ylab = "P",
     xlim = c(0, 180), ylim = c(0, 150))
lines(res_delay[, 2], res_delay[, 3], col = 4)
legend("topright", inset = 0.02, legend = c("no delay", "with delay"),
       col = c(2, 4), lty = 1, cex = 0.8)
```

### Python implementation {.impl}

```{python}
# Lotka-Volterra with delays; Xs is (N, P) now, Xs_old is (N, P) a delay ago
def derivs_LV_delays(t, Xs, Xs_old, a, b, c, d):
    return np.array([Xs[0] * (a - b * Xs_old[1]), Xs[1] * (c * Xs_old[0] - d)])

# Heun DDE integrator for a generic multi-variable system (vector state + history)
def DDE_Heun_generic(derivs, t0, X0, t_total, dt, tau, *args):
    t_all = np.arange(t0 - tau, t_total + dt, dt)
    nt = len(t_all)
    X_all = np.zeros((nt, len(X0)))
    n_delay = int(tau / dt)
    X_all[0:n_delay + 1] = X0
    for i in range(n_delay, nt - 1):
        k1 = dt * derivs(t_all[i], X_all[i], X_all[i - n_delay], *args)
        X_all[i + 1] = X_all[i] + k1
        k2 = dt * derivs(t_all[i + 1], X_all[i + 1], X_all[i + 1 - n_delay], *args)
        X_all[i + 1] = X_all[i] + (k1 + k2) / 2
    return np.column_stack((t_all, X_all))
```

```{python}
a, b, c, d = 1, 0.03, 0.02, 1
X0 = [30, 10]
res_nodelay = DDE_Heun_generic(derivs_LV_delays, 0, X0, 50, 0.01, 0,    a, b, c, d)
res_delay   = DDE_Heun_generic(derivs_LV_delays, 0, X0, 50, 0.01, 0.01, a, b, c, d)

# time courses without delay: the usual oscillation
plt.plot(res_nodelay[:, 0], res_nodelay[:, 1], color="red",   label="N")
plt.plot(res_nodelay[:, 0], res_nodelay[:, 2], color="green", label="P")
plt.xlabel("t"); plt.ylabel("Population"); plt.xlim(-1, 50); plt.ylim(0, 160)
plt.legend(loc="upper right"); plt.show()
```

```{python}
#| fig-width: 5
#| fig-height: 5
#| out-width: "55%"
# phase plane: the delay turns the closed orbit into an outward spiral
plt.plot(res_nodelay[:, 1], res_nodelay[:, 2], color="red",  label="no delay")
plt.plot(res_delay[:, 1],   res_delay[:, 2],   color="blue", label="with delay")
plt.xlabel("N"); plt.ylabel("P"); plt.xlim(0, 180); plt.ylim(0, 150)
plt.legend(loc="upper right"); plt.show()
```

The delayed Lotka-Volterra approach extends to unequal delays $\tau_1 \neq \tau_2$ with a little more bookkeeping (a separate delay index per variable), left as an exercise.
